### 2024-02-22 README.Rmd
add README.Rmd

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "An R project compendium for the Jokura et al 2024 ctenophore apical organ paper"
```

# About the project: 
- this is a repository of an R project containing code and source data for the ctenophore apical organ connectome paper. Contributed by [Jékely lab](https://www.cos.uni-heidelberg.de/en/research-groups/gaspar-jekely).

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The project directory contains:

  - [:file\_folder: text](/text/Manuscript.qmd): Quarto source document
    for manuscript. Includes text, references and inserts the latest version of the figures from the /figures folder. 
There is also a rendered html version,
    `Manuscript.html`, suitable for reading.

  - [:file\_folder: data](/data): Data used in the analysis.
  - [:file\_folder: figures](/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplements](/supplements):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.
  - [:file\_folder:
    source_data](/source_data):
    Source data files associated with each figure containing all data points that have been plotted in the figure.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

If you would like to use the template, you could open Rstudio in your computer, then go to File > New Project > Version Control > Git. Under repository URL add the URL of this page https://github.com/JekelyLab/new_paper_template and save it in a local folder.

The R project will live in the folder (as working directory) where the new_paper_template.Rproj file is saved. All other files and directories will be accessed relative to this working directory. This way the project is portable and self-contained.

The working directory contains this README file, a LICENCE file, a CITATION.cff file, the .gitignore file and the R project file. You can update these files to fit your project.

### Licenses

**Text, figures, code, data :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

### Acknowledgements

This project is modified after
https://annakrystalli.me/rrresearch/10_compendium.html.

### 2024-02-22 README.Rmd
reorganise folders

@@
 
-  - [:file\_folder: data](/data): Data used in the analysis.
-  - [:file\_folder: figures](/figures): Plots and other
+  - [:file\_folder: data](analysis/data): Data used in the analysis.
+  - [:file\_folder: data](analysis/scripts): scripts used in the analysis.
+  - [:file\_folder: figures](manuscript/figures): Plots and other
     illustrations
   - [:file\_folder:
-    supplements](/supplements):
+    supplements](manuscript/supplements):
     Supplementary materials including notes and other documents
@@
   - [:file\_folder:
-    source_data](/source_data):
+    source_data](/manuscript/source_data):
     Source data files associated with each figure containing all data points that have been plotted in the figure.

### 2024-02-22 analysis/scripts/Figure1.R

@@ -1,0 +1,126 @@
+# Code to generate Figure 1 of the Jokura et al 2024 Ctenophore apical organ connectome paper
+
+# source packages and functions -----------------
+source("code/packages_and_functions.R")
+
+# load neurons ----------------------
+
+Q1 <- nlapply(read.neurons.catmaid("Q1", pid = 35, conn = conn_http1),
+              function(x) smooth_neuron(x, sigma = 1000)
+              )
+Q2 <- nlapply(read.neurons.catmaid("Q2", pid = 35, conn = conn_http1),
+              function(x) smooth_neuron(x, sigma = 1000)
+              )
+Q3 <- nlapply(read.neurons.catmaid("Q3", pid = 35, conn = conn_http1),
+              function(x) smooth_neuron(x, sigma = 1000)
+              )
+Q4 <- nlapply(read.neurons.catmaid("Q4", pid = 35, conn = conn_http1),
+              function(x) smooth_neuron(x, sigma = 1000)
+              )
+balancer <- nlapply(read.neurons.catmaid("balancer_cell", pid = 35, conn = conn_http1),
+                     function(x) smooth_neuron(x, sigma = 1000)
+                    )
+LB <- nlapply(read.neurons.catmaid("lamellate body", pid = 35, conn = conn_http1),
+                    function(x) smooth_neuron(x, sigma = 1000)
+                    )
+syn_neuron <- nlapply(read.neurons.catmaid("pre-synaptic", pid = 35, conn = conn_http1),
+                    function(x) smooth_neuron(x, sigma = 1000)
+                    )
+monocilia <- nlapply(read.neurons.catmaid("monocilia", pid = 35, conn = conn_http1),
+                    function(x) smooth_neuron(x, sigma = 1000)
+                    )
+
+balancer_Q1 <- nlapply(
+  read.neurons.catmaid(skids_by_2annotations("balancer_cell", "Q1"),
+                       pid = 35, conn = conn_http1), 
+  function(x) smooth_neuron(x, sigma = 1000)
+  )
+balancer_Q2 <- nlapply(
+  read.neurons.catmaid(skids_by_2annotations("balancer_cell", "Q2"),
[…]

### 2024-02-22 analysis/scripts/Video1.R

@@ -1,0 +1,182 @@
+#Video1 of the Jokura et al ctenophore AO paper
+
+source("code/packages_and_functions.R")
+
+#create temp dir to store video frames
+mainDir = getwd()
+dir.create(file.path(mainDir, "videoframes"), showWarnings = FALSE)
+
+
+# load cell types from catmaid
+
+balancer <- nlapply(
+  read.neurons.catmaid(
+    "celltype:balancer", pid = 35
+    ),
+  function(x) smooth_neuron(x, sigma = 1000)
+)
+
+LB <- nlapply(
+  read.neurons.catmaid(
+    "celltype:lamellate", pid = 35
+    ),
+  function(x) smooth_neuron(x, sigma = 1000)
+)
+
+syn_neuron <- nlapply(
+  read.neurons.catmaid(
+    "celltype:neuron", pid = 35
+    ),
+  function(x) smooth_neuron(x, sigma = 1000)
+)
+
+monociliated <- nlapply(
+  read.neurons.catmaid(
+    "monociliated_cell", pid = 35
+  ),
+  function(x) smooth_neuron(x, sigma = 1000)
+)
+
+bridge <- nlapply(
+  read.neurons.catmaid(
+    "celltype:bridge", pid = 35
+  ),
+  function(x) smooth_neuron(x, sigma = 1000)
+)
+
+# cell types
+AO_celltypes <- list(
+  balancer, LB, syn_neuron, bridge, monociliated
+  )
+AO_celltype_names <- list(
+  "balancer", "LB", "syn_neuron", "bridge", "monociliated"
+  )
+
+# plot cells ----------------
+
+nopen3d() # opens a pannable 3d window
+mfrow3d(1, 2)  #defines the two scenes
+par3d(windowRect = c(0, 0, 1600, 800)) #to define the size of the rgl window
+par3d(zoom=0.75)
+nview3d("ventral", extramat=rotationMatrix(1.2, 0, 0, 1))
+
+plot3d(
+  balancer, soma = TRUE, color = Okabe_Ito[1], 
+  alpha = 0.6, lwd = 3
+  )
+
+texts3d(
+  30000, 32000, 1000, text = "balancer", col='black', cex = 2
+  )
+for(i in 101:120){
[…]

### 2024-02-22 analysis/scripts/cilium_lengths.R

@@ -1,0 +1,102 @@
+# code to measure length of cilia
+# WARNING: It ia absolutely essential that the root node is proximal to the basal body 
+# i.e., the order of nodes needs to be "root" -> "basal body" -> "cilium tip"
+# it should work fine if skeleton has multiple cilia
+
+library(catmaid)
+
+conn <- source("~/R/conn.R")
+
+tags <- catmaid_get_label_stats(pid = 35)
+
+# using "cilium tip" tag because that shows that cilium is complete, cut off cilia don't have that tag
+# measuring from tips because it's easy to get parent node, but harder to get child
+ctips <- tags[tags$labelName == "cilium tip", "treenodeID"]
+
+cilium_length <- function(ctip_id) {
+  # this will only work if root node is proximal to basal body tag
+  # treenode compact detail response:
+  # [ID, parent ID, x, y, z, confidence, radius, skeleton_id, edition_time, user_id]
+  path <- paste("/35/treenodes/", ctip_id, "/compact-detail", sep = "")
+  ctip_treenode <- catmaid_fetch(path = path)
+  skid <- ctip_treenode[[8]]
+  neuron_info <- read.neuron.catmaid(skid, pid = 35)
+  neuron_name <- catmaid_fetch(path = "/35/skeleton/neuronnames",
+                               body = list(skids=skid))[[1]]
+  # check which treenodes have basal body annotation, because this is the stop point
+  basalb_ids <- neuron_info$tags$`basal body`
+  treenode_id <- ctip_id
+  distance <- 0
+  while (!(treenode_id %in% basalb_ids)) {
+    index <- match(treenode_id, neuron_info$d$PointNo)
+    parent_id <- neuron_info$d$Parent[[index]]
+    if (parent_id == -1) {
+      print("root reached before basal body tag, from cilium tip")
+      print(ctip_id)
+      stop()
+    }
+    index_parent <- match(parent_id, neuron_info$d$PointNo)
+ 
+    distance_2nodes <- dist(rbind(c(neuron_info$d$X[[index]],
+                                neuron_info$d$Y[[index]],
+                                neuron_info$d$Z[[index]]),
+                              c(neuron_info$d$X[[index_parent]],
+                                neuron_info$d$Y[[index_parent]],
[…]

### 2024-02-22 analysis/scripts/crop_synapses_from_catmaid.R

@@ -1,0 +1,106 @@
+# crop synapses
+library(catmaid)
+
+conn <- source("~/R/conn.R")
+
+dir.create("synapse_tiff_stacks")
+
+# get all synapses from CATMAID
+all_syn_connectors <- catmaid_fetch(
+  path = "35/connectors/",
+  body = list(
+    relation_type = "presynaptic_to",
+    with_partners = "false"
+  )
+)
+
+# cut out volume is not isometric, because it looks weird
+half_bb_size_xy=1800
+half_bb_size_z=1000
+for (nconnector in seq_along(all_syn_connectors$connectors)) {
+#for (nconnector in seq(1,5)) {
+  print(nconnector)
+  id=all_syn_connectors$connectors[[nconnector]][[1]]
+  x=all_syn_connectors$connectors[[nconnector]][[2]]
+  y=all_syn_connectors$connectors[[nconnector]][[3]]
+  z=all_syn_connectors$connectors[[nconnector]][[4]]
+  catmaid_fetch(
+    path = "35/crop/",
+    body = list(
+      stack_ids=28,
+      min_x=x-half_bb_size_xy,
+      min_y=y-half_bb_size_xy,
+      min_z=z-half_bb_size_z,
+      max_x=x+half_bb_size_xy,
+      max_y=y+half_bb_size_xy,
+      max_z=z+half_bb_size_z,
+      zoom_level=0,
+      single_channel='true',
+      rotationcw=0
+    )
+  )
+}
+
+# server needs time to process this
+Sys.sleep(200)
+
+# download stacks ---------------------------------------------------------
+# the exact file names are needed for download,
+# to get them we need to check messages from the server
+# originally I was checking which messages are new, and downloading those
+# but this could lead to false positives
+# so I decided that's it's probably safe to download the last n messages
+# where n is the number of jobs
+
+# It would be cool if I can put the synapse ID as file name,
+# but it's not guaranteed that jobs will return in the same order
+
+setwd("synapse_tiff_stacks")
+
+cat_messages <- catmaid_fetch(
+  path = "messages/list"
+)
+
+for (i in seq_along(all_syn_connectors$connectors)) {
+  cat_message <- cat_messages[[i]]$text
+  link <- regmatches(cat_message, 
[…]

### 2024-02-22 analysis/scripts/network_graphs.R

@@ -1,0 +1,197 @@
+#network analysis and plotting
+
+source("code/packages_and_functions.R")
+
+# Create a random graph with 10 nodes
+g <- erdos.renyi.game(10, p = 0.3)
+
+# Assign names to the nodes
+V(g)$name <- c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J")
+#edges
+E(g)
+#vertices (nodes)
+V(g)
+
+g_tb <- g |> 
+  as_tbl_graph()
+
+g_tb |>
+  activate(nodes) |>
+  filter(name == "A")
+
+g_tb |>
+  activate(nodes) |>
+  mutate(color = "blue")
+
+plot(g_tb)
+
+# load cell types from catmaid
+
+balancer <- nlapply(
+  read.neurons.catmaid(
+    "celltype:balancer", pid = 35
+    ),
+  function(x) smooth_neuron(x, sigma = 1000)
+)
+
+LB <- nlapply(
+  read.neurons.catmaid(
+    "celltype:lamellate", pid = 35
+    ),
+  function(x) smooth_neuron(x, sigma = 1000)
+)
+
+syn_neuron <- nlapply(
+  read.neurons.catmaid(
+    "celltype:neuron", pid = 35
+    ),
+  function(x) smooth_neuron(x, sigma = 1000)
+)
+
+monociliated <- nlapply(
+  read.neurons.catmaid(
+    "monociliated_cell", pid = 35
+  ),
+  function(x) smooth_neuron(x, sigma = 1000)
+)
+
+bridge <- nlapply(
+  read.neurons.catmaid(
+    "celltype:bridge", pid = 35
+  ),
+  function(x) smooth_neuron(x, sigma = 1000)
+)
+
+# cell types
+AO_celltypes <- list(
+  balancer, LB, syn_neuron, bridge, monociliated
+  )
+AO_celltype_names <- list(
+  "balancer", "LB", "syn_neuron", "bridge", "monociliated"
+  )
+
+# iterate through cell group neuron lists and get connectivity
+# define empty synapse list with the right dimensions
+synapse_list <- c()
+
+for (i in 1:length(AO_celltypes)) {
+  for (j in 1:length(AO_celltypes)) {
+    # get connectors between two cell groups
+    presyn_skids <- attr(AO_celltypes[i][[1]], "df")$skid
+    postsyn_skids <- attr(AO_celltypes[j][[1]], "df")$skid
+    connectivity <- catmaid_get_connectors_between(
[…]

### 2024-02-22 analysis/scripts/number_of_postsynaptic_sites_per_synapse.R

@@ -1,0 +1,107 @@
+library(catmaid)
+library(data.table)
+library(tidyverse)
+
+conn <- source("~/R/conn.R")
+
+# calculate average number of post-synaptic sites per synapse  -----
+
+pre_connectors <- catmaid_fetch(
+  path = "35/connectors/",
+  body = list(
+    relation_type = "presynaptic_to",
+    with_partners = "false"
+  )
+)
+
+n_pre_post <- function(connector_id) {
+  connector_info <- catmaid_fetch(path = paste("35/connectors/", connector_id, sep = ""))
+  # pre should always be 1, but maybe it's good to check if there is any weird stuff going on
+  pre <- connector_info$partners %like% 'presynaptic_to'
+  n_pre <- table(pre)["TRUE"][[1]]
+  post <- connector_info$partners %like% 'postsynaptic_to'
+  n_post <- table(post)["TRUE"][[1]]
+  return(n_post)
+}
+
+pre_connector_IDs <- lapply(pre_connectors$connectors,'[[',1)
+
+n_post <- lapply(pre_connector_IDs, n_pre_post)
+
+summary(as.numeric(n_post))
+
+# calculate percentage of mitochondria which have vesicles or synapses associated with them -----
+
+synaptic_neurons <- read.neurons.catmaid("synaptic neuron", pid=35)
+
+mito_in_syn <- data.frame()
+for (i in seq_along(synaptic_neurons)) {
+  n_mito_vesicles_syn <- 0
+  n_mito_vesicles_no_syn <- 0
+  mito_vesicles <- synaptic_neurons[[i]]$tags$`mitochondrion vesicles`
+  # this doesn't check which type of connector is associated with the node,
+  # but in 99.5% cases it will be presynaptic, so it's probably not necessary to do extra checks
+  syn <- synaptic_neurons[[i]]$connectors$treenode_id
+  mito_syn <- list()
+  for (mito_vesicle in mito_vesicles) {
+    if (mito_vesicle %in% syn) {
+      n_mito_vesicles_syn <- n_mito_vesicles_syn + 1
+    }
+    else {
+      n_mito_vesicles_no_syn <- n_mito_vesicles_no_syn + 1
+    }
+  }
+  mito_unclear_vesicles <- synaptic_neurons[[i]]$tags$`mitochondrion unclear vesicles`
+  n_mito_unclear_vesicles <- length(mito_unclear_vesicles)
+  mito_no_vesicles <- synaptic_neurons[[i]]$tags$`mitochondrion no vesicles`
[…]

### 2024-02-22 analysis/scripts/packages_and_functions.R

@@ -1,0 +1,62 @@
+rm(list = ls(all.names = TRUE)) #will clear all objects includes hidden objects.
+gc() #free up memory and report the memory usage.
+
+# load some packages
+library(catmaid)
+library(tidyverse)
+library(cowplot)
+library(png)
+library(igraph)
+library(networkD3)
+library(visNetwork)
+library(webshot2)
+library(patchwork)
+library(RColorBrewer)
+library(igraph)
+library(tidygraph)
+library(av)
+
+# catmaid connection, needs username, password AND token - weird!
+{
+  # can run this separate file using source function
+  conn <- source("~/R/conn.R")
+  #for larger calls we need to use http/1, see https://www.gitmemory.com/issue/natverse/rcatmaid/158/641537466
+  #for this we configure to http/1.1
+  conn_http1 = catmaid_login(conn=conn, config=httr::config(ssl_verifypeer=0, http_version=1))
+}
+
+#define some colors
+#From Color Universal Design (CUD): https://jfly.uni-koeln.de/color/
+Okabe_Ito <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", 
+               "#CC79A7", "#000000")
+
+blues <- brewer.pal(9, 'Blues')
+bluepurple <- brewer.pal(9, 'BuPu')
+oranges <- brewer.pal(9, 'YlOrRd')
+
+#save session info and Rstudio version info for reproducibility
+writeLines(capture.output(sessionInfo()), "code/sessionInfo.txt")
+writeLines(capture.output(rstudioapi::versionInfo()), "code/versionInfo.txt")
+
+# functions ----------------------
+
+skids_by_2annotations <- function(annotation1,annotation2){
+  skids1 <- catmaid_skids(annotation1, pid = 35)
+  skids2 <- catmaid_skids(annotation2, pid = 35)
+  intersect <- intersect(skids1, skids2)
+  return(intersect)
+}
+
+plot_background <- function(x){
+  nopen3d() # opens a pannable 3d window
+  par3d(zoom=0.75)
+  nview3d("frontal", extramat=rotationMatrix(1.2, 0, 0, 1))
+  par3d(windowRect = c(0, 0, 800, 800)) #resize for frontal view
+}
+
+plot_background_ventral <- function(x){
+  nopen3d() # opens a pannable 3d window
+  par3d(zoom=0.75)
[…]

### 2024-02-22 code/Figure1.R
move files

@@ -1,126 +1,0 @@
-# Code to generate Figure 1 of the Jokura et al 2024 Ctenophore apical organ connectome paper
-
-# source packages and functions -----------------
-source("code/packages_and_functions.R")
-
-# load neurons ----------------------
-
-Q1 <- nlapply(read.neurons.catmaid("Q1", pid = 35, conn = conn_http1),
-              function(x) smooth_neuron(x, sigma = 1000)
-              )
-Q2 <- nlapply(read.neurons.catmaid("Q2", pid = 35, conn = conn_http1),
-              function(x) smooth_neuron(x, sigma = 1000)
-              )
-Q3 <- nlapply(read.neurons.catmaid("Q3", pid = 35, conn = conn_http1),
-              function(x) smooth_neuron(x, sigma = 1000)
-              )
-Q4 <- nlapply(read.neurons.catmaid("Q4", pid = 35, conn = conn_http1),
-              function(x) smooth_neuron(x, sigma = 1000)
-              )
-balancer <- nlapply(read.neurons.catmaid("balancer_cell", pid = 35, conn = conn_http1),
-                     function(x) smooth_neuron(x, sigma = 1000)
-                    )
-LB <- nlapply(read.neurons.catmaid("lamellate body", pid = 35, conn = conn_http1),
-                    function(x) smooth_neuron(x, sigma = 1000)
-                    )
-syn_neuron <- nlapply(read.neurons.catmaid("pre-synaptic", pid = 35, conn = conn_http1),
-                    function(x) smooth_neuron(x, sigma = 1000)
-                    )
-monocilia <- nlapply(read.neurons.catmaid("monocilia", pid = 35, conn = conn_http1),
-                    function(x) smooth_neuron(x, sigma = 1000)
-                    )
-
-balancer_Q1 <- nlapply(
-  read.neurons.catmaid(skids_by_2annotations("balancer_cell", "Q1"),
-                       pid = 35, conn = conn_http1), 
-  function(x) smooth_neuron(x, sigma = 1000)
-  )
-balancer_Q2 <- nlapply(
-  read.neurons.catmaid(skids_by_2annotations("balancer_cell", "Q2"),
[…]

### 2024-02-22 code/Video1.R

@@ -1,182 +1,0 @@
-#Video1 of the Jokura et al ctenophore AO paper
-
-source("code/packages_and_functions.R")
-
-#create temp dir to store video frames
-mainDir = getwd()
-dir.create(file.path(mainDir, "videoframes"), showWarnings = FALSE)
-
-
-# load cell types from catmaid
-
-balancer <- nlapply(
-  read.neurons.catmaid(
-    "celltype:balancer", pid = 35
-    ),
-  function(x) smooth_neuron(x, sigma = 1000)
-)
-
-LB <- nlapply(
-  read.neurons.catmaid(
-    "celltype:lamellate", pid = 35
-    ),
-  function(x) smooth_neuron(x, sigma = 1000)
-)
-
-syn_neuron <- nlapply(
-  read.neurons.catmaid(
-    "celltype:neuron", pid = 35
-    ),
-  function(x) smooth_neuron(x, sigma = 1000)
-)
-
-monociliated <- nlapply(
-  read.neurons.catmaid(
-    "monociliated_cell", pid = 35
-  ),
-  function(x) smooth_neuron(x, sigma = 1000)
-)
-
-bridge <- nlapply(
-  read.neurons.catmaid(
-    "celltype:bridge", pid = 35
-  ),
-  function(x) smooth_neuron(x, sigma = 1000)
-)
-
-# cell types
-AO_celltypes <- list(
-  balancer, LB, syn_neuron, bridge, monociliated
-  )
-AO_celltype_names <- list(
-  "balancer", "LB", "syn_neuron", "bridge", "monociliated"
-  )
-
-# plot cells ----------------
-
-nopen3d() # opens a pannable 3d window
-mfrow3d(1, 2)  #defines the two scenes
-par3d(windowRect = c(0, 0, 1600, 800)) #to define the size of the rgl window
-par3d(zoom=0.75)
-nview3d("ventral", extramat=rotationMatrix(1.2, 0, 0, 1))
-
-plot3d(
-  balancer, soma = TRUE, color = Okabe_Ito[1], 
-  alpha = 0.6, lwd = 3
-  )
-
-texts3d(
-  30000, 32000, 1000, text = "balancer", col='black', cex = 2
-  )
-for(i in 101:120){
[…]

### 2024-02-22 code/cell_type_clustering.R

@@ -1,85 +1,0 @@
-library(catmaid)
-library(tidyverse)  # data manipulation
-library(cluster)    # clustering algorithms
-library(factoextra) # clustering algorithms & visualization
-
-conn <- source("~/R/conn.R")
-
-skids <- unlist(
-  catmaid_fetch(path = "/35/skeletons/"))
-
-characters <- list("soma",
-                   "centriole",
-                   "basal body",
-                   "golgi",
-                   "peroxisome?",
-                   "multivesicular body",
-                   "SER",
-                   "RER",
-                   "plume",
-                   "large clear vesicles",
-                   "small clear vesicles",
-                   "large electrondense vesicles",
-                   "small dark dots")
-
-tag_stats_per_skid <- data.frame(matrix(ncol = length(characters)+1, nrow = 0))
-colnames(tag_stats_per_skid) <- c("skid", characters)
-tag_stats <- function(skid) {
-  # if skeleton has only one node read.neuron.catmaid gives an error
-  try(neuron_info <- read.neuron.catmaid(skid, pid = 35), silent=TRUE)
-  if("try-error" %in% class(t)) return(NULL)
-  tags <- neuron_info$tags
-  counts <- mapply(tag_count, list(tags), characters)
-  # ignore cells without a soma
-  if (counts[[1]] == 0) {
-    return(NULL)
-  }
-  else {
-    return(counts)
-  }
-}
-
-for (skid in skids) {
-  tag_counts <- tag_stats(skid)
-  if (!is.null(tag_counts)) {
-    tag_stats_per_skid[nrow(tag_stats_per_skid) + 1,] = c(skid, tag_counts)
-  }
-}
-
-tag_stats_noskid <- tag_stats_per_skid[,-1]
-
-km <- kmeans(tag_stats_noskid, centers = 20, nstart = 25)
-fviz_cluster(km, data = tag_stats_per_skid)
-
-
-fviz_cluster(km,# clustering result 
-             data = tag_stats_noskid, # data 
-             ellipse.type = "convex", 
-             star.plot = TRUE, 
-             repel = TRUE, 
-             ggtheme = theme_minimal()
-) 
-
-
-tag_stats_per_skid %>%
-  as_tibble() %>%
-  mutate(cluster = km$cluster,
-         state = row.names(skid)) %>%
[…]

### 2024-02-22 code/cilium_lengths.R

@@ -1,102 +1,0 @@
-# code to measure length of cilia
-# WARNING: It ia absolutely essential that the root node is proximal to the basal body 
-# i.e., the order of nodes needs to be "root" -> "basal body" -> "cilium tip"
-# it should work fine if skeleton has multiple cilia
-
-library(catmaid)
-
-conn <- source("~/R/conn.R")
-
-tags <- catmaid_get_label_stats(pid = 35)
-
-# using "cilium tip" tag because that shows that cilium is complete, cut off cilia don't have that tag
-# measuring from tips because it's easy to get parent node, but harder to get child
-ctips <- tags[tags$labelName == "cilium tip", "treenodeID"]
-
-cilium_length <- function(ctip_id) {
-  # this will only work if root node is proximal to basal body tag
-  # treenode compact detail response:
-  # [ID, parent ID, x, y, z, confidence, radius, skeleton_id, edition_time, user_id]
-  path <- paste("/35/treenodes/", ctip_id, "/compact-detail", sep = "")
-  ctip_treenode <- catmaid_fetch(path = path)
-  skid <- ctip_treenode[[8]]
-  neuron_info <- read.neuron.catmaid(skid, pid = 35)
-  neuron_name <- catmaid_fetch(path = "/35/skeleton/neuronnames",
-                               body = list(skids=skid))[[1]]
-  # check which treenodes have basal body annotation, because this is the stop point
-  basalb_ids <- neuron_info$tags$`basal body`
-  treenode_id <- ctip_id
-  distance <- 0
-  while (!(treenode_id %in% basalb_ids)) {
-    index <- match(treenode_id, neuron_info$d$PointNo)
-    parent_id <- neuron_info$d$Parent[[index]]
-    if (parent_id == -1) {
-      print("root reached before basal body tag, from cilium tip")
-      print(ctip_id)
-      stop()
-    }
-    index_parent <- match(parent_id, neuron_info$d$PointNo)
- 
-    distance_2nodes <- dist(rbind(c(neuron_info$d$X[[index]],
-                                neuron_info$d$Y[[index]],
-                                neuron_info$d$Z[[index]]),
-                              c(neuron_info$d$X[[index_parent]],
-                                neuron_info$d$Y[[index_parent]],
-                                neuron_info$d$Z[[index_parent]])
[…]

### 2024-02-22 code/crop_synapses_from_catmaid.R

@@ -1,106 +1,0 @@
-# crop synapses
-library(catmaid)
-
-conn <- source("~/R/conn.R")
-
-dir.create("synapse_tiff_stacks")
-
-# get all synapses from CATMAID
-all_syn_connectors <- catmaid_fetch(
-  path = "35/connectors/",
-  body = list(
-    relation_type = "presynaptic_to",
-    with_partners = "false"
-  )
-)
-
-# cut out volume is not isometric, because it looks weird
-half_bb_size_xy=1800
-half_bb_size_z=1000
-for (nconnector in seq_along(all_syn_connectors$connectors)) {
-#for (nconnector in seq(1,5)) {
-  print(nconnector)
-  id=all_syn_connectors$connectors[[nconnector]][[1]]
-  x=all_syn_connectors$connectors[[nconnector]][[2]]
-  y=all_syn_connectors$connectors[[nconnector]][[3]]
-  z=all_syn_connectors$connectors[[nconnector]][[4]]
-  catmaid_fetch(
-    path = "35/crop/",
-    body = list(
-      stack_ids=28,
-      min_x=x-half_bb_size_xy,
-      min_y=y-half_bb_size_xy,
-      min_z=z-half_bb_size_z,
-      max_x=x+half_bb_size_xy,
-      max_y=y+half_bb_size_xy,
-      max_z=z+half_bb_size_z,
-      zoom_level=0,
-      single_channel='true',
-      rotationcw=0
-    )
-  )
-}
-
-# server needs time to process this
-Sys.sleep(200)
-
-# download stacks ---------------------------------------------------------
-# the exact file names are needed for download,
-# to get them we need to check messages from the server
-# originally I was checking which messages are new, and downloading those
-# but this could lead to false positives
-# so I decided that's it's probably safe to download the last n messages
-# where n is the number of jobs
-
-# It would be cool if I can put the synapse ID as file name,
-# but it's not guaranteed that jobs will return in the same order
-
-setwd("synapse_tiff_stacks")
-
-cat_messages <- catmaid_fetch(
-  path = "messages/list"
-)
-
-for (i in seq_along(all_syn_connectors$connectors)) {
-  cat_message <- cat_messages[[i]]$text
-  link <- regmatches(cat_message, 
[…]

### 2024-02-22 code/network_graphs.R

@@ -1,197 +1,0 @@
-#network analysis and plotting
-
-source("code/packages_and_functions.R")
-
-# Create a random graph with 10 nodes
-g <- erdos.renyi.game(10, p = 0.3)
-
-# Assign names to the nodes
-V(g)$name <- c("A", "B", "C", "D", "E", "F", "G", "H", "I", "J")
-#edges
-E(g)
-#vertices (nodes)
-V(g)
-
-g_tb <- g |> 
-  as_tbl_graph()
-
-g_tb |>
-  activate(nodes) |>
-  filter(name == "A")
-
-g_tb |>
-  activate(nodes) |>
-  mutate(color = "blue")
-
-plot(g_tb)
-
-# load cell types from catmaid
-
-balancer <- nlapply(
-  read.neurons.catmaid(
-    "celltype:balancer", pid = 35
-    ),
-  function(x) smooth_neuron(x, sigma = 1000)
-)
-
-LB <- nlapply(
-  read.neurons.catmaid(
-    "celltype:lamellate", pid = 35
-    ),
-  function(x) smooth_neuron(x, sigma = 1000)
-)
-
-syn_neuron <- nlapply(
-  read.neurons.catmaid(
-    "celltype:neuron", pid = 35
-    ),
-  function(x) smooth_neuron(x, sigma = 1000)
-)
-
-monociliated <- nlapply(
-  read.neurons.catmaid(
-    "monociliated_cell", pid = 35
-  ),
-  function(x) smooth_neuron(x, sigma = 1000)
-)
-
-bridge <- nlapply(
-  read.neurons.catmaid(
-    "celltype:bridge", pid = 35
-  ),
-  function(x) smooth_neuron(x, sigma = 1000)
-)
-
-# cell types
-AO_celltypes <- list(
-  balancer, LB, syn_neuron, bridge, monociliated
-  )
-AO_celltype_names <- list(
-  "balancer", "LB", "syn_neuron", "bridge", "monociliated"
-  )
-
-# iterate through cell group neuron lists and get connectivity
-# define empty synapse list with the right dimensions
-synapse_list <- c()
-
-for (i in 1:length(AO_celltypes)) {
-  for (j in 1:length(AO_celltypes)) {
-    # get connectors between two cell groups
-    presyn_skids <- attr(AO_celltypes[i][[1]], "df")$skid
-    postsyn_skids <- attr(AO_celltypes[j][[1]], "df")$skid
-    connectivity <- catmaid_get_connectors_between(
[…]

### 2024-02-22 code/number_of_postsynaptic_sites_per_synapse.R

@@ -1,107 +1,0 @@
-library(catmaid)
-library(data.table)
-library(tidyverse)
-
-conn <- source("~/R/conn.R")
-
-# calculate average number of post-synaptic sites per synapse  -----
-
-pre_connectors <- catmaid_fetch(
-  path = "35/connectors/",
-  body = list(
-    relation_type = "presynaptic_to",
-    with_partners = "false"
-  )
-)
-
-n_pre_post <- function(connector_id) {
-  connector_info <- catmaid_fetch(path = paste("35/connectors/", connector_id, sep = ""))
-  # pre should always be 1, but maybe it's good to check if there is any weird stuff going on
-  pre <- connector_info$partners %like% 'presynaptic_to'
-  n_pre <- table(pre)["TRUE"][[1]]
-  post <- connector_info$partners %like% 'postsynaptic_to'
-  n_post <- table(post)["TRUE"][[1]]
-  return(n_post)
-}
-
-pre_connector_IDs <- lapply(pre_connectors$connectors,'[[',1)
-
-n_post <- lapply(pre_connector_IDs, n_pre_post)
-
-summary(as.numeric(n_post))
-
-# calculate percentage of mitochondria which have vesicles or synapses associated with them -----
-
-synaptic_neurons <- read.neurons.catmaid("synaptic neuron", pid=35)
-
-mito_in_syn <- data.frame()
-for (i in seq_along(synaptic_neurons)) {
-  n_mito_vesicles_syn <- 0
-  n_mito_vesicles_no_syn <- 0
-  mito_vesicles <- synaptic_neurons[[i]]$tags$`mitochondrion vesicles`
-  # this doesn't check which type of connector is associated with the node,
-  # but in 99.5% cases it will be presynaptic, so it's probably not necessary to do extra checks
-  syn <- synaptic_neurons[[i]]$connectors$treenode_id
-  mito_syn <- list()
-  for (mito_vesicle in mito_vesicles) {
-    if (mito_vesicle %in% syn) {
-      n_mito_vesicles_syn <- n_mito_vesicles_syn + 1
-    }
-    else {
-      n_mito_vesicles_no_syn <- n_mito_vesicles_no_syn + 1
-    }
-  }
-  mito_unclear_vesicles <- synaptic_neurons[[i]]$tags$`mitochondrion unclear vesicles`
-  n_mito_unclear_vesicles <- length(mito_unclear_vesicles)
-  mito_no_vesicles <- synaptic_neurons[[i]]$tags$`mitochondrion no vesicles`
[…]

### 2024-02-22 code/packages_and_functions.R

@@ -1,62 +1,0 @@
-rm(list = ls(all.names = TRUE)) #will clear all objects includes hidden objects.
-gc() #free up memory and report the memory usage.
-
-# load some packages
-library(catmaid)
-library(tidyverse)
-library(cowplot)
-library(png)
-library(igraph)
-library(networkD3)
-library(visNetwork)
-library(webshot2)
-library(patchwork)
-library(RColorBrewer)
-library(igraph)
-library(tidygraph)
-library(av)
-
-# catmaid connection, needs username, password AND token - weird!
-{
-  # can run this separate file using source function
-  conn <- source("~/R/conn.R")
-  #for larger calls we need to use http/1, see https://www.gitmemory.com/issue/natverse/rcatmaid/158/641537466
-  #for this we configure to http/1.1
-  conn_http1 = catmaid_login(conn=conn, config=httr::config(ssl_verifypeer=0, http_version=1))
-}
-
-#define some colors
-#From Color Universal Design (CUD): https://jfly.uni-koeln.de/color/
-Okabe_Ito <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", 
-               "#CC79A7", "#000000")
-
-blues <- brewer.pal(9, 'Blues')
-bluepurple <- brewer.pal(9, 'BuPu')
-oranges <- brewer.pal(9, 'YlOrRd')
-
-#save session info and Rstudio version info for reproducibility
-writeLines(capture.output(sessionInfo()), "code/sessionInfo.txt")
-writeLines(capture.output(rstudioapi::versionInfo()), "code/versionInfo.txt")
-
-# functions ----------------------
-
-skids_by_2annotations <- function(annotation1,annotation2){
-  skids1 <- catmaid_skids(annotation1, pid = 35)
-  skids2 <- catmaid_skids(annotation2, pid = 35)
-  intersect <- intersect(skids1, skids2)
-  return(intersect)
-}
-
-plot_background <- function(x){
-  nopen3d() # opens a pannable 3d window
-  par3d(zoom=0.75)
-  nview3d("frontal", extramat=rotationMatrix(1.2, 0, 0, 1))
-  par3d(windowRect = c(0, 0, 800, 800)) #resize for frontal view
-}
-
-plot_background_ventral <- function(x){
-  nopen3d() # opens a pannable 3d window
-  par3d(zoom=0.75)
[…]

### 2024-02-22 analysis/scripts/contactome.R
contactome: new script to get abutting contactome

@@ -1,0 +1,17 @@
+source("analysis/scripts/packages_and_functions.R")
+
+# get all abutting contacts from CATMAID ------------
+
+contacts <- catmaid_fetch(
+  path = "35/connectors/",
+  body = list(
+    relation_type = "abutting",
+    with_partners = "true"
+  )
+)
+length(contacts$connectors)
+
+skid1 <- contacts$partners[[1]][[1]][[3]]
+
+skid2 <- contacts$partners[[1]][[2]][[3]]
+

### 2024-02-22 manuscript/Manuscript.qmd
reorganise folders

---
title: "New paper template - title here"
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
max-width: 100%
number-sections: false
format:
  html: default
  pdf:
    pdf-engine: xelatex
    number-sections: true
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

\

Author 1^1,2,\*^ , Author 2^2^, Author 3^1,2^\

^1^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany\
^2^Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom\
^\*^Correspondence: corr_author\@email.com

## Abstract

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

**text in bold** *italic* [underline]{.underline}

## Introduction

You can add references either by referring to their id in the .bib file e.g., [@Marinkovi2020], or by switching to the visual editor (Cogwheel in the .Rmd menu -\> Use Visual Editor). The awesome paper by [@Jokura_2019].

In the visual editor mode, go to 'Insert' -\> \@ Citation

You can select a Zotero library, PubMed, CrossRef etc. and insert the citations.

The easiest way is to use the command line:

```{bash}
#| echo: true
#| eval: false

curl -LH "Accept: application/x-bibtex" https://doi.org/10.7554/eLife.91258.1 >> references.bib
```

*Platynereis dumerilii* is a marine annelid... [@ozpolat2021]

The references are stored in manuscript/references.bib (need to be defined in the Yaml header). This file will automatically updated when you insert a new reference through the Visual editor \> Insert \> Citations.

In this documents, references will be formatted in the style of eLife. This is defined in the Yaml header under: csl: elife.csl. The elife.csl file is saved in the /manuscript folder.

If you would like to use a different citation format, download the respective .csl file (e.g., from the Zotero style repository <https://www.zotero.org/styles>), save it in the /manuscript folder of the project and change the Yaml to csl: your_favourite_journal.csl.

## Results

Inserting Figures

You can add your figures into the rendered document. We saved the figures into /manuscript/figures or /manuscript/figure_supplements and can insert them from there. We use knitr::include_graphics for this. The title and legend can also be edited, as will as the width of the output figure.

```{r, out.width = "100%", fig.cap="**Figure 1. Title fig 1** (A) legend (B) legend. " }
#| echo: false

knitr::include_graphics("figures/Figure1.png")
```

## Equations

Equations can also be inserted, Insert -\> Display Math:

$$
\bar{X} = \frac{\sum_{i=1}^{n} x_{i}}{n}
$$

## Sourcing code and working with variable

```{r}
#| include: false
#| eval: true
#source(here::here("analysis/scripts/statistics_for_paper.R"))
```

The 'analysis/scripts/statistics_for_paper.R' script is sourced and it runs but the output is not included in the knitted output. But we can access the variables defined in the sourced script simply by adding \` r var_name \` between \` backticks, in this case max_PRC value is (now this number comes from our sourced script).

If we update the data, the script can recalculate the variable we want to refer to in the text and update the number.

## Materials and Methods

## Acknowledgements

We would like to thank the Jekely lab for the R project template (<https://github.com/JekelyLab/new_paper_template>) we used to write this paper. This work was funded by ...

## References

### 2024-02-22 manuscript/Manuscript.tex
reorganise folders

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Arial}
  \setmonofont[]{DejaVu Sans Mono}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[lmargin=0.8in,rmargin=0.8in,tmargin=0.8in,bmargin=0.8in]{geometry}
\ifLuaTeX
  \usepackage{luacolor}
  \usepackage[soul]{lua-ul}
\else
  \usepackage{soul}

\fi
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{lineno}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={New paper template - title here},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{New paper template - title here}
\author{}
\date{}

\begin{document}
\maketitle

\linenumbers

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}
\setstretch{1.2}
\hfill\break

Author 1\textsuperscript{1,2,*} , Author 2\textsuperscript{2}, Author
3\textsuperscript{1,2}\\

\textsuperscript{1}Heidelberg University, Centre for Organismal Studies
(COS), 69120 Heidelberg, Germany\\
\textsuperscript{2}Living Systems Institute, University of Exeter,
Exeter, EX4 4QD, United Kingdom\\
\textsuperscript{*}Correspondence: corr\_author@email.com

\section{Abstract}\label{abstract}

This is an R Markdown document. Markdown is a simple formatting syntax
for authoring HTML, PDF, and MS Word documents. For more details on
using R Markdown see \url{http://rmarkdown.rstudio.com}.

\textbf{text in bold} \emph{italic} \ul{underline}

\section{Introduction}\label{introduction}

You can add references either by referring to their id in the .bib file
e.g., (Marinković et al., 2019), or by switching to the visual editor
(Cogwheel in the .Rmd menu -\textgreater{} Use Visual Editor). The
awesome paper by (Jokura et al., 2019).

In the visual editor mode, go to `Insert' -\textgreater{} @ Citation

You can select a Zotero library, PubMed, CrossRef etc. and insert the
citations.

The easiest way is to use the command line:

\begin{Shaded}
\begin{Highlighting}[]
\ExtensionTok{curl} \AttributeTok{{-}LH} \StringTok{"Accept: application/x{-}bibtex"}\NormalTok{ https://doi.org/10.7554/eLife.91258.1 }\OperatorTok{\textgreater{}\textgreater{}}\NormalTok{ references.bib}
\end{Highlighting}
\end{Shaded}

\emph{Platynereis dumerilii} is a marine annelid\ldots{} (Ozpolat et
al., 2021)

The references are stored in manuscript/references.bib (need to be
defined in the Yaml header). This file will automatically updated when
you insert a new reference through the Visual editor \textgreater{}
Insert \textgreater{} Citations.

In this documents, references will be formatted in the style of eLife.
This is defined in the Yaml header under: csl: elife.csl. The elife.csl
file is saved in the /manuscript folder.

If you would like to use a different citation format, download the
respective .csl file (e.g., from the Zotero style repository
\url{https://www.zotero.org/styles}), save it in the /manuscript folder
of the project and change the Yaml to csl: your\_favourite\_journal.csl.

\section{Results}\label{results}

Inserting Figures

You can add your figures into the rendered document. We saved the
figures into /manuscript/figures or /manuscript/figure\_supplements and
can insert them from there. We use knitr::include\_graphics for this.
The title and legend can also be edited, as will as the width of the
output figure.

\begin{figure}[H]

{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure1.png}

}

\caption{\textbf{Figure 1. Title fig 1} (A) legend (B) legend.}

\end{figure}%

\section{Equations}\label{equations}

Equations can also be inserted, Insert -\textgreater{} Display Math:

\[
\bar{X} = \frac{\sum_{i=1}^{n} x_{i}}{n}
\]

\section{Sourcing code and working with
variable}\label{sourcing-code-and-working-with-variable}

The `analysis/scripts/statistics\_for\_paper.R' script is sourced and it
runs but the output is not included in the knitted output. But we can
access the variables defined in the sourced script simply by adding ` r
var\_name ` between ` backticks, in this case max\_PRC value is (now
this number comes from our sourced script).

If we update the data, the script can recalculate the variable we want
to refer to in the text and update the number.

\section{Materials and Methods}\label{materials-and-methods}

\section{Acknowledgements}\label{acknowledgements}

We would like to thank the Jekely lab for the R project template
(\url{https://github.com/JekelyLab/new_paper_template}) we used to write
this paper. This work was funded by \ldots{}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Jokura_2019}
Jokura K, Shibata D, Yamaguchi K, Shiba K, Makino Y, Shigenobu S, Inaba
K. 2019. CTENO64 is required for coordinated paddling of ciliary comb
plate in ctenophores. \emph{Current Biology} \textbf{29}:3510--3516.e4.
doi:\href{https://doi.org/10.1016/j.cub.2019.08.059}{10.1016/j.cub.2019.08.059}

\bibitem[\citeproctext]{ref-Marinkovi2020}
Marinković M, Berger J, Jékely G. 2019. Neuronal coordination of motile
cilia in locomotion and feeding. \emph{Philosophical Transactions of the
Royal Society B: Biological Sciences} \textbf{375}:20190165.
doi:\href{https://doi.org/10.1098/rstb.2019.0165}{10.1098/rstb.2019.0165}

\bibitem[\citeproctext]{ref-ozpolat2021}
Ozpolat BD, Randel N, Williams EA, Bezares-Calderón LA, Andreatta G,
Balavoine G, Bertucci PY, Ferrier DEK, Gambi MC, Gazave E,
Handberg-Thorsager M, Hardege J, Hird C, Hsieh Y-W, Hui J, Mutemi KN,
Schneider SQ, Simakov O, Vergara HM, Vervoort M, Jékely G,
Tessmar-Raible K, Raible F, Arendt D. 2021. The Nereid on the rise:
Platynereis as a model system. \emph{Zenodo}.
doi:\href{https://doi.org/10.5281/ZENODO.4907400}{10.5281/ZENODO.4907400}

\end{CSLReferences}

\end{document}

### 2024-02-22 analysis/scripts/Figure1.R

@@ -1,7 +1,7 @@
 # Code to generate Figure 1 of the Jokura et al 2024 Ctenophore apical organ connectome paper
 
 # source packages and functions -----------------
-source("code/packages_and_functions.R")
+source("analysis/scripts/packages_and_functions.R")
 
 # load neurons ----------------------
 
@@ -17,36 +17,36 @@ Q3 <- nlapply(read.neurons.catmaid("Q3", pid = 35, conn = conn_http1),
 Q4 <- nlapply(read.neurons.catmaid("Q4", pid = 35, conn = conn_http1),
               function(x) smooth_neuron(x, sigma = 1000)
               )
-balancer <- nlapply(read.neurons.catmaid("balancer_cell", pid = 35, conn = conn_http1),
+balancer <- nlapply(read.neurons.catmaid("celltype:balancer", pid = 35, conn = conn_http1),
                      function(x) smooth_neuron(x, sigma = 1000)
                     )
-LB <- nlapply(read.neurons.catmaid("lamellate body", pid = 35, conn = conn_http1),
+LB <- nlapply(read.neurons.catmaid("celltype:lamellate", pid = 35, conn = conn_http1),
                     function(x) smooth_neuron(x, sigma = 1000)
                     )
-syn_neuron <- nlapply(read.neurons.catmaid("pre-synaptic", pid = 35, conn = conn_http1),
+syn_neuron <- nlapply(read.neurons.catmaid("celltype:neuron", pid = 35, conn = conn_http1),
                     function(x) smooth_neuron(x, sigma = 1000)
                     )
-monocilia <- nlapply(read.neurons.catmaid("monocilia", pid = 35, conn = conn_http1),
+monocilia <- nlapply(read.neurons.catmaid("monociliated_cell", pid = 35, conn = conn_http1),
                     function(x) smooth_neuron(x, sigma = 1000)
                     )
 
 balancer_Q1 <- nlapply(
-  read.neurons.catmaid(skids_by_2annotations("balancer_cell", "Q1"),
+  read.neurons.catmaid(skids_by_2annotations("celltype:balancer", "Q1"),
                        pid = 35, conn = conn_http1), 
[…]

### 2024-02-22 analysis/scripts/network_graphs.R

@@ -1,6 +1,6 @@
 #network analysis and plotting
 
-source("code/packages_and_functions.R")
+source("analysis/scripts/packages_and_functions.R")
 
 # Create a random graph with 10 nodes
 g <- erdos.renyi.game(10, p = 0.3)

### 2024-02-22 analysis/scripts/packages_and_functions.R

@@ -35,8 +35,8 @@ bluepurple <- brewer.pal(9, 'BuPu')
 oranges <- brewer.pal(9, 'YlOrRd')
 
 #save session info and Rstudio version info for reproducibility
-writeLines(capture.output(sessionInfo()), "code/sessionInfo.txt")
-writeLines(capture.output(rstudioapi::versionInfo()), "code/versionInfo.txt")
+writeLines(capture.output(sessionInfo()), "analysis/scripts/sessionInfo.txt")
+writeLines(capture.output(rstudioapi::versionInfo()), "analysis/scripts/versionInfo.txt")
 
 # functions ----------------------
 

### 2024-02-22 analysis/scripts/contactome.R
contactome: add tag query

@@ -11,7 +11,68 @@ contacts <- catmaid_fetch(
 )
 length(contacts$connectors)
 
-skid1 <- contacts$partners[[1]][[1]][[3]]
+# get connector ids
+connector_ids <- c()
+for(i in 1:length(contacts$connectors)){
+  connector_ids[i] <- contacts$connectors[[i]][[1]]
+  }
+connector_ids
 
+partner1_skid <- c()
+for(i in 1:length(contacts$connectors)){
+  partner1_skid[i] <- contacts$partners[[i]][[1]][[3]]
+}
+partner1_skid
+
+partner2_skid <- c()
+for(i in 1:length(contacts$connectors)){
+  partner2_skid[i] <- contacts$partners[[i]][[2]][[3]]
+}
+partner2_skid
+
+node_id_of_contact_partner1 <- c()
+for(i in 1:length(contacts$connectors)){
+  node_id_of_contact_partner1[i] <- contacts$partners[[i]][[1]][[2]]
+}
+node_id_of_contact_partner1
+
+node_id_of_contact_partner2 <- c()
+for(i in 1:length(contacts$connectors)){
+  node_id_of_contact_partner2[i] <- contacts$partners[[i]][[2]][[2]]
+}
+node_id_of_contact_partner2
+
+
+
+connector_tb <- tibble(
+  connectors = (connector_ids),
+  skid1 = (partner1_skid),
+  skid2 = (partner2_skid)
+  )
+connector_tb
+
+#
+partner1_pre <- c()
+for(i in 1:length(contacts$connectors)){
+  neuron <- read.neuron.catmaid(partner1_skid[i], pid = 35)
+  partner1_pre[i] <- node_id_of_contact_partner1[i] %in% neuron$tags$enter_cil
+}
+partner1_pre
+node_id_of_contact_partner1[10]
+
+partner2_pre <- c()
+for(i in 1:length(contacts$connectors)){
+  neuron <- read.neuron.catmaid(partner2_skid[i], pid = 35)
+  partner2_pre[i] <- node_id_of_contact_partner2[i] %in% neuron$tags$enter_nocil
+}
+partner2_pre
+
+neuron <- read.neuron.catmaid(partner1_skid[1], pid = 35)
+neuron$tags$enter_cil
+node_id_of_contact_partner1[1] %in% neuron$tags$enter_cil
+
+neuron$`2476258`
+
+catmaid_t
 skid2 <- contacts$partners[[1]][[2]][[3]]
 

### 2024-02-22 analysis/scripts/contactome.R
contactome: finish connector tbl

@@ -43,36 +43,43 @@ for(i in 1:length(contacts$connectors)){
 node_id_of_contact_partner2
 
 
-
-connector_tb <- tibble(
-  connectors = (connector_ids),
-  skid1 = (partner1_skid),
-  skid2 = (partner2_skid)
-  )
-connector_tb
-
-#
-partner1_pre <- c()
+#check for tags enter_cil and enter_nocil
+partner1_pre_cil <- c()
+partner1_pre_nocil <- c()
 for(i in 1:length(contacts$connectors)){
   neuron <- read.neuron.catmaid(partner1_skid[i], pid = 35)
-  partner1_pre[i] <- node_id_of_contact_partner1[i] %in% neuron$tags$enter_cil
+  partner1_pre_cil[i] <- node_id_of_contact_partner1[i] %in% neuron$tags$enter_cil
+  partner1_pre_nocil[i] <- node_id_of_contact_partner1[i] %in% neuron$tags$enter_nocil
 }
-partner1_pre
-node_id_of_contact_partner1[10]
+partner1_pre_cil
+partner1_pre_nocil
 
-partner2_pre <- c()
+partner2_pre_cil <- c()
+partner2_pre_nocil <- c()
 for(i in 1:length(contacts$connectors)){
   neuron <- read.neuron.catmaid(partner2_skid[i], pid = 35)
-  partner2_pre[i] <- node_id_of_contact_partner2[i] %in% neuron$tags$enter_nocil
+  partner2_pre_cil[i] <- node_id_of_contact_partner2[i] %in% neuron$tags$enter_cil
+  partner2_pre_nocil[i] <- node_id_of_contact_partner2[i] %in% neuron$tags$enter_nocil
 }
-partner2_pre
+partner2_pre_cil
+partner2_pre_nocil
 
-neuron <- read.neuron.catmaid(partner1_skid[1], pid = 35)
-neuron$tags$enter_cil
-node_id_of_contact_partner1[1] %in% neuron$tags$enter_cil
+# define connectivity tibble
 
-neuron$`2476258`
+connector_tb <- tibble(
+  connectors = connector_ids,
+  skid1 = partner1_skid,
+  skid2 = partner2_skid,
+  partner1_pre_cil = partner1_pre_cil,
+  partner1_pre_nocil = partner1_pre_nocil,
+  partner2_pre_cil = partner2_pre_cil,
[…]

### 2024-03-08 manuscript/Manuscript.qmd
MS: add refs, update bib

@@
 ---
-title: "New paper template - title here"
+title: "Cell-type diversity and connectome of the ctenophore apical organ"
 mainfont: Arial
@@
 max-width: 100%
-number-sections: false
 format:
@@
     pdf-engine: xelatex
-    number-sections: true
+    number-sections: false
     margin-left: 0.8in
@@
     margin-bottom: 0.8in
-    toc: true
+    toc: false
     keep-tex: true
@@
 
-Author 1^1,2,\*^ , Author 2^2^, Author 3^1,2^\
+Kei Jokura^1,3,\*^, Sanja Jasek^1,2^, Lara Kewalow^2^, Pawel Burkhardt^4^, Gáspár Jékely^1,2,\*^\
 
-^1^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany\
-^2^Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom\
+^1^Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom\
+^2^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany\
+^3^National Institute for Basic Biology, Okazaki, ..., Japan
+^4^Sars Centre ... Berge, Norway
 ^\*^Correspondence: corr_author\@email.com
@@
 
-You can add references either by referring to their id in the .bib file e.g., [@Marinkovi2020], or by switching to the visual editor (Cogwheel in the .Rmd menu -\> Use Visual Editor).
-The awesome paper by [@Jokura_2019].
+First descriptions of AO: [@aronova1974electron]
 
-In the visual editor mode, go to 'Insert' -\> \@ Citation
+The lobate ctenophore *Mnemiopsis*
 
-You can select a Zotero library, PubMed, CrossRef etc. and insert the citations.
+The lithocytes have a large calcium carbonate concretion and... [@Tamm_2014]
+Living lithocytes are transported along the balancer cilia to the tip of the statocyst [@Noda_2014]
+The projections of a group of bridge cells located on two sides of the apical organ forms an arch and connects the balancers along the tentacular plane [@Tamm_2002].
 
@@
 
-curl -LH "Accept: application/x-bibtex" https://doi.org/10.7554/eLife.91258.1 >> references.bib
+#curl -LH "Accept: application/x-bibtex" https://doi.org/10.7554/eLife.91258.1 >> references.bib
 ```
 
-*Platynereis dumerilii* is a marine annelid... [@ozpolat2021]
+## Results
 
-The references are stored in manuscript/references.bib (need to be defined in the Yaml header).
-This file will automatically updated when you insert a new reference through the Visual editor \> Insert \> Citations.
+### Volume EM reconstruction of the *Mnemiopsis* apical organ
 
-In this documents, references will be formatted in the style of eLife. This is defined in the Yaml header under: csl: elife.csl. The elife.csl file is saved in the /manuscript folder.
+- specimen 
+- volume, resolution, fixation, quality
+- number of sections
+- skeletonisation in Catmaid
+- number of cells
+- general anatomy, regions, quadrants
 
-If you would like to use a different citation format, download the respective .csl file (e.g., from the Zotero style repository <https://www.zotero.org/styles>), save it in the /manuscript folder of the project and change the Yaml to csl: your_favourite_journal.csl.
+Figure1: (pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions)
 
-## Results
+### Classification of cell types
+
+- ultrastructural criteria (cilia, vesicles etc)
+- number of types, cells per type - table
+
+Figure2 - Balancer complex: balancer, lithocyte, groove 2 types, bridge - EM pics
+
+"the larval bridge has not yet developed forks at its ends, as seen in adult Mnemiopsis bridges"[@Tamm_2002]
+
+Figure3 - dome, bicil, monocil, no-cil, plumose, dense vesicle cells (rename), gland? cells, oligo/multiciliated cells
+
+Plumose cells [@Hernandez_Nicaise_1984] may be pressure sensors [@Tamm_2014b]
+
+Figure4 - lamellate body complex, LB, intra-multi-ciliated, relationship to balancer
+
+Figure5 - nerve net, different fragments/cells/syntytia, EM, relation to entire organ
+
+Figure 6 -  cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.
+
+## Synapses
+
+"synaptic regions with the characteristic presynaptic triad mor-
+phology of ctenophore nerves (i.e., a single layer of
+vesicles, smooth ER sac, and closely apposed mito-
+chondria; Horridge and Mackay, 1964; Hernandez-
+Nicaise, 1973)"
+
+Figure 7 - synapses, reconstruction, annotation, numbers, distribution of partners, example EM (figure supplement - lots of EM), mitochondria
+(Fig suppl - mitochondrion stats - Sanja)
+
+Gap juntions? Innexins are expressed in the apical organ in specific patches of cells in each quadrant [@Ortiz_2023]
+
+Ctenophore synapses [@Hernandez_Nicaise_1973]
+
+We found no synapses between bridge and balancer, in agreement with Tamm [@Tamm_2002].
+
+Figure 8 - connectome
+cell-type based, quadrant based, nerve net syntycia, bridge, quadrants
+(Fig suppl - cell-based network)
+
+Tamm&Tamm observed synapses from neurites to bridge cells (2002).
+and in Mnemiopsis larvae observed synapses from bridge cells onto other cells
 
@@
 We would like to thank the Jekely lab for the R project template (<https://github.com/JekelyLab/new_paper_template>) we used to write this paper. This work was funded by ...
+JSPS...
+ERC..
+Others ...
 

### 2024-03-08 analysis/scripts/Figure_balancer.R
Figure_balancer: start new fig

@@ -121,9 +121,15 @@ close3d()
 
 # crop EM pictures ---------------
 
-crop_catmaid("nice LB", 5000, 5000, 1, 0, "manuscript/pictures/")
+#crop_catmaid("nice LB", 5000, 5000, 1, 0, "manuscript/pictures/")
 
 #list all files starting with crop
 system("ls ./manuscript/pictures/crop_*")
 
-LB_EM1 <- ggdraw() + draw_image(magick::image_read("manuscript/pictures/crop_6l53xv.tiff"))
+# assemble figure --------------------
+
+LB_EM1 <- ggdraw() + draw_image(magick::image_read("manuscript/pictures/crop_3yr8pe_8.4um.tif"))
+
+
+
+

### 2024-03-10 manuscript/Manuscript.qmd
Ms: add ref

@@
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
-
 **text in bold** *italic* [underline]{.underline}
@@
 
-The easiest way is to use the command line:
+The easiest way is to use the command line to add new refs to the .bib file:
 
@@
 vesicles, smooth ER sac, and closely apposed mito-
-chondria; Horridge and Mackay, 1964; Hernandez-
+chondria [@Horridge_1964; @Hernandez_Nicaise_1973] Hernandez-
 Nicaise, 1973)"

### 2024-03-12 manuscript/Manuscript.qmd
Figure1

@@
 
-Figure1: (pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions)
+```{r, out.width = "100%", fig.cap="**Figure 1. Title fig 1** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
+#| echo: false
+
+knitr::include_graphics("figures/Figure1.png")
+```
 
@@
 
-Figure2 - Balancer complex: balancer, lithocyte, groove 2 types, bridge - EM pics
+```{r, out.width = "100%", fig.cap="**Figure 2. Title fig 2** Balancer complex: balancer, lithocyte, groove 2 types, bridge - EM pics, (A) legend (B) legend. " }
+#| echo: false
+
+knitr::include_graphics("figures/Figure2.png")
+```
 
@@
 
-```{r, out.width = "100%", fig.cap="**Figure 1. Title fig 1** (A) legend (B) legend. " }
-#| echo: false
-
-knitr::include_graphics("figures/Figure1.png")
-```
-
 ## Equations

### 2024-04-11 manuscript/Manuscript.qmd
add commenting functionality to html output with hypothes.is
Fig: celltype network

@@
 format:
-  html: default
+  html:
+    embed-resources: true 
   pdf:
@@
 
-knitr::include_graphics("figures/Figure2.png")
+knitr::include_graphics("figures/Figure_network.png")
 ```
@@
 ## References
+
+<!--
+## See https://github.com/hypothesis/client
+## Add commenting functionality with this script
+-->
+
+<script src="hypothes.is/embed.js" async></script>

### 2024-04-11 analysis/scripts/Figure_celltype_network.R

@@ -464,46 +464,46 @@ webshot2::webshot(url = "manuscript/pictures/visNetwork_celltype_connectome.html
 # assemble figure ---------------------------------------------------------
 
 panel_mono_oral <- ggdraw() + draw_image(readPNG("manuscript/pictures/mono_syn_ant.png")) +
-  draw_label("oral view", x = 0.25, y = 0.95, size = 10) +
-  draw_label("monociliated", x = 0.7, y = 0.95, size = 10)
+  draw_label("oral view", x = 0.75, y = 0.95, size = 10) +
+  draw_label("monociliated", x = 0.1, y = 0.95, size = 10, hjust = 0)
 panel_mono_side <- ggdraw() + draw_image(readPNG("manuscript/pictures/mono_syn_left.png")) +
-  draw_label("side view", x = 0.25, y = 0.95, size = 10)
+  draw_label("side view", x = 0.75, y = 0.95, size = 10)
 
 panel_bridge_oral <- ggdraw() + draw_image(readPNG("manuscript/pictures/bridge_syn_ant.png")) +
-  draw_label("oral view", x = 0.25, y = 0.95, size = 10) +
-  draw_label("bridge", x = 0.7, y = 0.95, size = 10)
+  draw_label("oral view", x = 0.75, y = 0.95, size = 10) +
+  draw_label("bridge", x = 0.1, y = 0.95, size = 10, hjust = 0)
 panel_bridge_side <- ggdraw() + draw_image(readPNG("manuscript/pictures/bridge_syn_left.png")) +
-  draw_label("side view", x = 0.25, y = 0.95, size = 10)
+  draw_label("side view", x = 0.75, y = 0.95, size = 10)
 
 panel_bal_oral <- ggdraw() + draw_image(readPNG("manuscript/pictures/bal_syn_ant.png")) +
-  draw_label("oral view", x = 0.25, y = 0.95, size = 10) +
-  draw_label("balancer", x = 0.7, y = 0.95, size = 10)
+  draw_label("oral view", x = 0.75, y = 0.95, size = 10) +
+  draw_label("balancer", x = 0.1, y = 0.95, size = 10, hjust = 0)
 panel_bal_side <- ggdraw() + draw_image(readPNG("manuscript/pictures/bal_syn_left.png")) +
-  draw_label("side view", x = 0.25, y = 0.95, size = 10)
+  draw_label("side view", x = 0.75, y = 0.95, size = 10)
 
[…]

### 2024-09-18 manuscript/Manuscript.tex
affiliation

@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % Make \paragraph and \subparagraph free-standing
@@
 \newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
@@
 \newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
 \newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
@@
 \hypersetup{
-  pdftitle={New paper template - title here},
+  pdftitle={Cell-type diversity and connectome of the ctenophore apical organ},
   colorlinks=true,
@@
 
-\title{New paper template - title here}
+\title{Cell-type diversity and connectome of the ctenophore apical
+organ}
 \author{}
@@
 
-\renewcommand*\contentsname{Table of contents}
-{
-\hypersetup{linkcolor=}
-\setcounter{tocdepth}{3}
-\tableofcontents
-}
 \setstretch{1.2}
@@
 
-Author 1\textsuperscript{1,2,*} , Author 2\textsuperscript{2}, Author
-3\textsuperscript{1,2}\\
+Kei Jokura\textsuperscript{1,3,*}, Sanja Jasek\textsuperscript{1,2},
+Lara Kewalow\textsuperscript{2}, Pawel Burkhardt\textsuperscript{4},
+Gáspár Jékely\textsuperscript{1,2,*}\\
 
-\textsuperscript{1}Heidelberg University, Centre for Organismal Studies
+\textsuperscript{1}Living Systems Institute, University of Exeter,
+Exeter, EX4 4QD, United Kingdom\\
+\textsuperscript{2}Heidelberg University, Centre for Organismal Studies
 (COS), 69120 Heidelberg, Germany\\
-\textsuperscript{2}Living Systems Institute, University of Exeter,
-Exeter, EX4 4QD, United Kingdom\\
-\textsuperscript{*}Correspondence: corr\_author@email.com
+\textsuperscript{3}Exploratory Research Center on Life and Living
+Systems (ExCELLS), National Institutes of Natural Sciences, Okazaki,
+Japan\\
+\textsuperscript{4}Michael Sars Centre, University of Bergen, Norway\\
+\textsuperscript{*}Correspondence: jokura@nibb.ac.jp,
+gaspar.jekely@cos.uni-heidelberg.de
 
-\section{Abstract}\label{abstract}
-
-This is an R Markdown document. Markdown is a simple formatting syntax
-for authoring HTML, PDF, and MS Word documents. For more details on
-using R Markdown see \url{http://rmarkdown.rstudio.com}.
+\subsection{Abstract}\label{abstract}
 
@@
 
-\section{Introduction}\label{introduction}
+\subsection{Introduction}\label{introduction}
 
-You can add references either by referring to their id in the .bib file
-e.g., (Marinković et al., 2019), or by switching to the visual editor
-(Cogwheel in the .Rmd menu -\textgreater{} Use Visual Editor). The
-awesome paper by (Jokura et al., 2019).
+First descriptions of AO: (Aronova, 1974)
 
-In the visual editor mode, go to `Insert' -\textgreater{} @ Citation
+The lobate ctenophore \emph{Mnemiopsis}
 
-You can select a Zotero library, PubMed, CrossRef etc. and insert the
-citations.
+The lithocytes have a large calcium carbonate concretion and\ldots{}
+(Tamm, 2014a) Living lithocytes are transported along the balancer cilia
+to the tip of the statocyst (Noda and Tamm, 2014) The projections of a
+group of bridge cells located on two sides of the apical organ forms an
+arch and connects the balancers along the tentacular plane (Tamm and
+Tamm, 2002).
 
-The easiest way is to use the command line:
+The easiest way is to use the command line to add new refs to the .bib
+file:
 
@@
 \begin{Highlighting}[]
-\ExtensionTok{curl} \AttributeTok{{-}LH} \StringTok{"Accept: application/x{-}bibtex"}\NormalTok{ https://doi.org/10.7554/eLife.91258.1 }\OperatorTok{\textgreater{}\textgreater{}}\NormalTok{ references.bib}
+\CommentTok{\#curl {-}LH "Accept: application/x{-}bibtex" https://doi.org/10.7554/eLife.91258.1 \textgreater{}\textgreater{} references.bib}
 \end{Highlighting}
@@
 
-\emph{Platynereis dumerilii} is a marine annelid\ldots{} (Ozpolat et
-al., 2021)
+\subsection{Results}\label{results}
 
-The references are stored in manuscript/references.bib (need to be
-defined in the Yaml header). This file will automatically updated when
-you insert a new reference through the Visual editor \textgreater{}
-Insert \textgreater{} Citations.
+\subsubsection{\texorpdfstring{Volume EM reconstruction of the
+\emph{Mnemiopsis} apical
+organ}{Volume EM reconstruction of the Mnemiopsis apical organ}}\label{volume-em-reconstruction-of-the-mnemiopsis-apical-organ}
 
-In this documents, references will be formatted in the style of eLife.
-This is defined in the Yaml header under: csl: elife.csl. The elife.csl
-file is saved in the /manuscript folder.
+\begin{itemize}
+\tightlist
+\item
+  specimen
+\item
+  volume, resolution, fixation, quality
+\item
+  number of sections
+\item
+  skeletonisation in Catmaid
+\item
+  number of cells
+\item
+  general anatomy, regions, quadrants
+\end{itemize}
 
-If you would like to use a different citation format, download the
-respective .csl file (e.g., from the Zotero style repository
-\url{https://www.zotero.org/styles}), save it in the /manuscript folder
-of the project and change the Yaml to csl: your\_favourite\_journal.csl.
+\begin{figure}[H]
 
-\section{Results}\label{results}
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure1.png}
+
+}
+
+\caption{\textbf{Figure 1. Title fig 1} pic of a larva, boxed region of
+AO, schematic, catmaid pic with slice, dimensions, quadrants, regions,
+(A) legend (B) legend.}
+
+\end{figure}%
+
+\subsubsection{Classification of cell
+types}\label{classification-of-cell-types}
+
+\begin{itemize}
+\tightlist
+\item
+  ultrastructural criteria (cilia, vesicles etc)
+\item
+  number of types, cells per type - table
+\end{itemize}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure_network.png}
+
+}
+
+\caption{\textbf{Figure 2. Title fig 2} Balancer complex: balancer,
+lithocyte, groove 2 types, bridge - EM pics, (A) legend (B) legend.}
+
+\end{figure}%
+
+``the larval bridge has not yet developed forks at its ends, as seen in
+adult Mnemiopsis bridges''(Tamm and Tamm, 2002)
+
+Figure3 - dome, bicil, monocil, no-cil, plumose, dense vesicle cells
+(rename), gland? cells, oligo/multiciliated cells
+
+Plumose cells (Hernandez-Nicaise, 1984) may be pressure sensors (Tamm,
+2014b)
+
+Figure4 - lamellate body complex, LB, intra-multi-ciliated, relationship
+to balancer
+
+Figure5 - nerve net, different fragments/cells/syntytia, EM, relation to
+entire organ
+
+Figure 6 - cilia? renderings of basal bodies, centrioles, length,
+axoneme structure (table), table about cell types, number of cilia etc.
+
+\subsection{Synapses}\label{synapses}
+
+``synaptic regions with the characteristic presynaptic triad mor-
+phology of ctenophore nerves (i.e., a single layer of vesicles, smooth
+ER sac, and closely apposed mito- chondria (Hernandez-Nicaise, 1973;
+Horridge and Mackay, 1964) Hernandez- Nicaise, 1973)''
+
+Figure 7 - synapses, reconstruction, annotation, numbers, distribution
+of partners, example EM (figure supplement - lots of EM), mitochondria
+(Fig suppl - mitochondrion stats - Sanja)
+
+Gap juntions? Innexins are expressed in the apical organ in specific
+patches of cells in each quadrant (Ortiz et al., 2023)
+
+Ctenophore synapses (Hernandez-Nicaise, 1973)
+
+We found no synapses between bridge and balancer, in agreement with Tamm
+(Tamm and Tamm, 2002).
+
+Figure 8 - connectome cell-type based, quadrant based, nerve net
+syntycia, bridge, quadrants (Fig suppl - cell-based network)
+
+Tamm\&Tamm observed synapses from neurites to bridge cells (2002). and
+in Mnemiopsis larvae observed synapses from bridge cells onto other
+cells
 
@@
 
-\begin{figure}[H]
-
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure1.png}
-
-}
-
-\caption{\textbf{Figure 1. Title fig 1} (A) legend (B) legend.}
-
-\end{figure}%
-
-\section{Equations}\label{equations}
+\subsection{Equations}\label{equations}
 
@@
 
-\section{Sourcing code and working with
+\subsection{Sourcing code and working with
 variable}\label{sourcing-code-and-working-with-variable}
@@
 
-\section{Materials and Methods}\label{materials-and-methods}
+\subsection{Materials and Methods}\label{materials-and-methods}
 
-\section{Acknowledgements}\label{acknowledgements}
+\subsection{Acknowledgements}\label{acknowledgements}
 
@@
 (\url{https://github.com/JekelyLab/new_paper_template}) we used to write
-this paper. This work was funded by \ldots{}
+this paper. This work was funded by \ldots{} JSPS\ldots{} ERC.. Others
+\ldots{}
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
+\subsection{References}\label{references}
 
@@
 \begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-Jokura_2019}
-Jokura K, Shibata D, Yamaguchi K, Shiba K, Makino Y, Shigenobu S, Inaba
-K. 2019. CTENO64 is required for coordinated paddling of ciliary comb
-plate in ctenophores. \emph{Current Biology} \textbf{29}:3510--3516.e4.
-doi:\href{https://doi.org/10.1016/j.cub.2019.08.059}{10.1016/j.cub.2019.08.059}
+\bibitem[\citeproctext]{ref-aronova1974electron}
+Aronova M. 1974. Electron microscopic observation of the aboral organ of
+ctenophora. I. The gravity receptor. \emph{Zeitschrift fur
+mikroskopisch-anatomische Forschung} \textbf{88}:401--412.
 
-\bibitem[\citeproctext]{ref-Marinkovi2020}
-Marinković M, Berger J, Jékely G. 2019. Neuronal coordination of motile
-cilia in locomotion and feeding. \emph{Philosophical Transactions of the
-Royal Society B: Biological Sciences} \textbf{375}:20190165.
-doi:\href{https://doi.org/10.1098/rstb.2019.0165}{10.1098/rstb.2019.0165}
+\bibitem[\citeproctext]{ref-Hernandez_Nicaise_1984}
+Hernandez-Nicaise M-L. 1984. CtenophoraBiology of the Integument.
+Springer Berlin Heidelberg. pp. 96--111.
+doi:\href{https://doi.org/10.1007/978-3-642-51593-4_9}{10.1007/978-3-642-51593-4\_9}
 
-\bibitem[\citeproctext]{ref-ozpolat2021}
-Ozpolat BD, Randel N, Williams EA, Bezares-Calderón LA, Andreatta G,
-Balavoine G, Bertucci PY, Ferrier DEK, Gambi MC, Gazave E,
-Handberg-Thorsager M, Hardege J, Hird C, Hsieh Y-W, Hui J, Mutemi KN,
-Schneider SQ, Simakov O, Vergara HM, Vervoort M, Jékely G,
-Tessmar-Raible K, Raible F, Arendt D. 2021. The Nereid on the rise:
-Platynereis as a model system. \emph{Zenodo}.
-doi:\href{https://doi.org/10.5281/ZENODO.4907400}{10.5281/ZENODO.4907400}
+\bibitem[\citeproctext]{ref-Hernandez_Nicaise_1973}
+Hernandez-Nicaise M-L. 1973. The nervous system of ctenophores III.
+Ultrastructure of synapses. \emph{Journal of Neurocytology}
+\textbf{2}:249--263.
+doi:\href{https://doi.org/10.1007/bf01104029}{10.1007/bf01104029}
+
+\bibitem[\citeproctext]{ref-Horridge_1964}
+Horridge GA, Mackay B. 1964. Neurociliary synapses in pleurobrachia
+(ctenophora). \emph{Journal of Cell Science} \textbf{S3-105}:163--174.
+doi:\href{https://doi.org/10.1242/jcs.s3-105.70.163}{10.1242/jcs.s3-105.70.163}
+
+\bibitem[\citeproctext]{ref-Noda_2014}
+Noda N, Tamm SL. 2014. Lithocytes are transported along the ciliary
+surface to build the statolith of ctenophores. \emph{Current Biology}
+\textbf{24}:R951--R952.
+doi:\href{https://doi.org/10.1016/j.cub.2014.08.045}{10.1016/j.cub.2014.08.045}
+
+\bibitem[\citeproctext]{ref-Ortiz_2023}
+Ortiz J, Bobkov YV, DeBiasse MB, Mitchell DG, Edgar A, Martindale MQ,
+Moss AG, Babonis LS, Ryan JF. 2023. Independent innexin radiation shaped
+signaling in ctenophores. \emph{Molecular Biology and Evolution}
+\textbf{40}.
+doi:\href{https://doi.org/10.1093/molbev/msad025}{10.1093/molbev/msad025}
+
+\bibitem[\citeproctext]{ref-Tamm_2014}
+Tamm SL. 2014a. Formation of the statolith in the ctenophore mnemiopsis
+leidyi. \emph{The Biological Bulletin} \textbf{227}:7--18.
+doi:\href{https://doi.org/10.1086/bblv227n1p7}{10.1086/bblv227n1p7}
+
+\bibitem[\citeproctext]{ref-Tamm_2014b}
+Tamm SL. 2014b. Cilia and the life of ctenophores. \emph{Invertebrate
+Biology} \textbf{133}:1--46.
+doi:\href{https://doi.org/10.1111/ivb.12042}{10.1111/ivb.12042}
+
+\bibitem[\citeproctext]{ref-Tamm_2002}
+Tamm SL, Tamm S. 2002. Novel bridge of axon‐like processes of epithelial
+cells in the aboral sense organ of ctenophores. \emph{Journal of
+Morphology} \textbf{254}:99--120.
+doi:\href{https://doi.org/10.1002/jmor.10019}{10.1002/jmor.10019}
 

### 2024-11-29 analysis/scripts/Figure1.R
Figure1

@@ -3,298 +3,35 @@
 # source packages and functions -----------------
 source("analysis/scripts/packages_and_functions.R")
 
-# load cell type ----------------------------------
-
-read_smooth_neuron <- function(annotation){
-  nlapply(read.neurons.catmaid(annotation, pid = 35),
-          function(x)
-            smooth_neuron(x, sigma = 1000))
-}
-
-balancer <- read_smooth_neuron("celltype:balancer")
-bridge <- read_smooth_neuron("celltype:bridge")
-bristle <- read_smooth_neuron("celltype:bristle")
-dome <- read_smooth_neuron("celltype:dome")
-Cgroove_tag <- read_smooth_cell("celltype:Cgroove-tag")
-Cgroove_sag <- read_smooth_cell("celltype:Cgroove-sag")
-intramulticilia <- read_smooth_neuron("celltype:intra-multi-ciliated")
-lamellate <- read_smooth_neuron("celltype:lamellate")
-lithocyte <- read_smooth_neuron("celltype:lithocyte")
-neuron <- read_smooth_neuron("celltype:neuron")
-plumose <- read_smooth_neuron("celltype:plumose")
-dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
-monocilia <- read_smooth_neuron("celltype:monociliated")
-bicilia <- read_smooth_neuron("celltype:biciliated")
-non_cilia <- read_smooth_neuron("celltype:nonciliated")
-
-
-Q1 <- nlapply(read.neurons.catmaid("Q1", pid = 35),
-              function(x)
-                smooth_neuron(x, sigma = 1000))
-Q2 <- nlapply(read.neurons.catmaid("Q2", pid = 35),
-              function(x)
-                smooth_neuron(x, sigma = 1000))
-Q3 <- nlapply(read.neurons.catmaid("Q3", pid = 35),
-              function(x)
-                smooth_neuron(x, sigma = 1000))
-Q4 <- nlapply(read.neurons.catmaid("Q4", pid = 35),
-              function(x)
-                smooth_neuron(x, sigma = 1000))
-
-
-balancer_Q1 <- nlapply(read.neurons.catmaid(get_skids_with_annot(pid = 35, c("celltype:balancer", "Q1")),
-                                            pid = 35),
-                       function(x)
[…]

### 2024-11-29 analysis/scripts/all_celltype_plot.R

@@ -1,0 +1,203 @@
+# Code to generate all cells
+
+# source packages and functions -----------------
+source("analysis/scripts/packages_and_functions.R")
+
+# load cell type ----------------------------------
+
+read_smooth_neuron <- function(annotation){
+  nlapply(read.neurons.catmaid(annotation, pid = 35),
+          function(x)
+            smooth_neuron(x, sigma = 1000))
+}
+
+balancer <- read_smooth_neuron("celltype:balancer")
+bridge <- read_smooth_neuron("celltype:bridge")
+bristle <- read_smooth_neuron("celltype:bristle")
+dome <- read_smooth_neuron("celltype:dome")
+Cgroove_tag <- read_smooth_neuron("celltype:Cgroove-tag")
+Cgroove_sag <- read_smooth_neuron("celltype:Cgroove-sag")
+intramulticilia <- read_smooth_neuron("celltype:intra-multi-ciliated")
+lamellate <- read_smooth_neuron("celltype:lamellate")
+lithocyte <- read_smooth_neuron("celltype:lithocyte")
+neuron <- read_smooth_neuron("celltype:neuron")
+plumose <- read_smooth_neuron("celltype:plumose")
+dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
+monocilia <- read_smooth_neuron("celltype:monociliated")
+bicilia <- read_smooth_neuron("celltype:biciliated")
+non_cilia <- read_smooth_neuron("celltype:nonciliated")
+
+
+Q1 <- nlapply(read.neurons.catmaid("Q1", pid = 35),
+              function(x)
+                smooth_neuron(x, sigma = 1000))
+Q2 <- nlapply(read.neurons.catmaid("Q2", pid = 35),
+              function(x)
+                smooth_neuron(x, sigma = 1000))
+Q3 <- nlapply(read.neurons.catmaid("Q3", pid = 35),
+              function(x)
+                smooth_neuron(x, sigma = 1000))
+Q4 <- nlapply(read.neurons.catmaid("Q4", pid = 35),
+              function(x)
+                smooth_neuron(x, sigma = 1000))
+
+
+balancer_Q1 <- nlapply(read.neurons.catmaid(get_skids_with_annot(pid = 35, c("celltype:balancer", "Q1")),
+                                            pid = 35),
+                       function(x)
[…]

### 2024-11-29 analysis/scripts/packages_and_functions.R

@@ -16,6 +16,8 @@ library(igraph)
 library(tidygraph)
 library(av)
 
+
+
 # catmaid connection, needs username, password AND token - weird!
 {
   # can run this separate file using source function
@@ -203,3 +205,4 @@ get_celltypes <- function(pid) {
     select(name) |> pull() |> sub(pattern="^celltype:", replacement="")
   return(celltypes)
 }
+

### 2024-12-02 analysis/scripts/Figure1.R
Figure2

@@ -32,6 +32,6 @@ ggsave("manuscript/figures/Figure1.png", limitsize = FALSE,
        units = c("px"), Figure1, width = 2600, height = 1000, bg='white')  
 
 
-ggsave("manuscript/figures/Fig1.pdf", limitsize = FALSE, 
+ggsave("manuscript/figures/Figure1.pdf", limitsize = FALSE, 
        units = c("px"), Figure1, width = 2600, height = 1000) 
 

### 2024-12-02 analysis/scripts/Figure2.R

@@ -3,448 +3,35 @@
 # source packages and functions ------------------------------------------------
 source("analysis/scripts/packages_and_functions.R")
 
-# load cell types from catmaid -------------------------------------------------
 
-read_smooth_cell <- function(annotation){
-  nlapply(read.neurons.catmaid(annotation, pid = 35),
-          function(x)
-            smooth_neuron(x, sigma = 1000))
-}
+# assemble figure -------------------------------------------------------------
 
 
-bridge_Q1Q2 <-
-  nlapply(read.neurons.catmaid(get_skids_with_annot(pid = 35, c("celltype:bridge", "Q1Q2")),
-                               pid = 35),
-          function(x)
-            smooth_neuron(x, sigma = 1000))
+panel_synapse <- ggdraw() + draw_image(readPNG("manuscript/pictures/Figure_mito_syn_ves_syn.png"))
+panel_SSN_Q1234 <- ggdraw() + draw_image(readPNG("manuscript/pictures/SSN_Q1234_panel.png"))
+panel_SSN_Q12_Q34 <- ggdraw() + draw_image(readPNG("manuscript/pictures/SSN_Q1Q2_Q3Q4_panel.png"))
+panel_mito <- ggdraw() + draw_image(readPNG("manuscript/pictures/mito_pos_SSN_anterior_trim.png"))
 
-bridge_Q3Q4 <-
-  nlapply(read.neurons.catmaid(get_skids_with_annot(pid = 35, c("celltype:bridge", "Q3Q4")),
-                               pid = 35),
-          function(x)
-            smooth_neuron(x, sigma = 1000))
+layout <- "
+AA#BBB
+######
+DD#CCC
+"
 
+Figure2 <- panel_synapse + 
+  panel_SSN_Q1234 + panel_SSN_Q12_Q34 +
+  panel_mito +
+  plot_layout(design = layout,
+              heights = c(1, 0.1, 1),
+              widths = c(1, 1, 0.1, 1, 1, 1)) + 
+  plot_annotation(tag_levels = "A") + 
+  theme(plot.tag = element_text(size = 12, face='plain', color='black'))
 
-balancer_Q1 <-
-  nlapply(read.neurons.catmaid(get_skids_with_annot(pid = 35, c("celltype:balancer", "Q2")),
-                               pid = 35),
-          function(x)
-            smooth_neuron(x, sigma = 1000))
-balancer_Q2 <-
[…]

### 2024-12-04 analysis/scripts/Figure3.R
Figure3

@@ -1,0 +1,36 @@
+# Code to generate Figure 3 of the Jokura et al 2024 Ctenophore apical organ connectome paper
+
+# source packages and functions ------------------------------------------------
+source("analysis/scripts/packages_and_functions.R")
+
+
+# assemble figure -------------------------------------------------------------
+
+
+panel_AO_matrix <- ggdraw() + draw_image(readPNG("manuscript/pictures/AO_neurons_matrix.png"))
+panel_bal_3d <- ggdraw() + draw_image(readPNG("manuscript/pictures/balancer_text.png"))
+panel_bal_syn <- ggdraw() + draw_image(readPNG("manuscript/pictures/bal_syn.png"))
+panel_grav_nc_graph <- ggdraw() + draw_image(readPNG("manuscript/pictures/gravity_neural_circuit_graph.png"))
+
+layout <- "
+AA#BBBBB
+########
+CCCCC#DD
+"
+
+Figure3 <- panel_AO_matrix + panel_bal_3d + 
+  panel_bal_syn + panel_grav_nc_graph +
+  plot_layout(design = layout,
+              heights = c(1, 0.1, 1),
+              widths = c(1, 1, 0.1, 1, 1, 0.1, 1, 1)) + 
+  plot_annotation(tag_levels = "A") + 
+  theme(plot.tag = element_text(size = 12, face='plain', color='black'))
+
+
+ggsave("manuscript/figures/Figure3.png", limitsize = FALSE, 
+       units = c("px"), Figure3, width = 2000, height = 1000, bg='white')  
+
+
+ggsave("manuscript/figures/Figure3.pdf", limitsize = FALSE, 
+       units = c("px"), Figure3, width = 2000, height = 1000) 
+

### 2024-12-06 analysis/scripts/Figure4.R
Figure4

@@ -1,0 +1,37 @@
+# Code to generate Figure 4 of the Jokura et al 2024 Ctenophore apical organ connectome paper
+
+# source packages and functions ------------------------------------------------
+source("analysis/scripts/packages_and_functions.R")
+
+
+# assemble figure -------------------------------------------------------------
+
+
+panel_bri_3d <- ggdraw() + draw_image(readPNG("manuscript/pictures/bridge_text.png"))
+panel_bri_mito <- ggdraw() + draw_image(readPNG("manuscript/pictures/mito_pos_bridge_text.png"))
+panel_bri_syn <- ggdraw() + draw_image(readPNG("manuscript/pictures/bridge_syn_text.png"))
+panel_grav_matrix <- ggdraw() + draw_image(readPNG("manuscript/pictures/mech_girdle_chaeMech_syn_matrix.png"))
+panel_grav_graph_bridge <- ggdraw() + draw_image(readPNG("manuscript/pictures/gravity_neural_circuit_graph_bridge.png"))
+
+layout <- "
+AAAABBB
+#######
+CCCDDEE
+"
+
+Figure4 <- panel_bri_3d + panel_bri_mito + 
+  panel_bri_syn + panel_grav_matrix + panel_grav_graph_bridge +
+  plot_layout(design = layout,
+              heights = c(1, 0.1, 1.2),
+              widths = c(1, 1, 1, 1, 1, 1, 0.5)) + 
+  plot_annotation(tag_levels = "A") + 
+  theme(plot.tag = element_text(size = 12, face='plain', color='black'))
+
+
+ggsave("manuscript/figures/Figure4.png", limitsize = FALSE, 
+       units = c("px"), Figure4, width = 2400, height = 1000, bg='white')  
+
+
+ggsave("manuscript/figures/Figure4.pdf", limitsize = FALSE, 
+       units = c("px"), Figure4, width = 2400, height = 1000) 
+

### 2024-12-17 manuscript/Manuscript.tex
Figure5

@@
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
 
@@
 \newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
 \newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
@@
 \newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
 \newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
@@
 \makeatother
+
 \ifLuaTeX
@@
 
-Kei Jokura\textsuperscript{1,3,*}, Sanja Jasek\textsuperscript{1,2},
-Lara Kewalow\textsuperscript{2}, Pawel Burkhardt\textsuperscript{4},
+Kei Jokura\textsuperscript{1,3,4,*}, Sanja Jasek\textsuperscript{1,2},
+Lara Kewalow\textsuperscript{2}, Pawel Burkhardt\textsuperscript{5},
 Gáspár Jékely\textsuperscript{1,2,*}\\
@@
 Japan\\
-\textsuperscript{4}Michael Sars Centre, University of Bergen, Norway\\
+\textsuperscript{4}National Institute for Basic Biology (NIBB), National
+Institutes of Natural Sciences, Okazaki, Japan\\
+\textsuperscript{5}Michael Sars Centre, University of Bergen, Norway\\
 \textsuperscript{*}Correspondence: jokura@nibb.ac.jp,
@@
 The lithocytes have a large calcium carbonate concretion and\ldots{}
-(Tamm, 2014a) Living lithocytes are transported along the balancer cilia
+(Tamm, 2014) Living lithocytes are transported along the balancer cilia
 to the tip of the statocyst (Noda and Tamm, 2014) The projections of a
@@
 
-\begin{itemize}
-\tightlist
-\item
-  specimen
-\item
-  volume, resolution, fixation, quality
-\item
-  number of sections
-\item
-  skeletonisation in Catmaid
-\item
-  number of cells
-\item
-  general anatomy, regions, quadrants
-\end{itemize}
+First, we embedded the entire body of the \emph{Mnemiopsis leidyi}
+cydippid-stage larvae, five days post-fertilization, in Epon resin
+following high-pressure freezing. To reconstruct all synaptic-resolution
+connections among the cells comprising the apical organ, also known as
+the aboral organ, we prepared approximately 1,000 ultra-thin serial
+sections (50 nm thick) from the aboral tip of the larval body embedded
+in the resin block. Using a scanning electron microscope (Gemini SEM
+500) with a resolution of 2.0 nm/pixel, we imaged only the region
+containing the apical organ. Subsequently, 620 of these images were
+stitched and aligned using TrakEM2. The resulting volumetric EM dataset
+had dimensions of 60 μm × 40 μm × 30 μm. We traced and annotated all
+cells in this dataset, ultimately reconstructing 909 cells, each with
+both nuclei and cell bodies intact.
+
+Cells containing cilia were traced down to the tips of the cilia, and
+basal bodies were annotated accordingly. For neuronal skeletonization,
+nodes were placed to interconnect the profiles of the same neuron's
+processes across layers, extending the skeleton until all branches were
+fully traced. Each node was tagged, and skeletons were named and
+assigned multi-level annotations. As described later, many neurons we
+identified formed loop-like structures (anastomosed neurons), wherein
+separated branches often rejoined either the main trunk or other
+branches. In such cases, branch nodes were placed near the closest
+existing node and annotated accordingly. The entire skeletonized volume
+was composed of 134,591 nodes. However, 88 fragments could not be
+attached to somata-associated skeletons. Most of these fragments
+represented short skeletal branches that could not be traced beyond gaps
+or low-quality layers.
+
+Next, we decided to divide the entire apical organ into four broad
+quadrants to facilitate grouping the identified cells. The general body
+plan of ctenophores, when viewed from the aboral side, exhibits biradial
+symmetry around the positions of the anal pores. This symmetry
+corresponds to the four blastomeres present at the four-cell stage
+during early embryonic development. We categorized the traced cells into
+four quadrants and designated these groups as Q1 through Q4 for clarity
+and consistency.
 
@@
 
-\subsubsection{Classification of cell
-types}\label{classification-of-cell-types}
+\subsubsection{Identification of Synaptic Structures in Syncytium
+Neurons}\label{identification-of-synaptic-structures-in-syncytium-neurons}
 
-\begin{itemize}
-\tightlist
-\item
-  ultrastructural criteria (cilia, vesicles etc)
-\item
-  number of types, cells per type - table
-\end{itemize}
+Classical neural staining techniques do not provide clear images of the
+neurons at the aboral pole. However, ultrastructural studies have
+provided morphological evidence of elements resembling neurons, based on
+synaptic structures, located on the epithelial floor of the apical
+organs (Hernandez-Nicaise, 1973; Horridge and Mackay, 1964). Based on
+these findings, we identified synaptic structures characteristic of
+ctenophore neurons in our data, using previously identified pre-synaptic
+triad morphological features, such as single-layered vesicles, smooth
+endoplasmic reticulum, and tightly packed mitochondria. In our study, we
+specifically identified synaptic sites and marked mitochondria (orange)
+as synaptic nodes. The regions where synaptic vesicles align were marked
+as connectors (light blue arrows) between cells across the membrane.
+These connectors link the synaptic nodes of the pre-synaptic
+cytoskeleton to the partner nodes of the post-synaptic cytoskeleton. As
+previously reported, the specialization of post-synaptic structures in
+ctenophores is not apparent, so we recognized synaptic vesicle clusters
+on the pre-synaptic membrane as the point of reference. Synapses were
+identified as either monoadic or polyadic, with one pre-synaptic neuron
+forming connections with one or multiple post-synaptic cells.
+
+As mentioned above, following the identification of the presynaptic
+structures, we reconstructed three major Syncytial neurons. Each of
+these Syncytial neurons was a cell with multiple nuclei, with membranes
+fused by continuous plasma membranes. These neurons are distinct in
+morphology from the Syncytial subepithelial nerve net (SNN) neurons with
+blebbed morphology previously reconstructed in 3D
+(\textbf{Burkhardt\_2023?}). Our findings represent the second
+documented discovery of Syncytial neurons in ctenophores. Furthermore,
+these neurons exhibited clear morphological differences when compared to
+other sensory cells reported in the same study, such as mesogleal
+neurons and sensory cells with presynaptic structures and cilia (types
+1-4)(\textbf{Burkhardt\_2023?}). Based on their distinct spatial
+relationships, we were able to classify these three Syncytial neurons
+into two categories. The first type is a larger ``AO neuron\_Q1234,''
+which possesses four (or possibly six?) nuclei spanning four quadrants.
+It contained X presynaptic structures. The second type consists of ``AO
+neuron\_Q12'' and ``AO neuron\_Q34,'' each having two nuclei spanning
+two quadrants, and each containing X presynaptic structures.
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure_network.png}
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure2.png}
 
@@
 
-\caption{\textbf{Figure 2. Title fig 2} Balancer complex: balancer,
-lithocyte, groove 2 types, bridge - EM pics, (A) legend (B) legend.}
+\caption{\textbf{Figure 2. Title fig 2} pic of a larva, boxed region of
+AO, schematic, catmaid pic with slice, dimensions, quadrants, regions,
+(A) legend (B) legend.}
 
@@
 
-``the larval bridge has not yet developed forks at its ends, as seen in
-adult Mnemiopsis bridges''(Tamm and Tamm, 2002)
+\subsubsection{Identification the Gravity-Sensitive Neural Circuit via
+the Syncytial Neurons
+Network}\label{identification-the-gravity-sensitive-neural-circuit-via-the-syncytial-neurons-network}
 
-Figure3 - dome, bicil, monocil, no-cil, plumose, dense vesicle cells
-(rename), gland? cells, oligo/multiciliated cells
+We discovered that AO neurons form synaptic connections with each other,
+and that AO neuron\_Q1234 forms self-synapses, or ``autapses.'' To our
+knowledge, previous reports have not identified synaptic connections
+between subepithelial nerve net (SNN) neurons. Thus, our results
+represent the first report of synaptic connectivity between neurons in
+the apical organ, forming a network in ctenophores (subject to
+confirmation). These AO neuron networks were found to form many
+presynaptic structures in relation to the gravity-sensing balancer
+cells.
 
-Plumose cells (Hernandez-Nicaise, 1984) may be pressure sensors (Tamm,
-2014b)
+Balancer cells are monociliated cells, and their cilia protrude longer
+than those of other monociliated cells. Moreover, the cilia of balancer
+cells are bundled into four groups at the center of the apical organ,
+forming a compound cilium. Based on these features, we identified and
+classified the balancer cells. The cellular arrangement differed clearly
+when viewed in the lateral view, sagittal plane, and tentacular plane,
+with the cell bodies gathering toward the apical organ in the tentacular
+plane. In each quadrant, the number of cells was as follows: Q1: 37, Q2:
+32, Q3: 32, Q4: 28. Each cell contained 3 to 10 ? mitochondria.
 
-Figure4 - lamellate body complex, LB, intra-multi-ciliated, relationship
-to balancer
+From AO neuron\_Q1Q2, synapses were formed with 6 of the 37 balancer
+cells in the Q1 region, and 8 of the 32 balancer cells in the Q2 region.
+Similarly, from AO neuron\_Q3Q4, synapses were formed with 1 of the 32
+balancer cells in the Q3 region, and 5 of the 28 balancer cells in the
+Q4 region. AO neuron\_Q1234 formed input synapses with balancer cells in
+the Q1 region (7 cells), Q2 region (11 cells), Q3 region (6 cells), and
+Q4 region (10 cells). Some balancer cells received inputs from both AO
+neuron\_Q1Q2 or AO neuron\_Q3Q4 and AO neuron\_Q1234. While previous
+studies have suggested the presence of afferent synapses from balancer
+cells to neurons (\textbf{Hernandez\_Nicaise\_1974?}), our data did not
+reveal any synaptic inputs from balancer cells to AO neurons. However,
+we identified ``bridge cells'' that actively formed input synapses with
+AO neurons, highlighting their potential role in the network.
 
-Figure5 - nerve net, different fragments/cells/syntytia, EM, relation to
-entire organ
+\begin{figure}[H]
 
-Figure 6 - cilia? renderings of basal bodies, centrioles, length,
-axoneme structure (table), table about cell types, number of cilia etc.
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure3.png}
 
-\subsection{Synapses}\label{synapses}
+}
 
-``synaptic regions with the characteristic presynaptic triad mor-
-phology of ctenophore nerves (i.e., a single layer of vesicles, smooth
-ER sac, and closely apposed mito- chondria (Hernandez-Nicaise, 1973;
-Horridge and Mackay, 1964) Hernandez- Nicaise, 1973)''
+\caption{\textbf{Figure 3. Title fig 3} pic of a larva, boxed region of
+AO, schematic, catmaid pic with slice, dimensions, quadrants, regions,
+(A) legend (B) legend.}
 
-Figure 7 - synapses, reconstruction, annotation, numbers, distribution
-of partners, example EM (figure supplement - lots of EM), mitochondria
-(Fig suppl - mitochondrion stats - Sanja)
+\end{figure}%
 
-Gap juntions? Innexins are expressed in the apical organ in specific
-patches of cells in each quadrant (Ortiz et al., 2023)
+\subsubsection{Identification of Neuron-like Bridge Cells Forming
+Afferent Synapses in the Gravity-Sensitive Neural
+Circuit}\label{identification-of-neuron-like-bridge-cells-forming-afferent-synapses-in-the-gravity-sensitive-neural-circuit}
 
-Ctenophore synapses (Hernandez-Nicaise, 1973)
+From our data, we identified several cell groups that form multiple
+afferent synapses with AO neurons. These cells, based on their
+morphological features, were found to be the bridge cells first
+described by Tamm et al.~in 2002 (\textbf{Tamm\_and\_Tamm\_2002?}).
+These cells are characterized by bundles of elongated processes filled
+with microtubules that arch over the epithelial layer, resembling a
+bridge. They originate from the base of paired balancer cells along the
+tentacle surface and extend across the sagittal plane toward the base of
+the opposite balancer cells. In regions where the mitochondria of the
+bridge cells are localized (approximately 30\%), a presynaptic triad
+structure, similar to that of AO neurons, was found, containing synaptic
+vesicles and smooth endoplasmic reticulum.
 
-We found no synapses between bridge and balancer, in agreement with Tamm
-(Tamm and Tamm, 2002).
+Our three-dimensional reconstruction data revealed that the bridge cells
+form two distinct cell groups across the sagittal plane, between the
+Q1Q2 and Q3Q4 regions. In the Q1Q2 region, 14 cells were identified,
+while in the Q3Q4 region, 12 cells were identified, totaling 26 bridge
+cells. Nearly all of these bridge cells (25 out of 26) exhibited
+afferent synapses from AO neurons. For bridge cells located in the Q1Q2
+region, synaptic inputs came primarily from AO neuron\_Q1Q2 (11 cells),
+AO neuron\_Q1234 (1 cell), or both (2 cells). For bridge cells located
+in the Q3Q4 region, synaptic inputs were received from AO neuron\_Q3Q4
+(1 cell), AO neuron\_Q1234 (7 cells), or both (1 cell). In other words,
+bridge cells in the Q1Q2 region mainly received input from AO
+neuron\_Q1Q2, while those in the Q3Q4 region received input primarily
+from AO neuron\_Q1234, showing a distinct pattern of input in the two
+regions.
 
-Figure 8 - connectome cell-type based, quadrant based, nerve net
-syntycia, bridge, quadrants (Fig suppl - cell-based network)
+Some bridge cells also formed afferent synapses with AO neurons. For
+example, bridge cells in the Q1Q2 region formed synapses with AO
+neuron\_Q1Q2 (3 cells) or both AO neuron\_Q1Q2 and AO neuron\_Q1234 (2
+cells). Bridge cells in the Q3Q4 region formed synapses with AO
+neuron\_Q3Q4 (1 cell) or AO neuron\_Q1234 (6 cells). A notable
+difference was observed between the Q1Q2 and Q3Q4 regions in the
+proportion of synaptic inputs from bridge cells to AO neurons.
 
-Tamm\&Tamm observed synapses from neurites to bridge cells (2002). and
-in Mnemiopsis larvae observed synapses from bridge cells onto other
-cells
+Interestingly, in both regions, bridge cells also formed synapses with
+adjacent bridge cells. However, no synaptic input was found from bridge
+cells across the sagittal plane to those in the opposite region. To
+analyze the grouped synaptic connectivity graph, we classified cell
+types within each region, collapsed cells of the same type into a single
+node, summed the number of synapses, and explored directed pathways to
+effectors (balancer cell groups). The results revealed a feedback
+pathway from AO neurons through the synaptic connections formed by
+bridge cells, thereby shedding light on the neural circuit structure
+involving these cells.
 
-Inserting Figures
+\begin{figure}[H]
 
-You can add your figures into the rendered document. We saved the
-figures into /manuscript/figures or /manuscript/figure\_supplements and
-can insert them from there. We use knitr::include\_graphics for this.
-The title and legend can also be edited, as will as the width of the
-output figure.
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure4.png}
 
-\subsection{Equations}\label{equations}
+}
 
-Equations can also be inserted, Insert -\textgreater{} Display Math:
+\caption{\textbf{Figure 4. Title fig 4} pic of a larva, boxed region of
+AO, schematic, catmaid pic with slice, dimensions, quadrants, regions,
+(A) legend (B) legend.}
 
-\[
-\bar{X} = \frac{\sum_{i=1}^{n} x_{i}}{n}
-\]
+\end{figure}%
+
+\subsection{Discussion}\label{discussion}
+
+\subsection{Materials and Methods}\label{materials-and-methods}
+
+\subsection{Acknowledgements}\label{acknowledgements}
+
+We would like to thank the Jekely lab for the R project template
+(\url{https://github.com/JekelyLab/new_paper_template}) we used to write
+this paper. This work was funded by \ldots{} JSPS\ldots{} ERC.. Others
+\ldots{}
+
+\subsection{References}\label{references}
 
@@
 
-\subsection{Materials and Methods}\label{materials-and-methods}
-
-\subsection{Acknowledgements}\label{acknowledgements}
-
-We would like to thank the Jekely lab for the R project template
-(\url{https://github.com/JekelyLab/new_paper_template}) we used to write
-this paper. This work was funded by \ldots{} JSPS\ldots{} ERC.. Others
-\ldots{}
-
-\subsection{References}\label{references}
-
 \phantomsection\label{refs}
@@
 mikroskopisch-anatomische Forschung} \textbf{88}:401--412.
-
-\bibitem[\citeproctext]{ref-Hernandez_Nicaise_1984}
-Hernandez-Nicaise M-L. 1984. CtenophoraBiology of the Integument.
-Springer Berlin Heidelberg. pp. 96--111.
-doi:\href{https://doi.org/10.1007/978-3-642-51593-4_9}{10.1007/978-3-642-51593-4\_9}
 
@@
 
-\bibitem[\citeproctext]{ref-Ortiz_2023}
-Ortiz J, Bobkov YV, DeBiasse MB, Mitchell DG, Edgar A, Martindale MQ,
-Moss AG, Babonis LS, Ryan JF. 2023. Independent innexin radiation shaped
-signaling in ctenophores. \emph{Molecular Biology and Evolution}
-\textbf{40}.
-doi:\href{https://doi.org/10.1093/molbev/msad025}{10.1093/molbev/msad025}
-
 \bibitem[\citeproctext]{ref-Tamm_2014}
-Tamm SL. 2014a. Formation of the statolith in the ctenophore mnemiopsis
+Tamm SL. 2014. Formation of the statolith in the ctenophore mnemiopsis
 leidyi. \emph{The Biological Bulletin} \textbf{227}:7--18.
 doi:\href{https://doi.org/10.1086/bblv227n1p7}{10.1086/bblv227n1p7}
-
-\bibitem[\citeproctext]{ref-Tamm_2014b}
-Tamm SL. 2014b. Cilia and the life of ctenophores. \emph{Invertebrate
-Biology} \textbf{133}:1--46.
-doi:\href{https://doi.org/10.1111/ivb.12042}{10.1111/ivb.12042}
 

### 2025-01-13 analysis/scripts/Figure5.R
Figure5

@@ -1,0 +1,39 @@
+# Code to generate Figure 5 of the Jokura et al 2024 Ctenophore apical organ connectome paper
+
+# source packages and functions ------------------------------------------------
+source("analysis/scripts/packages_and_functions.R")
+
+
+# assemble figure -------------------------------------------------------------
+
+panel_bal_s <- ggdraw() + draw_image(readPNG("manuscript/pictures/balancer_sagittal.png"))
+panel_bal_t <- ggdraw() + draw_image(readPNG("manuscript/pictures/balancer_tentacular.png"))
+panel_map_s <- ggdraw() + draw_image(readPNG("manuscript/pictures/map_sagittal.png"))
+panel_map_t <- ggdraw() + draw_image(readPNG("manuscript/pictures/map_tentacular.png"))
+panel_kymograph <- ggdraw() + draw_image(readPNG("manuscript/pictures/balacer_arrest_rebeat_kymograph.png"))
+panel_graph <- ggdraw() + draw_image(readPNG("manuscript/pictures/balacer_arrest_rebeat_graph.png"))
+
+layout <- "
+AA#BB#CC#DD
+###########
+EEEEEEEFFFF
+"
+
+
+
+Figure5 <- panel_bal_s + panel_map_s + panel_bal_t + panel_map_t +
+  panel_kymograph + panel_graph +
+  plot_layout(design = layout,
+              heights = c(1, 0.1, 1.5),
+              widths = c(1, 1, 0.1, 1, 1, 0.1, 1, 1, 0.1, 1, 1)) + 
+  plot_annotation(tag_levels = "A") + 
+  theme(plot.tag = element_text(size = 12, face='plain', color='black'))
+
+
+ggsave("manuscript/figures/Figure5.png", limitsize = FALSE, 
+       units = c("px"), Figure5, width = 2000, height = 1200, bg='white')  
+
+
+ggsave("manuscript/figures/Figure5.pdf", limitsize = FALSE, 
+       units = c("px"), Figure5, width = 2000, height = 1200) 
+

### 2025-01-29 analysis/scripts/Figure5.R
update fig5

@@ -6,6 +6,7 @@ source("analysis/scripts/packages_and_functions.R")
 
 # assemble figure -------------------------------------------------------------
 
+panel_ms <- ggdraw() + draw_image(readPNG("manuscript/pictures/tilt_microscope.png"))
 panel_bal_s <- ggdraw() + draw_image(readPNG("manuscript/pictures/balancer_sagittal.png"))
 panel_bal_t <- ggdraw() + draw_image(readPNG("manuscript/pictures/balancer_tentacular.png"))
 panel_map_s <- ggdraw() + draw_image(readPNG("manuscript/pictures/map_sagittal.png"))
@@ -14,26 +15,28 @@ panel_kymograph <- ggdraw() + draw_image(readPNG("manuscript/pictures/balacer_ar
 panel_graph <- ggdraw() + draw_image(readPNG("manuscript/pictures/balacer_arrest_rebeat_graph.png"))
 
 layout <- "
-AA#BB#CC#DD
-###########
-EEEEEEEFFFF
+AA#B#C
+AA####
+AA#D#E
+######
+FFFF#G
 "
 
 
 
-Figure5 <- panel_bal_s + panel_map_s + panel_bal_t + panel_map_t +
+Figure5 <- panel_ms + panel_bal_s + panel_bal_t + panel_map_s + panel_map_t +
   panel_kymograph + panel_graph +
   plot_layout(design = layout,
-              heights = c(1, 0.1, 1.5),
-              widths = c(1, 1, 0.1, 1, 1, 0.1, 1, 1, 0.1, 1, 1)) + 
+              heights = c(1.5, 0.1, 1, 0.1, 3),
+              widths = c(1, 1, 0.1, 1, 0.1, 1)) + 
   plot_annotation(tag_levels = "A") + 
   theme(plot.tag = element_text(size = 12, face='plain', color='black'))
 
 
 ggsave("manuscript/figures/Figure5.png", limitsize = FALSE, 
-       units = c("px"), Figure5, width = 2000, height = 1200, bg='white')  
+       units = c("px"), Figure5, width = 2500, height = 1800, bg='white')  
 
 
 ggsave("manuscript/figures/Figure5.pdf", limitsize = FALSE, 
-       units = c("px"), Figure5, width = 2000, height = 1200) 
+       units = c("px"), Figure5, width = 2500, height = 1800) 
 

### 2025-02-07 analysis/scripts/Figure6.R
Discussion

@@ -1,0 +1,31 @@
+# Code to generate Figure 6 of the Jokura et al 2024 Ctenophore apical organ connectome paper
+
+# source packages and functions ------------------------------------------------
+source("analysis/scripts/packages_and_functions.R")
+
+
+# assemble figure -------------------------------------------------------------
+
+panel_sum <- ggdraw() + draw_image(readPNG("manuscript/pictures/discussion.png"))
+
+layout <- "
+A
+"
+
+
+
+Figure6 <- panel_sum +
+  plot_layout(design = layout,
+              heights = c(1),
+              widths = c(1)) + 
+  plot_annotation(tag_levels = "A") + 
+  theme(plot.tag = element_text(size = 12, face='plain', color='black'))
+
+
+ggsave("manuscript/figures/Figure6.png", limitsize = FALSE, 
+       units = c("px"), Figure6, width = 2500, height = 1050, bg='white')  
+
+
+ggsave("manuscript/figures/Figure6.pdf", limitsize = FALSE, 
+       units = c("px"), Figure6, width = 2500, height = 1050) 
+

### 2025-02-11 manuscript/Manuscript.tex
Introduction

@@
 \makeatother
-
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{241,243,245}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
-\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
 Tamm, 2002).
-
-The easiest way is to use the command line to add new refs to the .bib
-file:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#curl {-}LH "Accept: application/x{-}bibtex" https://doi.org/10.7554/eLife.91258.1 \textgreater{}\textgreater{} references.bib}
-\end{Highlighting}
-\end{Shaded}
 
@@
 
+\subsubsection{Control of Balancer Ciliary Movements by
+Gravity-Sensitive Neural
+Circuits}\label{control-of-balancer-ciliary-movements-by-gravity-sensitive-neural-circuits}
+
+To investigate the function of gravity-sensitive circuits, we conducted
+a comparative analysis of balancer ciliary movements across different
+regions. Previous studies (Tamm, 1980, 1982; Lowe, 1997) have
+established that balancer cilia function as mechanoreceptors, with their
+beating frequency modulated by inclination. Furthermore, it has been
+suggested that differences in the asymmetry of statolith morphology and
+the elongation direction of balancer cilia between the tentacular and
+sagittal planes could result in variations in the force received from
+the statolith (Tamm, 2014, 2015). Building on these findings, we
+standardized conditions for observing balancer cilia by tilting the
+microscope by 90 degrees and fixing samples such that the aboral-oral
+axis was aligned parallel to the vertical stage, with the aboral side
+facing upward. Using connectome analysis, we aimed to evaluate the
+neural circuit differences by analyzing the control patterns of left and
+right balancer cilia in the tentacular and sagittal planes.
+Specifically, we hypothesized that differences in control between Q1
+(Q2) and Q3 (Q4) regions, as well as between Q1 (Q4) and Q2 (Q3)
+regions, would reveal distinct functions of AO neurons (e.g., AO neuron
+Q12/Q34 vs.~AO neuron Q1234). For the sagittal plane, we recorded
+balancer ciliary movements in the Q1(4) and Q2(3) regions (left and
+right sides of the image) and, for the tentacular plane, in the Q1(3)
+and Q2(4) regions, using a high-speed camera (100 fps) for 2 minutes
+(12,000 frames). Regions of interest (ROIs) were defined in areas where
+brightness changes indicated ciliary beating. We measured these
+brightness changes to generate dynamics graphs of balancer ciliary
+movements for both planes.
+
+During the 2-minute recordings, balancer cilia exhibited movements that
+could be fast, slow, or stop abruptly (arrest) and start moving again
+(re-beat). Occasionally, large contraction responses caused the entire
+cydippid to move out of frame, and data from these moments were
+excluded. We focused on two metrics for left-right comparison: (1) the
+timing of ciliary arrest and (2) the timing of re-beat initiation
+following an arrest. Due to the low baseline beating frequency of the
+cilia, rapid synaptic input responses were not evaluated, suggesting
+that ciliary beat regulation may involve non-synaptic control
+mechanisms. The analysis revealed that both arrest and re-beat events
+were synchronized between left and right balancers in both planes within
+a few milliseconds, suggesting that these events are governed by
+synaptic input from neurons. Interestingly, while the re-beat timing
+showed no significant differences between the sagittal and tentacular
+planes, arrest timing exhibited greater variability in the tentacular
+plane. Comparing these findings with neural circuit diagrams suggests
+that in the sagittal plane, the same AO neurons (Q12 or Q34) project to
+both balancer cell groups, transmitting signals simultaneously and
+enabling synchronization. In contrast, in the tentacular plane, separate
+AO neurons project to different regions, potentially introducing an
+additional synaptic step, which may account for the observed variability
+in arrest timing.
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure5.png}
+
+}
+
+\caption{\textbf{Figure 5. Title fig 5} pic of a larva, boxed region of
+AO, schematic, catmaid pic with slice, dimensions, quadrants, regions,
+(A) legend (B) legend.}
+
+\end{figure}%
+
 \subsection{Discussion}\label{discussion}
 
+In this study, we traced all the constituent cells of the aboral organ
+in five-day-old larvae of \emph{Mnemiopsis leidyi} and completed a
+comprehensive neural map. As a result, we identified a novel type of
+syncytial neuron with a previously unreported morphology. Furthermore,
+we found that three such syncytial neurons project to distinct
+quadrants, integrating with balancer cell groups to form a
+gravity-sensing neural circuit. This study represents the first complete
+neural circuit map reported in ctenophores. Additionally, our analysis
+of balancer ciliary movements, based on neural circuit data, revealed
+differences in ciliary arrest control arising from circuit variations.
+
+\subsubsection{Identification of a Previously Unreported Syncytial
+Neuron}\label{identification-of-a-previously-unreported-syncytial-neuron}
+
+We identified three syncytial neurons with multiple nuclei in the apical
+organ of five-day-old \emph{M. leidyi} cydippid larvae. Burkhurdt et
+al.~(2023) previously reported that in one-day-old larvae, the
+sub-epithelial nerve net (SNN) contains neurons that form syncytia.
+However, the newly discovered syncytial neurons embedded within the
+apical organ cell cluster did not exhibit the characteristic ``pearl
+necklace-like structure'' of the SNN, which consists of microtubules
+connecting neurons. Instead, their morphology was more reminiscent of
+fiber cells previously described in Placozoa (Mayorova et al., 2021),
+with cell bodies extending through intercellular spaces. These
+morphological characteristics suggest that the newly identified neurons
+represent a second distinct type of syncytial neuron in ctenophores.
+
+\subsubsection{Functional Diversification of the Three Syncytial
+Neurons}\label{functional-diversification-of-the-three-syncytial-neurons}
+
+The three identified neurons may be classified into two distinct
+functional types based on their spatial arrangement and their influence
+on the differential control of balancer ciliary movement between the
+quadrants. In metazoans, a single neuron can directly provide synaptic
+input to multiple ciliated cells at a distance, synchronizing their
+movements. For instance, in larvae of the annelid \emph{Platynereis
+dumerilii}, a cholinergic motor neuron (MC) innervates multiple
+prototroch ciliated cells, synchronizing their motion such that
+activation of the neuron leads to simultaneous arrest of ciliary beating
+(Verasztó et al., 2017). Additionally, a pair of serotonergic neurons,
+designated Ser-h1, extend axons that cross at the midline; the left
+Ser-h1\_l predominantly innervates right-sided ciliated cells, while the
+right Ser-h1\_r mainly projects to the left-side ciliated cells.
+Activation of Ser-h1 induces ciliary beating initiation and increases
+beat frequency.
+
+In the this study, we found that three neurons are closely positioned
+within the confined space of the apical organ, potentially enabling a
+single neuron to control distant ciliary movements in a synchronized
+manner. Moreover, the presence of neurons projecting to the same
+ciliated cells in distinct patterns suggests that these neurons may
+serve opposing functions---one inhibiting ciliary activity while the
+other promotes it, or one modulating ciliary beat frequency. While our
+synaptic ultrastructure analysis did not allow us to distinguish
+excitatory from inhibitory neurons, future electrophysiological
+investigations will be essential to elucidate their functional
+properties.
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure6.png}
+
+}
+
+\caption{\textbf{Figure 6. Title fig 6} pic of a larva, boxed region of
+AO, schematic, catmaid pic with slice, dimensions, quadrants, regions,
+(A) legend (B) legend.}
+
+\end{figure}%
+
+\subsubsection{Paracrine Signaling and Synaptic Transmission from
+Sensory to Neural
+Cells}\label{paracrine-signaling-and-synaptic-transmission-from-sensory-to-neural-cells}
+
+Previous electron microscopy studies have suggested the presence of
+various sensory cells, including photoreceptor and mechanoreceptor
+cells, within the aboral organ of ctenophores (Horridge, 1965; Krisch,
+1973; Aronova, 1974; Vinnikov, 1974). By comparing our dataset with
+previously reported ultrastructural data, we successfully identified
+several sensory cells within the apical organ. However, notably, we did
+not find any clear presynaptic structures between these sensory cells
+and either neural or other cellular targets. In prior studies,
+immunostaining for specific neuropeptides in ctenophores has revealed
+positive signals in distinct cell populations within the apical organ
+(Sachkova et al., 2021; Hayakawa et al., 2022). These findings suggest
+that many sensory cells in the apical organ may primarily rely on
+neuropeptide-mediated paracrine signaling rather than direct synaptic
+connections, supporting the chemical brain hypothesis (Jekely, 2021).
+While paracrine transmission is generally slower than synaptic
+transmission, it allows for a greater diversity of information
+processing.
+
+Conversely, previous research has demonstrated that mechanoreceptive
+sensory cells equipped with cilia (referred to as type 3 and type 4
+cells) in ctenophores form direct synaptic connections with the
+sub-epithelial nerve net (SNN) and mesogleal neurons (Burkhurdt et al.,
+2023). Based on their ultrastructural features, these sensory cells are
+believed to respond to physical stimuli such as water flow, vibrations,
+and direct contact, transmitting information rapidly to cilia and muscle
+cells (Hertwig, 1880; Chun, 1880; Eimer, 1880).
+
+Surprisingly, our study identified bridge cells as the primary cell type
+forming synaptic inputs onto neurons in the apical organ. Bridge cells
+were first described by Tamm et al.~(2002) based on their morphological
+characteristics, with cellular projections extending to the basal region
+of balancer cilia (Tamm \& Tamm, 2002). Our findings suggest that bridge
+cells are electrically coupled to balancer cilia and provide rapid
+feedback to neurons via synaptic transmission. Future studies analyzing
+differential responses to various sensory stimuli will not only
+elucidate the functional roles of bridge cells but also provide insights
+into the evolutionary relationship between transmission speed and
+synapse development.
+
+\subsubsection{Circuit Variability as a Driver of Behavioral
+Diversity}\label{circuit-variability-as-a-driver-of-behavioral-diversity}
+
+Our study suggests that differences in neural circuits contribute to
+variations in the timing of ciliary arrest, which may play a role in
+modulating swimming direction in ctenophores (Richard Satterlie, 2015).
+This finding provides evidence that increased circuit complexity
+contributes to behavioral diversification. By integrating multiple
+neurons and altering circuit structure, information can be branched,
+allowing for diverse motor patterns within the framework of fast
+synaptic transmission.
+
+The neural cells identified in the apical organ exhibit a divergent
+feedforward connectivity pattern, where a single neuron forms
+presynaptic connections with multiple downstream targets. Such a circuit
+design minimizes unwanted signal interference, reduces processing
+delays, and enables relatively fast response execution (Luo, 2021).
+Moreover, the presence of multiple divergent feedforward circuits
+suggests that slight variations in the synchronization and phase delay
+of ciliary beating (within the range of a few hundred milliseconds) may
+have evolved as a mechanism for generating behavioral diversity.
+
+From a broader neuroanatomical perspective, nervous system structures
+across various animals have been shaped through selective optimization
+to adapt to environmental constraints (Bullmore \& Sporns, 2012; Perin
+et al., 2011). For example, in nematodes, the spatial arrangement of
+ganglia is optimized to minimize the total length of neural wiring
+(White et al., 1986). Similarly, in ctenophores, understanding how
+transmission via syncytial neurons and synaptic connections has been
+selectively optimized at the circuit level will provide crucial insights
+into the evolutionary scaling of neural networks from early nervous
+systems to complex brains (Farnworth and Montgomery, 2024). In
+particular, this research may help elucidate how circuit modularization
+and layered structures emerged over evolutionary time.
+
 \subsection{Materials and Methods}\label{materials-and-methods}
+
+\subsubsection{Data and Code
+Availability}\label{data-and-code-availability}
+
+The following EM image stacks, including all traces and annotations, are
+available at https://catmaid.jekelylab.ex.ac.uk. These datasets can be
+queried either within CATMAID or through the CATMAID application
+programming interface (API) using the catmaid or natverse packages
+(Bates et al., 2020). The dataset encompasses all EM images (in JPG
+format), skeletons, meshes, node tags, connectors, and annotations.
+Additionally, we provide all R scripts used for data acquisition and
+figure generation (Jokura et al., 2025). All plots, figures (including
+anatomical renderings), and figure layouts can be fully reproduced using
+the provided R scripts. While the scripts are mostly organized by
+figure, some general-purpose scripts---for tasks such as loading
+libraries, accessing CATMAID data, and defining common functions---are
+shared across multiple figures.
+
+\subsubsection{Specimen Preparation, Transmission Electron Microscopy,
+and Image
+Processing}\label{specimen-preparation-transmission-electron-microscopy-and-image-processing}
+
+Larvae of Mnemiopsis leidyi (5 days old) were cryofixed using a
+high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) and
+immediately transferred to liquid nitrogen for storage. The frozen
+samples were processed in a substitution medium containing 2\% (w/v)
+osmium tetroxide and 0.5\% uranyl acetate in acetone, using a
+cryo-substitution device (EM AFS-2, Leica). Cryo-substitution was
+performed by gradually raising the temperature, and the samples were
+embedded in epoxy resin. Serial sections of 50 nm thickness were
+prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º
+DiATOME diamond knife. To enhance section adhesion and improve
+hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO
+Glass, UQG Optics) was treated with air glow discharge using the PELCO
+easiGlow system (Ted Pella, Inc.), rendering the carbon film surface
+negatively charged. Section ribbons were collected on the prepared glass
+slides, slowly dried to ensure proper stretching, and firmly adhered to
+the glass surface. The sections on glass were stained with UranyLess and
+lead citrate (Reynolds) using airless staining bottles (Delta
+Microscopies). The glass slides were mounted on STEM-specific stages
+(Zeiss) using Copper Foil EMI Shielding Tape (3M). Imaging was performed
+using a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5
+imaging software (Zeiss). The full dataset consisted of 620 serial
+sections, each 50 nm thick. The imaging resolution was 2.8 nm/pixel,
+using electron holography transmission at an acceleration voltage of 1.5
+kV (in-lens detector, dwell time: 3 µs).
+
+\subsubsection{Image-Stack Realignment and Spatial Data Transformation
+in
+CATMAID}\label{image-stack-realignment-and-spatial-data-transformation-in-catmaid}
+
+To process the image stack, we utilized the TrakEM2 plugin of FIJI
+(ImageJ) (version 2.0.0-rc-15/1.49k / Java 1.6.0\_24 (64-bit) -- 2014).
+A project was created, and all TIFF images were imported using the
+``import sequence as grid'' function. Subsequently, the following
+filters were applied sequentially: Invert, Equalize Histogram, and
+Gaussian Blur.
+
+The alignment process consisted of three stages, each progressively
+refining the spatial accuracy:
+
+Rigid Alignment Initially, a rigid alignment was performed with the
+following parameters: least squares mode (linear feature
+correspondences), encompassing the entire layer range with the first
+layer as the reference. Only visible images were used, without
+propagation. The alignment was executed with an initial Gaussian blur of
+1.6 pixels, three steps per scale octave, a minimum image size of 512
+pixels, and a maximum of 2048 pixels. Additional parameters included a
+feature descriptor size of 8, orientation bins of 8, and a closest ratio
+of 0.92. The alignment allowed clearing the cache, using 32 feature
+extraction threads, a maximal alignment error of 100 pixels, a minimal
+inlier ratio of 0.20, and a minimum of 12 inliers. The expected and
+desired transformations were set to rigid, with testing multiple
+hypotheses (tolerance: 5.00 pixels) and considering up to 5 neighboring
+layers, giving up after 5 failures. Regularization was performed with a
+maximal iteration of 1000, a maximal plateau width of 200, and a rigid
+lambda of 0.10.
+
+Affine Alignment Following the rigid alignment, an affine alignment was
+applied using similar parameters, except the expected and desired
+transformations were set to affine. The minimal image size was reduced
+to 64 pixels, while the other parameters (e.g., Gaussian blur, feature
+descriptor size, inliers, and testing hypotheses) remained unchanged to
+ensure consistent processing.
+
+Elastic Alignment Two iterations of elastic alignment were performed to
+fine-tune the spatial data. Key parameters included a block matching
+layer scale of 0.05, a search radius of 200 pixels, a block radius of
+2000 pixels, and a resolution of 60. Correlation filters were employed
+with a minimal PMCC r of 0.10, a maximal curvature ratio of 1000, and a
+maximal second-best r/best r of 0.90. A local smoothness filter was
+applied with the approximate local transformation set to affine, a local
+region sigma of 1000 pixels, and an absolute maximal local displacement
+of 10 pixels (relative maximal displacement: 3.00). Pre-aligned layers
+were tested for up to 4 neighboring layers. The elastic alignment used a
+rigid approximation, maximal iterations of 3000, a plateau width of 200,
+spring mesh stiffness of 0.01, and a maximal stretch of 2000 pixels. A
+legacy optimizer was employed to enhance performance.
+
+After each alignment stage, the project was saved as an XML file under a
+unique name to preserve iterative progress. Finally, the images were
+exported from FIJI using TrakEM2 in a format compatible with CATMAID.
+
+\subsubsection{Neuron Tracing, Synapse Annotation, and
+Review}\label{neuron-tracing-synapse-annotation-and-review}
+
+The reconstruction of cells and connectivity was performed using CATMAID
+(Saalfeld et al., 2009), analyzing a total of 620 serial sections. To
+digitally reconstruct all neurons within the serial TEM dataset, we
+utilized the collaborative web-based application CATMAID, installed on a
+local server (Saalfeld et al., 2009; Schneider-Mizell et al., 2016). To
+mark the locations of cell bodies, we placed tags at the approximate
+center of each nucleus within the dataset. At each nuclear center, the
+radius of the single node was adjusted according to the size of the cell
+body in that specific layer. All skeletons were rooted at their
+respective cell bodies, and nodes were tagged as ``soma.'' Synapses were
+identified based on four key structural features: the cell membrane,
+synaptic vesicles, the endoplasmic reticulum, and mitochondria. Most
+synapses could be verified across consecutive sections, ensuring
+accurate annotation and connectivity mapping.
+
+\subsubsection{Cell Nomenclature and
+Annotations}\label{cell-nomenclature-and-annotations}
+
+Each cell is assigned a specific type name based on its category
+(balancer, biC, bridge, bristle, dense vesicle cell, dome, EF, IcMC, LB,
+lithocyte, monoC, non-C, plumose cell, SNN, ssCG, SSN, stCG), resulting
+in a total of 17 cell types. Additionally, a quadrant number is appended
+to the cell type name following an underscore (``\_'') to indicate the
+cell body's location. For example, cells located in the first quadrant
+are annotated with ``Q1.'' If a cell is situated between the first and
+second quadrants, the annotation ``Q1Q2'' is applied.
+
+Cells of the same type within the same region are further distinguished
+by serial numbering. Each cell is also assigned multiple annotations,
+which can be utilized to query the database via CATMAID or the CATMAID
+API (e.g., using the R catmaid package). These annotations provide a
+structured and precise framework for identifying and analyzing specific
+cells, facilitating robust data integration and retrieval within the
+dataset.
+
+This system ensures consistent and accurate tracking of cellular data,
+enabling comprehensive analysis of cellular structures and functions
+across regions.
+
+\subsubsection{Analysis of Balancer
+Cilia}\label{analysis-of-balancer-cilia}
+
+We conducted the analysis using the cydippid stage of Mnemiopsis leidyi
+at five days post-fertilization. A coverslip with a thin layer of
+Vaseline applied to two edges was lightly placed on a slide glass.
+Filtered natural seawater was introduced under the coverslip, along with
+the cydippids. By gently moving the coverslip, the desired orientation
+for observation was adjusted, and once positioned, slight pressure was
+applied to immobilize the cydippids. To stabilize the statolith's
+position, the microscope was tilted 90 degrees, arranging the stage
+vertically. The movement of the balancer cilia was observed using a
+differential interference contrast (DIC) microscope (Zeiss Axio
+Imager.M2) equipped with a highly sensitive CMOS monochrome camera
+optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2,
+iDS). Observations were made using a 40x objective lens (Objective LD
+LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27) with glycerine immersion.
+For recording, we employed a custom-made NIR LED strobe illumination
+system (wavelength: 850 nm) synchronized with the camera's exposure
+signals. The camera and LED strobe system were operated using the Video
+Capture Software BohNavi. Images were recorded at a resolution of
+640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms
+pulses from the LED. To analyze the ciliary beating of the balancer
+cilia, we utilized the Multi Kymograph project in Fiji software.
 

### 2025-02-13 analysis/scripts/Figure1.R
Ms and code: edits

@@ -3,16 +3,169 @@
 # source packages and functions -----------------
 source("analysis/scripts/packages_and_functions.R")
 
+<<<<<<< Updated upstream
 
 # assemble figure -------------------------------------------------------------
 
 library(jpeg)
+=======
+# load cells ----------------------------------
+
+balancer <- read_smooth_neuron("celltype:balancer")
+
+Q1 <- nlapply(
+  read.neurons.catmaid("Q1", pid = 35),
+  function(x)
+  smooth_neuron(x, sigma = 1000)
+  )
+
+Q2 <- nlapply(
+  read.neurons.catmaid("Q2", pid = 35),
+  function(x)
+  smooth_neuron(x, sigma = 1000)
+  )
+
+Q3 <- nlapply(
+  read.neurons.catmaid("Q3", pid = 35),
+  function(x)
+  smooth_neuron(x, sigma = 1000)
+  )
+
+Q4 <- nlapply(
+  read.neurons.catmaid("Q4", pid = 35),
+  function(x)
+  smooth_neuron(x, sigma = 1000)
+  )
+
+with_soma <- nlapply(
+  read.neurons.catmaid("with_soma", pid = 35),
+  function(x)
+  smooth_neuron(x, sigma = 1000)
+  )
+
+
+# plot balancer and quadrants ----------------------
+
+
+
+# plot balancer -----------------------------------------------
+
+plot_background()
+close3d()
+# 3d plotting of cells
+nopen3d() 
+mfrow3d(1, 3)  #defines the two scenes
+#define the size of the rgl window, the view and zoom
+#par3d(windowRect = c(0, 0, 1200, 350))
+par3d(windowRect = c(0, 0, 2400, 700))
+
+#plot aboral view
+plot3d(balancer,
+       soma = T, lwd = 1, add = T, alpha = 0.5, col = Okabe_Ito[5],
+       WithConnectors = F, WithNodes = F)
+
+#for (object in objects) {
+#  plot3d(object, soma = TRUE, lwd = 1, add = TRUE, alpha = 0.05, col = Okabe_Ito[8])
+#}
+
+
+plot3d(
+  with_soma, soma = T, lwd = 1, add = T, 
+  alpha = 0.05, col = Okabe_Ito[8]
+  )
+
+#texts3d(58000,51000,5000, "INRGW", cex = 3, col = "#56B4E9")
+
+#aboral view
+nview3d("anterior", extramat = rotationMatrix(1.05, 250, -200, 1000))
[…]

### 2025-02-13 analysis/scripts/all_celltype_plot3d.R

@@ -6,20 +6,26 @@ source("analysis/scripts/packages_and_functions.R")
 # load cell type ----------------------------------
 
 balancer <- read_smooth_neuron("celltype:balancer")
+plumose <- read_smooth_neuron("celltype:plumose")
 bridge <- read_smooth_neuron("celltype:bridge")
-bristle <- read_smooth_neuron("celltype:bristle")
+lithocyte <- read_smooth_neuron("celltype:lithocyte")
+SER <- read_smooth_neuron("celltype:SER_epithelia")
 dome <- read_smooth_neuron("celltype:dome")
+bristle <- read_smooth_neuron("celltype:bristle")
+intramulticilia <- read_smooth_neuron("celltype:intra-multi-ciliated")
+bicilia <- read_smooth_neuron("celltype:biciliated")
+monocilia <- read_smooth_neuron("celltype:monociliated")
+non_cilia <- read_smooth_neuron("celltype:nonciliated")
+dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
 Cgroove_tag <- read_smooth_neuron("celltype:Cgroove-tag")
 Cgroove_sag <- read_smooth_neuron("celltype:Cgroove-sag")
-intramulticilia <- read_smooth_neuron("celltype:intra-multi-ciliated")
-lamellate <- read_smooth_neuron("celltype:lamellate")
-lithocyte <- read_smooth_neuron("celltype:lithocyte")
-neuron <- read_smooth_neuron("celltype:SSN")
-plumose <- read_smooth_neuron("celltype:plumose")
-dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
-monocilia <- read_smooth_neuron("celltype:monociliated")
-bicilia <- read_smooth_neuron("celltype:biciliated")
-non_cilia <- read_smooth_neuron("celltype:nonciliated")
+SSN_neuron <- read_smooth_neuron("celltype:SSN")
+basalbody <- read_smooth_neuron("celltype:basalbody")
+multicentriole <- read_smooth_neuron("celltype:multicentriole")
+SNN_neuron <- read_smooth_neuron("celltype:SNN")
+lamellateE <- read_smooth_neuron("celltype:lamellate-extra")
+lamellateI <- read_smooth_neuron("celltype:lamellate-intra")
+
 
 
 Q1 <- nlapply(read.neurons.catmaid("Q1", pid = 35),
[…]

### 2025-02-13 analysis/scripts/balancer_plot3d.R

@@ -5,27 +5,27 @@ source("analysis/scripts/packages_and_functions.R")
 
 # load cell type ----------------------------------
 
-read_smooth_neuron <- function(annotation){
-  nlapply(read.neurons.catmaid(annotation, pid = 35),
-          function(x)
-            smooth_neuron(x, sigma = 1000))
-}
-
 balancer <- read_smooth_neuron("celltype:balancer")
+plumose <- read_smooth_neuron("celltype:plumose")
 bridge <- read_smooth_neuron("celltype:bridge")
-bristle <- read_smooth_neuron("celltype:bristle")
+lithocyte <- read_smooth_neuron("celltype:lithocyte")
+SER <- read_smooth_neuron("celltype:SER_epithelia")
 dome <- read_smooth_neuron("celltype:dome")
+bristle <- read_smooth_neuron("celltype:bristle")
+intramulticilia <- read_smooth_neuron("celltype:intra-multi-ciliated")
+bicilia <- read_smooth_neuron("celltype:biciliated")
+monocilia <- read_smooth_neuron("celltype:monociliated")
+non_cilia <- read_smooth_neuron("celltype:nonciliated")
+dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
 Cgroove_tag <- read_smooth_neuron("celltype:Cgroove-tag")
 Cgroove_sag <- read_smooth_neuron("celltype:Cgroove-sag")
-intramulticilia <- read_smooth_neuron("celltype:intra-multi-ciliated")
-lamellate <- read_smooth_neuron("celltype:lamellate")
-lithocyte <- read_smooth_neuron("celltype:lithocyte")
-neuron <- read_smooth_neuron("celltype:SSN")
-plumose <- read_smooth_neuron("celltype:plumose")
-dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
-monocilia <- read_smooth_neuron("celltype:monociliated")
-bicilia <- read_smooth_neuron("celltype:biciliated")
-non_cilia <- read_smooth_neuron("celltype:nonciliated")
+SSN_neuron <- read_smooth_neuron("celltype:SSN")
+basalbody <- read_smooth_neuron("celltype:basalbody")
+multicentriole <- read_smooth_neuron("celltype:multicentriole")
+SNN_neuron <- read_smooth_neuron("celltype:SNN")
+lamellateE <- read_smooth_neuron("celltype:lamellate-extra")
[…]

### 2025-02-13 analysis/scripts/packages_and_functions.R

@@ -15,6 +15,7 @@ library(RColorBrewer)
 library(igraph)
 library(tidygraph)
 library(av)
+library(jpeg)
 
 
 

### 2025-02-13 manuscript/Manuscript.qmd
balancer_plot_3d
MS: add hypothes.is

--- 2024-03-12 manuscript/Manuscript.qmd
@@
 ---
-title: "Cell-type diversity and connectome of the ctenophore apical organ"
+title: "Synaptic connectome of the ctenophore apical organ"
+subtitle: |
+  \raggedright  Kei Jokura^1,2,4,5\*^, Sanja Jasek^1,3,4^,  Lara Kewalow^3^, Pawel Burkhardt^6^, Gáspár Jékely^1,3,4,\*^  \newline<br>^1^Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom \newline<br>^2^Exploratory Research Center on Life and Living Systems (ExCELLS), National Institutes of Natural Sciences, Okazaki, Japan\newline<br>^3^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany \newline<br>^4^BioSciences, University of Exeter, Stocker Road, Exeter EX4 4QD, Exeter, United Kingdom\newline<br> ^5^National Institute for Basic Biology (NIBB), National Institutes of Natural Sciences, Okazaki, Japan\newline<br>^6^Michael Sars Centre, University of Bergen, Norway \newline<br> ^\*^Correspondence:  jokura@nibb.ac.jp, gaspar.jekely@cos.uni-heidelberg.de
 mainfont: Arial
@@
 
-\
+```{r}
+#| include: false
+#| eval: true
+#source(here::here("analysis/scripts/statistics_for_paper.R"))
 
-Kei Jokura^1,3,\*^, Sanja Jasek^1,2^, Lara Kewalow^2^, Pawel Burkhardt^4^, Gáspár Jékely^1,2,\*^\
+#The 'analysis/scripts/statistics_for_paper.R' script is sourced and it runs but the output is not included in the knitted output.
+But we can access the variables defined in the sourced script simply by adding \` r var_name \` between \` backticks, in this case max_PRC value is (now this number comes from our sourced script).
 
-^1^Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom\
-^2^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany\
-^3^National Institute for Basic Biology, Okazaki, ..., Japan
-^4^Sars Centre ... Berge, Norway
-^\*^Correspondence: corr_author\@email.com
+```
 
@@
 
-First descriptions of AO: [@aronova1974electron]
+Understanding how neurons evolved and what their initial functions were is key to elucidating the processes by which animals developed flexible behaviors.
+Neurons do not function in isolation; rather, they form networks that integrate and process information, enabling the emergence of complex behaviors.
+Therefore, understanding the structure and function of neural circuits is essential for investigating the mechanisms underlying animal behavior.
 
-The lobate ctenophore *Mnemiopsis*
+Recent comparative genomic studies (Ryan et al., 2013; Whelan et al., 2015; Whelan et al., 2017; Li et al., 2021; Whelan and Halanych, 2023) and chromosome-level synteny analyses across different animal phyla (Schultz et al., 2023) suggest that ctenophores are the sister group to all other metazoans and were the first animals to evolve a nervous system.
+This has led to attempts to compare genes and molecular modules in ctenophores using single-cell analyses, in which neuropeptides have been employed as neuronal markers (Sebé-Pedrós et al., 2018; Sachkova et al., 2021; Hayakawa et al., 2022).
+However, the homologous relationships between ctenophore genes and those of other animals are highly divergent, limiting the feasibility of sequence-based comparisons.
+Given these challenges, we considered an alternative approach: linking neural circuits to physiological responses to gain insight into neural function.
 
-The lithocytes have a large calcium carbonate concretion and... [@Tamm_2014]
-Living lithocytes are transported along the balancer cilia to the tip of the statocyst [@Noda_2014]
-The projections of a group of bridge cells located on two sides of the apical organ forms an arch and connects the balancers along the tentacular plane [@Tamm_2002].
+Despite their phylogenetic position, ctenophores exhibit sophisticated and dynamic behavioral patterns in response to a wide range of external stimuli.
+These behaviors are thought to be regulated by multiple neural systems, including the subepithelial nerve net (SNN), mesogleal neurons, an apical organ, and tentacle nerves.
+However, the mechanisms by which these neural systems control behavior remain largely unexplored.
 
-The easiest way is to use the command line to add new refs to the .bib file:
+Among these, we focused on the apical organ of ctenophores.
+In many bilaterian and non-bilaterian animals, an apical nervous system is located at one pole of the body, forming a specialized sensory structure that governs physiological functions and motor activities.
+In ctenophores, ultrastructural analyses have revealed that the aboral pole contains a diverse array of sensory cells and neuronal components (Hernandez-Nicaise, 1968, 1973, 1974; Tamm, 1982).
+Additionally, immunostaining using an anti-tyrosylated α-tubulin antibody has identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control (Jager et al., 2011).
 
-```{bash}
-#| echo: true
-#| eval: false
-
-#curl -LH "Accept: application/x-bibtex" https://doi.org/10.7554/eLife.91258.1 >> references.bib
-```
+In this study, we specifically examined the gravity-sensing system within the apical organ.
+The apical organ of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014].
+The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014] (Tamm, 1982).
+Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron].
+Therefore, we conducted a connectome analysis to characterize the network structure of the apical organ and elucidate the neural circuits that link sensory input to behavioral responses in ctenophores.
 
@@
 
-- specimen 
-- volume, resolution, fixation, quality
-- number of sections
-- skeletonisation in Catmaid
-- number of cells
-- general anatomy, regions, quadrants
+First, we embedded the entire body of the *Mnemiopsis leidyi* cydippid-stage larvae, five days post-fertilization, in Epon resin following high-pressure freezing.
+To reconstruct all synaptic-resolution connections among the cells comprising the apical organ, also known as the aboral organ, we prepared approximately 1,000 ultra-thin serial sections (50 nm thick) from the aboral tip of the larval body embedded in the resin block.
+Using a scanning electron microscope (Gemini SEM 500) with a resolution of 2.0 nm/pixel, we imaged only the region containing the apical organ.
+Subsequently, 620 of these images were stitched and aligned using TrakEM2.
+The resulting volumetric EM dataset had dimensions of 60 μm × 40 μm × 30 μm.
+We traced and annotated all cells in this dataset, ultimately reconstructing 909 cells, each with both nuclei and cell bodies intact.
+
+Cells containing cilia were traced down to the tips of the cilia, and basal bodies were annotated accordingly.
+For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced.
+Each node was tagged, and skeletons were named and assigned multi-level annotations.
+As described later, many neurons we identified formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main trunk or other branches.
+In such cases, branch nodes were placed near the closest existing node and annotated accordingly.
+The entire skeletonized volume was composed of 134,591 nodes.
+However, 88 fragments could not be attached to somata-associated skeletons.
+Most of these fragments represented short skeletal branches that could not be traced beyond gaps or low-quality layers.
+
+Next, we decided to divide the entire apical organ into four broad quadrants to facilitate grouping the identified cells.
+The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the positions of the anal pores.
+This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development.
+We categorized the traced cells into four quadrants and designated these groups as Q1 through Q4 for clarity and consistency.
 
@@
 
-### Classification of cell types
+### Identification of Synaptic Structures in Syncytium Neurons
 
-- ultrastructural criteria (cilia, vesicles etc)
-- number of types, cells per type - table
+Classical neural staining techniques do not provide clear images of the neurons at the aboral pole.
+However, ultrastructural studies have provided morphological evidence of elements resembling neurons, based on synaptic structures, located on the epithelial floor of the apical organs [@Horridge_1964; @Hernandez_Nicaise_1973].
+Based on these findings, we identified synaptic structures characteristic of ctenophore neurons in our data, using previously identified pre-synaptic triad morphological features, such as single-layered vesicles, smooth endoplasmic reticulum, and tightly packed mitochondria.
+In our study, we specifically identified synaptic sites and marked mitochondria (orange) as synaptic nodes.
+The regions where synaptic vesicles align were marked as connectors (light blue arrows) between cells across the membrane.
+These connectors link the synaptic nodes of the pre-synaptic cytoskeleton to the partner nodes of the post-synaptic cytoskeleton.
+As previously reported, the specialization of post-synaptic structures in ctenophores is not apparent, so we recognized synaptic vesicle clusters on the pre-synaptic membrane as the point of reference.
+Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells.
 
-```{r, out.width = "100%", fig.cap="**Figure 2. Title fig 2** Balancer complex: balancer, lithocyte, groove 2 types, bridge - EM pics, (A) legend (B) legend. " }
+As mentioned above, following the identification of the presynaptic structures, we reconstructed three major Syncytial neurons.
+Each of these Syncytial neurons was a cell with multiple nuclei, with membranes fused by continuous plasma membranes.
+These neurons are distinct in morphology from the Syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously reconstructed in 3D [@Burkhardt_2023].
+Our findings represent the second documented discovery of Syncytial neurons in ctenophores.
+Furthermore, these neurons exhibited clear morphological differences when compared to other sensory cells reported in the same study, such as mesogleal neurons and sensory cells with presynaptic structures and cilia (types 1-4)[@Burkhardt_2023].
+Based on their distinct spatial relationships, we were able to classify these three Syncytial neurons into two categories.
+The first type is a larger "AO neuron_Q1234," which possesses four (or possibly six?) nuclei spanning four quadrants.
+It contained X presynaptic structures.
+The second type consists of "AO neuron_Q12" and "AO neuron_Q34," each having two nuclei spanning two quadrants, and each containing X presynaptic structures.
+
+```{r, out.width = "100%", fig.cap="**Figure 2. Title fig 2** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
 #| echo: false
@@
 
-"the larval bridge has not yet developed forks at its ends, as seen in adult Mnemiopsis bridges"[@Tamm_2002]
+### Identification the Gravity-Sensitive Neural Circuit via the Syncytial Neurons Network
 
-Figure3 - dome, bicil, monocil, no-cil, plumose, dense vesicle cells (rename), gland? cells, oligo/multiciliated cells
+We discovered that AO neurons form synaptic connections with each other, and that AO neuron_Q1234 forms self-synapses, or “autapses.” To our knowledge, previous reports have not identified synaptic connections between subepithelial nerve net (SNN) neurons.
+Thus, our results represent the first report of synaptic connectivity between neurons in the apical organ, forming a network in ctenophores (subject to confirmation).
+These AO neuron networks were found to form many presynaptic structures in relation to the gravity-sensing balancer cells.
 
-Plumose cells [@Hernandez_Nicaise_1984] may be pressure sensors [@Tamm_2014b]
+Balancer cells are monociliated cells, and their cilia protrude longer than those of other monociliated cells.
+Moreover, the cilia of balancer cells are bundled into four groups at the center of the apical organ, forming a compound cilium.
+Based on these features, we identified and classified the balancer cells.
+The cellular arrangement differed clearly when viewed in the lateral view, sagittal plane, and tentacular plane, with the cell bodies gathering toward the apical organ in the tentacular plane.
+In each quadrant, the number of cells was as follows: Q1: 37, Q2: 32, Q3: 32, Q4: 28.
+Each cell contained 3 to 10 ?
+mitochondria.
 
-Figure4 - lamellate body complex, LB, intra-multi-ciliated, relationship to balancer
+From AO neuron_Q1Q2, synapses were formed with 6 of the 37 balancer cells in the Q1 region, and 8 of the 32 balancer cells in the Q2 region.
+Similarly, from AO neuron_Q3Q4, synapses were formed with 1 of the 32 balancer cells in the Q3 region, and 5 of the 28 balancer cells in the Q4 region.
+AO neuron_Q1234 formed input synapses with balancer cells in the Q1 region (7 cells), Q2 region (11 cells), Q3 region (6 cells), and Q4 region (10 cells).
+Some balancer cells received inputs from both AO neuron_Q1Q2 or AO neuron_Q3Q4 and AO neuron_Q1234.
+While previous studies have suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez_Nicaise_1974], our data did not reveal any synaptic inputs from balancer cells to AO neurons.
+However, we identified "bridge cells" that actively formed input synapses with AO neurons, highlighting their potential role in the network.
 
-Figure5 - nerve net, different fragments/cells/syntytia, EM, relation to entire organ
+```{r, out.width = "100%", fig.cap="**Figure 3. Title fig 3** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
+#| echo: false
 
+knitr::include_graphics("figures/Figure3.png")
+```
+
+### Identification of Neuron-like Bridge Cells Forming Afferent Synapses in the Gravity-Sensitive Neural Circuit
+
+From our data, we identified several cell groups that form multiple afferent synapses with AO neurons.
+These cells, based on their morphological features, were found to be the bridge cells first described by Tamm et al.
+in 2002 [@Tamm_and_Tamm_2002].
+These cells are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge.
+They originate from the base of paired balancer cells along the tentacle surface and extend across the sagittal plane toward the base of the opposite balancer cells.
+In regions where the mitochondria of the bridge cells are localized (approximately 30%), a presynaptic triad structure, similar to that of AO neurons, was found, containing synaptic vesicles and smooth endoplasmic reticulum.
+
+Our three-dimensional reconstruction data revealed that the bridge cells form two distinct cell groups across the sagittal plane, between the Q1Q2 and Q3Q4 regions.
+In the Q1Q2 region, 14 cells were identified, while in the Q3Q4 region, 12 cells were identified, totaling 26 bridge cells.
+Nearly all of these bridge cells (25 out of 26) exhibited afferent synapses from AO neurons.
+For bridge cells located in the Q1Q2 region, synaptic inputs came primarily from AO neuron_Q1Q2 (11 cells), AO neuron_Q1234 (1 cell), or both (2 cells).
+For bridge cells located in the Q3Q4 region, synaptic inputs were received from AO neuron_Q3Q4 (1 cell), AO neuron_Q1234 (7 cells), or both (1 cell).
+In other words, bridge cells in the Q1Q2 region mainly received input from AO neuron_Q1Q2, while those in the Q3Q4 region received input primarily from AO neuron_Q1234, showing a distinct pattern of input in the two regions.
+
+Some bridge cells also formed afferent synapses with AO neurons.
+For example, bridge cells in the Q1Q2 region formed synapses with AO neuron_Q1Q2 (3 cells) or both AO neuron_Q1Q2 and AO neuron_Q1234 (2 cells).
+Bridge cells in the Q3Q4 region formed synapses with AO neuron_Q3Q4 (1 cell) or AO neuron_Q1234 (6 cells).
+A notable difference was observed between the Q1Q2 and Q3Q4 regions in the proportion of synaptic inputs from bridge cells to AO neurons.
+
+Interestingly, in both regions, bridge cells also formed synapses with adjacent bridge cells.
+However, no synaptic input was found from bridge cells across the sagittal plane to those in the opposite region.
+To analyze the grouped synaptic connectivity graph, we classified cell types within each region, collapsed cells of the same type into a single node, summed the number of synapses, and explored directed pathways to effectors (balancer cell groups).
+The results revealed a feedback pathway from AO neurons through the synaptic connections formed by bridge cells, thereby shedding light on the neural circuit structure involving these cells.
+
+```{r, out.width = "100%", fig.cap="**Figure 4. Title fig 4** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
+#| echo: false
+
+knitr::include_graphics("figures/Figure4.png")
+```
+
+### Control of Balancer Ciliary Movements by Gravity-Sensitive Neural Circuits
+
+To investigate the function of gravity-sensitive circuits, we conducted a comparative analysis of balancer ciliary movements across different regions.
+Previous studies (Tamm, 1980, 1982; Lowe, 1997) have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination.
+Furthermore, it has been suggested that differences in the asymmetry of statolith morphology and the elongation direction of balancer cilia between the tentacular and sagittal planes could result in variations in the force received from the statolith (Tamm, 2014, 2015).
+Building on these findings, we standardized conditions for observing balancer cilia by tilting the microscope by 90 degrees and fixing samples such that the aboral-oral axis was aligned parallel to the vertical stage, with the aboral side facing upward.
+Using connectome analysis, we aimed to evaluate the neural circuit differences by analyzing the control patterns of left and right balancer cilia in the tentacular and sagittal planes.
+Specifically, we hypothesized that differences in control between Q1 (Q2) and Q3 (Q4) regions, as well as between Q1 (Q4) and Q2 (Q3) regions, would reveal distinct functions of AO neurons (e.g., AO neuron Q12/Q34 vs.
+AO neuron Q1234).
+For the sagittal plane, we recorded balancer ciliary movements in the Q1(4) and Q2(3) regions (left and right sides of the image) and, for the tentacular plane, in the Q1(3) and Q2(4) regions, using a high-speed camera (100 fps) for 2 minutes (12,000 frames).
+Regions of interest (ROIs) were defined in areas where brightness changes indicated ciliary beating.
+We measured these brightness changes to generate dynamics graphs of balancer ciliary movements for both planes.
+
+During the 2-minute recordings, balancer cilia exhibited movements that could be fast, slow, or stop abruptly (arrest) and start moving again (re-beat).
+Occasionally, large contraction responses caused the entire cydippid to move out of frame, and data from these moments were excluded.
+We focused on two metrics for left-right comparison: (1) the timing of ciliary arrest and (2) the timing of re-beat initiation following an arrest.
+Due to the low baseline beating frequency of the cilia, rapid synaptic input responses were not evaluated, suggesting that ciliary beat regulation may involve non-synaptic control mechanisms.
+The analysis revealed that both arrest and re-beat events were synchronized between left and right balancers in both planes within a few milliseconds, suggesting that these events are governed by synaptic input from neurons.
+Interestingly, while the re-beat timing showed no significant differences between the sagittal and tentacular planes, arrest timing exhibited greater variability in the tentacular plane.
+Comparing these findings with neural circuit diagrams suggests that in the sagittal plane, the same AO neurons (Q12 or Q34) project to both balancer cell groups, transmitting signals simultaneously and enabling synchronization.
+In contrast, in the tentacular plane, separate AO neurons project to different regions, potentially introducing an additional synaptic step, which may account for the observed variability in arrest timing.
+
+```{r, out.width = "100%", fig.cap="**Figure 5. Title fig 5** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
+#| echo: false
+
+knitr::include_graphics("figures/Figure5.png")
+```
+
+## Discussion
+
+In this study, we traced all the constituent cells of the aboral organ in five-day-old larvae of *Mnemiopsis leidyi* and completed a comprehensive neural map.
+As a result, we identified a novel type of syncytial neuron with a previously unreported morphology.
+Furthermore, we found that three such syncytial neurons project to distinct quadrants, integrating with balancer cell groups to form a gravity-sensing neural circuit.
+This study represents the first complete neural circuit map reported in ctenophores.
+Additionally, our analysis of balancer ciliary movements, based on neural circuit data, revealed differences in ciliary arrest control arising from circuit variations.
+
+### Identification of a Previously Unreported Syncytial Neuron
+
+We identified three syncytial neurons with multiple nuclei in the apical organ of five-day-old *M.
+leidyi* cydippid larvae.
+Burkhurdt et al.
+(2023) previously reported that in one-day-old larvae, the sub-epithelial nerve net (SNN) contains neurons that form syncytia.
+However, the newly discovered syncytial neurons embedded within the apical organ cell cluster did not exhibit the characteristic "pearl necklace-like structure" of the SNN, which consists of microtubules connecting neurons.
+Instead, their morphology was more reminiscent of fiber cells previously described in Placozoa (Mayorova et al., 2021), with cell bodies extending through intercellular spaces.
+These morphological characteristics suggest that the newly identified neurons represent a second distinct type of syncytial neuron in ctenophores.
+
+### Functional Diversification of the Three Syncytial Neurons
+
+The three identified neurons may be classified into two distinct functional types based on their spatial arrangement and their influence on the differential control of balancer ciliary movement between the quadrants.
+In metazoans, a single neuron can directly provide synaptic input to multiple ciliated cells at a distance, synchronizing their movements.
+For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating (Verasztó et al., 2017).
+Additionally, a pair of serotonergic neurons, designated Ser-h1, extend axons that cross at the midline; the left Ser-h1_l predominantly innervates right-sided ciliated cells, while the right Ser-h1_r mainly projects to the left-side ciliated cells.
+Activation of Ser-h1 induces ciliary beating initiation and increases beat frequency.
+
+In the this study, we found that three neurons are closely positioned within the confined space of the apical organ, potentially enabling a single neuron to control distant ciliary movements in a synchronized manner.
+Moreover, the presence of neurons projecting to the same ciliated cells in distinct patterns suggests that these neurons may serve opposing functions—one inhibiting ciliary activity while the other promotes it, or one modulating ciliary beat frequency.
+While our synaptic ultrastructure analysis did not allow us to distinguish excitatory from inhibitory neurons, future electrophysiological investigations will be essential to elucidate their functional properties.
+
+```{r, out.width = "100%", fig.cap="**Figure 6. Title fig 6** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
+#| echo: false
+
+knitr::include_graphics("figures/Figure6.png")
+```
+
+### Paracrine Signaling and Synaptic Transmission from Sensory to Neural Cells
+
+Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores (Horridge, 1965; Krisch, 1973; Aronova, 1974; Vinnikov, 1974).
+By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the apical organ.
+However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets.
+In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the apical organ (Sachkova et al., 2021; Hayakawa et al., 2022).
+These findings suggest that many sensory cells in the apical organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis (Jekely, 2021).
+While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.
+
+Conversely, previous research has demonstrated that mechanoreceptive sensory cells equipped with cilia (referred to as type 3 and type 4 cells) in ctenophores form direct synaptic connections with the sub-epithelial nerve net (SNN) and mesogleal neurons (Burkhurdt et al., 2023).
+Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells (Hertwig, 1880; Chun, 1880; Eimer, 1880).
+
+Surprisingly, our study identified bridge cells as the primary cell type forming synaptic inputs onto neurons in the apical organ.
+Bridge cells were first described by Tamm et al.
+(2002) based on their morphological characteristics, with cellular projections extending to the basal region of balancer cilia [@Tamm_2002] (Tamm & Tamm, 2002).
+Our findings suggest that bridge cells are electrically coupled to balancer cilia and provide rapid feedback to neurons via synaptic transmission.
+Future studies analyzing differential responses to various sensory stimuli will not only elucidate the functional roles of bridge cells but also provide insights into the evolutionary relationship between transmission speed and synapse development.
+
+### Circuit Variability as a Driver of Behavioral Diversity
+
+Our study suggests that differences in neural circuits contribute to variations in the timing of ciliary arrest, which may play a role in modulating swimming direction in ctenophores (Richard Satterlie, 2015).
+This finding provides evidence that increased circuit complexity contributes to behavioral diversification.
+By integrating multiple neurons and altering circuit structure, information can be branched, allowing for diverse motor patterns within the framework of fast synaptic transmission.
+
+The neural cells identified in the apical organ exhibit a divergent feedforward connectivity pattern, where a single neuron forms presynaptic connections with multiple downstream targets.
+Such a circuit design minimizes unwanted signal interference, reduces processing delays, and enables relatively fast response execution (Luo, 2021).
+Moreover, the presence of multiple divergent feedforward circuits suggests that slight variations in the synchronization and phase delay of ciliary beating (within the range of a few hundred milliseconds) may have evolved as a mechanism for generating behavioral diversity.
+
+From a broader neuroanatomical perspective, nervous system structures across various animals have been shaped through selective optimization to adapt to environmental constraints (Bullmore & Sporns, 2012; Perin et al., 2011).
+For example, in nematodes, the spatial arrangement of ganglia is optimized to minimize the total length of neural wiring (White et al., 1986).
+Similarly, in ctenophores, understanding how transmission via syncytial neurons and synaptic connections has been selectively optimized at the circuit level will provide crucial insights into the evolutionary scaling of neural networks from early nervous systems to complex brains (Farnworth and Montgomery, 2024).
+In particular, this research may help elucidate how circuit modularization and layered structures emerged over evolutionary time.
+
+## Materials and Methods
+
+### Data and Code Availability
+
+The following EM image stacks, including all traces and annotations, are available at https://catmaid.jekelylab.ex.ac.uk.
+These datasets can be queried either within CATMAID or through the CATMAID application programming interface (API) using the catmaid or natverse packages (Bates et al., 2020).
+The dataset encompasses all EM images (in JPG format), skeletons, meshes, node tags, connectors, and annotations.
+Additionally, we provide all R scripts used for data acquisition and figure generation (Jokura et al., 2025).
+All plots, figures (including anatomical renderings), and figure layouts can be fully reproduced using the provided R scripts.
+While the scripts are mostly organized by figure, some general-purpose scripts—for tasks such as loading libraries, accessing CATMAID data, and defining common functions—are shared across multiple figures.
+
+### Specimen Preparation, Transmission Electron Microscopy, and Image Processing
+
+Larvae of Mnemiopsis leidyi (5 days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) and immediately transferred to liquid nitrogen for storage.
+The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica).
+Cryo-substitution was performed by gradually raising the temperature, and the samples were embedded in epoxy resin.
+Serial sections of 50 nm thickness were prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º DiATOME diamond knife.
+To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the carbon film surface negatively charged.
+Section ribbons were collected on the prepared glass slides, slowly dried to ensure proper stretching, and firmly adhered to the glass surface.
+The sections on glass were stained with UranyLess and lead citrate (Reynolds) using airless staining bottles (Delta Microscopies).
+The glass slides were mounted on STEM-specific stages (Zeiss) using Copper Foil EMI Shielding Tape (3M).
+Imaging was performed using a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5 imaging software (Zeiss).
+The full dataset consisted of 620 serial sections, each 50 nm thick.
+The imaging resolution was 2.8 nm/pixel, using electron holography transmission at an acceleration voltage of 1.5 kV (in-lens detector, dwell time: 3 µs).
+
+### Image-Stack Realignment and Spatial Data Transformation in CATMAID
+
+To process the image stack, we utilized the TrakEM2 plugin of FIJI (ImageJ) (version 2.0.0-rc-15/1.49k / Java 1.6.0_24 (64-bit) – 2014).
+A project was created, and all TIFF images were imported using the "import sequence as grid" function.
+Subsequently, the following filters were applied sequentially: Invert, Equalize Histogram, and Gaussian Blur.
+
+The alignment process consisted of three stages, each progressively refining the spatial accuracy:
+
+Rigid Alignment Initially, a rigid alignment was performed with the following parameters: least squares mode (linear feature correspondences), encompassing the entire layer range with the first layer as the reference.
+Only visible images were used, without propagation.
+The alignment was executed with an initial Gaussian blur of 1.6 pixels, three steps per scale octave, a minimum image size of 512 pixels, and a maximum of 2048 pixels.
+Additional parameters included a feature descriptor size of 8, orientation bins of 8, and a closest ratio of 0.92.
+The alignment allowed clearing the cache, using 32 feature extraction threads, a maximal alignment error of 100 pixels, a minimal inlier ratio of 0.20, and a minimum of 12 inliers.
+The expected and desired transformations were set to rigid, with testing multiple hypotheses (tolerance: 5.00 pixels) and considering up to 5 neighboring layers, giving up after 5 failures.
+Regularization was performed with a maximal iteration of 1000, a maximal plateau width of 200, and a rigid lambda of 0.10.
+
+Affine Alignment Following the rigid alignment, an affine alignment was applied using similar parameters, except the expected and desired transformations were set to affine.
+The minimal image size was reduced to 64 pixels, while the other parameters (e.g., Gaussian blur, feature descriptor size, inliers, and testing hypotheses) remained unchanged to ensure consistent processing.
+
+Elastic Alignment Two iterations of elastic alignment were performed to fine-tune the spatial data.
+Key parameters included a block matching layer scale of 0.05, a search radius of 200 pixels, a block radius of 2000 pixels, and a resolution of 60.
+Correlation filters were employed with a minimal PMCC r of 0.10, a maximal curvature ratio of 1000, and a maximal second-best r/best r of 0.90.
+A local smoothness filter was applied with the approximate local transformation set to affine, a local region sigma of 1000 pixels, and an absolute maximal local displacement of 10 pixels (relative maximal displacement: 3.00).
+Pre-aligned layers were tested for up to 4 neighboring layers.
+The elastic alignment used a rigid approximation, maximal iterations of 3000, a plateau width of 200, spring mesh stiffness of 0.01, and a maximal stretch of 2000 pixels.
+A legacy optimizer was employed to enhance performance.
+
+After each alignment stage, the project was saved as an XML file under a unique name to preserve iterative progress.
+Finally, the images were exported from FIJI using TrakEM2 in a format compatible with CATMAID.
+
+### Neuron Tracing, Synapse Annotation, and Review
+
+The reconstruction of cells and connectivity was performed using CATMAID (Saalfeld et al., 2009), analyzing a total of 620 serial sections.
+To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server (Saalfeld et al., 2009; Schneider-Mizell et al., 2016).
+To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset.
+At each nuclear center, the radius of the single node was adjusted according to the size of the cell body in that specific layer.
+All skeletons were rooted at their respective cell bodies, and nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria.
+Most synapses could be verified across consecutive sections, ensuring accurate annotation and connectivity mapping.
+
+### Cell Nomenclature and Annotations
+
+Each cell is assigned a specific type name based on its category (balancer, biC, bridge, bristle, dense vesicle cell, dome, EF, IcMC, LB, lithocyte, monoC, non-C, plumose cell, SNN, ssCG, SSN, stCG), resulting in a total of 17 cell types.
+Additionally, a quadrant number is appended to the cell type name following an underscore ("\_") to indicate the cell body’s location.
+For example, cells located in the first quadrant are annotated with "Q1." If a cell is situated between the first and second quadrants, the annotation "Q1Q2" is applied.
+
+Cells of the same type within the same region are further distinguished by serial numbering.
+Each cell is also assigned multiple annotations, which can be utilized to query the database via CATMAID or the CATMAID API (e.g., using the R catmaid package).
+These annotations provide a structured and precise framework for identifying and analyzing specific cells, facilitating robust data integration and retrieval within the dataset.
+
+This system ensures consistent and accurate tracking of cellular data, enabling comprehensive analysis of cellular structures and functions across regions.
+
+### Analysis of Balancer Cilia
+
+We conducted the analysis using the cydippid stage of Mnemiopsis leidyi at five days post-fertilization.
+A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass.
+Filtered natural seawater was introduced under the coverslip, along with the cydippids.
+By gently moving the coverslip, the desired orientation for observation was adjusted, and once positioned, slight pressure was applied to immobilize the cydippids.
+To stabilize the statolith's position, the microscope was tilted 90 degrees, arranging the stage vertically.
+The movement of the balancer cilia was observed using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS).
+Observations were made using a 40x objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27) with glycerine immersion.
+For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals.
+The camera and LED strobe system were operated using the Video Capture Software BohNavi.
+Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED.
+To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph project in Fiji software.
+=======
 Figure 6 - cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.
 
-## Synapses
+### Synapses
 
-"synaptic regions with the characteristic presynaptic triad mor-
-phology of ctenophore nerves (i.e., a single layer of
-vesicles, smooth ER sac, and closely apposed mito-
-chondria [@Horridge_1964; @Hernandez_Nicaise_1973] Hernandez-
-Nicaise, 1973)"
+"synaptic regions with the characteristic presynaptic triad mor- phology of ctenophore nerves (i.e., a single layer of vesicles, smooth ER sac, and closely apposed mito- chondria [@Horridge_1964; @Hernandez_Nicaise_1973] Hernandez- Nicaise, 1973)"
 
-Figure 7 - synapses, reconstruction, annotation, numbers, distribution of partners, example EM (figure supplement - lots of EM), mitochondria
-(Fig suppl - mitochondrion stats - Sanja)
+Figure 7 - synapses, reconstruction, annotation, numbers, distribution of partners, example EM (figure supplement - lots of EM), mitochondria (Fig suppl - mitochondrion stats - Sanja)
 
@@
 
-Figure 8 - connectome
-cell-type based, quadrant based, nerve net syntycia, bridge, quadrants
-(Fig suppl - cell-based network)
+Figure 8 - connectome cell-type based, quadrant based, nerve net syntycia, bridge, quadrants (Fig suppl - cell-based network)
 
-Tamm&Tamm observed synapses from neurites to bridge cells (2002).
-and in Mnemiopsis larvae observed synapses from bridge cells onto other cells
+Tamm&Tamm observed synapses from neurites to bridge cells (2002). and in Mnemiopsis larvae observed synapses from bridge cells onto other cells
 
@@
 The title and legend can also be edited, as will as the width of the output figure.
+
+### Connectivity
+
+### Physiology?
+
+### Discussion
 
@@
 
-## Sourcing code and working with variable
+## Materials and Methods
 
-```{r}
-#| include: false
-#| eval: true
-#source(here::here("analysis/scripts/statistics_for_paper.R"))
-```
+### Volume electron microscopy
 
-The 'analysis/scripts/statistics_for_paper.R' script is sourced and it runs but the output is not included in the knitted output.
-But we can access the variables defined in the sourced script simply by adding \` r var_name \` between \` backticks, in this case max_PRC value is (now this number comes from our sourced script).
+### Imaging of balancer activity
 
-If we update the data, the script can recalculate the variable we want to refer to in the text and update the number.
+### Cell-type annotation and connectome analysis
 
-## Materials and Methods
+### Data and code availability
+
+>>>>>>> Stashed changes
 
@@
 
-We would like to thank the Jekely lab for the R project template (<https://github.com/JekelyLab/new_paper_template>) we used to write this paper. This work was funded by ...
+This work was funded by ...
 JSPS...
-ERC..
-Others ...
+This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 101020792).
 
-## References
+## References {.unnumbered}
+
+::: {#refs}
+:::
+
+```{=html}
+<!--
+## See https://github.com/hypothesis/client
+## Add commenting functionality with this script
+-->
+```
+```{=html}
+<script src="hypothes.is/embed.js" async></script>
+```
+\newpage
+
+## Supplementary material {.unnumbered}

### 2025-02-13 manuscript/Manuscript.qmd
MS: update refs

@@
 
-Recent comparative genomic studies (Ryan et al., 2013; Whelan et al., 2015; Whelan et al., 2017; Li et al., 2021; Whelan and Halanych, 2023) and chromosome-level synteny analyses across different animal phyla (Schultz et al., 2023) suggest that ctenophores are the sister group to all other metazoans and were the first animals to evolve a nervous system.
-This has led to attempts to compare genes and molecular modules in ctenophores using single-cell analyses, in which neuropeptides have been employed as neuronal markers (Sebé-Pedrós et al., 2018; Sachkova et al., 2021; Hayakawa et al., 2022).
+Recent comparative genomic studies [@Ryan2013; @Whelan2015a; @Whelan2017; @Whelan2023; @Li2021] and chromosome-level synteny analyses across different animal phyla [@Schultz2023] suggest that ctenophores are the sister group to all other metazoans and were the first animals to evolve a nervous system.
+This has led to attempts to compare genes and molecular modules in ctenophores using single-cell analyses, in which neuropeptides have been employed as neuronal markers [@Sebé-Pedrós2018; @Sachkova2021; @Hayakawa2022].
 However, the homologous relationships between ctenophore genes and those of other animals are highly divergent, limiting the feasibility of sequence-based comparisons.
@@
 In many bilaterian and non-bilaterian animals, an apical nervous system is located at one pole of the body, forming a specialized sensory structure that governs physiological functions and motor activities.
-In ctenophores, ultrastructural analyses have revealed that the aboral pole contains a diverse array of sensory cells and neuronal components (Hernandez-Nicaise, 1968, 1973, 1974; Tamm, 1982).
-Additionally, immunostaining using an anti-tyrosylated α-tubulin antibody has identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control (Jager et al., 2011).
+In ctenophores, ultrastructural analyses have revealed that the aboral pole contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974] <!--# (Tamm, 1982 ?) -->.
+Additionally, immunostaining using an anti-tyrosylated α-tubulin antibody has identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].
 
@@
 The apical organ of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014].
-The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014] (Tamm, 1982).
+The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014] <!--# (Tamm, 1982 ?) -->.
 Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron].
@@
 To reconstruct all synaptic-resolution connections among the cells comprising the apical organ, also known as the aboral organ, we prepared approximately 1,000 ultra-thin serial sections (50 nm thick) from the aboral tip of the larval body embedded in the resin block.
-Using a scanning electron microscope (Gemini SEM 500) with a resolution of 2.0 nm/pixel, we imaged only the region containing the apical organ.
+Using a scanning electron microscope (Zeiss Gemini 500) with a resolution of 2.0 nm/pixel, we imaged only the region containing the apical organ.
 Subsequently, 620 of these images were stitched and aligned using TrakEM2.
@@
 
-Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores (Horridge, 1965; Krisch, 1973; Aronova, 1974; Vinnikov, 1974).
+Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores [@HORRIDGE1965; @vinnikov1974e] <!--#  Krisch, 1973; Aronova, 1974 ?
+--> ).
 By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the apical organ.
 However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets.
-In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the apical organ (Sachkova et al., 2021; Hayakawa et al., 2022).
+In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the apical organ [@Sachkova2021; @Hayakawa2022].
 These findings suggest that many sensory cells in the apical organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis (Jekely, 2021).
@@
 
-Conversely, previous research has demonstrated that mechanoreceptive sensory cells equipped with cilia (referred to as type 3 and type 4 cells) in ctenophores form direct synaptic connections with the sub-epithelial nerve net (SNN) and mesogleal neurons (Burkhurdt et al., 2023).
-Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells (Hertwig, 1880; Chun, 1880; Eimer, 1880).
+Conversely, previous research has demonstrated that mechanoreceptive sensory cells equipped with cilia (referred to as type 3 and type 4 cells) in ctenophores form direct synaptic connections with the sub-epithelial nerve net (SNN) and mesogleal neurons [@Burkhardt2023].
+Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells <!--# add these to bib file -->(Hertwig, 1880; Chun, 1880; Eimer, 1880).
 
@@
 Bridge cells were first described by Tamm et al.
-(2002) based on their morphological characteristics, with cellular projections extending to the basal region of balancer cilia [@Tamm_2002] (Tamm & Tamm, 2002).
+based on their morphological characteristics, with cellular projections extending to the basal region of balancer cilia [@Tamm_2002].
 Our findings suggest that bridge cells are electrically coupled to balancer cilia and provide rapid feedback to neurons via synaptic transmission.
@@
 
-Our study suggests that differences in neural circuits contribute to variations in the timing of ciliary arrest, which may play a role in modulating swimming direction in ctenophores (Richard Satterlie, 2015).
+Our study suggests that differences in neural circuits contribute to variations in the timing of ciliary arrest, which may play a role in modulating swimming direction in ctenophores [@Satterlie2015a].
 This finding provides evidence that increased circuit complexity contributes to behavioral diversification.
@@
 For example, in nematodes, the spatial arrangement of ganglia is optimized to minimize the total length of neural wiring (White et al., 1986).
-Similarly, in ctenophores, understanding how transmission via syncytial neurons and synaptic connections has been selectively optimized at the circuit level will provide crucial insights into the evolutionary scaling of neural networks from early nervous systems to complex brains (Farnworth and Montgomery, 2024).
+Similarly, in ctenophores, understanding how transmission via syncytial neurons and synaptic connections has been selectively optimized at the circuit level will provide crucial insights into the evolutionary scaling of neural networks from early nervous systems to complex brains [@Farnworth2024].
 In particular, this research may help elucidate how circuit modularization and layered structures emerged over evolutionary time.
@@
 The following EM image stacks, including all traces and annotations, are available at https://catmaid.jekelylab.ex.ac.uk.
-These datasets can be queried either within CATMAID or through the CATMAID application programming interface (API) using the catmaid or natverse packages (Bates et al., 2020).
+These datasets can be queried either within CATMAID or through the CATMAID application programming interface (API) using the catmaid or natverse packages [@Bates2020].
 The dataset encompasses all EM images (in JPG format), skeletons, meshes, node tags, connectors, and annotations.
@@
 The reconstruction of cells and connectivity was performed using CATMAID (Saalfeld et al., 2009), analyzing a total of 620 serial sections.
-To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server (Saalfeld et al., 2009; Schneider-Mizell et al., 2016).
+To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server [@Saalfeld2009; @Schneider-Mizell2016].
 To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset.
@@
 
-We conducted the analysis using the cydippid stage of Mnemiopsis leidyi at five days post-fertilization.
+We conducted the analysis using the cydippid stage of *Mnemiopsis leidyi* at five days post-fertilization.
 A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass.
@@
 To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph project in Fiji software.
-=======
+
 Figure 6 - cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.
@@
 
-"synaptic regions with the characteristic presynaptic triad mor- phology of ctenophore nerves (i.e., a single layer of vesicles, smooth ER sac, and closely apposed mito- chondria [@Horridge_1964; @Hernandez_Nicaise_1973] Hernandez- Nicaise, 1973)"
+"synaptic regions with the characteristic presynaptic triad mor- phology of ctenophore nerves (i.e., a single layer of vesicles, smooth ER sac, and closely apposed mito- chondria [@Horridge_1964; @Hernandez_Nicaise_1973] "
 
@@
 
->>>>>>> Stashed changes
-
 ## Acknowledgements
@@
 ```
+
 ```{=html}
@@
 ```
+
 \newpage

### 2025-02-13 analysis/scripts/Figure1.R
Fig1: add quadrants

@@ -3,120 +3,149 @@
 # source packages and functions -----------------
 source("analysis/scripts/packages_and_functions.R")
 
-<<<<<<< Updated upstream
-
-# assemble figure -------------------------------------------------------------
-
-library(jpeg)
-=======
 # load cells ----------------------------------
 
 balancer <- read_smooth_neuron("celltype:balancer")
-
-Q1 <- nlapply(
-  read.neurons.catmaid("Q1", pid = 35),
-  function(x)
-  smooth_neuron(x, sigma = 1000)
-  )
-
-Q2 <- nlapply(
-  read.neurons.catmaid("Q2", pid = 35),
-  function(x)
-  smooth_neuron(x, sigma = 1000)
-  )
-
-Q3 <- nlapply(
-  read.neurons.catmaid("Q3", pid = 35),
-  function(x)
-  smooth_neuron(x, sigma = 1000)
-  )
-
-Q4 <- nlapply(
-  read.neurons.catmaid("Q4", pid = 35),
-  function(x)
-  smooth_neuron(x, sigma = 1000)
-  )
-
-with_soma <- nlapply(
-  read.neurons.catmaid("with_soma", pid = 35),
-  function(x)
-  smooth_neuron(x, sigma = 1000)
-  )
-
+Q1 <- read_smooth_neuron("Q1")
+Q2 <- read_smooth_neuron("Q2")
+Q3 <- read_smooth_neuron("Q3")
+Q4 <- read_smooth_neuron("Q4")
+with_soma <- read_smooth_neuron("with_soma")
+lithocyte <- read_smooth_neuron("celltype:lithocyte")
 
 # plot balancer and quadrants ----------------------
 
+plot_background()
 
+plot3d(
+  balancer, soma = T, lwd = 1, add = T, 
+  alpha = 0.9, col = Okabe_Ito[5],
+  WithConnectors = F
+  )
+plot3d(
+  lithocyte, soma = T, lwd = 1, add = T, 
+  alpha = 0.8, col = Okabe_Ito[8],
+  WithConnectors = F
+)
+plot3d(
+  Q1, soma = T, lwd = 1, add = T, 
+  alpha = 0.15, col = Okabe_Ito[1]
+)
+plot3d(
+  Q2, soma = T, lwd = 1, add = T, 
+  alpha = 0.15, col = Okabe_Ito[2]
+)
+plot3d(
+  Q3, soma = T, lwd = 1, add = T, 
+  alpha = 0.15, col = Okabe_Ito[6]
+)
+plot3d(
[…]

### 2025-02-13 analysis/scripts/packages_and_functions.R

@@ -16,7 +16,7 @@ library(igraph)
 library(tidygraph)
 library(av)
 library(jpeg)
-
+library(magick)
 
 
 # catmaid connection, needs username, password AND token - weird!
@@ -27,7 +27,7 @@ library(jpeg)
   #for this we configure to http/1.1
   conn_http1 = catmaid_login(conn=conn, config=httr::config(ssl_verifypeer=0, http_version=1))
 }
-
+"manuscript/pictures/balancer_Q1_4_aboral_view.png"
 system("git submodule init")
 system("git submodule update")
 source("rcatmaid_functions_library/functions.R")
@@ -54,6 +54,13 @@ plot_background <- function(x){
   par3d(zoom=0.75)
   nview3d("frontal", extramat=rotationMatrix(1.2, 0, 0, 1))
   par3d(windowRect = c(0, 0, 800, 800)) #resize for frontal view
+  nview3d("anterior", extramat = rotationMatrix(1.05, 250, -200, 1000))
+  par3d(zoom=0.7)
+  #y-axis clip
+  clipplanes3d(1, 0, 0, -11500)
+  #x-axis clip
+  clipplanes3d(0, 1, 0, -24000)
+  
 }
 
 plot_background_ventral <- function(x){
@@ -145,7 +152,13 @@ crop_catmaid <- function(tagname,
           body = list(
             stack_ids=28,
             min_x=x-half_bb_size_x,
-            min_y=y-half_bb_size_y,
+            min_y=ynview3d("anterior", extramat = rotationMatrix(1.05, 250, -200, 1000))
+            par3d(zoom=0.7)
+            #y-axis clip
+            clipplanes3d(1, 0, 0, -11500)
+            #x-axis clip
+            clipplanes3d(0, 1, 0, -24000)
+            -half_bb_size_y,
             min_z=z-half_bb_size_z,
             max_x=x+half_bb_size_x,
             max_y=y+half_bb_size_y,
@@ -207,3 +220,35 @@ get_celltypes <- function(pid) {
   return(celltypes)
 }
 
+
+#function to add a text label to a neuron and a line to one of the neurons of the neuronlist starting from the soma
+add_label_with_line <- function(neuronlist, offsetx, offsety, offsetz, side, label){
+  cords_left = unlist(as_tibble(soma(neuronlist)) %>%
+                        arrange(X) %>% slice(1) %>% as.list()
+  )
[…]

### 2025-02-14 manuscript/Manuscript.qmd
Figure legends

@@
 
-```{r, out.width = "100%", fig.cap="**Figure 1. Title fig 1** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
-#| echo: false
-
-knitr::include_graphics("figures/Figure1.png")
-```
+![**Figure 1. Morphology of the apical organ in *Mnemiopsis leidyi***\
+(A) Whole-body image of a 5-day-old M. leidyi cydippid larva in lateral view, observing the tentacular plane. The boxed region indicates the apical organ. Scale bar: 100 µm.\
+(B) Enlarged views of the apical organ from different perspectives in a 5-day-old M.
+leidyi cydippid larva.
+The left panel shows the apical organ from a aboral view.
+The middle panel presents a sagittal view of the apical organ.
+The right panel provides a tentacular plane view of the apical organ.
+Scale bar: 10 µm.\
+(C) Schematic diagram of the serial sectioning process.
+The entire cydippid larva was embedded in resin, trimmed around the apical organ, and sectioned into serial ultrathin sections, which were collected on glass slides in a ribbon-like arrangement.\
+(D) Example of cell tracing using the collaborative annotation toolkit CATMAID for large-scale electron microscopy image datasets.
+The spherical objects indicate nuclear positions, while the lines represent the traced cell centers.
+Scale bar: 10 µm.\
+(E) Three-dimensional reconstruction of all cells composing the apical organ, displayed from different perspectives.
+The left panel shows a aboral view of the apical organ.
+The middle panel presents a sagittal plane view.
+The right panel provides a tentacular plane view.
+Cells are color-coded according to their types.
+Scale bar: 10 µm.](figures/Figure1.png)
 
@@
 
-```{r, out.width = "100%", fig.cap="**Figure 2. Title fig 2** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
-#| echo: false
-
-knitr::include_graphics("figures/Figure2.png")
-```
+![**Figure 2. Reconstruction of Three Syncytial Neurons in the Apical Organ**\
+(A) Representative electron micrographs of presynaptic triad structures within the dataset.
+The positions of mitochondria are marked in orange.
+Postsynaptic target cells were inferred based on the locations of synaptic vesicles and the endoplasmic reticulum, with light blue arrows indicating the direction across the cell membrane.
+The left panel illustrates a monadic structure, while the right panel shows a polyadic (or dyadic) structure.
+Scale bar: 500 nm.\
+(B) 3D reconstruction of the syncytial apical organ (AO) neuron Q1Q2Q3Q4, which spans all four quadrants and contains multiple nuclei (light blue).
+The spheres represent the positions of individual nuclei.
+The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+(C) 3D reconstruction of two syncytial apical organ neurons, AO neuron Q1Q2 (pink) and AO neuron Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei.
+The spheres indicate the positions of the nuclei.
+The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+(D) Localization of mitochondria within AO neuron Q1Q2Q3Q4.
+Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.](figures/Figure2.png)
 
@@
 
-```{r, out.width = "100%", fig.cap="**Figure 3. Title fig 3** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
-#| echo: false
-
-knitr::include_graphics("figures/Figure3.png")
-```
+![**Figure 3. Reconstruction of Balancer Ciliated Cells in the Apical Organ**\
+(A) Connectivity matrix of AO neurons.
+Columns represent presynaptic AO neurons, rows represent postsynaptic AO neurons, and each number indicates the number of synapses.
+AO neurons are interconnected via synapses, and AO neuron Q1234 forms a self-synapse, autapse.\
+(B) Three-dimensional reconstruction of balancer ciliated cells in all four quadrant regions.
+Each quadrant contains approximately 30 cells.
+All balancer cilia are monocilia.
+The spheres indicate the positions of the nuclei.
+The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+(C) Synaptic connections from AO neurons to balancer ciliated cells.
+The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view.
+Balancer ciliated cells are depicted in light gray, AO neuron Q1234 in black, AO neuron Q12 in orange, and AO neuron Q34 in blue.\
+(D) Connectivity map of three AO neurons (Q1234 in blue, Q12 in pink, and Q34 in orange) and balancer ciliated cells (light blue).
+The thickness of the arrows and the numbers correspond to the number of synapses.
+Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.](figures/Figure3.png)
 
@@
 
-```{r, out.width = "100%", fig.cap="**Figure 4. Title fig 4** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
-#| echo: false
-
-knitr::include_graphics("figures/Figure4.png")
-```
+![**Figure 4. Reconstruction of bridge cells and gravity-sensing neural circuits in the apical organ**\
+(A) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants.
+The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color.
+The morphology of individual bridge cells extending across opposite quadrant regions is depicted.
+The spheres represent the positions of individual nuclei.
+The left panel shows an aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+(B) Mitochondrial localization within bridge cells and associated presynaptic triad structures.
+Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.
+The left panel shows a dorsal view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+(C) Synaptic connections between AO neurons and bridge cells.
+The positions of synapses from AO neurons to bridge cells (red) and from bridge cells to AO neurons (light blue) are indicated.
+The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view.
+Bridge cells are shown in light gray, AO neurons in Q1234 are in black, AO neurons in Q12 are in orange, and AO neurons in Q34 are in blue.\
+(D) Connectivity matrix of the gravity-sensing neural circuit.
+Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups.
+The numbers and varying shades of blue correspond to the number of synapses.\
+(E) Complete connectivity map of the gravity-sensing neural circuit.
+Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels.
+The thickness of the arrows and the numerical values indicate the number of synapses.
+AO neurons in Q1234 are shown in black, AO neurons in Q12 in yellow, AO neurons in Q34 in blue, balancer cell groups in gray, and bridge cell groups in pink.](figures/Figure4.png)
 
@@
 
-```{r, out.width = "100%", fig.cap="**Figure 5. Title fig 5** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
-#| echo: false
-
-knitr::include_graphics("figures/Figure5.png")
-```
+![**Figure 5. Differences in Balancer Cilia Movement Control Due to Variations in the Gravity-Sensing Neural Circuit Pathways**\
+(A) Schematic diagram of differential interference contrast (DIC) microscopy setup for imaging balancer cilia movement.
+The microscope was tilted 90 degrees so that the stage was positioned vertically.
+A monochrome CMOS camera sensitive to near-infrared (NIR) light was used, synchronized with an 850 nm strobe light source.
+The movement of the balancer cilia was recorded using a 40× objective lens.\
+(B, C) Sagittal plane view (B) and tentacular plane view (C) of the apical organ under the microscope.
+The direction of gravity is always oriented downward in the images.
+The two balancer cilia being compared are indicated by arrowheads.\
+(D, E) Connectivity maps of the gravity-sensing neural circuit corresponding to Figure 3E.
+These maps emphasize the differences in neural circuits in the sagittal plane (D) and tentacular plane (E), focusing on the balancer cell groups (blue hexagons).
+AO neuron Q1234 is shown in green, while AO neurons Q12 and Q34 are shown in pink.\
+(F) Kymograph patterns comparing the arrest and re-beat of balancer cilia movement.
+The left and right balancer cilia movements are compared in both the sagittal plane (left) and tentacular plane (right).
+The direction and length of the arrows indicate time.\
+(G) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia.
+S and T represent the sagittal plane and tentacular plane, respectively.](figures/Figure5.png)
 

### 2025-02-17 manuscript/Manuscript.qmd
Materials and Methods
Fig6 legend

@@
 
-```{r, out.width = "100%", fig.cap="**Figure 6. Title fig 6** pic of a larva, boxed region of AO, schematic, catmaid pic with slice, dimensions, quadrants, regions, (A) legend (B) legend. " }
-#| echo: false
-
-knitr::include_graphics("figures/Figure6.png")
-```
+![**Figure 6. Comparison of Neural Circuits Regulating Ciliary Movement in *M. leidyi* (Ctenophore) and *P. dumerilii* (Annelid)**\
+Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation.
+Ciliated cells are shown in gray.
+Synapses are indicated by arrows, with magenta representing synapses that induce ciliary arrest and blue representing synapses that induce ciliary re-beat or an increase in ciliary beating frequency.\
+(A) Gravity-sensing neural circuit of *M.
+leidyi*, as revealed in this study.
+Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying their potential involvement in feedback mechanisms between neurons and ciliated cells.\
+(B) Neural circuit regulating prototroch ciliary movement in *P.
+dumerilii*.
+Serotonergic neurons (Ser-h1 neurons, blue) activate ciliary movement, while cholinergic neurons (MC neurons, magenta) induce ciliary arrest.
+Arrows directed toward ciliated cells in the head prototroch are color-coded accordingly.](figures/Figure6.png)
 
@@
 
+Figure 6 - cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.
+
+### Synapses
+
+"synaptic regions with the characteristic presynaptic triad mor- phology of ctenophore nerves (i.e., a single layer of vesicles, smooth ER sac, and closely apposed mito- chondria [@Horridge_1964; @Hernandez_Nicaise_1973] "
+
+Figure 7 - synapses, reconstruction, annotation, numbers, distribution of partners, example EM (figure supplement - lots of EM), mitochondria (Fig suppl - mitochondrion stats - Sanja)
+
+Gap juntions? Innexins are expressed in the apical organ in specific patches of cells in each quadrant [@Ortiz_2023]
+
+Ctenophore synapses [@Hernandez_Nicaise_1973]
+
+We found no synapses between bridge and balancer, in agreement with Tamm [@Tamm_2002].
+
+Figure 8 - connectome cell-type based, quadrant based, nerve net syntycia, bridge, quadrants (Fig suppl - cell-based network)
+
+Tamm&Tamm observed synapses from neurites to bridge cells (2002). and in Mnemiopsis larvae observed synapses from bridge cells onto other cells
+
+Inserting Figures
+
+You can add your figures into the rendered document.
+We saved the figures into /manuscript/figures or /manuscript/figure_supplements and can insert them from there.
+We use knitr::include_graphics for this.
+The title and legend can also be edited, as will as the width of the output figure.
+
+### Connectivity
+
+### Physiology?
+
+### Discussion
+
+## Equations
+
+Equations can also be inserted, Insert -\> Display Math:
+
+$$
+\bar{X} = \frac{\sum_{i=1}^{n} x_{i}}{n}
+$$
+
 ## Materials and Methods
-
-### Data and Code Availability
-
-The following EM image stacks, including all traces and annotations, are available at https://catmaid.jekelylab.ex.ac.uk.
-These datasets can be queried either within CATMAID or through the CATMAID application programming interface (API) using the catmaid or natverse packages [@Bates2020].
-The dataset encompasses all EM images (in JPG format), skeletons, meshes, node tags, connectors, and annotations.
-Additionally, we provide all R scripts used for data acquisition and figure generation (Jokura et al., 2025).
-All plots, figures (including anatomical renderings), and figure layouts can be fully reproduced using the provided R scripts.
-While the scripts are mostly organized by figure, some general-purpose scripts—for tasks such as loading libraries, accessing CATMAID data, and defining common functions—are shared across multiple figures.
 
@@
 
-### Cell Nomenclature and Annotations
+### Cell-type Nomenclature, Annotations and Connectome Analysis
 
@@
 
-### Analysis of Balancer Cilia
+### Imaging of Balancer Cilia activity
 
@@
 
-Figure 6 - cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.
+### Data and Code Availability
 
-### Synapses
-
-"synaptic regions with the characteristic presynaptic triad mor- phology of ctenophore nerves (i.e., a single layer of vesicles, smooth ER sac, and closely apposed mito- chondria [@Horridge_1964; @Hernandez_Nicaise_1973] "
-
-Figure 7 - synapses, reconstruction, annotation, numbers, distribution of partners, example EM (figure supplement - lots of EM), mitochondria (Fig suppl - mitochondrion stats - Sanja)
-
-Gap juntions? Innexins are expressed in the apical organ in specific patches of cells in each quadrant [@Ortiz_2023]
-
-Ctenophore synapses [@Hernandez_Nicaise_1973]
-
-We found no synapses between bridge and balancer, in agreement with Tamm [@Tamm_2002].
-
-Figure 8 - connectome cell-type based, quadrant based, nerve net syntycia, bridge, quadrants (Fig suppl - cell-based network)
-
-Tamm&Tamm observed synapses from neurites to bridge cells (2002). and in Mnemiopsis larvae observed synapses from bridge cells onto other cells
-
-Inserting Figures
-
-You can add your figures into the rendered document.
-We saved the figures into /manuscript/figures or /manuscript/figure_supplements and can insert them from there.
-We use knitr::include_graphics for this.
-The title and legend can also be edited, as will as the width of the output figure.
-
-### Connectivity
-
-### Physiology?
-
-### Discussion
-
-## Equations
-
-Equations can also be inserted, Insert -\> Display Math:
-
-$$
-\bar{X} = \frac{\sum_{i=1}^{n} x_{i}}{n}
-$$
-
-## Materials and Methods
-
-### Volume electron microscopy
-
-### Imaging of balancer activity
-
-### Cell-type annotation and connectome analysis
-
-### Data and code availability
+The following EM image stacks, including all traces and annotations, are available at https://catmaid.jekelylab.ex.ac.uk.
+These datasets can be queried either within CATMAID or through the CATMAID application programming interface (API) using the catmaid or natverse packages [@Bates2020].
+The dataset encompasses all EM images (in JPG format), skeletons, meshes, node tags, connectors, and annotations.
+Additionally, we provide all R scripts used for data acquisition and figure generation (Jokura et al., 2025).
+All plots, figures (including anatomical renderings), and figure layouts can be fully reproduced using the provided R scripts.
+While the scripts are mostly organized by figure, some general-purpose scripts—for tasks such as loading libraries, accessing CATMAID data, and defining common functions—are shared across multiple figures.
 

### 2025-02-17 analysis/scripts/Figure2_sup.R

@@ -946,11 +946,11 @@ Figure1 <- panel_balancer + panel_bridge + panel_bristle +
                                          face='plain', color='black'))
 
 
-ggsave("manuscript/figures/Figure1.png", limitsize = FALSE, 
+ggsave("manuscript/figures/Fig2_sup1.png", limitsize = FALSE, 
        units = c("px"), Figure1, width = 3800, height = 2400, bg='white')  
 
 
-ggsave("manuscript/figures/Fig1.pdf", limitsize = FALSE, 
+ggsave("manuscript/figures/Fig2_sup1.pdf", limitsize = FALSE, 
        units = c("px"), Figure1, width = 3800, height = 2400) 
 
 
@@ -1103,12 +1103,12 @@ Figure1 <-
   ggplot2::theme(plot.tag = element_text(size = 12, face='plain', color='black'))
 
 
-ggsave("manuscript/figures/Figure1.png", limitsize = FALSE, 
+ggsave("manuscript/figures/Fig2_sup2.png", limitsize = FALSE, 
        units = c("px"), Figure1, width = 3200, height = 3200, bg = "white")
 
 
 ggsave(
-  "manuscript/figures/Figure1.pdf",
+  "manuscript/figures/Fig2_sup2.pdf",
   limitsize = FALSE,
   units = c("px"),
   Figure1,

### 2025-02-17 analysis/scripts/packages_and_functions.R

@@ -152,12 +152,12 @@ crop_catmaid <- function(tagname,
           body = list(
             stack_ids=28,
             min_x=x-half_bb_size_x,
-            min_y=ynview3d("anterior", extramat = rotationMatrix(1.05, 250, -200, 1000))
-            par3d(zoom=0.7)
+            min_y=ynview3d("anterior", extramat = rotationMatrix(1.05, 250, -200, 1000)),
+            par3d(zoom=0.7),
             #y-axis clip
-            clipplanes3d(1, 0, 0, -11500)
+            clipplanes3d(1, 0, 0, -11500),
             #x-axis clip
-            clipplanes3d(0, 1, 0, -24000)
+            clipplanes3d(0, 1, 0, -24000),
             -half_bb_size_y,
             min_z=z-half_bb_size_z,
             max_x=x+half_bb_size_x,

### 2025-02-17 manuscript/Manuscript.qmd
Subtitle

@@
 title: "Synaptic connectome of the ctenophore apical organ"
+
 subtitle: |
-  \raggedright  Kei Jokura^1,2,4,5\*^, Sanja Jasek^1,3,4^,  Lara Kewalow^3^, Pawel Burkhardt^6^, Gáspár Jékely^1,3,4,\*^  \newline<br>^1^Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom \newline<br>^2^Exploratory Research Center on Life and Living Systems (ExCELLS), National Institutes of Natural Sciences, Okazaki, Japan\newline<br>^3^Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany \newline<br>^4^BioSciences, University of Exeter, Stocker Road, Exeter EX4 4QD, Exeter, United Kingdom\newline<br> ^5^National Institute for Basic Biology (NIBB), National Institutes of Natural Sciences, Okazaki, Japan\newline<br>^6^Michael Sars Centre, University of Bergen, Norway \newline<br> ^\*^Correspondence:  jokura@nibb.ac.jp, gaspar.jekely@cos.uni-heidelberg.de
+  **Kei Jokura**$^{1,2,3,*}$, **Sanja Jasek**$^{1,4}$, **Lara Kewalow**$^{4}$, **Pawel Burkhardt**$^{5}$, **Gáspár Jékely**$^{1,4,*}$  
+  <br><br>  
+  $^{1}$ Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom  
+  <br>  
+  $^{2}$ Exploratory Research Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan  
+  <br>  
+  $^{3}$ National Institute for Basic Biology (NIBB), Okazaki, 444-8585, Japan  
+  <br>  
+  $^{4}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  
+  <br>  
+  $^{5}$ Michael Sars Centre, University of Bergen, Norway  
+  <br><br>  
+  **$^{*}$ Correspondence:**  
+  [jokura@nibb.ac.jp](mailto:jokura@nibb.ac.jp),  
+  [gaspar.jekely@cos.uni-heidelberg.de](mailto:gaspar.jekely@cos.uni-heidelberg.de)
+
 mainfont: Arial

### 2025-02-17 manuscript/Manuscript.qmd
discussion and discussion figure (Fig.6)

@@
 ---
-
-```{r}
-#| include: false
-#| eval: true
-#source(here::here("analysis/scripts/statistics_for_paper.R"))
-
-#The 'analysis/scripts/statistics_for_paper.R' script is sourced and it runs but the output is not included in the knitted output.
-But we can access the variables defined in the sourced script simply by adding \` r var_name \` between \` backticks, in this case max_PRC value is (now this number comes from our sourced script).
-
-```
 
@@
 
+Figure 6 - cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.
+
+### Synapses
+
+"synaptic regions with the characteristic presynaptic triad mor- phology of ctenophore nerves (i.e., a single layer of vesicles, smooth ER sac, and closely apposed mito- chondria [@Horridge_1964; @Hernandez_Nicaise_1973] "
+
+Figure 7 - synapses, reconstruction, annotation, numbers, distribution of partners, example EM (figure supplement - lots of EM), mitochondria (Fig suppl - mitochondrion stats - Sanja)
+
+Gap juntions? Innexins are expressed in the apical organ in specific patches of cells in each quadrant [@Ortiz_2023]
+
+Ctenophore synapses [@Hernandez_Nicaise_1973]
+
+We found no synapses between bridge and balancer, in agreement with Tamm [@Tamm_2002].
+
+Figure 8 - connectome cell-type based, quadrant based, nerve net syntycia, bridge, quadrants (Fig suppl - cell-based network)
+
+Tamm&Tamm observed synapses from neurites to bridge cells (2002). and in Mnemiopsis larvae observed synapses from bridge cells onto other cells
+
+Inserting Figures
+
+You can add your figures into the rendered document.
+We saved the figures into /manuscript/figures or /manuscript/figure_supplements and can insert them from there.
+We use knitr::include_graphics for this.
+The title and legend can also be edited, as will as the width of the output figure.
+
+### Connectivity
+
+### Physiology?
+
 ## Discussion
@@
 In particular, this research may help elucidate how circuit modularization and layered structures emerged over evolutionary time.
-
-Figure 6 - cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.
-
-### Synapses
-
-"synaptic regions with the characteristic presynaptic triad mor- phology of ctenophore nerves (i.e., a single layer of vesicles, smooth ER sac, and closely apposed mito- chondria [@Horridge_1964; @Hernandez_Nicaise_1973] "
-
-Figure 7 - synapses, reconstruction, annotation, numbers, distribution of partners, example EM (figure supplement - lots of EM), mitochondria (Fig suppl - mitochondrion stats - Sanja)
-
-Gap juntions? Innexins are expressed in the apical organ in specific patches of cells in each quadrant [@Ortiz_2023]
-
-Ctenophore synapses [@Hernandez_Nicaise_1973]
-
-We found no synapses between bridge and balancer, in agreement with Tamm [@Tamm_2002].
-
-Figure 8 - connectome cell-type based, quadrant based, nerve net syntycia, bridge, quadrants (Fig suppl - cell-based network)
-
-Tamm&Tamm observed synapses from neurites to bridge cells (2002). and in Mnemiopsis larvae observed synapses from bridge cells onto other cells
-
-Inserting Figures
-
-You can add your figures into the rendered document.
-We saved the figures into /manuscript/figures or /manuscript/figure_supplements and can insert them from there.
-We use knitr::include_graphics for this.
-The title and legend can also be edited, as will as the width of the output figure.
-
-### Connectivity
-
-### Physiology?
-
-### Discussion
-
-## Equations
-
-Equations can also be inserted, Insert -\> Display Math:
-
-$$
-\bar{X} = \frac{\sum_{i=1}^{n} x_{i}}{n}
-$$
 

### 2025-02-17 analysis/scripts/Figure1.R
Fig1 scale bar

@@ -184,8 +184,18 @@ panel_balancer <- ggdraw() + draw_image(readPNG("manuscript/pictures/balancer.pn
                arrow.fill = "black", linewidth = 0.175)
 
 
-panel_larva_pic <- ggdraw() + draw_image(image_read("manuscript/pictures/Mnemiopsis_larva_5dpf.png"))
-panel_AO_pic <- ggdraw() + draw_image(image_read("manuscript/pictures/AO_mag_pics.png"))
+panel_larva_pic <- ggdraw() + draw_image(image_read("manuscript/pictures/Mnemiopsis_larva_5dpf.png")) +
+  draw_label("lateral view of TA plane", x = 0.5, y = 0.99, color="black", size = 8) +
+  draw_line(x = c(0.85, 0.95), y = c(0.1, 0.1), color = "black", size = 0.5) +
+  draw_label(expression(paste("500 ", mu, " m")), x = 0.9, y = 0.14, color = "black", size = 7, hjust = 0.5)
+
+panel_AO_pic <- ggdraw() + draw_image(image_read("manuscript/pictures/AO_mag_pics.png")) +
+  draw_label("aboral view", x = 0.2, y = 0.99, color="black", size = 8) +
+  draw_label("lateral view of PA plane", x = 0.5, y = 0.99, color="black", size = 8) +
+  draw_label("lateral view of TA plane", x = 0.85, y = 0.99, color="black", size = 8) +
+  draw_line(x = c(0.85, 0.95), y = c(0.05, 0.05), color = "black", size = 0.5) +
+  draw_label(expression(paste("50 ", mu, " m")), x = 0.9, y = 0.1, color = "black", size = 7, hjust = 0.5)
+
 panel_serial_sectioning <- ggdraw() + draw_image(readPNG("manuscript/pictures/serial sectioning.png"))
 panel_catmaid_overview <- ggdraw() + draw_image(readPNG("manuscript/pictures/overview.png")) +
   draw_label("serial EM volume", x = 0.3, y = 0.98, size = 10, fontface="plain") +

### 2025-02-17 analysis/scripts/plot_each_cell_types.R
edit file name for the plot each cell types

@@ -1,4 +1,4 @@
-# Code to generate Figure 1 of the Jokura et al 2024 Ctenophore apical organ connectome paper
+
 
 # source packages and functions -----------------
 source("analysis/scripts/packages_and_functions.R")

### 2025-03-03 analysis/scripts/flashcards.R
flashcards: visualize cut cilia too

@@ -33,6 +33,8 @@ celltype_flashcard(celltype) {
   ### morphology for single cell, with cilia
   cell <- cells[[1]]
   cilia <- segments_between_tags(cell, "cilium tip", "basal body")
+  # here we also want cut cilia
+  cilia_cut <- segments_between_tags(cell, "cilium cut", "basal body")
   # visualize part of cilium in ciliary pocket
   # this is a bit tricky, because it has "exit_ciliary_pocket" if it exits half way,
   # but "cilium tip intracellular" if the whole cilium is inside the cell
@@ -52,11 +54,14 @@ celltype_flashcard(celltype) {
   
   cell_smooth <- smooth_neuron(cell, sigma = 500)
   cilia_smooth <- lapply(cilia, smooth_neuron, sigma = 500) %>% as.neuronlist()
+  cilia_cut_smooth <- lapply(cilia_cut, smooth_neuron, sigma = 500) %>% as.neuronlist()
   in_pocket_smooth <- lapply(in_pocket, smooth_neuron, sigma = 500) %>% as.neuronlist()
-  
   plot3d(cell_smooth, soma = TRUE, lwd = 1, color = "grey")
   if (length(cilia_smooth) > 0) {
     plot3d(cilia_smooth, soma = TRUE, lwd = 2, color = "orange")
+  }
+  if (length(cilia_cut_smooth) > 0) {
+    plot3d(cilia_cut_smooth, soma = TRUE, lwd = 2, color = "orange")
   }
   if (length(in_pocket_smooth) > 0) {
     plot3d(in_pocket_smooth, soma = TRUE, lwd = 5, color = "purple")

### 2025-03-03 analysis/scripts/Figure2.R
edit Figure

@@ -4,10 +4,6 @@
 source("analysis/scripts/packages_and_functions.R")
 
 
-
-
-
-
 # load cell type ----------------------------------
 
 balancer <- read_smooth_neuron("celltype:balancer")
@@ -51,6 +47,8 @@ all_celltypes <- list(balancer,
                       biciliated,
                       multiciliated,
                       nonciliated)
+
+
 
 # plot SSN neuron -----------------------------------------------
 
@@ -146,6 +144,162 @@ close3d()
 
 
 
+# retrieve connectors ----------------------------------------------------------
+
+conn_syn <- connectors(SSN)
+presyn_syn <- subset(conn_syn, prepost == 0)
+postsyn_syn <- subset(conn_syn, prepost == 1)
+
+
+
+
+# plot SSN synapses ---------------------------------------------------------------------
+
+plot_background()
+close3d()
+# 3d plotting of cells
+nopen3d() 
+mfrow3d(1, 3)  #defines the two scenes
+#define the size of the rgl window, the view and zoom
+#par3d(windowRect = c(0, 0, 1200, 350))
+par3d(windowRect = c(0, 0, 2400, 700))
+
+#plot aboral view
+plot3d(SSN,
+       soma = FALSE, lwd = 2, add = TRUE, alpha = 0.5, 
+       col = Okabe_Ito[c(1, 5, 8)],
+       WithConnectors = F, WithNodes = F)
+
+# plot presynapses
+plot3d(
+  presyn_syn$x, 
+  presyn_syn$y, 
+  presyn_syn$z, 
+  size = 5, alpha = 1, col = "magenta2", 
+  add = TRUE,
+  point_antialias = TRUE
+)
+
+# plot postsynapses
+plot3d(
+  postsyn_syn$x, 
+  postsyn_syn$y, 
+  postsyn_syn$z, 
+  size = 5, alpha = 1, col = "cyan2", 
+  add = TRUE,
+  point_antialias = TRUE
+)
+
+for (i in 1:length(all_celltypes)) {
+  print(i)
+  plot3d(
+    all_celltypes[[i]], soma = TRUE, lwd = 1, add = TRUE, 
+    alpha = 0.05, col = Okabe_Ito[8]
+  )
+}
+
+#texts3d(58000,51000,5000, "INRGW", cex = 3, col = "#56B4E9")
+
[…]

### 2025-03-03 analysis/scripts/flashcards.R
flashcards: stats - cilia, mito, polyadicity

@@ -1,6 +1,9 @@
 source("analysis/scripts/packages_and_functions.R")
 
 celltype <- "balancer"
+cilium_lengths <- read_csv("analysis/data/cilium_lengths.csv")
+organelle_stats <- read_csv("analysis/data/organelle_stats.csv")
+stats_master <- read_csv("analysis/data/stats_master.csv")
 
 celltype_flashcard(celltype) {
   # things to show:
@@ -21,6 +24,7 @@ celltype_flashcard(celltype) {
   ### location by quadrant
   # read, smooth and plot skeletons
   cells <- read.neurons.catmaid(paste("celltype:", celltype, sep=""), pid=35)
+  skids <- names(cells)
   
   sigma = 2000
   lapply(cells, plot_multinucleated_cell, sigma=sigma, color="darkblue", lwd=1)
@@ -114,45 +118,55 @@ celltype_flashcard(celltype) {
                       35, 28)
   
   
-  
-  
-
-  for (cell in cells) {
-    sskid <- cell$skid
-    
-    ########## cilia and pocket lengths
-    cilium_lengths <- tibble(cil_length=integer(),
-                             pocket_length=integer())
-    cilia <- segments_between_tags(cell, "cilium tip", "basal body")
-    for (cilium in cilia) {
-      cil_length <- smooth_neuron(cilium, sigma=sigma) %>%
-        summary() %>% select(cable.length)
-      # check if it has exit_ciliary_pocket tag
-      #print(cilium$tags)
-      exit_cil_pocket <- cilium$tags[["exit_ciliary_pocket"]]
-      ctip_intra <- cilium$tags[["cilium tip intracellular"]]
-      if (length(exit_cil_pocket) > 0) {
-        pocket <- segments_between_tags(cilium, "exit_ciliary_pocket", "basal body")
-        pocket_length <- smooth_neuron(pocket[[1]], sigma = sigma) %>%
-          summary() %>% select(cable.length)
-      }
-      else if (length(ctip_intra) > 0) {
-        pocket_length <- cil_length
-      }
-      else {
-        pocket_length <- 0
-      }
-    }
-    cilium_lengths <- tibble(cil_length=integer(),
-                             pocket_length=integer())
-    cilium_lengths <- bind_rows(cilium_lengths, cil_len_tb)
-    
-    ########## polyadicity 
-    polyadicities <- list()
[…]

### 2025-03-03 manuscript/Manuscript.qmd

@@
 ---
-title: "Synaptic connectome of the ctenophore apical organ"
+title: "Neural Circuit Organization of the Ctenophore Apical Organ and Its Role in Gravity Sensing"
 
@@
 
-**text in bold** *italic* [underline]{.underline}
+Understanding the evolution and function of neurons is crucial for elucidating the origins of animal behavior.
+Ctenophores, as the earliest-diverging lineage with a nervous system, provide a unique model for investigating neural circuit organization.
+In this study, we performed volumetric electron microscopy (EM) reconstruction of the apical organ in Mnemiopsis leidyi to characterize its neural network involved in gravity sensing.
+We reconstructed 909 cells, including novel syncytial neurons spanning multiple quadrants, which form complex synaptic networks with balancer cells—key mechanosensory components controlling ciliary movement.
+Our connectome analysis revealed that these syncytial neurons establish direct and indirect pathways to balancer cells, with "bridge cells" acting as afferent relay neurons.
+High-speed imaging of balancer cilia movements further demonstrated differential regulation between sagittal and tentacular planes, suggesting distinct functional roles of apical organ neurons in modulating posture and locomotion.
+These findings provide new insights into the structure and function of ctenophore neural circuits, shedding light on early nervous system evolution and mechanosensory integration in basal metazoans.
 
@@
 
-![**Figure 3. Reconstruction of Balancer Ciliated Cells in the Apical Organ**\
-(A) Connectivity matrix of AO neurons.
-Columns represent presynaptic AO neurons, rows represent postsynaptic AO neurons, and each number indicates the number of synapses.
-AO neurons are interconnected via synapses, and AO neuron Q1234 forms a self-synapse, autapse.\
-(B) Three-dimensional reconstruction of balancer ciliated cells in all four quadrant regions.
-Each quadrant contains approximately 30 cells.
-All balancer cilia are monocilia.
-The spheres indicate the positions of the nuclei.
-The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
-(C) Synaptic connections from AO neurons to balancer ciliated cells.
-The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view.
-Balancer ciliated cells are depicted in light gray, AO neuron Q1234 in black, AO neuron Q12 in orange, and AO neuron Q34 in blue.\
-(D) Connectivity map of three AO neurons (Q1234 in blue, Q12 in pink, and Q34 in orange) and balancer ciliated cells (light blue).
-The thickness of the arrows and the numbers correspond to the number of synapses.
-Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.](figures/Figure3.png)
-
 ### Identification of Neuron-like Bridge Cells Forming Afferent Synapses in the Gravity-Sensitive Neural Circuit
@@
 
-![**Figure 4. Reconstruction of bridge cells and gravity-sensing neural circuits in the apical organ**\
+![**Figure 3. Reconstruction of Balancer Ciliated Cells in the Apical Organ Reconstruction of bridge cells and gravity-sensing neural circuits in the apical organ**\
+(A) Connectivity matrix of AO neurons.
+Columns represent presynaptic AO neurons, rows represent postsynaptic AO neurons, and each number indicates the number of synapses.
+AO neurons are interconnected via synapses, and AO neuron Q1234 forms a self-synapse, autapse.\
+(B) Three-dimensional reconstruction of balancer ciliated cells in all four quadrant regions.
+Each quadrant contains approximately 30 cells.
+All balancer cilia are monocilia.
+The spheres indicate the positions of the nuclei.
+The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+(C) Synaptic connections from AO neurons to balancer ciliated cells.
+The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view.
+Balancer ciliated cells are depicted in light gray, AO neuron Q1234 in black, AO neuron Q12 in orange, and AO neuron Q34 in blue.\
+(D) Connectivity map of three AO neurons (Q1234 in blue, Q12 in pink, and Q34 in orange) and balancer ciliated cells (light blue).
+The thickness of the arrows and the numbers correspond to the number of synapses.
+Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.
 (A) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants.
@@
 The thickness of the arrows and the numerical values indicate the number of synapses.
-AO neurons in Q1234 are shown in black, AO neurons in Q12 in yellow, AO neurons in Q34 in blue, balancer cell groups in gray, and bridge cell groups in pink.](figures/Figure4.png)
+AO neurons in Q1234 are shown in black, AO neurons in Q12 in yellow, AO neurons in Q34 in blue, balancer cell groups in gray, and bridge cell groups in pink.](figures/Figure3.png)
 
@@
 
-![**Figure 5. Differences in Balancer Cilia Movement Control Due to Variations in the Gravity-Sensing Neural Circuit Pathways**\
+![**Figure 4.
+Differences in Balancer Cilia Movement Control Due to Variations in the Gravity-Sensing Neural Circuit Pathways Comparison of Neural Circuits Regulating Ciliary Movement in *M.
+leidyi* (Ctenophore) and *P.
+dumerilii* (Annelid)**\
 (A) Schematic diagram of differential interference contrast (DIC) microscopy setup for imaging balancer cilia movement.
@@
 The direction and length of the arrows indicate time.\
-(G) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia.
-S and T represent the sagittal plane and tentacular plane, respectively.](figures/Figure5.png)
-
-Figure 6 - cilia? renderings of basal bodies, centrioles, length, axoneme structure (table), table about cell types, number of cilia etc.
-
-### Synapses
-
-"synaptic regions with the characteristic presynaptic triad mor- phology of ctenophore nerves (i.e., a single layer of vesicles, smooth ER sac, and closely apposed mito- chondria [@Horridge_1964; @Hernandez_Nicaise_1973] "
-
-Figure 7 - synapses, reconstruction, annotation, numbers, distribution of partners, example EM (figure supplement - lots of EM), mitochondria (Fig suppl - mitochondrion stats - Sanja)
-
-Gap juntions? Innexins are expressed in the apical organ in specific patches of cells in each quadrant [@Ortiz_2023]
-
-Ctenophore synapses [@Hernandez_Nicaise_1973]
-
-We found no synapses between bridge and balancer, in agreement with Tamm [@Tamm_2002].
-
-Figure 8 - connectome cell-type based, quadrant based, nerve net syntycia, bridge, quadrants (Fig suppl - cell-based network)
-
-Tamm&Tamm observed synapses from neurites to bridge cells (2002). and in Mnemiopsis larvae observed synapses from bridge cells onto other cells
-
-Inserting Figures
-
-You can add your figures into the rendered document.
-We saved the figures into /manuscript/figures or /manuscript/figure_supplements and can insert them from there.
-We use knitr::include_graphics for this.
-The title and legend can also be edited, as will as the width of the output figure.
-
-### Connectivity
-
-### Physiology?
+(G) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia. S and T represent the sagittal plane and tentacular plane, respectively.
+Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation.
+Ciliated cells are shown in gray.
+Synapses are indicated by arrows, with magenta representing synapses that induce ciliary arrest and blue representing synapses that induce ciliary re-beat or an increase in ciliary beating frequency.\
+(A) Gravity-sensing neural circuit of *M.
+leidyi*, as revealed in this study.
+Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying their potential involvement in feedback mechanisms between neurons and ciliated cells.\
+(B) Neural circuit regulating prototroch ciliary movement in *P.
+dumerilii*.
+Serotonergic neurons (Ser-h1 neurons, blue) activate ciliary movement, while cholinergic neurons (MC neurons, magenta) induce ciliary arrest.
+Arrows directed toward ciliated cells in the head prototroch are color-coded accordingly.](figures/Figure4.png)
 
@@
 While our synaptic ultrastructure analysis did not allow us to distinguish excitatory from inhibitory neurons, future electrophysiological investigations will be essential to elucidate their functional properties.
-
-![**Figure 6. Comparison of Neural Circuits Regulating Ciliary Movement in *M. leidyi* (Ctenophore) and *P. dumerilii* (Annelid)**\
-Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation.
-Ciliated cells are shown in gray.
-Synapses are indicated by arrows, with magenta representing synapses that induce ciliary arrest and blue representing synapses that induce ciliary re-beat or an increase in ciliary beating frequency.\
-(A) Gravity-sensing neural circuit of *M.
-leidyi*, as revealed in this study.
-Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying their potential involvement in feedback mechanisms between neurons and ciliated cells.\
-(B) Neural circuit regulating prototroch ciliary movement in *P.
-dumerilii*.
-Serotonergic neurons (Ser-h1 neurons, blue) activate ciliary movement, while cholinergic neurons (MC neurons, magenta) induce ciliary arrest.
-Arrows directed toward ciliated cells in the head prototroch are color-coded accordingly.](figures/Figure6.png)
 

### 2025-03-03 analysis/scripts/number_of_postsynaptic_sites_per_mitochondria.R
mitochondria: change views of snapshots
- aboral, sagittal, tentacular

@@ -109,6 +109,10 @@ get_mito_pos <- function(neur) {
 mito_vesicle_info <- lapply(mito_done, get_mito_pos) |>
   bind_rows()
 
+#for (mito_done_neu in mito_done) {
+#  print(mito_done_neu[1])
+#  get_mito_pos(mito_done_neu)
+#}
 
 # create tibble for graph ---------------------------------
 
@@ -194,7 +198,7 @@ plot_mito_stats <- ggplot(mito_means_tidy, aes(
 # I'm not sure it makes sense to do it for other celltypes because not all cells
 # for them had annotated mito
 for (cell_type in c("SSN", "bridge")) {
-  plot_background()
+  nopen3d()
   # plot3d(outline,
   #  WithNodes = F, add = T, alpha = 0.07, col = "#E2E2E2"
   # )
@@ -251,10 +255,19 @@ for (cell_type in c("SSN", "bridge")) {
     size = 5,
     alpha = 1
   )
-  anterior()
-  paste("manuscript/pictures/mito_pos_", cell_type, "_anterior.png", sep = "") %>% rgl.snapshot()
-  left()
-  paste("manuscript/pictures/mito_pos_", cell_type, "_left.png", sep = "") %>% rgl.snapshot()
+  par3d(zoom=0.50)
+  # anterior
+  nview3d("anterior", extramat = rotationMatrix(1.05, 250, -200, 1000))
+  par3d(windowRect = c(0, 0, 400, 800))
+  paste("manuscript/pictures/mito_pos_", cell_type, "_aboral.png", sep = "") %>% rgl.snapshot()
+  # sagittal
+  nview3d("left", extramat = rotationMatrix(300, 4200, 1800, 800))
+  par3d(windowRect = c(0, 0, 450, 450))
+  paste("manuscript/pictures/mito_pos_", cell_type, "_sagittal.png", sep = "") %>% rgl.snapshot()
+  # tentacular
+  nview3d("left", extramat = rotationMatrix(-1.7, 190, -120, -140))
+  par3d(windowRect = c(0, 0, 800, 450))
+  paste("manuscript/pictures/mito_pos_", cell_type, "_tentacular.png", sep = "") %>% rgl.snapshot()
   close3d()
 }
 

### 2025-03-03 analysis/scripts/number_of_postsynaptic_sites_per_mitochondria.R
mito_positions: show all somas; update rcatmaid lib

@@ -209,7 +209,7 @@ for (cell_type in c("SSN", "bridge")) {
 
   for (skid in skids) {
     cell <- mito_done[[as.character(skid)]] %>% smooth_neuron(sigma = 1000)
-    plot3d(cell, color = "grey", lwd = 2, soma = TRUE, alpha = 0.6)
+    plot_multinucleated_cell(cell, color = "grey", lwd = 2, alpha = 0.6)
   }
 
   pos_ves_syn <- mito_vesicle_info |>

### 2025-03-16 manuscript/Manuscript.qmd
edit ref

@@
 In many bilaterian and non-bilaterian animals, an apical nervous system is located at one pole of the body, forming a specialized sensory structure that governs physiological functions and motor activities.
-In ctenophores, ultrastructural analyses have revealed that the aboral pole contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974] <!--# (Tamm, 1982 ?) -->.
+In ctenophores, ultrastructural analyses have revealed that the aboral pole contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974; @Tamm1982].
 Additionally, immunostaining using an anti-tyrosylated α-tubulin antibody has identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].
@@
 The apical organ of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014].
-The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014] <!--# (Tamm, 1982 ?) -->.
+The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014; @Tamm1982].
 Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron].
@@
 Based on their distinct spatial relationships, we were able to classify these three Syncytial neurons into two categories.
-The first type is a larger "AO neuron_Q1234," which possesses four (or possibly six?) nuclei spanning four quadrants.
+The first type is a larger "AO neuron_Q1234," which possesses four (or possibly five?) nuclei spanning four quadrants.
 It contained X presynaptic structures.
@@
 To investigate the function of gravity-sensitive circuits, we conducted a comparative analysis of balancer ciliary movements across different regions.
-Previous studies (Tamm, 1980, 1982; Lowe, 1997) have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination.
-Furthermore, it has been suggested that differences in the asymmetry of statolith morphology and the elongation direction of balancer cilia between the tentacular and sagittal planes could result in variations in the force received from the statolith (Tamm, 2014, 2015).
+Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination.
+Furthermore, it has been suggested that differences in the asymmetry of statolith morphology and the elongation direction of balancer cilia between the tentacular and sagittal planes could result in variations in the force received from the statolith [@Tamm_2014; @Tamm2015].
 Building on these findings, we standardized conditions for observing balancer cilia by tilting the microscope by 90 degrees and fixing samples such that the aboral-oral axis was aligned parallel to the vertical stage, with the aboral side facing upward.
@@
 The direction and length of the arrows indicate time.\
-(G) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia. S and T represent the sagittal plane and tentacular plane, respectively.
+(G) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia.
+S and T represent the sagittal plane and tentacular plane, respectively.
 Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation.
@@
 Burkhurdt et al.
-(2023) previously reported that in one-day-old larvae, the sub-epithelial nerve net (SNN) contains neurons that form syncytia.
+previously reported that in one-day-old larvae, the sub-epithelial nerve net (SNN) contains neurons that form syncytia [@Burkhardt2023].
 However, the newly discovered syncytial neurons embedded within the apical organ cell cluster did not exhibit the characteristic "pearl necklace-like structure" of the SNN, which consists of microtubules connecting neurons.
-Instead, their morphology was more reminiscent of fiber cells previously described in Placozoa (Mayorova et al., 2021), with cell bodies extending through intercellular spaces.
+Instead, their morphology was more reminiscent of fiber cells previously described in Placozoa [@Mayorova2021], with cell bodies extending through intercellular spaces.
 These morphological characteristics suggest that the newly identified neurons represent a second distinct type of syncytial neuron in ctenophores.
@@
 In metazoans, a single neuron can directly provide synaptic input to multiple ciliated cells at a distance, synchronizing their movements.
-For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating (Verasztó et al., 2017).
+For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating [@Verasztó2017].
 Additionally, a pair of serotonergic neurons, designated Ser-h1, extend axons that cross at the midline; the left Ser-h1_l predominantly innervates right-sided ciliated cells, while the right Ser-h1_r mainly projects to the left-side ciliated cells.
@@
 
-Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores [@HORRIDGE1965; @vinnikov1974e] <!--#  Krisch, 1973; Aronova, 1974 ?
---> ).
+Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores [@HORRIDGE1965; @vinnikov1974e; @Krisch1973; Aronova1974].
 By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the apical organ.
@@
 In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the apical organ [@Sachkova2021; @Hayakawa2022].
-These findings suggest that many sensory cells in the apical organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis (Jekely, 2021).
+These findings suggest that many sensory cells in the apical organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis [@Jékely2021].
 While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.
@@
 Conversely, previous research has demonstrated that mechanoreceptive sensory cells equipped with cilia (referred to as type 3 and type 4 cells) in ctenophores form direct synaptic connections with the sub-epithelial nerve net (SNN) and mesogleal neurons [@Burkhardt2023].
-Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells <!--# add these to bib file -->(Hertwig, 1880; Chun, 1880; Eimer, 1880).
+Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells [@Hertwig1880; @Chun1880; @Eimer1880].
 
@@
 The neural cells identified in the apical organ exhibit a divergent feedforward connectivity pattern, where a single neuron forms presynaptic connections with multiple downstream targets.
-Such a circuit design minimizes unwanted signal interference, reduces processing delays, and enables relatively fast response execution (Luo, 2021).
+Such a circuit design minimizes unwanted signal interference, reduces processing delays, and enables relatively fast response execution [@Luo2021].
 Moreover, the presence of multiple divergent feedforward circuits suggests that slight variations in the synchronization and phase delay of ciliary beating (within the range of a few hundred milliseconds) may have evolved as a mechanism for generating behavioral diversity.
 
-From a broader neuroanatomical perspective, nervous system structures across various animals have been shaped through selective optimization to adapt to environmental constraints (Bullmore & Sporns, 2012; Perin et al., 2011).
-For example, in nematodes, the spatial arrangement of ganglia is optimized to minimize the total length of neural wiring (White et al., 1986).
+From a broader neuroanatomical perspective, nervous system structures across various animals have been shaped through selective optimization to adapt to environmental constraints [@Bullmore2012; @Perin2011].
+For example, in nematodes, the spatial arrangement of ganglia is optimized to minimize the total length of neural wiring [@White1986].
 Similarly, in ctenophores, understanding how transmission via syncytial neurons and synaptic connections has been selectively optimized at the circuit level will provide crucial insights into the evolutionary scaling of neural networks from early nervous systems to complex brains [@Farnworth2024].
@@
 
-The reconstruction of cells and connectivity was performed using CATMAID (Saalfeld et al., 2009), analyzing a total of 620 serial sections.
+The reconstruction of cells and connectivity was performed using CATMAID [@Saalfeld2009], analyzing a total of 620 serial sections.
 To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server [@Saalfeld2009; @Schneider-Mizell2016].

### 2025-03-17 analysis/scripts/Supplement_Figures.R
coding 3d plots of all individual cell types and all cell types together

@@ -1,0 +1,254 @@
+# Code to generate Supplement Figures of the Jokura et al 2024 Ctenophore apical organ connectome paper
+# not sure yet which figure this will be used to supplement, so store everything in this file for now
+
+# source packages and functions ------------------------------------------------
+source("analysis/scripts/packages_and_functions.R")
+
+
+# load all cell types to plot individually -------------------------------------
+#twelve celltypes
+balancer <- read_smooth_neuron("celltype:balancer")
+bridge <- read_smooth_neuron("celltype:bridge")
+bristle <- read_smooth_neuron("celltype:bristle")
+Cgroove <- read_smooth_neuron("celltype:Cgroove")
+dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
+dome <- read_smooth_neuron("celltype:dome")
+intra_multi_ciliated <- read_smooth_neuron("celltype:intra-multi-ciliated")
+lamellate <- read_smooth_neuron("celltype:lamellate")
+lithocyte <- read_smooth_neuron("celltype:lithocyte")
+plumose <- read_smooth_neuron("celltype:plumose")
+SSN <- read_smooth_neuron("celltype:SSN")
+epithelial_floor <- read_smooth_neuron("celltype:epithelial_floor")
+
+with_soma <- read_smooth_neuron("with_soma")
+
+# function individual plot process --------------------------------------------------------
+
+plot_neuron_views <- function(neuron_name, neuron_data) {
+  close3d()
+  nopen3d()
+  mfrow3d(1, 3)
+  par3d(windowRect = c(0, 0, 2400, 700))
+  
+  for (view in c("aboral", "sagittal", "tentacular")) {
+    plot3d(neuron_data, soma = TRUE, lwd = 1, add = TRUE, alpha = 0.8, col = Okabe_Ito[5], WithConnectors = FALSE, WithNodes = FALSE)
+    plot3d(with_soma, soma = TRUE, lwd = 1, add = TRUE, alpha = 0.05, col = Okabe_Ito[8])
+    plot3d(outline, add = TRUE, alpha = 0.025, col = Okabe_Ito[8])
+    
+    switch(view,
+           "aboral" = aboral(),
+           "sagittal" = sagittal(),
+           "tentacular" = tentacular())
+    
+    par3d(zoom = 0.61)
[…]

### 2025-03-17 analysis/scripts/all_celltype_plot.R

@@ -1,256 +1,0 @@
-# Code to generate all cells
-
-# source packages and functions -----------------
-source("analysis/scripts/packages_and_functions.R")
-
-# load cell type ----------------------------------
-
-read_smooth_neuron <- function(annotation) {
-  nlapply(
-    read.neurons.catmaid(annotation, pid = 35),
-    function(x) {
-      smooth_neuron(x, sigma = 1000)
-    }
-  )
-}
-
-balancer <- read_smooth_neuron("celltype:balancer")
-bridge <- read_smooth_neuron("celltype:bridge")
-bristle <- read_smooth_neuron("celltype:bristle")
-dome <- read_smooth_neuron("celltype:dome")
-Cgroove_tag <- read_smooth_neuron("celltype:Cgroove-tag")
-Cgroove_sag <- read_smooth_neuron("celltype:Cgroove-sag")
-intramulticilia <- read_smooth_neuron("celltype:intra-multi-ciliated")
-lamellate <- read_smooth_neuron("celltype:lamellate")
-lithocyte <- read_smooth_neuron("celltype:lithocyte")
-SSN <- read_smooth_neuron("celltype:SSN")
-plumose <- read_smooth_neuron("celltype:plumose")
-dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
-monocilia <- read_smooth_neuron("celltype:monociliated")
-bicilia <- read_smooth_neuron("celltype:biciliated")
-non_cilia <- read_smooth_neuron("celltype:nonciliated")
-
-
-Q1 <- nlapply(
-  read.neurons.catmaid("Q1", pid = 35),
-  function(x) {
-    smooth_neuron(x, sigma = 1000)
-  }
-)
-Q2 <- nlapply(
-  read.neurons.catmaid("Q2", pid = 35),
-  function(x) {
-    smooth_neuron(x, sigma = 1000)
-  }
-)
-Q3 <- nlapply(
-  read.neurons.catmaid("Q3", pid = 35),
-  function(x) {
-    smooth_neuron(x, sigma = 1000)
-  }
-)
-Q4 <- nlapply(
-  read.neurons.catmaid("Q4", pid = 35),
-  function(x) {
-    smooth_neuron(x, sigma = 1000)
-  }
-)
-
-# plot neuron types -------------------
-
-all_celltypes <- list(
-  balancer, bridge, bristle, dome, Cgroove_tag, Cgroove_sag,
[…]

### 2025-03-17 analysis/scripts/plot3d_all_assemble.R

@@ -1,1544 +1,0 @@
-
-
-# source packages and functions -----------------
-source("analysis/scripts/packages_and_functions.R")
-
-# load cell type ----------------------------------
-
-#thirteen celltypes
-balancer <- read_smooth_neuron("celltype:balancer")
-bridge <- read_smooth_neuron("celltype:bridge")
-bristle <- read_smooth_neuron("celltype:bristle")
-Cgroove_sag <- read_smooth_neuron("celltype:Cgroove-sag")
-Cgroove_tag <- read_smooth_neuron("celltype:Cgroove-tag")
-dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
-dome <- read_smooth_neuron("celltype:dome")
-epithelial_floor <- read_smooth_neuron("celltype:epithelial_floor")
-intra_multi_ciliated <- read_smooth_neuron("celltype:intra-multi-ciliated")
-lamellate <- read_smooth_neuron("celltype:lamellate")
-lithocyte <- read_smooth_neuron("celltype:lithocyte")
-plumose <- read_smooth_neuron("celltype:plumose")
-SSN <- read_smooth_neuron("celltype:SSN")
-
-#four ciliated celltypes
-monociliated <- read_smooth_neuron("celltype:monociliated")
-biciliated <- read_smooth_neuron("celltype:biciliated")
-multiciliated <- read_smooth_neuron("celltype:multiciliated")
-nonciliated <- read_smooth_neuron("celltype:nonciliated")
-
-
-# plot balancer -----------------------------------------------
-
-close3d()
-# 3d plotting of cells
-nopen3d() 
-mfrow3d(1, 3)  #defines the two scenes
-#define the size of the rgl window, the view and zoom
-#par3d(windowRect = c(0, 0, 1200, 350))
-par3d(windowRect = c(0, 0, 2400, 700))
-
-#plot aboral view
-plot3d(balancer,
-       soma = T, lwd = 1, add = T, alpha = 0.5, col = Okabe_Ito[5],
-       WithConnectors = F, WithNodes = F)
-
-
-for (i in 1:length(all_celltypes)) {
-  print(i)
-  plot3d(
[…]

### 2025-03-17 analysis/scripts/plot3d_all_celltypes.R

@@ -1,70 +1,0 @@
-# Code to generate all cell types in 3d plot and 
-#save the views from three different directions as png files
-
-# source packages and functions -----------------
-source("analysis/scripts/packages_and_functions.R")
-
-# get celltypes
-celltypes <- get_celltypes(35)
-celltypes_list <- list()
-for (celltype in celltypes) {
-  celltypes_list[[celltype]] <- read.neurons.catmaid(paste("celltype:", celltype, sep = ""), pid = 35)
-}
-
-colour_palettes <- c(Okabe_Ito, bluepurple, oranges)
-
-nopen3d() 
-mfrow3d(1, 3)  #defines the two scenes
-par3d(windowRect = c(0, 0, 2400, 700))
-
-
-start.time <- Sys.time()
-#plot aboral view
-for (i in seq_along(celltypes_list)) {
-  # using i because counter is needed for color palette
-  # only use plot_multinucleated_cell for SSN, because it's much slower than normal plotting
-  if (names(celltypes_list)[[i]] == "SSN") {
-    plot_multinucleated_cell(celltypes_list[[i]], alpha = 0.6, color = colour_palettes[i])
-  } else {
-    plot3d(celltypes_list[[i]], alpha = 0.6, color = colour_palettes[i], soma = TRUE)
-  }
-  
-  i = i + 1
-}
-end.time <- Sys.time()
-time.taken <- end.time - start.time
-time.taken
-
-
-anterior()
-par3d(zoom=0.61)
-next3d(clear=F)
-
-#plot lateral view of Sagittal plane
-for (i in 1:length(all_celltypes)) {
-  print(i)
-  plot3d(
-    all_celltypes[[i]], soma = TRUE, lwd = 1, add = TRUE, 
-    alpha = 0.6, col = colour_palettes[i]
-  )
-}
-sagittal()
-par3d(zoom=0.61)
-
-next3d(clear=F)
-
-# plot screen 3
-for (i in 1:length(all_celltypes)) {
-  print(i)
-  plot3d(
-    all_celltypes[[i]], soma = TRUE, lwd = 1, add = TRUE, 
-    alpha = 0.6, col = colour_palettes[i]
-  )
-}
-tentacular()
-par3d(zoom=0.61)
-
-#make a snapshot
-rgl.snapshot("manuscript/pictures/all_cells_3_views_alt.png")
-close3d()
-

### 2025-03-17 analysis/scripts/Supplement_Figures.R
add supplement

@@ -1,254 +1,0 @@
-# Code to generate Supplement Figures of the Jokura et al 2024 Ctenophore apical organ connectome paper
-# not sure yet which figure this will be used to supplement, so store everything in this file for now
-
-# source packages and functions ------------------------------------------------
-source("analysis/scripts/packages_and_functions.R")
-
-
-# load all cell types to plot individually -------------------------------------
-#twelve celltypes
-balancer <- read_smooth_neuron("celltype:balancer")
-bridge <- read_smooth_neuron("celltype:bridge")
-bristle <- read_smooth_neuron("celltype:bristle")
-Cgroove <- read_smooth_neuron("celltype:Cgroove")
-dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
-dome <- read_smooth_neuron("celltype:dome")
-intra_multi_ciliated <- read_smooth_neuron("celltype:intra-multi-ciliated")
-lamellate <- read_smooth_neuron("celltype:lamellate")
-lithocyte <- read_smooth_neuron("celltype:lithocyte")
-plumose <- read_smooth_neuron("celltype:plumose")
-SSN <- read_smooth_neuron("celltype:SSN")
-epithelial_floor <- read_smooth_neuron("celltype:epithelial_floor")
-
-with_soma <- read_smooth_neuron("with_soma")
-
-# function individual plot process --------------------------------------------------------
-
-plot_neuron_views <- function(neuron_name, neuron_data) {
-  close3d()
-  nopen3d()
-  mfrow3d(1, 3)
-  par3d(windowRect = c(0, 0, 2400, 700))
-  
-  for (view in c("aboral", "sagittal", "tentacular")) {
-    plot3d(neuron_data, soma = TRUE, lwd = 1, add = TRUE, alpha = 0.8, col = Okabe_Ito[5], WithConnectors = FALSE, WithNodes = FALSE)
-    plot3d(with_soma, soma = TRUE, lwd = 1, add = TRUE, alpha = 0.05, col = Okabe_Ito[8])
-    plot3d(outline, add = TRUE, alpha = 0.025, col = Okabe_Ito[8])
-    
-    switch(view,
-           "aboral" = aboral(),
-           "sagittal" = sagittal(),
-           "tentacular" = tentacular())
-    
-    par3d(zoom = 0.61)
[…]

### 2025-03-17 analysis/scripts/Supplemental_figure.R

@@ -1,0 +1,292 @@
+# Code to generate Supplement Figures of the Jokura et al 2024 Ctenophore apical organ connectome paper
+# not sure yet which figure this will be used to supplement, so store everything in this file for now
+
+# source packages and functions ------------------------------------------------
+source("analysis/scripts/packages_and_functions.R")
+
+
+# load all cell types to plot individually -------------------------------------
+#twelve celltypes
+balancer <- read_smooth_neuron("celltype:balancer")
+bridge <- read_smooth_neuron("celltype:bridge")
+bristle <- read_smooth_neuron("celltype:bristle")
+Cgroove <- read_smooth_neuron("celltype:Cgroove")
+dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
+dome <- read_smooth_neuron("celltype:dome")
+intra_multi_ciliated <- read_smooth_neuron("celltype:intra-multi-ciliated")
+lamellate <- read_smooth_neuron("celltype:lamellate")
+lithocyte <- read_smooth_neuron("celltype:lithocyte")
+plumose <- read_smooth_neuron("celltype:plumose")
+SSN <- read_smooth_neuron("celltype:SSN")
+epithelial_floor <- read_smooth_neuron("celltype:epithelial_floor")
+
+with_soma <- read_smooth_neuron("with_soma")
+
+# function individual plot process --------------------------------------------------------
+
+plot_neuron_views <- function(neuron_name, neuron_data) {
+  close3d()
+  nopen3d()
+  mfrow3d(1, 3)
+  par3d(windowRect = c(0, 0, 2400, 700))
+  
+  for (view in c("aboral", "sagittal", "tentacular")) {
+    plot3d(neuron_data, soma = TRUE, lwd = 1, add = TRUE, alpha = 0.8, col = Okabe_Ito[5], WithConnectors = FALSE, WithNodes = FALSE)
+    plot3d(with_soma, soma = TRUE, lwd = 1, add = TRUE, alpha = 0.05, col = Okabe_Ito[8])
+    plot3d(outline, add = TRUE, alpha = 0.025, col = Okabe_Ito[8])
+    
+    switch(view,
+           "aboral" = aboral(),
+           "sagittal" = sagittal(),
+           "tentacular" = tentacular())
+    
[…]

### 2025-03-17 manuscript/Manuscript.qmd
add supplement
edit figure legends

@@
 Each of these Syncytial neurons was a cell with multiple nuclei, with membranes fused by continuous plasma membranes.
-These neurons are distinct in morphology from the Syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously reconstructed in 3D [@Burkhardt_2023].
+These neurons are distinct in morphology from the Syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously reconstructed in 3D [@Burkhardt2023].
 Our findings represent the second documented discovery of Syncytial neurons in ctenophores.
-Furthermore, these neurons exhibited clear morphological differences when compared to other sensory cells reported in the same study, such as mesogleal neurons and sensory cells with presynaptic structures and cilia (types 1-4)[@Burkhardt_2023].
+Furthermore, these neurons exhibited clear morphological differences when compared to other sensory cells reported in the same study, such as mesogleal neurons and sensory cells with presynaptic structures and cilia (types 1-4)[@Burkhardt2023].
 Based on their distinct spatial relationships, we were able to classify these three Syncytial neurons into two categories.
@@
 ![**Figure 2. Reconstruction of Three Syncytial Neurons in the Apical Organ**\
+(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (light blue).
+The spheres represent the positions of individual nuclei.
+The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei.
+The spheres indicate the positions of the nuclei.
+The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+(C) The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).
+(D) Localization of mitochondria within AO neuron Q1Q2Q3Q4.
+Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
 (A) Representative electron micrographs of presynaptic triad structures within the dataset.
@@
 The left panel illustrates a monadic structure, while the right panel shows a polyadic (or dyadic) structure.
-Scale bar: 500 nm.\
-(B) 3D reconstruction of the syncytial apical organ (AO) neuron Q1Q2Q3Q4, which spans all four quadrants and contains multiple nuclei (light blue).
-The spheres represent the positions of individual nuclei.
-The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
-(C) 3D reconstruction of two syncytial apical organ neurons, AO neuron Q1Q2 (pink) and AO neuron Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei.
-The spheres indicate the positions of the nuclei.
-The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
-(D) Localization of mitochondria within AO neuron Q1Q2Q3Q4.
-Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.](figures/Figure2.png)
+Scale bar: 500 nm.](figures/Figure2.png)
 
@@
 Some balancer cells received inputs from both AO neuron_Q1Q2 or AO neuron_Q3Q4 and AO neuron_Q1234.
-While previous studies have suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez_Nicaise_1974], our data did not reveal any synaptic inputs from balancer cells to AO neurons.
+While previous studies have suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez-Nicaise1974], our data did not reveal any synaptic inputs from balancer cells to AO neurons.
 However, we identified "bridge cells" that actively formed input synapses with AO neurons, highlighting their potential role in the network.
@@
 These cells, based on their morphological features, were found to be the bridge cells first described by Tamm et al.
-in 2002 [@Tamm_and_Tamm_2002].
+in 2002 [@Tamm_2002].
 These cells are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge.
@@
 
-![**Figure 3. Reconstruction of Balancer Ciliated Cells in the Apical Organ Reconstruction of bridge cells and gravity-sensing neural circuits in the apical organ**\
-(A) Connectivity matrix of AO neurons.
-Columns represent presynaptic AO neurons, rows represent postsynaptic AO neurons, and each number indicates the number of synapses.
-AO neurons are interconnected via synapses, and AO neuron Q1234 forms a self-synapse, autapse.\
-(B) Three-dimensional reconstruction of balancer ciliated cells in all four quadrant regions.
-Each quadrant contains approximately 30 cells.
-All balancer cilia are monocilia.
-The spheres indicate the positions of the nuclei.
-The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
-(C) Synaptic connections from AO neurons to balancer ciliated cells.
-The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view.
-Balancer ciliated cells are depicted in light gray, AO neuron Q1234 in black, AO neuron Q12 in orange, and AO neuron Q34 in blue.\
-(D) Connectivity map of three AO neurons (Q1234 in blue, Q12 in pink, and Q34 in orange) and balancer ciliated cells (light blue).
+![**Figure 3. Reconstruction of the balancers and bridges that receives synapses from the ANN.**\
+(A) Graph showing the number of synaptic outputs from the ANN circuit for each cell type.
+The balancer and bridge cell types receive relatively high outputs from the ANN circuit.
+ANN represents the number of synapses within the ANN circuit.
+The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
+(B) Connectivity map of three AO neurons (Q1234 in blue, Q12 in pink, and Q34 in orange) and balancer ciliated cells (light blue).
 The thickness of the arrows and the numbers correspond to the number of synapses.
-Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.
-(A) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants.
+Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.\
+(C) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants.
 The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color.
@@
 The left panel shows an aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
-(B) Mitochondrial localization within bridge cells and associated presynaptic triad structures.
+(D) Mitochondrial localization within bridge cells and associated presynaptic triad structures.
 Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.
 The left panel shows a dorsal view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
-(C) Synaptic connections between AO neurons and bridge cells.
+(E) Synaptic connections from AO neurons to balancer ciliated cells.
+The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view.
+Balancer ciliated cells are depicted in light gray, AO neuron Q1234 in black, AO neuron Q12 in orange, and AO neuron Q34 in blue.\
+(F) Synaptic connections between AO neurons and bridge cells.
 The positions of synapses from AO neurons to bridge cells (red) and from bridge cells to AO neurons (light blue) are indicated.
@@
 Bridge cells are shown in light gray, AO neurons in Q1234 are in black, AO neurons in Q12 are in orange, and AO neurons in Q34 are in blue.\
-(D) Connectivity matrix of the gravity-sensing neural circuit.
+(G) Connectivity matrix of the gravity-sensing neural circuit.
 Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups.
 The numbers and varying shades of blue correspond to the number of synapses.\
-(E) Complete connectivity map of the gravity-sensing neural circuit.
+(H) Complete connectivity map of the gravity-sensing neural circuit.
 Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels.
@@
 
-![**Figure 4.
-Differences in Balancer Cilia Movement Control Due to Variations in the Gravity-Sensing Neural Circuit Pathways Comparison of Neural Circuits Regulating Ciliary Movement in *M.
-leidyi* (Ctenophore) and *P.
-dumerilii* (Annelid)**\
+![**Figure 4. Differences in Balancer Cilia Movement Control Due to Variations in the Neural Circuit Pathways and Comparison of Neural Circuits Regulating Ciliary Movement**\
 (A) Schematic diagram of differential interference contrast (DIC) microscopy setup for imaging balancer cilia movement.
@@
 The movement of the balancer cilia was recorded using a 40× objective lens.\
-(B, C) Sagittal plane view (B) and tentacular plane view (C) of the apical organ under the microscope.
-The direction of gravity is always oriented downward in the images.
-The two balancer cilia being compared are indicated by arrowheads.\
-(D, E) Connectivity maps of the gravity-sensing neural circuit corresponding to Figure 3E.
-These maps emphasize the differences in neural circuits in the sagittal plane (D) and tentacular plane (E), focusing on the balancer cell groups (blue hexagons).
-AO neuron Q1234 is shown in green, while AO neurons Q12 and Q34 are shown in pink.\
-(F) Kymograph patterns comparing the arrest and re-beat of balancer cilia movement.
+(B) Kymograph patterns comparing the arrest and re-beat of balancer cilia movement.
 The left and right balancer cilia movements are compared in both the sagittal plane (left) and tentacular plane (right).
 The direction and length of the arrows indicate time.\
-(G) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia.
-S and T represent the sagittal plane and tentacular plane, respectively.
-Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation.
+(C) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia. S and T represent the sagittal plane and tentacular plane, respectively.\
+(D) Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation.
 Ciliated cells are shown in gray.
 Synapses are indicated by arrows, with magenta representing synapses that induce ciliary arrest and blue representing synapses that induce ciliary re-beat or an increase in ciliary beating frequency.\
-(A) Gravity-sensing neural circuit of *M.
+(Left) Gravity-sensing neural circuit of *M.
 leidyi*, as revealed in this study.
 Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying their potential involvement in feedback mechanisms between neurons and ciliated cells.\
-(B) Neural circuit regulating prototroch ciliary movement in *P.
+(Righr) Neural circuit regulating prototroch ciliary movement in *P.
 dumerilii*.
@@
 ## Supplementary material {.unnumbered}
+
+![**Supplemental figure**\
+(A)...](figures/Supplemental_figure.png)

### 2025-03-17 analysis/scripts/Figure4.R

@@ -13,7 +13,7 @@ df_time_dif <- read_csv("analysis/data/arrest_rebeat_time_differences.csv")
 
 plot_arrest_rebeat <- 
   ggplot(df_time_dif) +
-  aes(x = interaction(response, plane), y = time, fill = response) +
+  aes(x = response, y = time, fill = response) +  
   geom_boxplot(size = 0.5, outlier.shape = NA) +
   geom_beeswarm(aes(color = time),
                 size = 2,
@@ -23,18 +23,21 @@ plot_arrest_rebeat <-
   labs(y = "Time difference (sec)") +
   xlab("") +
   theme_minimal() +
-  theme(#axis.text.x = element_blank(),
-        legend.title = element_blank(),
-        legend.position = "inside",
-        legend.position.inside = c(0.3, 0.75)
-        ) +
+  theme(
+    legend.title = element_blank(),
+    legend.position = "inside",
+    legend.position.inside = c(0.3, 0.75),
+    strip.text = element_text(size = 11),  
+    strip.placement = "outside",  
+    strip.background = element_blank(),  
+    axis.text.x = element_blank(),  
+    axis.ticks.x = element_blank()  
+  ) +
   scale_fill_manual(
     values = c("re-beat" = "#28A8FF", 
                "arrest" = "#f56EBA")) +
-  scale_x_discrete(labels = c("arrest.sagittal" = "sagittal",
-                              "re-beat.sagittal" = "sagittal",
-                              "arrest.tentacular" = "tentacular",
-                              "re-beat.tentacular" = "tentacular"))
+  facet_grid(~ plane, scales = "free_x", space = "free_x", switch = "x")  
+
 
 plot_arrest_rebeat
 

### 2025-03-18 manuscript/Manuscript.qmd
edit figure legends

@@
 ---
-title: "Neural Circuit Organization of the Ctenophore Apical Organ and Its Role in Gravity Sensing"
+title: "Neural Circuit Organization of the Ctenophore Apical Organ"
 
@@
 ![**Figure 1. Morphology of the apical organ in *Mnemiopsis leidyi***\
-(A) Whole-body image of a 5-day-old M. leidyi cydippid larva in lateral view, observing the tentacular plane. The boxed region indicates the apical organ. Scale bar: 100 µm.\
-(B) Enlarged views of the apical organ from different perspectives in a 5-day-old M.
-leidyi cydippid larva.
-The left panel shows the apical organ from a aboral view.
-The middle panel presents a sagittal view of the apical organ.
-The right panel provides a tentacular plane view of the apical organ.
-Scale bar: 10 µm.\
-(C) Schematic diagram of the serial sectioning process.
-The entire cydippid larva was embedded in resin, trimmed around the apical organ, and sectioned into serial ultrathin sections, which were collected on glass slides in a ribbon-like arrangement.\
-(D) Example of cell tracing using the collaborative annotation toolkit CATMAID for large-scale electron microscopy image datasets.
+(A) Whole-body image of a 5-day-old *M. leidyi* cydippid larva in lateral view. The boxed region indicates the apical organ (ao). Scale bar: 100 µm.\
+(B) Schematic diagram of the aboral view of a cydippid larva.
+The two-tone coloration represents the biradial symmetry of the body.
+For convenience, the body is divided into four quadrants, designated as the first (Q1) through fourth (Q4) quadrants.
+The two primary viewing angles are referred to as the sagittal plane (S) and the tentacular plane (T).
+Abbreviations: ao, apical organ; cg, ciliated groove; cr, comb plates; tb, tentacle bulbs.\
+(C) Enlarged views of the apical organ from different perspectives in a cydippid larva using a DIC microscope.
+The left panel presents an aboral view of the apical organ.
+The middle panel shows a lateral view of the apical organ in the sagittal (S) plane, while the right panel displays a lateral view in the tentacular (T) plane.
+Abbreviations: bal, balancer; cg, ciliated grooves; do, dome cilia; li, lithocyte.
+Scale bar: 50 µm.\
+(D) Schematic representations of the apical organ from three different perspectives.
+The left panel illustrates the aboral view, the middle panel shows the lateral view in the sagittal plane, and the right panel presents the lateral view in the tentacular plane.
+The quadrant colors correspond to those in panel (B).\
+(E) Example of cell tracing using the collaborative annotation toolkit CATMAID for large-scale electron microscopy image datasets.
 The spherical objects indicate nuclear positions, while the lines represent the traced cell centers.
 Scale bar: 10 µm.\
-(E) Three-dimensional reconstruction of all cells composing the apical organ, displayed from different perspectives.
-The left panel shows a aboral view of the apical organ.
-The middle panel presents a sagittal plane view.
-The right panel provides a tentacular plane view.
-Cells are color-coded according to their types.
-Scale bar: 10 µm.](figures/Figure1.png)
+(F) Three-dimensional reconstruction of cells composing the apical organ, displayed from different perspectives.
+The left panel presents an aboral view of the apical organ.
+The middle panel shows a sagittal plane view, while the right panel provides a tentacular plane view.
+Cells are color-coded according to their respective quadrants.
+Lithocytes (li) are represented as three gray spheres, while balancers (bal) are depicted as gray lines.
+Scale bar: 25 µm.](figures/Figure1.png)
 
@@
 
-![**Figure 2. Reconstruction of Three Syncytial Neurons in the Apical Organ**\
-(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (light blue).
+![**Figure 2. Identification of Three Syncytial Neurons in the Apical Organ**\
+(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (blue).
 The spheres represent the positions of individual nuclei.
@@
 The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
-(C) The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).
+(C) Graph showing the number of mitochondria per cell and the proportion of mitochondria involved in synapse formation for each cell type.
+Green represents mitochondria associated with synapses, while gray represents mitochondria not involved in synapse formation.
+ANN has a significantly higher number of mitochondria compared to other cell types, with most of them forming synapses.
+The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
 (D) Localization of mitochondria within AO neuron Q1Q2Q3Q4.
 Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
-(A) Representative electron micrographs of presynaptic triad structures within the dataset.
-The positions of mitochondria are marked in orange.
-Postsynaptic target cells were inferred based on the locations of synaptic vesicles and the endoplasmic reticulum, with light blue arrows indicating the direction across the cell membrane.
-The left panel illustrates a monadic structure, while the right panel shows a polyadic (or dyadic) structure.
-Scale bar: 500 nm.](figures/Figure2.png)
+(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams.
+The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4).
+The abbreviations are as follows: mi (mitochondrion), er (endoplasmic reticulum), and sv (synaptic vesicles).\
+(F) Plot showing the three-dimensional positions of synapses.
+Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan.
+Blue dots indicate the locations of autapses within ANN Q1-4.\
+(G) Neural circuit map of ANN connectivity.
+The blue, orange, and pink circles represent individual neurons in ANN Q1-4, Q2Q3, and Q3Q4, respectively.
+Magenta arrows indicate synaptic connections from the center ANN to the lateral ANN, while cyan arrows indicate connections from the lateral ANN to the center ANN.
+Numbers denote the number of synapses.
+Blue arrows represent autapses in ANN Q1-4 along with their synapse counts.](figures/Figure2.png)
 
@@
 
-Each cell is assigned a specific type name based on its category (balancer, biC, bridge, bristle, dense vesicle cell, dome, EF, IcMC, LB, lithocyte, monoC, non-C, plumose cell, SNN, ssCG, SSN, stCG), resulting in a total of 17 cell types.
+Each cell is assigned a specific type name based on its category (balancer, biC, bridge, bristle, dense vesicle cell, dome, EF, IcMC, LB, lithocyte, monoC, non-C, plumose cell, SNN, ssCG, ANN, stCG), resulting in a total of 17 cell types.
 Additionally, a quadrant number is appended to the cell type name following an underscore ("\_") to indicate the cell body’s location.

### 2025-03-18 analysis/scripts/Figure1.R

@@ -1,197 +1,45 @@
 # Code to generate Figure 1 of the Jokura et al 2024 Ctenophore apical organ connectome paper
 
-# source packages and functions -----------------
+# source packages and functions ------------------------------------------------
 source("analysis/scripts/packages_and_functions.R")
 
-# load cells ----------------------------------
+# load cells -------------------------------------------------------------------
 
 balancer <- read_smooth_neuron("celltype:balancer")
+cgroove <- read_smooth_neuron("celltype:Cgroove")
+statolith <- read_smooth_neuron("statolith")
+
 Q1 <- read_smooth_neuron("Q1")
 Q2 <- read_smooth_neuron("Q2")
 Q3 <- read_smooth_neuron("Q3")
 Q4 <- read_smooth_neuron("Q4")
-with_soma <- read_smooth_neuron("with_soma")
-lithocyte <- read_smooth_neuron("celltype:lithocyte")
 
-# plot quadrants ----------------------
+# plot quadrants ----------------------------------------------------------------
+
+plot_views <- function(view_func) {
+  plot3d(balancer, soma = FALSE, lwd = 2.5, add = TRUE, alpha = 0.75, col = "gray40", WithConnectors = FALSE)
+  plot3d(statolith, soma = TRUE, lwd = 0, add = TRUE, alpha = 1, col = "gray", WithConnectors = FALSE)
+  plot3d(Q1, soma = TRUE, lwd = 1, add = TRUE, alpha = 0.4, col = Okabe_Ito[1])
+  plot3d(Q2, soma = TRUE, lwd = 1, add = TRUE, alpha = 0.4, col = Okabe_Ito[2])
+  plot3d(Q3, soma = TRUE, lwd = 1, add = TRUE, alpha = 0.4, col = Okabe_Ito[6])
+  plot3d(Q4, soma = TRUE, lwd = 1, add = TRUE, alpha = 0.4, col = Okabe_Ito[7])
+  plot3d(outline, add = TRUE, alpha = 0.04, col = Okabe_Ito[8])
+  view_func()
+  par3d(zoom = 0.61)
+  next3d(clear = FALSE)
+}
 
 close3d()
-# 3d plotting of cells
[…]

### 2025-03-18 analysis/scripts/Figure2.R

@@ -225,20 +225,30 @@ close3d()
 
 
 
-# load of mitochondrial location information-------------------------------
+# load of mitochondrial tidy csv file ------------------------------------------
 
-mito_done <- read.neurons.catmaid("mitochondria done", pid = 35)
-mito_vesicle_info <- read.csv("analysis/data/mito_vesicle_info.csv")
 mito_means_tidy <- read.csv("analysis/data/mito_means_tidy.csv")
 
-# bar graph of mito vesicles ratio by celltype ---------------------------------
+# bar graph of mito vesicles ratio by celltypes ---------------------------------
+
+mito_means_tidy <- mito_means_tidy %>%
+  mutate(
+    celltype = ifelse(celltype %in% c("Cgroove-sag", "Cgroove-tag"), "cg", celltype),
+    characteristic = ifelse(characteristic %in% c("mean_vesicles_none", "mean_vesicles_unclear"), 
+                            "mean_vesicles_no_syn", 
+                            characteristic)
+  ) %>%
+  filter(!celltype %in% c("nonciliated", "multiciliated", "monociliated", "biciliated"))
 
 label_mapping <- c(
-  "balancer" = "bal", "bridge" = "brg", "bristle" = "bsl", 
-  "ciliated_groove" = "cg", "dense_vesicle" = "dv", "dome" = "do", 
-  "intra-multi-ciliated" = "imc", "lamellate" = "la", "lithocyte" = "li", 
-  "plumose" = "pl", "SSN" = "ANN", "epithelial_floor" = "ef"
+  "balancer" = "bal", "bridge" = "brg", "bristle" = "bsl", "cg" = "cg",
+  "dense_vesicle" = "dv", "dome" = "do", "intra-multi-ciliated" = "imc", 
+  "lamellate" = "la", "lithocyte" = "li", "plumose" = "pl", 
+  "SSN" = "ANN", "epithelial_floor" = "ef"
 )
+
+mito_means_tidy <- mito_means_tidy %>%
+  mutate(celltype = factor(celltype, levels = names(label_mapping)))
 
 plot_mito_stats <- ggplot(mito_means_tidy, aes(
   fill = factor(characteristic, levels = c("mean_vesicles_syn", "mean_vesicles_no_syn")),
[…]

### 2025-03-18 analysis/scripts/Supplemental_figure.R

@@ -272,7 +272,7 @@ panel_all_cells <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/al
 panel_SNN <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/plot_SNN.png"))
 
 
-Suppl_fig <- panel_all_assemble + panel_all_cells + panel_SSN
+Suppl_fig <- panel_all_assemble + panel_all_cells + panel_SNN
 
 ggsave("manuscript/figures/Supplemental_figure.png", limitsize = FALSE, 
        units = c("px"), Suppl_fig, width = 2700, height = 500, bg='white')  

### 2025-03-18 manuscript/Manuscript.qmd
edit main text

@@
 Based on their distinct spatial relationships, we were able to classify these three Syncytial neurons into two categories.
-The first type is a larger "AO neuron_Q1234," which possesses four (or possibly five?) nuclei spanning four quadrants.
+The first type is a larger "ANN_Q1-4," which possesses four (or possibly five?) nuclei spanning four quadrants.
 It contained X presynaptic structures.
-The second type consists of "AO neuron_Q12" and "AO neuron_Q34," each having two nuclei spanning two quadrants, and each containing X presynaptic structures.
+The second type consists of "ANN_Q1Q2" and "ANN_Q3Q4," each having two nuclei spanning two quadrants, and each containing X presynaptic structures.
 
@@
 The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
-(D) Localization of mitochondria within AO neuron Q1Q2Q3Q4.
+(D) Localization of mitochondria within ANN Q1-4.
 Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
@@
 
-We discovered that AO neurons form synaptic connections with each other, and that AO neuron_Q1234 forms self-synapses, or “autapses.” To our knowledge, previous reports have not identified synaptic connections between subepithelial nerve net (SNN) neurons.
+We discovered that ANN neurons form synaptic connections with each other, and that ANN_Q1-4 forms self-synapses, or “autapses.” To our knowledge, previous reports have not identified synaptic connections between subepithelial nerve net (SNN) neurons.
 Thus, our results represent the first report of synaptic connectivity between neurons in the apical organ, forming a network in ctenophores (subject to confirmation).
-These AO neuron networks were found to form many presynaptic structures in relation to the gravity-sensing balancer cells.
+These ANN networks were found to form many presynaptic structures in relation to the gravity-sensing balancer cells.
 
@@
 
-From AO neuron_Q1Q2, synapses were formed with 6 of the 37 balancer cells in the Q1 region, and 8 of the 32 balancer cells in the Q2 region.
-Similarly, from AO neuron_Q3Q4, synapses were formed with 1 of the 32 balancer cells in the Q3 region, and 5 of the 28 balancer cells in the Q4 region.
-AO neuron_Q1234 formed input synapses with balancer cells in the Q1 region (7 cells), Q2 region (11 cells), Q3 region (6 cells), and Q4 region (10 cells).
-Some balancer cells received inputs from both AO neuron_Q1Q2 or AO neuron_Q3Q4 and AO neuron_Q1234.
-While previous studies have suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez-Nicaise1974], our data did not reveal any synaptic inputs from balancer cells to AO neurons.
-However, we identified "bridge cells" that actively formed input synapses with AO neurons, highlighting their potential role in the network.
+From ANN_Q1Q2, synapses were formed with 6 of the 37 balancer cells in the Q1 region, and 8 of the 32 balancer cells in the Q2 region.
+Similarly, from ANN_Q3Q4, synapses were formed with 1 of the 32 balancer cells in the Q3 region, and 5 of the 28 balancer cells in the Q4 region.
+ANN_Q1-4 formed input synapses with balancer cells in the Q1 region (7 cells), Q2 region (11 cells), Q3 region (6 cells), and Q4 region (10 cells).
+Some balancer cells received inputs from both ANN_Q1Q2 or ANN_Q3Q4 and ANN_Q1-4.
+While previous studies have suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez-Nicaise1974], our data did not reveal any synaptic inputs from balancer cells to ANN neurons.
+However, we identified "bridge cells" that actively formed input synapses with ANN neurons, highlighting their potential role in the network.
 
@@
 
-From our data, we identified several cell groups that form multiple afferent synapses with AO neurons.
+From our data, we identified several cell groups that form multiple afferent synapses with ANNs.
 These cells, based on their morphological features, were found to be the bridge cells first described by Tamm et al.
@@
 They originate from the base of paired balancer cells along the tentacle surface and extend across the sagittal plane toward the base of the opposite balancer cells.
-In regions where the mitochondria of the bridge cells are localized (approximately 30%), a presynaptic triad structure, similar to that of AO neurons, was found, containing synaptic vesicles and smooth endoplasmic reticulum.
+In regions where the mitochondria of the bridge cells are localized (approximately 30%), a presynaptic triad structure, similar to that of ANNs, was found, containing synaptic vesicles and smooth endoplasmic reticulum.
 
@@
 In the Q1Q2 region, 14 cells were identified, while in the Q3Q4 region, 12 cells were identified, totaling 26 bridge cells.
-Nearly all of these bridge cells (25 out of 26) exhibited afferent synapses from AO neurons.
-For bridge cells located in the Q1Q2 region, synaptic inputs came primarily from AO neuron_Q1Q2 (11 cells), AO neuron_Q1234 (1 cell), or both (2 cells).
-For bridge cells located in the Q3Q4 region, synaptic inputs were received from AO neuron_Q3Q4 (1 cell), AO neuron_Q1234 (7 cells), or both (1 cell).
-In other words, bridge cells in the Q1Q2 region mainly received input from AO neuron_Q1Q2, while those in the Q3Q4 region received input primarily from AO neuron_Q1234, showing a distinct pattern of input in the two regions.
+Nearly all of these bridge cells (25 out of 26) exhibited afferent synapses from ANNs.
+For bridge cells located in the Q1Q2 region, synaptic inputs came primarily from ANN_Q1Q2 (11 cells), ANN_Q1-4 (1 cell), or both (2 cells).
+For bridge cells located in the Q3Q4 region, synaptic inputs were received from ANN_Q3Q4 (1 cell), ANN_Q1-4 (7 cells), or both (1 cell).
+In other words, bridge cells in the Q1Q2 region mainly received input from ANN_Q1Q2, while those in the Q3Q4 region received input primarily from ANN_Q1-4, showing a distinct pattern of input in the two regions.
 
-Some bridge cells also formed afferent synapses with AO neurons.
-For example, bridge cells in the Q1Q2 region formed synapses with AO neuron_Q1Q2 (3 cells) or both AO neuron_Q1Q2 and AO neuron_Q1234 (2 cells).
-Bridge cells in the Q3Q4 region formed synapses with AO neuron_Q3Q4 (1 cell) or AO neuron_Q1234 (6 cells).
-A notable difference was observed between the Q1Q2 and Q3Q4 regions in the proportion of synaptic inputs from bridge cells to AO neurons.
+Some bridge cells also formed afferent synapses with ANN neurons.
+For example, bridge cells in the Q1Q2 region (3 cells) formed synapses with ANN_Q1Q2 or (2 cells) both ANN_Q1Q2 and ANN_Q1-4.
+Bridge cells in the Q3Q4 region formed synapses with ANN_Q3Q4 (1 cell) or ANN_Q1-4 (6 cells).
+A notable difference was observed between the Q1Q2 and Q3Q4 regions in the proportion of synaptic inputs from bridge cells to ANN neurons.
 
@@
 To analyze the grouped synaptic connectivity graph, we classified cell types within each region, collapsed cells of the same type into a single node, summed the number of synapses, and explored directed pathways to effectors (balancer cell groups).
-The results revealed a feedback pathway from AO neurons through the synaptic connections formed by bridge cells, thereby shedding light on the neural circuit structure involving these cells.
+The results revealed a feedback pathway from ANN neurons through the synaptic connections formed by bridge cells, thereby shedding light on the neural circuit structure involving these cells.
 
@@
 The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
-(B) Connectivity map of three AO neurons (Q1234 in blue, Q12 in pink, and Q34 in orange) and balancer ciliated cells (light blue).
+(B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink, and Q3Q4 in orange) and balancer ciliated cells (light blue).
 The thickness of the arrows and the numbers correspond to the number of synapses.
@@
 The left panel shows a dorsal view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
-(E) Synaptic connections from AO neurons to balancer ciliated cells.
+(E) Synaptic connections from ANN neurons to balancer ciliated cells.
+The positions of synapses from ANNs to balancers (magenta) are indicated.
 The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view.
-Balancer ciliated cells are depicted in light gray, AO neuron Q1234 in black, AO neuron Q12 in orange, and AO neuron Q34 in blue.\
-(F) Synaptic connections between AO neurons and bridge cells.
-The positions of synapses from AO neurons to bridge cells (red) and from bridge cells to AO neurons (light blue) are indicated.
+Balancer ciliated cells are depicted in light gray.\
+(F) Synaptic connections between ANNs and bridge cells.
+The positions of synapses from ANNs to bridge cells (magenta) and from bridge cells to ANNs (light blue) are indicated.
 The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view.
-Bridge cells are shown in light gray, AO neurons in Q1234 are in black, AO neurons in Q12 are in orange, and AO neurons in Q34 are in blue.\
+Bridge cells are shown in light gray.\
 (G) Connectivity matrix of the gravity-sensing neural circuit.
@@
 The thickness of the arrows and the numerical values indicate the number of synapses.
-AO neurons in Q1234 are shown in black, AO neurons in Q12 in yellow, AO neurons in Q34 in blue, balancer cell groups in gray, and bridge cell groups in pink.](figures/Figure3.png)
+ANN Q1-4 are shown in blue, ANN Q1Q2 in orange, ANNs in Q3Q4 in pink, balancer cell groups in gray, and bridge cell groups in yellow.](figures/Figure3.png)
 
@@
 Using connectome analysis, we aimed to evaluate the neural circuit differences by analyzing the control patterns of left and right balancer cilia in the tentacular and sagittal planes.
-Specifically, we hypothesized that differences in control between Q1 (Q2) and Q3 (Q4) regions, as well as between Q1 (Q4) and Q2 (Q3) regions, would reveal distinct functions of AO neurons (e.g., AO neuron Q12/Q34 vs.
-AO neuron Q1234).
+Specifically, we hypothesized that differences in control between Q1 (Q2) and Q3 (Q4) regions, as well as between Q1 (Q4) and Q2 (Q3) regions, would reveal distinct functions of ANNs (e.g., ANN Q1Q2/Q3Q4 vs.
+ANN Q1-4).
 For the sagittal plane, we recorded balancer ciliary movements in the Q1(4) and Q2(3) regions (left and right sides of the image) and, for the tentacular plane, in the Q1(3) and Q2(4) regions, using a high-speed camera (100 fps) for 2 minutes (12,000 frames).
@@
 Interestingly, while the re-beat timing showed no significant differences between the sagittal and tentacular planes, arrest timing exhibited greater variability in the tentacular plane.
-Comparing these findings with neural circuit diagrams suggests that in the sagittal plane, the same AO neurons (Q12 or Q34) project to both balancer cell groups, transmitting signals simultaneously and enabling synchronization.
-In contrast, in the tentacular plane, separate AO neurons project to different regions, potentially introducing an additional synaptic step, which may account for the observed variability in arrest timing.
+Comparing these findings with neural circuit diagrams suggests that in the sagittal plane, the same ANNs (Q1Q2 or Q3Q4) project to both balancer cell groups, transmitting signals simultaneously and enabling synchronization.
+In contrast, in the tentacular plane, separate ANNs project to different regions, potentially introducing an additional synaptic step, which may account for the observed variability in arrest timing.
 
@@
 
-Larvae of Mnemiopsis leidyi (5 days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) and immediately transferred to liquid nitrogen for storage.
+Larvae of *Mnemiopsis leidyi* (5 days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) and immediately transferred to liquid nitrogen for storage.
 The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica).
@@
 
-The reconstruction of cells and connectivity was performed using CATMAID [@Saalfeld2009], analyzing a total of 620 serial sections.
+The reconstruction of cells and connectivity was performed using CATMAID [@Saalfeld2009], analyzing a total of 619 serial sections.
 To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server [@Saalfeld2009; @Schneider-Mizell2016].
@@
 
-Each cell is assigned a specific type name based on its category (balancer, biC, bridge, bristle, dense vesicle cell, dome, EF, IcMC, LB, lithocyte, monoC, non-C, plumose cell, SNN, ssCG, ANN, stCG), resulting in a total of 17 cell types.
+Each cell is assigned a specific type name based on its category (balancer, bridge, bristle, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types, or those that could not be classified: presence and number of cilia (biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)) in 4 ciliated cell types.
 Additionally, a quadrant number is appended to the cell type name following an underscore ("\_") to indicate the cell body’s location.
@@
 
-We conducted the analysis using the cydippid stage of *Mnemiopsis leidyi* at five days post-fertilization.
+We conducted the analysis using the cydippid stage of *M.
+leidyi* at five days post-fertilization.
 A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass.

### 2025-03-20 manuscript/Manuscript.qmd
MS: edit intro, results

@@
 subtitle: |
-  **Kei Jokura**$^{1,2,3,*}$, **Sanja Jasek**$^{1,4}$, **Lara Kewalow**$^{4}$, **Pawel Burkhardt**$^{5}$, **Gáspár Jékely**$^{1,4,*}$  
+  **Kei Jokura**$^{1,2,3,4*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewalow**$^{6}$, **Pawel Burkhardt**$^{5}$, **Gáspár Jékely**$^{1,2,6,*}$  
   <br><br>  
@@
   <br>
-  $^{2}$ Exploratory Research Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan  
+  $^{2}$ Biosciences, Faculty of Health and Life Sciences, University of Exeter, Exeter EX4 4QD, UK
   <br>
-  $^{3}$ National Institute for Basic Biology (NIBB), Okazaki, 444-8585, Japan  
+  $^{3}$ Exploratory Research Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan  
   <br>  
-  $^{4}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  
+  $^{4}$ National Institute for Basic Biology (NIBB), Okazaki, 444-8585, Japan  
   <br>  
   $^{5}$ Michael Sars Centre, University of Bergen, Norway  
-  <br><br>  
+  <br>
+  $^{6}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  <br><br>  
   **$^{*}$ Correspondence:**  
@@
   [gaspar.jekely@cos.uni-heidelberg.de](mailto:gaspar.jekely@cos.uni-heidelberg.de)
-
 mainfont: Arial
@@
 ## Abstract
-
-Understanding the evolution and function of neurons is crucial for elucidating the origins of animal behavior.
-Ctenophores, as the earliest-diverging lineage with a nervous system, provide a unique model for investigating neural circuit organization.
-In this study, we performed volumetric electron microscopy (EM) reconstruction of the apical organ in Mnemiopsis leidyi to characterize its neural network involved in gravity sensing.
-We reconstructed 909 cells, including novel syncytial neurons spanning multiple quadrants, which form complex synaptic networks with balancer cells—key mechanosensory components controlling ciliary movement.
-Our connectome analysis revealed that these syncytial neurons establish direct and indirect pathways to balancer cells, with "bridge cells" acting as afferent relay neurons.
-High-speed imaging of balancer cilia movements further demonstrated differential regulation between sagittal and tentacular planes, suggesting distinct functional roles of apical organ neurons in modulating posture and locomotion.
-These findings provide new insights into the structure and function of ctenophore neural circuits, shedding light on early nervous system evolution and mechanosensory integration in basal metazoans.
+<!--# I rewrote a bit the abstract, please check -->
+Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth.
+During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns.
+To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the apical organ of the ctenophore *Mnemiopsis leidyi*.
+We reconstructed 909 cells, including syncytial neurons that form a nerve net and synapse on the balancer cells.
+Our connectome analysis revealed that these syncytial neurons establish direct and indirect pathways to balancer cells, with "bridge cells" acting as relay neurons.
+High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation.
+Given that ctenophores branched off early from other animals, these results inform our understanding of the early divergence of nervous systems.
 
@@
 
-Understanding how neurons evolved and what their initial functions were is key to elucidating the processes by which animals developed flexible behaviors.
-Neurons do not function in isolation; rather, they form networks that integrate and process information, enabling the emergence of complex behaviors.
-Therefore, understanding the structure and function of neural circuits is essential for investigating the mechanisms underlying animal behavior.
+<!--# the paper is not so much about evolution, so I suggest to start by introducing the cteno balancer organ -->
 
-Recent comparative genomic studies [@Ryan2013; @Whelan2015a; @Whelan2017; @Whelan2023; @Li2021] and chromosome-level synteny analyses across different animal phyla [@Schultz2023] suggest that ctenophores are the sister group to all other metazoans and were the first animals to evolve a nervous system.
-This has led to attempts to compare genes and molecular modules in ctenophores using single-cell analyses, in which neuropeptides have been employed as neuronal markers [@Sebé-Pedrós2018; @Sachkova2021; @Hayakawa2022].
-However, the homologous relationships between ctenophore genes and those of other animals are highly divergent, limiting the feasibility of sequence-based comparisons.
-Given these challenges, we considered an alternative approach: linking neural circuits to physiological responses to gain insight into neural function.
+<!--# the things that need to be mentioned are -->
 
-Despite their phylogenetic position, ctenophores exhibit sophisticated and dynamic behavioral patterns in response to a wide range of external stimuli.
-These behaviors are thought to be regulated by multiple neural systems, including the subepithelial nerve net (SNN), mesogleal neurons, an apical organ, and tentacle nerves.
+Ctenophores are gelatinous ...
+ ciliary comb rows ... 
+ active swimming and reorientation ...
+
+<!--# then introduce the nervous system of ctenos -->
+
+Ctenophores show sophisticated and dynamic behavioral patterns and respond to a wide range of external stimuli.
+These behaviors are thought to be regulated by multiple neural systems, including the subepithelial nerve net (SNN), mesogleal neurons, an apical organ, and tentacular nerves.
 However, the mechanisms by which these neural systems control behavior remain largely unexplored.
+Molecular phylogeny and chromosome-level synteny analyses suggest that ctenophores may be the sister group to all other animals [@Ryan2013; @Whelan2015a; @Whelan2017; @Whelan2023; @Li2021; @Schultz2023].
+In agreement with this early divergence from all other animals with a nervous system, the ctenophore nervous system shows several unique features that suggest independent evolution from other animals <!--# cite some Moroz papers -->.
 
-Among these, we focused on the apical organ of ctenophores.
-In many bilaterian and non-bilaterian animals, an apical nervous system is located at one pole of the body, forming a specialized sensory structure that governs physiological functions and motor activities.
-In ctenophores, ultrastructural analyses have revealed that the aboral pole contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974; @Tamm1982].
-Additionally, immunostaining using an anti-tyrosylated α-tubulin antibody has identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].
+... syncytial nerve net
+... highly peptidergic
 
-In this study, we specifically examined the gravity-sensing system within the apical organ.
-The apical organ of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014].
+The molecular makeup of ctenophore neurons is so divergent, that the first single-cell analyses did not identify a neuronal cluster [@Sebe-Pedros2018].
+Only through the analysis of proneuropeptide markers were neuronal clusters identified [@Sachkova2021; @Hayakawa2022].
+... organisation largely unknown
+... synapse structure
+
+<!--# the phylogeny is not really relevant here, ctenos are interesting in their own right, so I would cut this a bit and we don't need to cite all the same genomic and tree papers, rater cite all the nice Tamm papers and EM papers etc -->
+
+Another unique feature of ctenophores is their aboral organ that contains the statocyst.
+Ultrastructural analyses have revealed that the aboral organ contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974; @Tamm1982].
+Immunostaining with an antibody labelling tyrosylated α-tubulin identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].
+
+<!--# then the makeup and function of the balancer -->
+
+...balancer in the aboral organ
+
+The apical organ, also known as aboral organ, of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014].
 The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014; @Tamm1982].
 Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron].
-Therefore, we conducted a connectome analysis to characterize the network structure of the apical organ and elucidate the neural circuits that link sensory input to behavioral responses in ctenophores.
+
+... balancer cilia, statolyth
+... tuning of frequency
+... pacemaker function
+... reorientation
+
+In this study, we combine volume EM and high-speed imaging to examine the gravity-sensing system within the apical organ of the ctenophore *Mnemiopsis leidyi*.
+Through connectome analysis, we uncovered the structure and synaptic organisation of the aboral nerve net.
+Our imaging experiments revealed correlated changes in the activity of balancer cilia, suggesting a coordinating activity for the aboral nerve net.
 
@@
 
-First, we embedded the entire body of the *Mnemiopsis leidyi* cydippid-stage larvae, five days post-fertilization, in Epon resin following high-pressure freezing.
-To reconstruct all synaptic-resolution connections among the cells comprising the apical organ, also known as the aboral organ, we prepared approximately 1,000 ultra-thin serial sections (50 nm thick) from the aboral tip of the larval body embedded in the resin block.
-Using a scanning electron microscope (Zeiss Gemini 500) with a resolution of 2.0 nm/pixel, we imaged only the region containing the apical organ.
-Subsequently, 620 of these images were stitched and aligned using TrakEM2.
-The resulting volumetric EM dataset had dimensions of 60 μm × 40 μm × 30 μm.
-We traced and annotated all cells in this dataset, ultimately reconstructing 909 cells, each with both nuclei and cell bodies intact.
+For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*.
+To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon).
+The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons.
+The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy).
+We imaged only the region containing the apical organ.
+Subsequently, 620 of these images were stitched and aligned in TrakEM2 <!--# add reference-->.
+The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm.
+We skeletonised and annotated  all cells in this dataset in CATMAID <!--#  ref-->, resulting in a reconstruction of 909 cells with nuclei and cell bodies intact.
+Most of the cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.
 
-Cells containing cilia were traced down to the tips of the cilia, and basal bodies were annotated accordingly.
+In cells containing cilia, we also traced cilia along their length and annotated all basal bodies.
 For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced.
-Each node was tagged, and skeletons were named and assigned multi-level annotations.
-As described later, many neurons we identified formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main trunk or other branches.
-In such cases, branch nodes were placed near the closest existing node and annotated accordingly.
+Each node was tagged <!--# is this true?
+not only a subset?
+-->, and skeletons were named and assigned multi-level annotations.
+As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches <!--#  ref Burkhardt-->.
+In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees).
 The entire skeletonized volume was composed of 134,591 nodes.
-However, 88 fragments could not be attached to somata-associated skeletons.
-Most of these fragments represented short skeletal branches that could not be traced beyond gaps or low-quality layers.
+88 fragments could not be attached to somata-associated skeletons.
+Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.
 
-Next, we decided to divide the entire apical organ into four broad quadrants to facilitate grouping the identified cells.
-The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the positions of the anal pores.
-This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development.
-We categorized the traced cells into four quadrants and designated these groups as Q1 through Q4 for clarity and consistency.
+Next, we divided the entire apical organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F)<!--#  add reference to figures also elsewhere-->.
+The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores <!--#  ref-->.
+This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development<!--# ref -->.
+We assigned each of the 909 skeletonised cells to one of the four quadrants.
 
@@
 
-Classical neural staining techniques do not provide clear images of the neurons at the aboral pole.
-However, ultrastructural studies have provided morphological evidence of elements resembling neurons, based on synaptic structures, located on the epithelial floor of the apical organs [@Horridge_1964; @Hernandez_Nicaise_1973].
-Based on these findings, we identified synaptic structures characteristic of ctenophore neurons in our data, using previously identified pre-synaptic triad morphological features, such as single-layered vesicles, smooth endoplasmic reticulum, and tightly packed mitochondria.
-In our study, we specifically identified synaptic sites and marked mitochondria (orange) as synaptic nodes.
-The regions where synaptic vesicles align were marked as connectors (light blue arrows) between cells across the membrane.
-These connectors link the synaptic nodes of the pre-synaptic cytoskeleton to the partner nodes of the post-synaptic cytoskeleton.
-As previously reported, the specialization of post-synaptic structures in ctenophores is not apparent, so we recognized synaptic vesicle clusters on the pre-synaptic membrane as the point of reference.
-Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells.
+Classical neural staining techniques did not provide clear images of neurons at the aboral pole.
+However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor <!--# what is this exactly?
+are there epithelial cells?
+-->of the apical organ [@Horridge_1964; @Hernandez_Nicaise_1973].
+In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion.
+At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) <!--# figure..
+-->.
+We could not detect specialized post-synaptic structures, in agreement with previous studies <!--# refs -->.
+Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells <!--# figure of the distribution -->.
 
-As mentioned above, following the identification of the presynaptic structures, we reconstructed three major Syncytial neurons.
-Each of these Syncytial neurons was a cell with multiple nuclei, with membranes fused by continuous plasma membranes.
-These neurons are distinct in morphology from the Syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously reconstructed in 3D [@Burkhardt2023].
-Our findings represent the second documented discovery of Syncytial neurons in ctenophores.
-Furthermore, these neurons exhibited clear morphological differences when compared to other sensory cells reported in the same study, such as mesogleal neurons and sensory cells with presynaptic structures and cilia (types 1-4)[@Burkhardt2023].
-Based on their distinct spatial relationships, we were able to classify these three Syncytial neurons into two categories.
-The first type is a larger "ANN_Q1-4," which possesses four (or possibly five?) nuclei spanning four quadrants.
-It contained X presynaptic structures.
-The second type consists of "ANN_Q1Q2" and "ANN_Q3Q4," each having two nuclei spanning two quadrants, and each containing X presynaptic structures.
+We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology.
+Each ANN had multiple nuclei <!--# add numbers e.g.
+(2-5 nuclei) -->, with anastomosing membranes <!--# show example EM pictures -->.
+These neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023].
+Our results reveal a second type of syncytial neurons in ctenophores.
+<!--# describe the differences and also mention the classical SNNs that Sanja reconstructed -->
+
+Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synalse on the SNN (types 1-4)[@Burkhardt2023].
+Based on their distinct size and position, we classified the ANNs into two types.
+The first type is a single large neuron (ANN_Q1-4) that has four (possibly five) nuclei and spans all four quadrants (Q1-Q4).
+This neuron has X <!--# add number --> presynaptic sites.
+The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each with two nuclei spanning two quadrants, and each containing X presynaptic structures.
+<!--#  -->
 

### 2025-03-20 manuscript/Manuscript.qmd
MS: edits

@@
 
-### Identification of Synaptic Structures in Syncytium Neurons
+### An aboral synaptic nerve net of syncytial neurons
 
@@
 Based on their distinct size and position, we classified the ANNs into two types.
-The first type is a single large neuron (ANN_Q1-4) that has four (possibly five) nuclei and spans all four quadrants (Q1-Q4).
-This neuron has X <!--# add number --> presynaptic sites.
-The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each with two nuclei spanning two quadrants, and each containing X presynaptic structures.
+The first type is a single large neuron (ANN_Q1-4) spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei This neuron has X <!--# add number --> presynaptic sites.
+The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and with two nuclei, and each containing X presynaptic structures.
 <!--#  -->
 
-![**Figure 2. Identification of Three Syncytial Neurons in the Apical Organ**\
+![**Figure 2. Organisation of the aboral synaptic nerve net**\
 (A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (blue).
@@
 
-### Identification the Gravity-Sensitive Neural Circuit via the Syncytial Neurons Network
+### Synaptic connectome of the gravisensory organ
 
-We discovered that ANN neurons form synaptic connections with each other, and that ANN_Q1-4 forms self-synapses, or “autapses.” To our knowledge, previous reports have not identified synaptic connections between subepithelial nerve net (SNN) neurons.
-Thus, our results represent the first report of synaptic connectivity between neurons in the apical organ, forming a network in ctenophores (subject to confirmation).
-These ANN networks were found to form many presynaptic structures in relation to the gravity-sensing balancer cells.
+Our synapse annotation revealed synapses between the ANN neurons and synapses that ANN_Q1-4 forms on itsels (autapses). The ANN neurons also form synapses on the gravity-sensing balancer cells.
 
-Balancer cells are monociliated cells, and their cilia protrude longer than those of other monociliated cells.
-Moreover, the cilia of balancer cells are bundled into four groups at the center of the apical organ, forming a compound cilium.
-Based on these features, we identified and classified the balancer cells.
-The cellular arrangement differed clearly when viewed in the lateral view, sagittal plane, and tentacular plane, with the cell bodies gathering toward the apical organ in the tentacular plane.
-In each quadrant, the number of cells was as follows: Q1: 37, Q2: 32, Q3: 32, Q4: 28.
-Each cell contained 3 to 10 ?
-mitochondria.
+Balancer cells are monociliated cells with long motile cilia that form four bundles of compound cilia, one in each quadrant.
+These bundles come together at the centre of the apical organ to support the cellular mass of the statolyth.
+The bending of the cilia and the position of somata differed clearly when viewed laterally from the sagittal or the tentacular plane (Figure 3B).
+In each quadrant, there were between 28-37 balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28).
+Each cell contained 3 to 10 <!-- check --> mitochondria.
 

### 2025-03-23 manuscript/Manuscript.qmd
MS: edit balancer imaging section

@@
 
-Our synapse annotation revealed synapses between the ANN neurons and synapses that ANN_Q1-4 forms on itsels (autapses). The ANN neurons also form synapses on the gravity-sensing balancer cells.
+Our synapse annotation revealed synapses between the ANN neurons and synapses that ANN_Q1-4 forms on itself (autapses). The ANN neurons also form synapses on the gravity-sensing balancer cells.
 
@@
 
-From ANN_Q1Q2, synapses were formed with 6 of the 37 balancer cells in the Q1 region, and 8 of the 32 balancer cells in the Q2 region.
-Similarly, from ANN_Q3Q4, synapses were formed with 1 of the 32 balancer cells in the Q3 region, and 5 of the 28 balancer cells in the Q4 region.
-ANN_Q1-4 formed input synapses with balancer cells in the Q1 region (7 cells), Q2 region (11 cells), Q3 region (6 cells), and Q4 region (10 cells).
-Some balancer cells received inputs from both ANN_Q1Q2 or ANN_Q3Q4 and ANN_Q1-4.
-While previous studies have suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez-Nicaise1974], our data did not reveal any synaptic inputs from balancer cells to ANN neurons.
-However, we identified "bridge cells" that actively formed input synapses with ANN neurons, highlighting their potential role in the network.
+ANN_Q1-4 formed synapses on balancer cells in all four quadrants (on 7 cells in Q1, 11 cells in Q2, 6 cells in Q3 and 10 cells in Q4) while ANN_Q1Q2 and ANN_Q3Q4 synapsed to balancer cells in their respective quadrants (ANN_Q1Q2 on 6/37 cells in Q1 and 8/32 cells in Q2; ANN_Q3Q4 on 1/32 cells in Q3 and 5/28 cells in Q4).
+Some balancer cells received inputs from both ANN_Q1-4 and either of ANN_Q1Q2 or ANN_Q3Q4.
+While previous studies suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez-Nicaise1974], our data did not reveal any presynaptic sites in balancer cells.
 
-### Identification of Neuron-like Bridge Cells Forming Afferent Synapses in the Gravity-Sensitive Neural Circuit
+The second group of cells that form synaptic contact with the ANN were the bridge cells.
+Bridge cells, first described by Tamm et al.
+in 2002 [@Tamm_2002], are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge.
+Their somata are located at the base of the paired balancer-cell clusters along the tentacle surface and extend across the sagittal plane towards the base of the opposite balancer cells.
+Bridge cells form two distinct cell groups across the sagittal plane, in the Q1Q2 and Q3Q4 regions.
+In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12 cells.
 
-From our data, we identified several cell groups that form multiple afferent synapses with ANNs.
-These cells, based on their morphological features, were found to be the bridge cells first described by Tamm et al.
-in 2002 [@Tamm_2002].
-These cells are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge.
-They originate from the base of paired balancer cells along the tentacle surface and extend across the sagittal plane toward the base of the opposite balancer cells.
-In regions where the mitochondria of the bridge cells are localized (approximately 30%), a presynaptic triad structure, similar to that of ANNs, was found, containing synaptic vesicles and smooth endoplasmic reticulum.
+Bridge cells had presynaptic sites with the typical presynaptic triad structure near 30% of their mitochondria.
+These bridge-cell synapses were formed on ANN neurons or other bridge cells.
+Bridge cells in the Q1Q2 region formed synapses on ANN_Q1Q2 (3 cells) or on both ANN_Q1Q2 and ANN_Q1-4 (2 cells).
+Bridge cells in Q3Q4 synapsed on ANN_Q3Q4 (1 cell) or ANN_Q1-4 (6 cells).
 
-Our three-dimensional reconstruction data revealed that the bridge cells form two distinct cell groups across the sagittal plane, between the Q1Q2 and Q3Q4 regions.
-In the Q1Q2 region, 14 cells were identified, while in the Q3Q4 region, 12 cells were identified, totaling 26 bridge cells.
-Nearly all of these bridge cells (25 out of 26) exhibited afferent synapses from ANNs.
-For bridge cells located in the Q1Q2 region, synaptic inputs came primarily from ANN_Q1Q2 (11 cells), ANN_Q1-4 (1 cell), or both (2 cells).
-For bridge cells located in the Q3Q4 region, synaptic inputs were received from ANN_Q3Q4 (1 cell), ANN_Q1-4 (7 cells), or both (1 cell).
-In other words, bridge cells in the Q1Q2 region mainly received input from ANN_Q1Q2, while those in the Q3Q4 region received input primarily from ANN_Q1-4, showing a distinct pattern of input in the two regions.
+Nearly all bridge cells (25/26) also received synaptic input from ANNs.
+Bridge cells in Q1Q2 received inputs from ANN_Q1Q2 (11 cells), ANN_Q1-4 (1 cell), or both (2 cells).
+In Q3Q4, synapses on bridge cells were from ANN_Q3Q4 (1 cell), ANN_Q1-4 (7 cells), or both (1 cell).
 
-Some bridge cells also formed afferent synapses with ANN neurons.
-For example, bridge cells in the Q1Q2 region (3 cells) formed synapses with ANN_Q1Q2 or (2 cells) both ANN_Q1Q2 and ANN_Q1-4.
-Bridge cells in the Q3Q4 region formed synapses with ANN_Q3Q4 (1 cell) or ANN_Q1-4 (6 cells).
-A notable difference was observed between the Q1Q2 and Q3Q4 regions in the proportion of synaptic inputs from bridge cells to ANN neurons.
+In both The Q1Q2 and the Q3Q4 regions, bridge cells also formed synapses with adjacent bridge cells. However, we found no synapses across the sagittal plane to bridge cells in the opposite region.
 
-Interestingly, in both regions, bridge cells also formed synapses with adjacent bridge cells.
-However, no synaptic input was found from bridge cells across the sagittal plane to those in the opposite region.
-To analyze the grouped synaptic connectivity graph, we classified cell types within each region, collapsed cells of the same type into a single node, summed the number of synapses, and explored directed pathways to effectors (balancer cell groups).
-The results revealed a feedback pathway from ANN neurons through the synaptic connections formed by bridge cells, thereby shedding light on the neural circuit structure involving these cells.
+To analyze the synaptic connectivity graph of the balancer organ and the ANN, we grouped cells of the same type and in the same region into a single node, summed the number of synapses, and laid out the network to reflect the anatomy of the four quadrants (Figure 3H).
+The network is characterised by feedback connections among ANN neurons and between ANN and bridge cells with a clear rotational (or mirror) symmetry.
+Q1Q2 and Q3Q4 also form separate local circuits centered around their regional ANN neurons while ANN_Q1-4 mediates global connectivity across the entire balancer organ.
+Notably, there were no synapses from the mechanosensory balancer cells to the ANN or from ANN to motor cells (e.g.
+ciliary groove), contrary to our initial expectation of a sensory-motor or input-output model of balancer function.
+
+The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, EF cells and several non-ciliated, monociliated or biciliated cells.
+These cell types and synaptic connections are outside the gravisensory organ and will be described elsewhere.
 
@@
 
-### Control of Balancer Ciliary Movements by Gravity-Sensitive Neural Circuits
+### Dynamics of balancer cilia imply a coordinating function for the nerve net
 
-To investigate the function of gravity-sensitive circuits, we conducted a comparative analysis of balancer ciliary movements across different regions.
+To investigate the regulation of balancer function, we carried out a high-speed video-microscope analysis of balancer cilia.
 Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination.
-Furthermore, it has been suggested that differences in the asymmetry of statolith morphology and the elongation direction of balancer cilia between the tentacular and sagittal planes could result in variations in the force received from the statolith [@Tamm_2014; @Tamm2015].
-Building on these findings, we standardized conditions for observing balancer cilia by tilting the microscope by 90 degrees and fixing samples such that the aboral-oral axis was aligned parallel to the vertical stage, with the aboral side facing upward.
-Using connectome analysis, we aimed to evaluate the neural circuit differences by analyzing the control patterns of left and right balancer cilia in the tentacular and sagittal planes.
-Specifically, we hypothesized that differences in control between Q1 (Q2) and Q3 (Q4) regions, as well as between Q1 (Q4) and Q2 (Q3) regions, would reveal distinct functions of ANNs (e.g., ANN Q1Q2/Q3Q4 vs.
-ANN Q1-4).
-For the sagittal plane, we recorded balancer ciliary movements in the Q1(4) and Q2(3) regions (left and right sides of the image) and, for the tentacular plane, in the Q1(3) and Q2(4) regions, using a high-speed camera (100 fps) for 2 minutes (12,000 frames).
-Regions of interest (ROIs) were defined in areas where brightness changes indicated ciliary beating.
-We measured these brightness changes to generate dynamics graphs of balancer ciliary movements for both planes.
+Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014; @Tamm2015].
+To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector.
+We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage.
 
-During the 2-minute recordings, balancer cilia exhibited movements that could be fast, slow, or stop abruptly (arrest) and start moving again (re-beat).
-Occasionally, large contraction responses caused the entire cydippid to move out of frame, and data from these moments were excluded.
-We focused on two metrics for left-right comparison: (1) the timing of ciliary arrest and (2) the timing of re-beat initiation following an arrest.
-Due to the low baseline beating frequency of the cilia, rapid synaptic input responses were not evaluated, suggesting that ciliary beat regulation may involve non-synaptic control mechanisms.
-The analysis revealed that both arrest and re-beat events were synchronized between left and right balancers in both planes within a few milliseconds, suggesting that these events are governed by synaptic input from neurons.
-Interestingly, while the re-beat timing showed no significant differences between the sagittal and tentacular planes, arrest timing exhibited greater variability in the tentacular plane.
-Comparing these findings with neural circuit diagrams suggests that in the sagittal plane, the same ANNs (Q1Q2 or Q3Q4) project to both balancer cell groups, transmitting signals simultaneously and enabling synchronization.
-In contrast, in the tentacular plane, separate ANNs project to different regions, potentially introducing an additional synaptic step, which may account for the observed variability in arrest timing.
+In larvae with their sagittal plane facing the objective, we could compared balancer-cilia movements between Q1(4) and Q2(3).
+In other larvae oriented in the tentacular plane, we could simultaneously image Q1(3) and Q2(4).
+We used a high-speed camera and recorded at 100 fps for 2 minutes (12,000 frames).
+To analyse ciliary beating, we selected regions of interest (ROIs) in areas where brightness changes indicated ciliary beating.
+Ciliary beating was both manually quantified and plotted as kymographs.
+
+During the two-minute recordings, balancer cilia could beat fast, slow, or exhibit abrupt stops (arrest) and start moving again (re-beat).
+Occasionally, large body-contractions moved the entire cydippid out of frame, and data from these episods were excluded.
+We focused on three metrics for inter-quadrant comparison: (1) the timing of ciliary arrest, (2) the timing of re-beat after an arrested phase, and (3) ciliary beat frequency.
+
+We found that arrest and re-beat events were synchronized between balancers across the sagittal plane.
+However, while the timing of re-beat was also near-simultaneous along the tentacular plane, arrest timing was offset by up to 2.xy <!-- add max value--> seconds (Figure 4) <!-- save source data for all plots in manuscript/source_data with the naming convention e.g.
+Figure 1A--source data 1 -->.
+Overall, these data reveal that arrests are only coordinated between Q1-Q2 and Q3-Q4, whereas re-beat is coordinated over all four quadrants.
+
+Evaluating these data in light of the circuit diagram suggests that shared ANN inputs (ANN_Q1Q2 or ANN_Q3Q4) to a pair of balancers along the sagittal plane may underlie their synchronized arrests.
+In contrast, in the tentacular plane, separate ANNs innervate the balancer pairs.
+At the same time, the ANN_Q1-4 neuron synapses on all four balancers, hinting at a neural substrate for their synchronised re-beat.
 

### 2025-03-24 manuscript/Manuscript.qmd
MS: main text

@@
 ---
-title: "Neural Circuit Organization of the Ctenophore Apical Organ"
+title: "Neural Circuit Organization of the Ctenophore Aboral Organ"
 
@@
 ## Abstract
+
 <!--# I rewrote a bit the abstract, please check -->
+
 Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth.
 During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns.
-To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the apical organ of the ctenophore *Mnemiopsis leidyi*.
+To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*.
 We reconstructed 909 cells, including syncytial neurons that form a nerve net and synapse on the balancer cells.
@@
 
-Ctenophores are gelatinous ...
- ciliary comb rows ... 
- active swimming and reorientation ...
+Ctenophores are gelatinous ... ciliary comb rows ... active swimming and reorientation ...
 
@@
 Ctenophores show sophisticated and dynamic behavioral patterns and respond to a wide range of external stimuli.
-These behaviors are thought to be regulated by multiple neural systems, including the subepithelial nerve net (SNN), mesogleal neurons, an apical organ, and tentacular nerves.
+These behaviors are thought to be regulated by multiple neural systems, including the subepithelial nerve net (SNN), mesogleal neurons, an aboral organ, and tentacular nerves.
 However, the mechanisms by which these neural systems control behavior remain largely unexplored.
@@
 
-... syncytial nerve net
-... highly peptidergic
+... syncytial nerve net ... highly peptidergic
 
@@
 Only through the analysis of proneuropeptide markers were neuronal clusters identified [@Sachkova2021; @Hayakawa2022].
-... organisation largely unknown
-... synapse structure
+...
+organisation largely unknown ...
+synapse structure
 
@@
 Ultrastructural analyses have revealed that the aboral organ contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974; @Tamm1982].
-Immunostaining with an antibody labelling tyrosylated α-tubulin identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].
+Immunostaining with an antibody labelling tyrosylated α-tubulin identified a neural structure called the deep nerve net within the aboral organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].
 
@@
 
-The apical organ, also known as aboral organ, of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014].
+The aboral organ, also known as aboral organ, of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014].
 The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014; @Tamm1982].
@@
 
-... balancer cilia, statolyth
-... tuning of frequency
-... pacemaker function
-... reorientation
+... balancer cilia, statolyth ... tuning of frequency ... pacemaker function ... reorientation
 
-In this study, we combine volume EM and high-speed imaging to examine the gravity-sensing system within the apical organ of the ctenophore *Mnemiopsis leidyi*.
+In this study, we combine volume EM and high-speed imaging to examine the gravity-sensing system within the aboral organ of the ctenophore *Mnemiopsis leidyi*.
 Through connectome analysis, we uncovered the structure and synaptic organisation of the aboral nerve net.
@@
 
-### Volume EM reconstruction of the *Mnemiopsis* apical organ
+### Volume EM reconstruction of the *Mnemiopsis* aboral organ
 
@@
 The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy).
-We imaged only the region containing the apical organ.
+We imaged only the region containing the aboral organ.
 Subsequently, 620 of these images were stitched and aligned in TrakEM2 <!--# add reference-->.
@@
 
-Next, we divided the entire apical organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F)<!--#  add reference to figures also elsewhere-->.
+Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F)<!--#  add reference to figures also elsewhere-->.
 The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores <!--#  ref-->.
@@
 
-![**Figure 1. Morphology of the apical organ in *Mnemiopsis leidyi***\
-(A) Whole-body image of a 5-day-old *M. leidyi* cydippid larva in lateral view. The boxed region indicates the apical organ (ao). Scale bar: 100 µm.\
+![**Figure 1. Morphology of the aboral organ in *Mnemiopsis leidyi***\
+(A) Whole-body image of a 5-day-old *M. leidyi* cydippid larva in lateral view. The boxed region indicates the aboral organ (ao). Scale bar: 100 µm.\
 (B) Schematic diagram of the aboral view of a cydippid larva.
@@
 The two primary viewing angles are referred to as the sagittal plane (S) and the tentacular plane (T).
-Abbreviations: ao, apical organ; cg, ciliated groove; cr, comb plates; tb, tentacle bulbs.\
-(C) Enlarged views of the apical organ from different perspectives in a cydippid larva using a DIC microscope.
-The left panel presents an aboral view of the apical organ.
-The middle panel shows a lateral view of the apical organ in the sagittal (S) plane, while the right panel displays a lateral view in the tentacular (T) plane.
+Abbreviations: ao, aboral organ; cg, ciliated groove; cr, comb plates; tb, tentacle bulbs.\
+(C) Enlarged views of the aboral organ from different perspectives in a cydippid larva using a DIC microscope.
+The left panel presents an aboral view of the aboral organ.
+The middle panel shows a lateral view of the aboral organ in the sagittal (S) plane, while the right panel displays a lateral view in the tentacular (T) plane.
 Abbreviations: bal, balancer; cg, ciliated grooves; do, dome cilia; li, lithocyte.
 Scale bar: 50 µm.\
-(D) Schematic representations of the apical organ from three different perspectives.
+(D) Schematic representations of the aboral organ from three different perspectives.
 The left panel illustrates the aboral view, the middle panel shows the lateral view in the sagittal plane, and the right panel presents the lateral view in the tentacular plane.
@@
 Scale bar: 10 µm.\
-(F) Three-dimensional reconstruction of cells composing the apical organ, displayed from different perspectives.
-The left panel presents an aboral view of the apical organ.
+(F) Three-dimensional reconstruction of cells composing the aboral organ, displayed from different perspectives.
+The left panel presents an aboral view of the aboral organ.
 The middle panel shows a sagittal plane view, while the right panel provides a tentacular plane view.
@@
 are there epithelial cells?
--->of the apical organ [@Horridge_1964; @Hernandez_Nicaise_1973].
+-->of the aboral organ [@Horridge_1964; @Hernandez_Nicaise_1973].
 In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion.
@@
 The spheres represent the positions of individual nuclei.
-The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
 (B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei.
 The spheres indicate the positions of the nuclei.
-The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
 (C) Graph showing the number of mitochondria per cell and the proportion of mitochondria involved in synapse formation for each cell type.
@@
 Balancer cells are monociliated cells with long motile cilia that form four bundles of compound cilia, one in each quadrant.
-These bundles come together at the centre of the apical organ to support the cellular mass of the statolyth.
+These bundles come together at the centre of the aboral organ to support the cellular mass of the statolyth.
 The bending of the cilia and the position of somata differed clearly when viewed laterally from the sagittal or the tentacular plane (Figure 3B).
@@
 The spheres represent the positions of individual nuclei.
-The left panel shows an aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+The left panel shows an aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
 (D) Mitochondrial localization within bridge cells and associated presynaptic triad structures.
 Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.
-The left panel shows a dorsal view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
+The left panel shows a dorsal view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
 (E) Synaptic connections from ANN neurons to balancer ciliated cells.
@@
 
-We identified three syncytial neurons with multiple nuclei in the apical organ of five-day-old *M.
+We identified three syncytial neurons with multiple nuclei in the aboral organ of five-day-old *M.
 leidyi* cydippid larvae.
@@
 previously reported that in one-day-old larvae, the sub-epithelial nerve net (SNN) contains neurons that form syncytia [@Burkhardt2023].
-However, the newly discovered syncytial neurons embedded within the apical organ cell cluster did not exhibit the characteristic "pearl necklace-like structure" of the SNN, which consists of microtubules connecting neurons.
+However, the newly discovered syncytial neurons embedded within the aboral organ cell cluster did not exhibit the characteristic "pearl necklace-like structure" of the SNN, which consists of microtubules connecting neurons.
 Instead, their morphology was more reminiscent of fiber cells previously described in Placozoa [@Mayorova2021], with cell bodies extending through intercellular spaces.
@@
 
-In the this study, we found that three neurons are closely positioned within the confined space of the apical organ, potentially enabling a single neuron to control distant ciliary movements in a synchronized manner.
+In the this study, we found that three neurons are closely positioned within the confined space of the aboral organ, potentially enabling a single neuron to control distant ciliary movements in a synchronized manner.
 Moreover, the presence of neurons projecting to the same ciliated cells in distinct patterns suggests that these neurons may serve opposing functions—one inhibiting ciliary activity while the other promotes it, or one modulating ciliary beat frequency.
@@
 Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores [@HORRIDGE1965; @vinnikov1974e; @Krisch1973; Aronova1974].
-By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the apical organ.
+By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the aboral organ.
 However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets.
-In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the apical organ [@Sachkova2021; @Hayakawa2022].
-These findings suggest that many sensory cells in the apical organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis [@Jékely2021].
+In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the aboral organ [@Sachkova2021; @Hayakawa2022].
+These findings suggest that many sensory cells in the aboral organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis [@Jékely2021].
 While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.
@@
 
-Surprisingly, our study identified bridge cells as the primary cell type forming synaptic inputs onto neurons in the apical organ.
+Surprisingly, our study identified bridge cells as the primary cell type forming synaptic inputs onto neurons in the aboral organ.
 Bridge cells were first described by Tamm et al.
@@
 
-The neural cells identified in the apical organ exhibit a divergent feedforward connectivity pattern, where a single neuron forms presynaptic connections with multiple downstream targets.
+The neural cells identified in the aboral organ exhibit a divergent feedforward connectivity pattern, where a single neuron forms presynaptic connections with multiple downstream targets.
 Such a circuit design minimizes unwanted signal interference, reduces processing delays, and enables relatively fast response execution [@Luo2021].

### 2025-03-28 manuscript/Manuscript.qmd
ed main text

@@
 We imaged only the region containing the aboral organ.
-Subsequently, 620 of these images were stitched and aligned in TrakEM2 <!--# add reference-->.
+Subsequently, 620 of these images were stitched and aligned in TrakEM2 [@cardona2012].
 The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm.
-We skeletonised and annotated all cells in this dataset in CATMAID <!--#  ref-->, resulting in a reconstruction of 909 cells with nuclei and cell bodies intact.
+We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009], resulting in a reconstruction of 909 cells with nuclei and cell bodies intact.
 Most of the cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.

### 2025-03-30 manuscript/Manuscript.qmd
ed main text

@@
 -->, and skeletons were named and assigned multi-level annotations.
-As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches <!--#  ref Burkhardt-->.
+As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023].
 In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees).
@@
 
-Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F)<!--#  add reference to figures also elsewhere-->.
-The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores <!--#  ref-->.
-This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development<!--# ref -->.
+Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F) [@Martindale1999].
+The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999].
+This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999].
 We assigned each of the 909 skeletonised cells to one of the four quadrants.
@@
 
-Classical neural staining techniques did not provide clear images of neurons at the aboral pole.
-However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor <!--# what is this exactly?
-are there epithelial cells?
--->of the aboral organ [@Horridge_1964; @Hernandez_Nicaise_1973].
+Classical neural staining techniques did not provide clear images of neurons at the aboral organ.
+However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise_1973].
 In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion.
@@
 -->.
-We could not detect specialized post-synaptic structures, in agreement with previous studies <!--# refs -->.
+We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise_1973].
 Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells <!--# figure of the distribution -->.

### 2025-04-02 manuscript/Manuscript.qmd
edit main text

@@
 In each quadrant, there were between 28-37 balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28).
-Each cell contained 3 to 10 <!-- check --> mitochondria.
+Each cell contained 3 to 10 <!--# check --> mitochondria.
 
@@
 We found that arrest and re-beat events were synchronized between balancers across the sagittal plane.
-However, while the timing of re-beat was also near-simultaneous along the tentacular plane, arrest timing was offset by up to 2.xy <!-- add max value--> seconds (Figure 4) <!-- save source data for all plots in manuscript/source_data with the naming convention e.g.
+However, while the timing of re-beat was also near-simultaneous along the tentacular plane, arrest timing was offset by up to 2.xy <!--# add max value--> seconds (Figure 4) <!--# save source data for all plots in manuscript/source_data with the naming convention e.g.
 Figure 1A--source data 1 -->.

### 2025-04-04 manuscript/Manuscript.qmd
MS: edits
MS and correlation plot: edits

@@
 ---
-title: "Neural Circuit Organization of the Ctenophore Aboral Organ"
-
+title: "Neural Connectome of the Ctenophore Statocyst"
 subtitle: |
@@
 To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*.
-We reconstructed 909 cells, including syncytial neurons that form a nerve net and synapse on the balancer cells.
-Our connectome analysis revealed that these syncytial neurons establish direct and indirect pathways to balancer cells, with "bridge cells" acting as relay neurons.
+We reconstructed 909 cells, including syncytial neurons that form a nerve net.
+The syncytial neurons synapse on the balancer cells and also form reciprocal connection with the bridge cells that span the statocyst.
 High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation.
-Given that ctenophores branched off early from other animals, these results inform our understanding of the early divergence of nervous systems.
+Our results suggest a coordinating rather than effector function for the nerve net and inform our understanding of the diversity of nervous-systems organisation across animals.
 
@@
 
-Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synalse on the SNN (types 1-4)[@Burkhardt2023].
+Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023].
 Based on their distinct size and position, we classified the ANNs into two types.
@@
 
+<!--# discuss frequency data -->
+
 Evaluating these data in light of the circuit diagram suggests that shared ANN inputs (ANN_Q1Q2 or ANN_Q3Q4) to a pair of balancers along the sagittal plane may underlie their synchronized arrests.
@@
 Additionally, our analysis of balancer ciliary movements, based on neural circuit data, revealed differences in ciliary arrest control arising from circuit variations.
+
+<!--#  discuss that it is a complete circuit, with no neuronal connections in or out - effector output is likely by hydrodynamics, as suggested by Tamm -->
 
@@
 A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass.
-Filtered natural seawater was introduced under the coverslip, along with the cydippids.
+Filtered natural seawater was introduced under the coverslip, along with the cy3dippids.
 By gently moving the coverslip, the desired orientation for observation was adjusted, and once positioned, slight pressure was applied to immobilize the cydippids.

### 2025-04-04 analysis/scripts/Rolling Pearson Correlation.R

@@ -148,8 +148,9 @@ plot_boxplot <- function(results, window) {
   }
 
   p <- ggplot(results, aes(x = plane, y = mean_correlation, color = plane)) +
-    geom_boxplot(fill = "lightgray", alpha = 0.5, outlier.shape = NA) +
+  #  geom_boxplot(fill = "lightgray", alpha = 0.5, outlier.shape = NA) +
     geom_jitter(width = 0.2, size = 3, alpha = 0.6) +
+    geom_violin(size = 1, alpha = 0.4, trim = FALSE, scale = "count") +
     scale_color_manual(values = c("S" = "#0F52BA", "T" = "#F28500")) +
     labs(
       x = "Plane",
@@ -166,24 +167,8 @@ plot_boxplot <- function(results, window) {
   print(test)
 }
 
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-# Run
-window_sizes <- c(10, 20, 30)
+# Run ------------
+window_sizes <- c(3, 10, 20, 30)
 for (window in window_sizes) {
   results <- data.frame(file_name = character(), plane = character(), mean_correlation = numeric())
   for (i in 1:nrow(filtered_files)) {

### 2025-04-07 manuscript/Manuscript.qmd
MS: edit discussion, refs

@@
 
-In this study, we traced all the constituent cells of the aboral organ in five-day-old larvae of *Mnemiopsis leidyi* and completed a comprehensive neural map.
-As a result, we identified a novel type of syncytial neuron with a previously unreported morphology.
-Furthermore, we found that three such syncytial neurons project to distinct quadrants, integrating with balancer cell groups to form a gravity-sensing neural circuit.
-This study represents the first complete neural circuit map reported in ctenophores.
-Additionally, our analysis of balancer ciliary movements, based on neural circuit data, revealed differences in ciliary arrest control arising from circuit variations.
+In this study, we generated a comprehensive neural map of the aboral organ of a five-day-old *Mnemiopsis leidyi*.
+We described a novel type of syncytial neuron and analysed its synaptic connections to balancer-cell groups across the four quadrants of the aboral organ.
+Given that all neurons are intrinsic to the aboral organ and our EM volume covers the entire organ, we could recover a complete circuit map, the first for a ctenophore.
+Contrary to our expectations, the nerve net has no motor connections to effector organs, such as the ciliary grooves or comb rows.
+Our high-speed imaging suggests that the nerve net instead coordinates the activity of balancer cilia.
+This agrees with a hydrodynamic model of graviorientation <!-- add ref to Tamm -->, with balancer cilia directly responding to statolyth load and acting as pacemakers for the comb rows via the ciliated grooves.
+The nerve net could in turn tune or reset sensitivites and valence of balancer cells, in a coordinatory role, as we will discuss below.
 
-<!--#  discuss that it is a complete circuit, with no neuronal connections in or out - effector output is likely by hydrodynamics, as suggested by Tamm -->
+### A New Type of Syncytial Neuron
 
-### Identification of a Previously Unreported Syncytial Neuron
+We identified three syncytial neurons with multiple nuclei in the aboral organ of the cydippid larva.
+Burkhurdt et al.
+previously reported syncytial neurons in the sub-epithelial nerve net (SNN) of a one-day-old larva [@Burkhardt2023].
+The syncytial neurons in the aboral organ did not exhibit the characteristic "pearl necklace-like structure" of the SNN neurons.
+Instead, their morphology was more reminiscent of fiber cells found in Placozoa [@Mayorova2021], with cell projections extending through intercellular spaces.
+In our volume, we also identified two SNN-like neurons with a pearl necklace-like morphology, indicating that the differences between ANN and SNN are not due to differences in stage or sample preparation.
+This confirms the ANN neurons as a second distinct type of syncytial neuron in ctenophores.
 
-We identified three syncytial neurons with multiple nuclei in the aboral organ of five-day-old *M.
-leidyi* cydippid larvae.
-Burkhurdt et al.
-previously reported that in one-day-old larvae, the sub-epithelial nerve net (SNN) contains neurons that form syncytia [@Burkhardt2023].
-However, the newly discovered syncytial neurons embedded within the aboral organ cell cluster did not exhibit the characteristic "pearl necklace-like structure" of the SNN, which consists of microtubules connecting neurons.
-Instead, their morphology was more reminiscent of fiber cells previously described in Placozoa [@Mayorova2021], with cell bodies extending through intercellular spaces.
-These morphological characteristics suggest that the newly identified neurons represent a second distinct type of syncytial neuron in ctenophores.
+### Functional Diversification of the Aboral Syncytial Neurons
 
-### Functional Diversification of the Three Syncytial Neurons
-
-The three identified neurons may be classified into two distinct functional types based on their spatial arrangement and their influence on the differential control of balancer ciliary movement between the quadrants.
-In metazoans, a single neuron can directly provide synaptic input to multiple ciliated cells at a distance, synchronizing their movements.
-For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating [@Verasztó2017].
-Additionally, a pair of serotonergic neurons, designated Ser-h1, extend axons that cross at the midline; the left Ser-h1_l predominantly innervates right-sided ciliated cells, while the right Ser-h1_r mainly projects to the left-side ciliated cells.
-Activation of Ser-h1 induces ciliary beating initiation and increases beat frequency.
+The three ANN neurons form two functional types based on their spatial arrangement and differential innervation of balancer cells.
 
@@
 While our synaptic ultrastructure analysis did not allow us to distinguish excitatory from inhibitory neurons, future electrophysiological investigations will be essential to elucidate their functional properties.
+
+In metazoans, a single neuron can directly provide synaptic input to multiple ciliated cells at a distance, synchronizing their movements.
+For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating [@Veraszto2017].
+Additionally, a pair of serotonergic neurons, designated Ser-h1, extend axons that cross at the midline; the left Ser-h1_l predominantly innervates right-sided ciliated cells, while the right Ser-h1_r mainly projects to the left-side ciliated cells.
+Activation of Ser-h1 induces ciliary beating initiation and increases beat frequency.
 

### 2025-04-09 manuscript/Manuscript.qmd
edit correlation analysis

@@
 subtitle: |
-  **Kei Jokura**$^{1,2,3,4*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewalow**$^{6}$, **Pawel Burkhardt**$^{5}$, **Gáspár Jékely**$^{1,2,6,*}$  
+  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewalow**$^{6}$, **Pawel Burkhardt**$^{7}$, **Gáspár Jékely**$^{1,2,6,*}$  
   <br><br>  
@@
   <br>  
-  $^{3}$ Exploratory Research Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan  
+  $^{3}$ Grass Laboratory, Marine Biological Laboratory, Woods Hole, MA 02543, USA  
   <br> 
-  $^{4}$ National Institute for Basic Biology (NIBB), Okazaki, 444-8585, Japan  
+  $^{4}$ Exploratory Research Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan  
   <br>  
-  $^{5}$ Michael Sars Centre, University of Bergen, Norway  
+  $^{5}$ National Institute for Basic Biology (NIBB), Okazaki, 444-8585, Japan  
   <br> 
-  $^{6}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  <br><br>  
+  $^{6}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  
+  <br>
+  $^{7}$ Michael Sars Centre, University of Bergen, Norway
+  <br><br>  
   **$^{*}$ Correspondence:**  
@@
 
-This work was funded by ...
-JSPS...
-This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement No 101020792).
+This work was supported by the Japan Society for the Promotion of Science (JSPS) Overseas Research Fellowships, the Grass Foundation, the Kavli Foundation, and the Marine Biological Laboratory (MBL).
+This project also received funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement No.
+101020792).
+We thank Dr.
+Réza Shahidi for sectioning and imaging, and Paulina Cherek for high-pressure freezing and sample preparation.
+We are also grateful to Dr.
+Chris Bjornsson, the MBL Central Microscopy Facility, Dr.
+Shoji A.
+Baba, and Dr.
+Kogiku Shiba (Shimoda Marine Research Center) for their assistance with microscopy.
 

### 2025-04-09 analysis/scripts/Rolling Pearson Correlation.R

@@ -9,10 +9,12 @@ library(zoo)
 # Paths
 balancer_info_path <- "analysis/data/balancer_CBF_Pearson_correlation_analysis/csv/balancer_info.csv"
 csv_folder <- "analysis/data/balancer_CBF_Pearson_correlation_analysis/csv"
+preproc_folder <- "analysis/data/balancer_CBF_Pearson_correlation_analysis/preprocessed"
 output_barplot <- "analysis/data/balancer_CBF_Pearson_correlation_analysis/output/CBF_barplot"
 output_corrplot <- "analysis/data/balancer_CBF_Pearson_correlation_analysis/output/pearson_rolling"
 output_boxplot <- "analysis/data/balancer_CBF_Pearson_correlation_analysis/output/pearson_boxplot"
 
+dir.create(preproc_folder, recursive = TRUE, showWarnings = FALSE)
 dir.create(output_barplot, recursive = TRUE, showWarnings = FALSE)
 dir.create(output_corrplot, recursive = TRUE, showWarnings = FALSE)
 dir.create(output_boxplot, recursive = TRUE, showWarnings = FALSE)
@@ -20,73 +22,79 @@ dir.create(output_boxplot, recursive = TRUE, showWarnings = FALSE)
 # Read info
 balancer_info <- read_csv(balancer_info_path)
 
-# Filter by abs_angle
+# Filter target
 filtered_files <- balancer_info %>%
   filter(abs_angle >= 0 & abs_angle < 20) %>%
   select(file_name, plane)
 
-# Barplot-----------------------------------------------------------------------
-plot_bar <- function(df, file_name, plane) {
-  df <- df %>%
-    filter(!is.na(left) & !is.na(right)) %>% # remove NA
-    filter(left != 0 & right != 0) %>% # remove 0
-    filter(is.finite(left) & is.finite(right)) %>% # remove Inf/NaN
-    mutate(frame = row_number())
+# STEP 1: Preprocess and save -------------------------------------------------------
+for (i in 1:nrow(filtered_files)) {
+  file_name <- filtered_files$file_name[i]
+  plane <- filtered_files$plane[i]
+  file_path <- file.path(csv_folder, file_name)
+  
+  if (file.exists(file_path)) {
+    df <- read_csv(file_path, show_col_types = FALSE) %>%
+      filter(!is.na(left) & !is.na(right)) %>%
+      filter(left != 0 & right != 0) %>%
[…]

### 2025-04-15 manuscript/Manuscript.qmd
MS: edit methods section

@@
 
-### Specimen Preparation, Transmission Electron Microscopy, and Image Processing
+### Specimen Preparation, Volume Electron Microscopy, and Image Processing
 
-Larvae of *Mnemiopsis leidyi* (5 days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) and immediately transferred to liquid nitrogen for storage.
+Larvae of *Mnemiopsis leidyi* (five days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) <!-- was this done in Exeter or Bergen?
+If Exeter, it is not the Baltec --> and immediately transferred to liquid nitrogen for storage.
 The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica).
-Cryo-substitution was performed by gradually raising the temperature, and the samples were embedded in epoxy resin.
-Serial sections of 50 nm thickness were prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º DiATOME diamond knife.
-To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the carbon film surface negatively charged.
+Cryo-substitution was carried out by gradually raising the temperature  <!-- steps?
+solutions?
+--> , and the samples were embedded in epoxy resin <!--type?
+-->.
+Serial untrithin sections of 50 nm thickness were prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º DiATOME diamond knife <!-- wasn't it our Leica UC7?
+-->.
+To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the carbon film <!-- was there carbon film?
+--> surface negatively charged.
 Section ribbons were collected on the prepared glass slides, slowly dried to ensure proper stretching, and firmly adhered to the glass surface.
@@
 The glass slides were mounted on STEM-specific stages (Zeiss) using Copper Foil EMI Shielding Tape (3M).
-Imaging was performed using a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5 imaging software (Zeiss).
+Imaging was done on a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5 imaging software (Zeiss).
 The full dataset consisted of 620 serial sections, each 50 nm thick.
-The imaging resolution was 2.8 nm/pixel, using electron holography transmission at an acceleration voltage of 1.5 kV (in-lens detector, dwell time: 3 µs).
+The imaging resolution was 2.8 nm/pixel, using electron holography transmission <!-- is this true?
+--> at an acceleration voltage of 1.5 kV (in-lens detector, dwell time: 3 µs  <!-- sounds very low -->).
 
-### Image-Stack Realignment and Spatial Data Transformation in CATMAID
+### Image-Stack Alignment and Export for CATMAID
 
@@
 
-The alignment process consisted of three stages, each progressively refining the spatial accuracy:
+The alignment process consisted of three stages, each progressively refining the spatial accuracy (rigid, affine, elastic).
 
-Rigid Alignment Initially, a rigid alignment was performed with the following parameters: least squares mode (linear feature correspondences), encompassing the entire layer range with the first layer as the reference.
+Initially, a rigid alignment was run with the following parameters: least squares mode (linear feature correspondences), encompassing the entire layer range with the first layer as the reference.
 Only visible images were used, without propagation.
@@
 Additional parameters included a feature descriptor size of 8, orientation bins of 8, and a closest ratio of 0.92.
-The alignment allowed clearing the cache, using 32 feature extraction threads, a maximal alignment error of 100 pixels, a minimal inlier ratio of 0.20, and a minimum of 12 inliers.
+The alignment allowed clearing the cache, using 32 feature-extraction threads, a maximal alignment error of 100 pixels, a minimal inlier ratio of 0.20, and a minimum of 12 inliers.
 The expected and desired transformations were set to rigid, with testing multiple hypotheses (tolerance: 5.00 pixels) and considering up to 5 neighboring layers, giving up after 5 failures.
-Regularization was performed with a maximal iteration of 1000, a maximal plateau width of 200, and a rigid lambda of 0.10.
+Regularization was done with a maximal iteration of 1000, a maximal plateau width of 200, and a rigid lambda of 0.10.
 
-Affine Alignment Following the rigid alignment, an affine alignment was applied using similar parameters, except the expected and desired transformations were set to affine.
-The minimal image size was reduced to 64 pixels, while the other parameters (e.g., Gaussian blur, feature descriptor size, inliers, and testing hypotheses) remained unchanged to ensure consistent processing.
+Next we applied an affine alignment step with similar parameters, except the expected and desired transformations were set to affine.
+The minimal image size was reduced to 64 pixels, while the other parameters (Gaussian blur, feature descriptor size, inliers, and testing hypotheses) remained unchanged to ensure consistent processing.
 
-Elastic Alignment Two iterations of elastic alignment were performed to fine-tune the spatial data.
-Key parameters included a block matching layer scale of 0.05, a search radius of 200 pixels, a block radius of 2000 pixels, and a resolution of 60.
+Finally, we ran two iterations of elastic alignment to fine-tune the spatial data.
+Key parameters included a block-matching layer scale of 0.05, a search radius of 200 pixels, a block radius of 2000 pixels, and a resolution of 60.
 Correlation filters were employed with a minimal PMCC r of 0.10, a maximal curvature ratio of 1000, and a maximal second-best r/best r of 0.90.
@@
 
-The reconstruction of cells and connectivity was performed using CATMAID [@Saalfeld2009], analyzing a total of 619 serial sections.
-To digitally reconstruct all neurons within the serial TEM dataset, we utilized the collaborative web-based application CATMAID, installed on a local server [@Saalfeld2009; @Schneider-Mizell2016].
+For skeletonisation, annotation and tagging we used CATMAID installed on a local server [@Saalfeld2009; @Schneider-Mizell2016].
 To mark the locations of cell bodies, we placed tags at the approximate center of each nucleus within the dataset.
 At each nuclear center, the radius of the single node was adjusted according to the size of the cell body in that specific layer.
-All skeletons were rooted at their respective cell bodies, and nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria.
+All skeletons were rooted at their respective cell bodies, and the root nodes were tagged as "soma." Synapses were identified based on four key structural features: the cell membrane, synaptic vesicles, the endoplasmic reticulum, and mitochondria.
 Most synapses could be verified across consecutive sections, ensuring accurate annotation and connectivity mapping.
@@
 
-Each cell is assigned a specific type name based on its category (balancer, bridge, bristle, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types, or those that could not be classified: presence and number of cilia (biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)) in 4 ciliated cell types.
-Additionally, a quadrant number is appended to the cell type name following an underscore ("\_") to indicate the cell body’s location.
-For example, cells located in the first quadrant are annotated with "Q1." If a cell is situated between the first and second quadrants, the annotation "Q1Q2" is applied.
+We assigned each cell a specific cell-type name based on its category (balancer, bridge, bristle, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types.
+Some cells only had more general features and were classified based on the presence/absence and number of cilia into four ciliated cell types:(biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)).
+Additionally, we appended the quadrant number to the cell type name after an underscore ("\_") to indicate the cell-body’s location.
+For example, cells located in the first quadrant are named with "_Q1." If a cell is situated between the first and second quadrants, we appended "Q1Q2" to the cell-type name.
 
-Cells of the same type within the same region are further distinguished by serial numbering.
-Each cell is also assigned multiple annotations, which can be utilized to query the database via CATMAID or the CATMAID API (e.g., using the R catmaid package).
-These annotations provide a structured and precise framework for identifying and analyzing specific cells, facilitating robust data integration and retrieval within the dataset.
+We further distinguished cells of the same type within the same region by serial numbering.
+This way each cell in the volume has a unique name string.
+Each cell was also assigned multiple annotations, which can be utilized to query the database via CATMAID or the CATMAID API (e.g., using the R catmaid package [@Bates2020]).
+These annotations provide a structured and precise framework for identifying and analyzing specific cells, facilitating robust data integration and retrieval from the dataset.
 
-This system ensures consistent and accurate tracking of cellular data, enabling comprehensive analysis of cellular structures and functions across regions.
+### Imaging the Activity of Balancer Cilia
 
-### Imaging of Balancer Cilia activity
-
-We conducted the analysis using the cydippid stage of *M.
+For ciliary imaging, we used the cydippid stage of *M.
 leidyi* at five days post-fertilization.
-A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass.
-Filtered natural seawater was introduced under the coverslip, along with the cy3dippids.
-By gently moving the coverslip, the desired orientation for observation was adjusted, and once positioned, slight pressure was applied to immobilize the cydippids.
-To stabilize the statolith's position, the microscope was tilted 90 degrees, arranging the stage vertically.
-The movement of the balancer cilia was observed using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS).
-Observations were made using a 40x objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27) with glycerine immersion.
+We gently placed a coverslip with a thin layer of Vaseline applied to its two edges on a slide glass.
+Filtered natural seawater was introduced under the coverslip, along with the cydippids.
+By gently moving the coverslip, we could adjust the orientation of the larva, and once positioned, we applied slight pressure to immobilize the cydippids.
+We imaged the movement of balancer cilia with a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS) and a 40x glycerine-immersion objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27).
+To stabilize the statolith's position, we tilted the microscope 90 degrees, arranging the stage vertically.
 For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals.
@@
 Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED.
-To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph project in Fiji software.
+To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph plugin in Fiji.
 
@@
 
-The following EM image stacks, including all traces and annotations, are available at https://catmaid.jekelylab.ex.ac.uk.
-These datasets can be queried either within CATMAID or through the CATMAID application programming interface (API) using the catmaid or natverse packages [@Bates2020].
+The EM image stacks, including all traces and annotations, are available at https://catmaid.jekelylab.ex.ac.uk.
 The dataset encompasses all EM images (in JPG format), skeletons, meshes, node tags, connectors, and annotations.

### 2025-04-22 manuscript/Manuscript.qmd
edit MS

@@
 
-<!--# the paper is not so much about evolution, so I suggest to start by introducing the cteno balancer organ -->
+Ctenophores, or comb jellies, are gelatinous marine animals that actively swim by beating rows of fused cilia known as comb plates.
+These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column.
+Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioral patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system.
 
-<!--# the things that need to be mentioned are -->
-
-Ctenophores are gelatinous ... ciliary comb rows ... active swimming and reorientation ...
-
-<!--# then introduce the nervous system of ctenos -->
-
-Ctenophores show sophisticated and dynamic behavioral patterns and respond to a wide range of external stimuli.
 These behaviors are thought to be regulated by multiple neural systems, including the subepithelial nerve net (SNN), mesogleal neurons, an aboral organ, and tentacular nerves.
@@
 Molecular phylogeny and chromosome-level synteny analyses suggest that ctenophores may be the sister group to all other animals [@Ryan2013; @Whelan2015a; @Whelan2017; @Whelan2023; @Li2021; @Schultz2023].
-In agreement with this early divergence from all other animals with a nervous system, the ctenophore nervous system shows several unique features that suggest independent evolution from other animals <!--# cite some Moroz papers -->.
+In agreement with this early divergence from all other animals with a nervous system, the ctenophore nervous system shows several unique features that suggest independent evolution from other animals
+
+<!--# cite some Moroz papers -->.
 

### 2025-04-23 manuscript/Manuscript.qmd
edit Introduction part

@@
 
-<!--# I rewrote a bit the abstract, please check -->
-
 Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth.
@@
 Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioral patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system.
+Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig; @Hernandez-Nicaise; @Jager].
+These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022].
+Notably, early single-cell transcriptomic studies failed to identify canonical neuronal clusters using conventional markers [@Sebe-Pedros2018].
+It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were subsequently identified [@Sachkova2021; @Hayakawa2022].
+In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@Moroz].
+This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems.
+However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system gives rise to behavior.
+Clarifying this relationship between neural architecture and behavioral control is essential to uncovering the logic of ctenophore neural function.
 
-These behaviors are thought to be regulated by multiple neural systems, including the subepithelial nerve net (SNN), mesogleal neurons, an aboral organ, and tentacular nerves.
-However, the mechanisms by which these neural systems control behavior remain largely unexplored.
-Molecular phylogeny and chromosome-level synteny analyses suggest that ctenophores may be the sister group to all other animals [@Ryan2013; @Whelan2015a; @Whelan2017; @Whelan2023; @Li2021; @Schultz2023].
-In agreement with this early divergence from all other animals with a nervous system, the ctenophore nervous system shows several unique features that suggest independent evolution from other animals
+To address the lack of structural understanding of how ctenophores control behavior, we focused on a unique sensory structure—the aboral organ.
+This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@Horiddge1965; @Tamm1980].
+Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez-Nicaise1968; @Jager2011].
+A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency.
+Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@Horridge1965; @Tamm1982; @Tamm2014].
+Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997].
+In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997].
+Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997].
+These findings support the hypothesis that neural modulation plays a critical role in shaping balancer cell output [@Lowe1997].
 
-<!--# cite some Moroz papers -->.
-
-... syncytial nerve net ... highly peptidergic
-
-The molecular makeup of ctenophore neurons is so divergent, that the first single-cell analyses did not identify a neuronal cluster [@Sebe-Pedros2018].
-Only through the analysis of proneuropeptide markers were neuronal clusters identified [@Sachkova2021; @Hayakawa2022].
-...
-organisation largely unknown ...
-synapse structure
-
-<!--# the phylogeny is not really relevant here, ctenos are interesting in their own right, so I would cut this a bit and we don't need to cite all the same genomic and tree papers, rater cite all the nice Tamm papers and EM papers etc -->
-
-Another unique feature of ctenophores is their aboral organ that contains the statocyst.
-Ultrastructural analyses have revealed that the aboral organ contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974; @Tamm1982].
-Immunostaining with an antibody labelling tyrosylated α-tubulin identified a neural structure called the deep nerve net within the aboral organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].
-
-<!--# then the makeup and function of the balancer -->
-
-...balancer in the aboral organ
-
-The aboral organ, also known as aboral organ, of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014].
-The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014; @Tamm1982].
-Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron].
-
-... balancer cilia, statolyth ... tuning of frequency ... pacemaker function ... reorientation
-
-In this study, we combine volume EM and high-speed imaging to examine the gravity-sensing system within the aboral organ of the ctenophore *Mnemiopsis leidyi*.
-Through connectome analysis, we uncovered the structure and synaptic organisation of the aboral nerve net.
-Our imaging experiments revealed correlated changes in the activity of balancer cilia, suggesting a coordinating activity for the aboral nerve net.
+To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy and high-speed imaging in the ctenophore *Mnemiopsis leidyi*.
+Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, and functional imaging revealed correlated activity patterns in balancer cilia.
+These results suggest that the aboral nerve net plays a central role in integrating sensory input and coordinating motor output.
 

### 2025-04-25 manuscript/Manuscript.qmd
edit fig1

@@
 For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced.
-Each node was tagged <!--# is this true?
-not only a subset?
--->, and skeletons were named and assigned multi-level annotations.
+All nodes relevant to synaptic sites were tagged, and skeletons were named and assigned multi-level annotations.
 As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023].
@@
 Abbreviations: ao, aboral organ; cg, ciliated groove; cr, comb plates; tb, tentacle bulbs.\
-(C) Enlarged views of the aboral organ from different perspectives in a cydippid larva using a DIC microscope.
-The left panel presents an aboral view of the aboral organ.
-The middle panel shows a lateral view of the aboral organ in the sagittal (S) plane, while the right panel displays a lateral view in the tentacular (T) plane.
-Abbreviations: bal, balancer; cg, ciliated grooves; do, dome cilia; li, lithocyte.
-Scale bar: 50 µm.\
-(D) Schematic representations of the aboral organ from three different perspectives.
-The left panel illustrates the aboral view, the middle panel shows the lateral view in the sagittal plane, and the right panel presents the lateral view in the tentacular plane.
-The quadrant colors correspond to those in panel (B).\
-(E) Example of cell tracing using the collaborative annotation toolkit CATMAID for large-scale electron microscopy image datasets.
+(C) Aboral views of the aboral organ highlighting its spatial organization.
+The left panel presents a schematic representation of the aboral organ, illustrating the four body quadrants (Q1–Q4), color-coded consistently with panel (B).
+The right panel shows a corresponding differential interference contrast (DIC) image from the same perspective.
+The region enclosed by the dotted line delineates the boundary of the aboral organ.
+The sagittal (S) and tentacular (T) arrows indicate the viewing directions corresponding to those planes.
+Scale bar: 50 µm.
+bal, balancer; cg, ciliated grooves; li, lithocyte.\
+(D) Lateral views of the aboral organ in two orthogonal planes captured using DIC microscopy.
+The left panel shows a lateral view in the sagittal (S) plane, while the right panel displays a view in the tentacular (T) plane.
+Balancer cilia (bal) and lithocytes (li) are enclosed within the dome (do).
+Scale bar: 50 µm.
+A–O indicates the aboral–oral body axis.\
+(E) Schematic representations of the aboral organ in sagittal and tentacular planes.
+The left panel shows a lateral schematic in the sagittal plane (S), and the right panel shows the same in the tentacular plane (T), corresponding to the views in panel (D).
+Colors match the quadrant scheme shown in panels (B) and (C).\
+(F) Example of cell tracing using the collaborative annotation toolkit CATMAID for large-scale electron microscopy image datasets.
 The spherical objects indicate nuclear positions, while the lines represent the traced cell centers.
 Scale bar: 10 µm.\
-(F) Three-dimensional reconstruction of cells composing the aboral organ, displayed from different perspectives.
+(G) Three-dimensional reconstruction of cells composing the aboral organ, displayed from different perspectives.
 The left panel presents an aboral view of the aboral organ.
@@
 
-Larvae of *Mnemiopsis leidyi* (five days old) were cryofixed using a high-pressure freezing apparatus (BAL-TEC HPM 010, Balzers) <!-- was this done in Exeter or Bergen?
-If Exeter, it is not the Baltec --> and immediately transferred to liquid nitrogen for storage.
+Larvae of *Mnemiopsis leidyi* (five days old) were cryofixed using a high-pressure freezing apparatus (HPM Live μ, CryoCapCell, https://www.cryocapcell.com/technologies/high-pressure-freezer) and immediately transferred to liquid nitrogen for storage.
 The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica).

### 2025-04-25 analysis/scripts/Figure1.R

@@ -102,7 +102,9 @@ close3d()
 panel_catmaid_overview <- ggdraw() + draw_image(readPNG("manuscript/pictures/overview.png")) +
   draw_label("serial EM volume", x = 0.5, y = 0.925, size = 11, fontface = "plain", hjust = 0.5) +
   draw_label("619 sections", x = 0.025, y = 0.81, color = "black", size = 9, hjust = 0) +
-  draw_label("927 cells", x = 0.025, y = 0.75, color = "black", size = 9, hjust = 0)
+  draw_label("927 cells", x = 0.025, y = 0.75, color = "black", size = 9, hjust = 0) +
+  draw_line(x = c(0.75, 0.9), y = c(0.05, 0.05), color = "black", size = 0.5) +
+  draw_label(expression(paste("10 ", mu, "m")), x = 0.825, y = 0.09, color = "black", size = 8, hjust = 0.5)
 
 #panel_3d_all_cells <- ggdraw() + draw_image(readPNG("manuscript/pictures/all_cells_3_views.png"))
 
@@ -138,7 +140,8 @@ panel_quadrants <- ggdraw() + draw_image(readPNG("manuscript/pictures/quadrants.
   draw_label("Q2", x = 0.06, y = 0.79, color = Okabe_Ito[2], size = 13, hjust = 0.5, alpha = 1) +
   draw_label("Q3", x = 0.03, y = 0.08, color = Okabe_Ito[6], size = 13, hjust = 0.5, alpha = 1) +
   draw_label("Q4", x = 0.275, y = 0.11, color = Okabe_Ito[7], size = 13, hjust = 0.5, alpha = 1) +
-  draw_label("aboral view", x = 0.17, y = 1, color = "black", size = 11, hjust = 0.5) +
+  draw_label("aboral view", x = 0.17, y = 1.1, color = "black", size = 12, hjust = 0.5) +
+  draw_label("lateral view", x = 0.66, y = 1.1, color = "black", size = 12, hjust = 0.5) +
   draw_label("sagittal plane", x = 0.5, y = 1, color = "black", size = 11, hjust = 0.5) +
   draw_label("tentacular plane", x = 0.83, y = 1, color = "black", size = 11, hjust = 0.5)  +
   draw_label(expression(paste("25 ", mu, "m")), x = 0.95, y = 0.1, color = "black", size = 10, hjust = 0.5) +
[…]

### 2025-04-28 manuscript/Manuscript.qmd
fig2 and supple fig S2

@@
 In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion.
-At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) <!--# figure..
--->.
+At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Supplementary Figure S2A).
 We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise_1973].
-Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells <!--# figure of the distribution -->.
+Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Supplementary Figure S2B).
 
-We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology.
-Each ANN had multiple nuclei <!--# add numbers e.g.
-(2-5 nuclei) -->, with anastomosing membranes <!--# show example EM pictures -->.
+We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology (Figure 2A and B).
+Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Supplementary Figure S2C).
 These neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023].
-Our results reveal a second type of syncytial neurons in ctenophores.
-<!--# describe the differences and also mention the classical SNNs that Sanja reconstructed -->
+Our results reveal a second type of syncytial neurons in ctenophores (Supplementary Figure S2D).
 
@@
 Based on their distinct size and position, we classified the ANNs into two types.
-The first type is a single large neuron (ANN_Q1-4) spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei This neuron has X <!--# add number --> presynaptic sites.
-The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and with two nuclei, and each containing X presynaptic structures.
-<!--#  -->
+The first type is a single large neuron (ANN_Q1-4) spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei.
+This neuron has 150 presynaptic sites.
+The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and containing two nuclei.
+ANN_Q1Q2 had 119 presynaptic structures, whereas ANN_Q3Q4 had 68 presynaptic structures.
 
@@
 ANN has a significantly higher number of mitochondria compared to other cell types, with most of them forming synapses.
+Except for lithocytes, each cell type has a sample size of at least three cells.
 The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
@@
 Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan.
-Blue dots indicate the locations of autapses within ANN Q1-4.\
-(G) Neural circuit map of ANN connectivity.
-The blue, orange, and pink circles represent individual neurons in ANN Q1-4, Q2Q3, and Q3Q4, respectively.
-Magenta arrows indicate synaptic connections from the center ANN to the lateral ANN, while cyan arrows indicate connections from the lateral ANN to the center ANN.
-Numbers denote the number of synapses.
-Blue arrows represent autapses in ANN Q1-4 along with their synapse counts.](figures/Figure2.png)
+Blue dots indicate the locations of autapses within ANN Q1-4.](figures/Figure2.png)
 

### 2025-04-28 analysis/scripts/Figure2.R

@@ -224,6 +224,90 @@ close3d()
 
 
 
+# calculate percentage of mitochondria which have vesicles or synapses associated with them -----
+
+mito_done <- read.neurons.catmaid("mitochondria done", pid = 35)
+
+get_pos_of_tags_in_neuron <- function(neur, tagname) {
+  # argument is neuron (data type), not skid
+  tag_positions <- tibble(
+    treenodeid = character(),
+    x = double(),
+    y = double(),
+    z = double()
+  )
+  tags <- neur$tags[[tagname]]
+  for (tag in tags) {
+    pos <- neur$d |>
+      filter(PointNo == tag) |>
+      select(X, Y, Z)
+    tag <- as.character(tag)
+    tag_positions <- tag_positions |> bind_rows(list(
+      treenodeid = tag,
+      x = pos$X, y = pos$Y, z = pos$Z
+    ))
+  }
+  return(tag_positions)
+}
+
+get_mito_pos <- function(neur) {
+  # accepts neuron as input
+  sskid <- neur$NeuronName
+  celltype <- catmaid_get_annotations_for_skeletons(sskid, pid = 35) |>
+    select(annotation) |>
+    filter(grepl("celltype:", annotation)) |>
+    pull()
+  celltype <- gsub(".*:", "", celltype)
+  if (length(celltype) == 0) {
+    celltype <- "NA"
+  }
+  
+  ves_none <- get_pos_of_tags_in_neuron(neur, "mitochondrion no vesicles")
+  ves_none <- bind_cols(mito_type = "vesicles_none", ves_none)
+  
+  if (nrow(ves_none) == 0) {
+    ves_none <- get_pos_of_tags_in_neuron(neur, "mitochondrion")
+    ves_none <- bind_cols(mito_type = "vesicles_none", ves_none)
+  }
+  
+  ves_unc <- get_pos_of_tags_in_neuron(neur, "mitochondrion unclear vesicles")
+  ves_unc <- bind_cols(mito_type = "vesicles_unc", ves_unc)
+  
+  ves_yes <- get_pos_of_tags_in_neuron(neur, "mitochondrion vesicles")
+  ves_yes <- bind_cols(mito_type = "vesicles", ves_yes)
+  # treenodes with outgoing synapses
+  connectors <- neur$connectors
+  if (!is.null(connectors)) {
+    syn_out_treenodes <- connectors |>
+      filter(prepost == 0) |>
+      select(treenode_id) |>
+      pull()
+  } else {
[…]

### 2025-04-28 analysis/scripts/Supplemental_figure.R

@@ -173,12 +173,61 @@ ggsave("manuscript/pictures/3d_plot/all_assemble.png", limitsize = FALSE,
 
 
 # load Subepithelial nerve net -------------------------------------------------
+
 SNN <- read_smooth_neuron("celltype:SNN")
 
 # 3d plot Subepithelial nerve net -----------------------------------------------
 
-plot_multinucleated_views("SNN", SNN)
+close3d()
+# 3d plotting of cells
+nopen3d() 
+mfrow3d(1, 3)  
+#define the size of the rgl window, the view and zoom
+par3d(windowRect = c(0, 0, 1200, 350))
 
+#plot aboral view
+plot_multinucleated_cell(SNN, lwd = 1.25, alpha = 0.8, col = Okabe_Ito[5])
+plot3d(outline, add = TRUE, alpha = 0.025, col = Okabe_Ito[8])
+
+#aboral view
+aboral()
+par3d(zoom = 0.7)
+
+#move to next panel in rgl window
+next3d(clear=F)
+
+
+
+#plot lateral view of Sagittal plane
+plot_multinucleated_cell(SNN, lwd = 1.25, alpha = 0.8, col = Okabe_Ito[5])
+plot3d(outline, add = TRUE, alpha = 0.025, col = Okabe_Ito[8])
+
+#lateral view of Sagittal plane
+sagittal()
+par3d(zoom = 0.7)
+
+#move to next panel in rgl window
+next3d(clear=F)
+
+
+
+#plot lateral view of Tentacular plane
+plot_multinucleated_cell(SNN, lwd = 1.25, alpha = 0.8, col = Okabe_Ito[5])
+plot3d(outline, add = TRUE, alpha = 0.025, col = Okabe_Ito[8])
+
+#lateral view of Tentacular plane
+tentacular()
+par3d(zoom = 0.7)
+
+#move to next panel in rgl window
+next3d(clear=F)
+
+
+#make a snapshot to the working directory
+rgl.snapshot("manuscript/pictures/3d_plot/plot_SNN.png")
+
+
+close3d()
 
 
 
@@ -269,13 +318,33 @@ panel_all_assemble <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot
 
 panel_all_cells <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/all_cells_3_views_alt.png"))
 
[…]

### 2025-04-28 analysis/scripts/number_of_postsynaptic_sites_per_mitochondria.R

@@ -1,6 +1,6 @@
 source("analysis/scripts/packages_and_functions.R")
 
-# calculate average number of post-synaptic sites per synapse  -----
+# calculate average number of post-synaptic sites per synapse  -----------------
 
 pre_connectors <- catmaid_fetch(
   path = "35/connectors/",
@@ -117,7 +117,7 @@ mito_vesicle_info <- lapply(mito_done, get_mito_pos) |>
 write.csv(mito_vesicle_info, "analysis/data/mito_vesicle_info.csv")
 mito_vesicle_info <- read.csv("analysis/data/mito_vesicle_info.csv")
 
-# create tibble for graph ---------------------------------
+# create tibble for graph ------------------------------------------------------
 
 mito_stats <- mito_vesicle_info %>%
   group_by(skid, mito_type) %>%

### 2025-04-29 manuscript/Manuscript.qmd
edit fig3

@@
 ![**Figure 3. Reconstruction of the balancers and bridges that receives synapses from the ANN.**\
-(A) Graph showing the number of synaptic outputs from the ANN circuit for each cell type.
-The balancer and bridge cell types receive relatively high outputs from the ANN circuit.
-ANN represents the number of synapses within the ANN circuit.
+(A) Graph showing the number of synaptic outputs from the ANNs (ANN Q1-4: blue, ANN Q1Q2: orange, ANN Q3Q4: pink) circuit for each cell type.
+The balancer and bridge cell types receive relatively high outputs from the ANNs circuit.
+ANN represents the number of synapses within the ANNs circuit.
 The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\

### 2025-04-29 analysis/scripts/Figure2.R

@@ -900,9 +900,14 @@ panel_SSN_Q12_Q34 <- ggdraw() + draw_image(readPNG("manuscript/pictures/SSN_Q12_
 
 panel_mito_bar_graph <- ggdraw() + draw_image(readPNG("manuscript/pictures/plot_mito_stats.png"))
 
+circle_g <- ggdraw() + draw_image(readPNG("manuscript/pictures/synapse_green.png"))
+circle_gy <- ggdraw() + draw_image(readPNG("manuscript/pictures/synapse_gray.png"))
+
 panel_mito_pos <- ggdraw() + draw_image(readPNG("manuscript/pictures/mito_pos_SSN.png")) +
-  draw_label("mitochondria with synapses", x = 0.05, y = 0.98, color = "green4", size = 10.5, hjust = 0) +
-  draw_label("not forming synapses", x = 0.05, y = 0.9, color = "black", size = 10.5, hjust = 0, alpha = 0.8) +
+  draw_plot(circle_g, x = 0.06, y = 0.955, width = 0.05, height = 0.05) +
+  draw_label("Mitochondria with synapses", x = 0.1, y = 0.98, color = "green4", size = 10.5, hjust = 0) +
+  draw_plot(circle_gy, x = 0.06, y = 0.875, width = 0.05, height = 0.05) +
+  draw_label("Not forming synapses", x = 0.1, y = 0.9, color = "black", size = 10.5, hjust = 0, alpha = 0.7) +
   draw_label("ANN Q1-4", x = 0.85, y = 1, color = Okabe_Ito[5], size = 8, hjust = 0) +
   draw_label("ANN Q1Q2", x = 0.85, y = 0.94, color = Okabe_Ito[6], size = 8, hjust = 0) +
   draw_label("ANN Q3Q4", x = 0.85, y = 0.88, color = Okabe_Ito[7], size = 8, hjust = 0)
@@ -914,25 +919,25 @@ circle_lb <- ggdraw() + draw_image(readPNG("manuscript/pictures/synapse_lightblu
 circle_db <- ggdraw() + draw_image(readPNG("manuscript/pictures/synapse_darkblue.png"))
 
 panel_SSN_prepost_synapse <- ggdraw() + draw_image(readPNG("manuscript/pictures/SSN_prepost_synapse.png")) +
[…]

### 2025-04-29 analysis/scripts/Figure3.R

@@ -19,12 +19,12 @@ balancer_Q4 <- read_smooth_neuron(get_skids_with_annot(pid = 35, c("celltype:bal
 bridge_Q1Q2 <- read_smooth_neuron(get_skids_with_annot(pid = 35, c("celltype:bridge", "Q1Q2")))
 bridge_Q3Q4 <- read_smooth_neuron(get_skids_with_annot(pid = 35, c("celltype:bridge", "Q3Q4")))
 
+
+# bar graph of outputs from SNNs -----------------------------------------------
+
 SSN_Q1Q2 <- read_smooth_neuron("SSN_Q1Q2")[[1]]
 SSN_Q3Q4 <- read_smooth_neuron("SSN_Q3Q4")[[1]]
 SSN_Q1Q2Q3Q4 <- read_smooth_neuron("SSN_Q1Q2Q3Q4")[[1]]
-
-
-# bar graph of outputs from SNNs -----------------------------------------------
 
 SSN_Q1Q2_skid <- SSN_Q1Q2$skid
 SSN_Q3Q4_skid <- SSN_Q3Q4$skid
@@ -111,36 +111,83 @@ label_mapping <- c(
 )
 
 
+# pre-synapse (take output-side connector)
+syn_out_SSN_Q1Q2 <- stats_synapse %>%
+  filter(prepost == 0, skid == SSN_Q1Q2_skid) %>%
+  select(connector_id) %>%
+  mutate(SSN_source = "SSN_Q1Q2")
+
+syn_out_SSN_Q3Q4 <- stats_synapse %>%
+  filter(prepost == 0, skid == SSN_Q3Q4_skid) %>%
+  select(connector_id) %>%
+  mutate(SSN_source = "SSN_Q3Q4")
+
+syn_out_SSN_Q1Q2Q3Q4 <- stats_synapse %>%
+  filter(prepost == 0, skid == SSN_Q1Q2Q3Q4_skid) %>%
+  select(connector_id) %>%
+  mutate(SSN_source = "SSN_Q1Q2Q3Q4")
+
+# Putting together all the connector_id and SSN_source
+connector_source_table <- bind_rows(
+  syn_out_SSN_Q1Q2,
+  syn_out_SSN_Q3Q4,
+  syn_out_SSN_Q1Q2Q3Q4
+)
+
+# join to connector_id when taking postsynapse
+SSN_downstream <- stats_synapse %>%
+  filter(prepost == 1) %>%
[…]

### 2025-04-29 analysis/scripts/Figure4.R
fig4

@@ -4,22 +4,23 @@
 source("analysis/scripts/packages_and_functions.R")
 
 
-# read data---------------------------------------------------------------------
+# read arrest re-beat data------------------------------------------------------
 
 df_time_dif <- read_csv("analysis/data/arrest_rebeat_time_differences.csv")
-
 
 # plot arrest re-beat graph ----------------------------------------------------
 
 plot_arrest_rebeat <- 
   ggplot(df_time_dif) +
-  aes(x = response, y = time, fill = response, color = response) +  
-  geom_boxplot(aes(color = plane), size = 0.5, outlier.shape = NA) +
-  geom_beeswarm(aes(color = time),
-                size = 2,
+  aes(x = response, y = time, fill = response) +  
+  geom_boxplot(aes(color = plane), 
+               size = 0.5, 
+               outlier.shape = NA) +
+  geom_beeswarm(aes(color = plane),    
+                size = 1.75,
                 cex = 3.2,
-                alpha = 0.3,
-                color = "gray5") +
+                alpha = 0.5,
+                show.legend = FALSE) +
   labs(y = "Time difference (sec)") +
   xlab("") +
   theme_minimal() +
@@ -34,12 +35,13 @@ plot_arrest_rebeat <-
     axis.ticks.x = element_blank()  
   ) +
   scale_fill_manual(
-    values = c("re-beat" = "gray", 
+    values = c("re-beat" = "gray70", 
                "arrest" = "white")) +
   scale_color_manual(
-    values = c("sagittal" = "#0097b7", "tentacular" = "#f39500")) +
-  facet_grid(~ plane, scales = "free_x", space = "free_x", switch = "x")  
-
+    values = c("sagittal" = "#0097b7", 
+               "tentacular" = "#f39500"),
+    guide = "none") +
+  facet_grid(~ plane, scales = "free_x", space = "free_x", switch = "x")
 
 plot_arrest_rebeat
 
@@ -53,48 +55,167 @@ ggsave(
   bg = "white"
 )
 
+# CBF barplot sagittal vs tentacular -------------------------------------------
+
+# Lists files to be processed and Plane information
[…]

### 2025-04-29 analysis/scripts/Rolling Pearson Correlation.R

@@ -186,3 +186,7 @@ for (window in window_sizes) {
   
   plot_boxplot(results, window)
 }
+
+# Use if you want to save the result as a csv file just in case ----------------
+
+write_csv(results, "analysis/data/balancer_CBF_Pearson_correlation_analysis/output/csv/correlation_results.csv")

### 2025-04-29 manuscript/Manuscript.qmd
Fig5

@@
 
-![**Figure 4. Differences in Balancer Cilia Movement Control Due to Variations in the Neural Circuit Pathways and Comparison of Neural Circuits Regulating Ciliary Movement**\
+![**Figure 4. Differences in Balancer Cilia Movement Control Due to Variations in the Neural Circuit Pathways**\
 (A) Schematic diagram of differential interference contrast (DIC) microscopy setup for imaging balancer cilia movement.
@@
 The direction and length of the arrows indicate time.\
-(C) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia. S and T represent the sagittal plane and tentacular plane, respectively.\
-(D) Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation.
-Ciliated cells are shown in gray.
-Synapses are indicated by arrows, with magenta representing synapses that induce ciliary arrest and blue representing synapses that induce ciliary re-beat or an increase in ciliary beating frequency.\
-(Left) Gravity-sensing neural circuit of *M.
-leidyi*, as revealed in this study.
-Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying their potential involvement in feedback mechanisms between neurons and ciliated cells.\
-(Righr) Neural circuit regulating prototroch ciliary movement in *P.
-dumerilii*.
-Serotonergic neurons (Ser-h1 neurons, blue) activate ciliary movement, while cholinergic neurons (MC neurons, magenta) induce ciliary arrest.
-Arrows directed toward ciliated cells in the head prototroch are color-coded accordingly.](figures/Figure4.png)
+(C) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia.
+S and T represent the sagittal plane and tentacular plane, respectively.](figures/Figure4.png)
 
@@
 The nerve net could in turn tune or reset sensitivites and valence of balancer cells, in a coordinatory role, as we will discuss below.
+
+![**Figure 5. Comparison of Neural Circuits Regulating Ciliary Movement**\
+(A) Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation.
+Ciliated cells are shown in gray.
+Synapses are indicated by arrows, with magenta representing synapses that induce ciliary arrest and blue representing synapses that induce ciliary re-beat or an increase in ciliary beating frequency.\
+(Left) Gravity-sensing neural circuit of *M.
+leidyi*, as revealed in this study.
+Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying their potential involvement in feedback mechanisms between neurons and ciliated cells.\
+(Righr) Neural circuit regulating prototroch ciliary movement in *P.
+dumerilii*.
+Serotonergic neurons (Ser-h1 neurons, blue) activate ciliary movement, while cholinergic neurons (MC neurons, magenta) induce ciliary arrest.
+Arrows directed toward ciliated cells in the head prototroch are color-coded accordingly.](figures/Figure5.png)
 

### 2025-04-29 analysis/scripts/Figure5.R

@@ -1,0 +1,29 @@
+# Code to generate Figure 5 of the Jokura et al 2024 Ctenophore apical organ connectome paper
+
+# source packages and functions ------------------------------------------------
+source("analysis/scripts/packages_and_functions.R")
+
+# assemble figure -------------------------------------------------------------
+
+panel_comparison <- ggdraw() + draw_image(readPNG("manuscript/pictures/map_comparison.png"))
+
+layout <- "
+A
+"
+
+Figure5 <- panel_comparison +
+  plot_layout(design = layout,
+              heights = c(1),
+              widths = c(1)) + 
+  patchwork::plot_annotation(tag_levels = "A") &  
+  ggplot2::theme(plot.tag = element_text(size = 12, 
+                                         face='plain', color='black'))
+
+ggsave("manuscript/figures/Figure5.png", limitsize = FALSE, 
+       units = c("px"), Figure5, width = 2100, height = 1100, bg='white')  
+
+
+ggsave("manuscript/figures/Figure5.pdf", limitsize = FALSE, 
+       units = c("px"), Figure5, width = 2100, height = 1100) 
+
+

### 2025-04-30 manuscript/Manuscript.qmd
Fig4

@@
 We found that arrest and re-beat events were synchronized between balancers across the sagittal plane.
-However, while the timing of re-beat was also near-simultaneous along the tentacular plane, arrest timing was offset by up to 2.xy <!--# add max value--> seconds (Figure 4) <!--# save source data for all plots in manuscript/source_data with the naming convention e.g.
-Figure 1A--source data 1 -->.
+However, while the timing of re-beat was also near-simultaneous along the tentacular plane, arrest timing was offset by up to 2.16 seconds (Figure 4D).
 Overall, these data reveal that arrests are only coordinated between Q1-Q2 and Q3-Q4, whereas re-beat is coordinated over all four quadrants.
 
-<!--# discuss frequency data -->
+As shown in Figure 4E, beat frequencies of left and right balancers fluctuated substantially in both orientations.
+However, while left-right activity was largely synchronized in sagittal pairs (Figure 4E, left), tentacular pairs (Figure 4E, right) showed strikingly periodic alternations between left and right balancer activity.
+This pattern suggested an offset in timing during arrest or re-beat events between tentacular balancers.
+
+To quantify this, we calculated a rolling Pearson correlation between left and right balancer activity in each plane.
+The resulting mean correlations were significantly higher in the tentacular plane compared to the sagittal plane (Figure 4F; t-test, p = 0.014).
+This result indicates that tentacular balancers beat in a more coordinated, mirror-symmetric manner, whereas sagittal balancers exhibit greater variability between sides.
 
@@
 At the same time, the ANN_Q1-4 neuron synapses on all four balancers, hinting at a neural substrate for their synchronised re-beat.
+
+Notably, bridge neurons associated with each quadrant region (Q1Q2 or Q3Q4) extend their processes across the midline, forming connections that span between opposite quadrant domains when viewed along the tentacular plane.
+This morphology suggests a potential role for bridge neurons in modulating inter-quadrant coordination.
+By linking ANN activity across orthogonal planes, these neurons may contribute to the dynamic regulation of CBF patterns across the whole system.
 
@@
 The movement of the balancer cilia was recorded using a 40× objective lens.\
-(B) Kymograph patterns comparing the arrest and re-beat of balancer cilia movement.
-The left and right balancer cilia movements are compared in both the sagittal plane (left) and tentacular plane (right).
-The direction and length of the arrows indicate time.\
-(C) Graph showing the time differences in the arrest and re-beat timing of the left and right balancer cilia.
-S and T represent the sagittal plane and tentacular plane, respectively.](figures/Figure4.png)
+(B) Representative DIC images of the aboral organ viewed along the sagittal (left) and tentacular (right) planes.
+Arrowheads indicate the balancer cilia selected for kymograph-based analysis in each orientation.\
+(C) Representative kymographs showing arrest and re-beat events of balancer cilia.
+Left and right balancer cilia were simultaneously recorded in either the sagittal (left) or tentacular plane (right).
+Arrowheads mark the onset of arrest (open) and re-beat (filled); horizontal arrows represent 1-second intervals.\
+(D) Boxplots showing time differences in the onset of arrest (white boxes) and re-beat (gray boxes) between left and right balancer cilia.
+Colors of the box outlines indicate the imaging plane: sagittal (blue) or tentacular (orange).
+Each dot represents a single larva.\
+(E) Bar plots showing ciliary beat frequency (CBF) dynamics of left and right balancers (top and bottom, respectively) recorded from sagittal (left) and tentacular (right) planes.
+Positive and negative values represent opposite sides within each plane.\
+(F) Mean rolling Pearson correlation of left–right balancer CBF calculated with a 20-frame window.
+Each dot represents an individual larva.
+Balancers on the tentacular plane showed significantly higher correlation than those on the sagittal plane (t-test, p = 0.014).](figures/Figure4.png)
 

### 2025-04-30 analysis/scripts/Figure4.R

@@ -118,6 +118,15 @@ for (info in files_info) {
   )
 }
 
+panel_cbf_bar_sag <- ggdraw() + draw_image(readPNG("manuscript/pictures/CBF_barplot_23_08_04_WT1_030_Rec_at_100.73fps_dark_f.png"))
+panel_cbf_bar_tent <- ggdraw() + draw_image(readPNG("manuscript/pictures/CBF_barplot_23_08_01_WT1_007_Rec_at_100.20fps_dark.png"))
+
+panel_cbf_bars_combined <- panel_cbf_bar_sag + panel_cbf_bar_tent +
+  plot_layout(ncol = 2, widths = c(1, 1))
+
+ggsave("manuscript/pictures/panel_cbf_bars_combined.png",
+       plot = panel_cbf_bars_combined,
+       width = 8, height = 3, dpi = 300, bg = "white")
 
 # boxplot of pearson comparison S vs T (windows = 20)---------------------------
 
@@ -185,25 +194,23 @@ panel_balancer <- ggdraw() + draw_image(readPNG("manuscript/pictures/balancer_cl
 panel_kymograph <- ggdraw() + draw_image(readPNG("manuscript/pictures/balacer_arrest_rebeat_kymograph.png"))
 panel_graph <- ggdraw() + draw_image(readPNG("manuscript/pictures/arrest_rebeat_graph.png"))
 
-panel_cbf_bar_sag <- ggdraw() + draw_image(readPNG("manuscript/pictures/CBF_barplot_23_08_04_WT1_030_Rec_at_100.73fps_dark_f.png")) +
-  draw_label("sagittal plane", x = 0.5, y = 0.95, size = 10, hjust = 0.5)
-panel_cbf_bar_tent <- ggdraw() + draw_image(readPNG("manuscript/pictures/CBF_barplot_23_08_01_WT1_007_Rec_at_100.20fps_dark.png")) +
-  draw_label("tentacular plane", x = 0.5, y = 0.95, size = 10, hjust = 0.5)
-panel_cor_graph <- ggdraw() + draw_image(readPNG("manuscript/pictures/boxplot_correlation_angle_0-20_win_20.png"))
+panel_cbf_bars <- ggdraw() + draw_image(readPNG("manuscript/pictures/panel_cbf_bars_combined.png")) +
+  draw_label("sagittal plane", x = 0.25, y = 0.975, size = 10, hjust = 0.5) +
[…]

### 2025-05-01 manuscript/Manuscript.qmd
MS M&M

@@
 The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica).
-Cryo-substitution was carried out by gradually raising the temperature  <!-- steps?
-solutions?
---> , and the samples were embedded in epoxy resin <!--type?
--->.
-Serial untrithin sections of 50 nm thickness were prepared using a Reichert Jung Ultracut E ultramicrotome and a 45º DiATOME diamond knife <!-- wasn't it our Leica UC7?
--->.
-To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the carbon film <!-- was there carbon film?
---> surface negatively charged.
-Section ribbons were collected on the prepared glass slides, slowly dried to ensure proper stretching, and firmly adhered to the glass surface.
-The sections on glass were stained with UranyLess and lead citrate (Reynolds) using airless staining bottles (Delta Microscopies).
-The glass slides were mounted on STEM-specific stages (Zeiss) using Copper Foil EMI Shielding Tape (3M).
-Imaging was done on a Gemini SEM 500 (Zeiss) equipped with SmartSEM and Atlas 5 imaging software (Zeiss).
-The full dataset consisted of 620 serial sections, each 50 nm thick.
-The imaging resolution was 2.8 nm/pixel, using electron holography transmission <!-- is this true?
---> at an acceleration voltage of 1.5 kV (in-lens detector, dwell time: 3 µs  <!-- sounds very low -->).
+Cryo-substitution was carried out by gradually raising the temperature from -90°C to -70°C over 4 hours, then returning to -20°C over 2 hours, and finally to room temperature over 2 hours.
+The samples were then embedded in epoxy resin (EMbed 812, Electron Microscopy Sciences).
+Serial ultrathin sections of 50 nm thickness were prepared using a Leica UC7 ultramicrotome and a 45º DiATOME diamond knife.
+To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the surface negatively charged.
+Section ribbons were collected on prepared glass slides, gently dried to ensure proper stretching, and firmly adhered to the glass surface.
+The sections were stained with UranyLess and lead citrate (Reynolds) using airless staining procedures.
+The glass slides were mounted on a STEM-specific stage (Zeiss) using Copper Foil EMI Shielding Tape (3M).
+Imaging was performed using a Gemini SEM 500 (Zeiss), equipped with SmartSEM and Atlas 5 imaging software (Zeiss), using the Inlens detector.
+A total of 620 serial sections, each 50 nm thick, were analyzed.
+The imaging resolution was 2.8 nm/pixel, and the acceleration voltage was set to 1.5 kV.
+The imaging time for each tile was 54 minutes and 21 seconds, with a pixel size of 2.8 nm, a tile size of 91.8 × 91.8 µm (32768 × 32768 pixels), and a dwell time of 3.0 µs.
 
@@
 leidyi* at five days post-fertilization.
-We gently placed a coverslip with a thin layer of Vaseline applied to its two edges on a slide glass.
-Filtered natural seawater was introduced under the coverslip, along with the cydippids.
-By gently moving the coverslip, we could adjust the orientation of the larva, and once positioned, we applied slight pressure to immobilize the cydippids.
-We imaged the movement of balancer cilia with a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS) and a 40x glycerine-immersion objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27).
-To stabilize the statolith's position, we tilted the microscope 90 degrees, arranging the stage vertically.
-For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals.
-The camera and LED strobe system were operated using the Video Capture Software BohNavi.
-Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED.
-To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph plugin in Fiji.
+A coverslip with a thin layer of Vaseline applied to its two edges was gently placed over a glass slide.
+Filtered natural seawater and the cydippids were introduced beneath the coverslip.
+The orientation of the larvae was adjusted by carefully sliding the coverslip, and slight pressure was then applied to immobilize the larvae.
+
+Balancer cilia movement was imaged using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a high-sensitivity monochrome CMOS camera optimized for the near-infrared range (UI-3360CP-NIR-GL Rev.2, iDS), and a 40x glycerine-immersion objective lens (LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27).
+To stabilize the statolith, the microscope was tilted by 90° to achieve vertical stage alignment.
+Illumination was provided by a custom-made NIR LED strobe system (850 nm) synchronized with the camera’s exposure signals.
+Image acquisition was controlled using BohNavi software.
+Recordings were acquired at 640 × 480 pixels, 100 frames per second (fps), for a duration of 2 minutes, using 0.05 ms pulse illumination from the LED.
+
+For beat frequency analysis, kymographs were generated using the Multi Kymograph plugin in Fiji.
+One-second segments were manually extracted from the kymographs, and ciliary beat frequencies were counted visually for each ROI.
+These manually determined frequencies were then used for subsequent computational analysis.
+
+To process the beat frequency time series data, left and right cilia traces were extracted, filtered to remove missing or non-finite values, and assigned frame indices.
+Data were saved as RDS objects for downstream analysis.
+Mean values were plotted as bar graphs with the left and right cilia shown in positive and negative directions, respectively.
+
+To assess coordination between left and right cilia, we computed rolling Pearson correlation coefficients over a window of 20 frames.
+For each dataset, we excluded windows with insufficient valid data or no variation.
+Time series of correlation coefficients were plotted and summarized as mean values.
+These mean correlations were then compared across planes (sagittal vs.
+tentacular) using Welch's t-test.
+The resulting statistics were visualized as boxplots with jittered data points.
 

### 2025-05-06 manuscript/Manuscript.qmd
Discussion

@@
 
-In this study, we generated a comprehensive neural map of the aboral organ of a five-day-old *Mnemiopsis leidyi*.
-We described a novel type of syncytial neuron and analysed its synaptic connections to balancer-cell groups across the four quadrants of the aboral organ.
-Given that all neurons are intrinsic to the aboral organ and our EM volume covers the entire organ, we could recover a complete circuit map, the first for a ctenophore.
-Contrary to our expectations, the nerve net has no motor connections to effector organs, such as the ciliary grooves or comb rows.
-Our high-speed imaging suggests that the nerve net instead coordinates the activity of balancer cilia.
-This agrees with a hydrodynamic model of graviorientation <!-- add ref to Tamm -->, with balancer cilia directly responding to statolyth load and acting as pacemakers for the comb rows via the ciliated grooves.
-The nerve net could in turn tune or reset sensitivites and valence of balancer cells, in a coordinatory role, as we will discuss below.
+### Functional Nerve Net for Ciliary Coordination
+
+In this study, we generated the first high-resolution connectome of a ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of five-day-old *Mnemiopsis leidyi* cydippids using volumetric electron microscopy.
+Among these, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve nets (ANNs)—which exhibit multiple nuclei and a morphology distinct from the “beads-on-a-string” organization reported for the subepithelial nerve net (SNN) in one-day-old cydippids.
+Instead of forming discrete nodal swellings, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion, suggesting a unique neuron subtype specific to the aboral organ.
+
+Remarkably, ANNs form no synaptic output to classical effector organs such as muscles or comb plates.
+All observed output synapses were directed to balancer ciliary cells.
+Moreover, the projection patterns of individual ANNs were asymmetric, selectively targeting different regions of the balancer system.
+These findings indicate that ANNs are unlikely to serve as motor drivers in the conventional sense.
+Rather, they may be involved in coordinating or modulating the timing and dynamics of ciliary activity.
+
+Notably, anatomical differences in ANN projection patterns across sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics.
+This correlation supports the idea that individual ANNs may play opposing or complementary roles in regulating ciliary beating—potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns.
+Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs.
+inhibitory) of these neurons remains unresolved.
+However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.
+
+Strikingly, this organization resembles circuit motifs found in other animals.
+For instance, in Platynereis dumerilii larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands, while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields.
+Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits.
+
+Taken together, our findings suggest that the ANN network represents a non-motor, coordination-based neural system—one whose primary function is to tune and synchronize locomotor effectors rather than activate them directly.
+This departs fundamentally from reflex-arc models of early nervous systems and instead supports the idea that coordination, not command, may have been a foundational role of ancestral neurons.
 
@@
 
-### A New Type of Syncytial Neuron
+### Bridge Cells as Feedback Regulators of Ciliary Rhythms
 
-We identified three syncytial neurons with multiple nuclei in the aboral organ of the cydippid larva.
-Burkhurdt et al.
-previously reported syncytial neurons in the sub-epithelial nerve net (SNN) of a one-day-old larva [@Burkhardt2023].
-The syncytial neurons in the aboral organ did not exhibit the characteristic "pearl necklace-like structure" of the SNN neurons.
-Instead, their morphology was more reminiscent of fiber cells found in Placozoa [@Mayorova2021], with cell projections extending through intercellular spaces.
-In our volume, we also identified two SNN-like neurons with a pearl necklace-like morphology, indicating that the differences between ANN and SNN are not due to differences in stage or sample preparation.
-This confirms the ANN neurons as a second distinct type of syncytial neuron in ctenophores.
+In our volumetric reconstruction, bridge cells were identified as the only cell type providing synaptic input to the ANN network, positioning them as key upstream regulators of the ciliary coordination system in the aboral organ.
+Originally described morphologically by Tamm et al.
+(2002), bridge cells extend thin projections toward the base of the balancer cilia.
+Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes.
+These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed toward the ANN circuit.
 
-### Functional Diversification of the Aboral Syncytial Neurons
+As revealed by our high-speed imaging, ciliary beat frequency (CBF) exhibited rhythmic changes over time that differed between anatomical planes, indicating spatially distinct coordination dynamics.
+These CBF rhythms are likely generated by the ANN circuit, while bridge cells may function to stabilize or adjust this rhythmic output in accordance with the real-time state of the cilia.
+In this way, bridge cells could provide feedback about the execution target—i.e., the balancer cilia—back into the coordination circuit, serving as a core element of a self-regulating rhythm control system.
 
-The three ANN neurons form two functional types based on their spatial arrangement and differential innervation of balancer cells.
-
-In the this study, we found that three neurons are closely positioned within the confined space of the aboral organ, potentially enabling a single neuron to control distant ciliary movements in a synchronized manner.
-Moreover, the presence of neurons projecting to the same ciliated cells in distinct patterns suggests that these neurons may serve opposing functions—one inhibiting ciliary activity while the other promotes it, or one modulating ciliary beat frequency.
-While our synaptic ultrastructure analysis did not allow us to distinguish excitatory from inhibitory neurons, future electrophysiological investigations will be essential to elucidate their functional properties.
-
-In metazoans, a single neuron can directly provide synaptic input to multiple ciliated cells at a distance, synchronizing their movements.
-For instance, in larvae of the annelid *Platynereis dumerilii*, a cholinergic motor neuron (MC) innervates multiple prototroch ciliated cells, synchronizing their motion such that activation of the neuron leads to simultaneous arrest of ciliary beating [@Veraszto2017].
-Additionally, a pair of serotonergic neurons, designated Ser-h1, extend axons that cross at the midline; the left Ser-h1_l predominantly innervates right-sided ciliated cells, while the right Ser-h1_r mainly projects to the left-side ciliated cells.
-Activation of Ser-h1 induces ciliary beating initiation and increases beat frequency.
-
-### Paracrine Signaling and Synaptic Transmission from Sensory to Neural Cells
-
-Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores [@HORRIDGE1965; @vinnikov1974e; @Krisch1973; Aronova1974].
-By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the aboral organ.
-However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets.
-In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the aboral organ [@Sachkova2021; @Hayakawa2022].
-These findings suggest that many sensory cells in the aboral organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis [@Jékely2021].
-While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.
-
-Conversely, previous research has demonstrated that mechanoreceptive sensory cells equipped with cilia (referred to as type 3 and type 4 cells) in ctenophores form direct synaptic connections with the sub-epithelial nerve net (SNN) and mesogleal neurons [@Burkhardt2023].
-Based on their ultrastructural features, these sensory cells are believed to respond to physical stimuli such as water flow, vibrations, and direct contact, transmitting information rapidly to cilia and muscle cells [@Hertwig1880; @Chun1880; @Eimer1880].
-
-Surprisingly, our study identified bridge cells as the primary cell type forming synaptic inputs onto neurons in the aboral organ.
-Bridge cells were first described by Tamm et al.
-based on their morphological characteristics, with cellular projections extending to the basal region of balancer cilia [@Tamm_2002].
-Our findings suggest that bridge cells are electrically coupled to balancer cilia and provide rapid feedback to neurons via synaptic transmission.
-Future studies analyzing differential responses to various sensory stimuli will not only elucidate the functional roles of bridge cells but also provide insights into the evolutionary relationship between transmission speed and synapse development.
-
-### Circuit Variability as a Driver of Behavioral Diversity
-
-Our study suggests that differences in neural circuits contribute to variations in the timing of ciliary arrest, which may play a role in modulating swimming direction in ctenophores [@Satterlie2015a].
-This finding provides evidence that increased circuit complexity contributes to behavioral diversification.
-By integrating multiple neurons and altering circuit structure, information can be branched, allowing for diverse motor patterns within the framework of fast synaptic transmission.
-
-The neural cells identified in the aboral organ exhibit a divergent feedforward connectivity pattern, where a single neuron forms presynaptic connections with multiple downstream targets.
-Such a circuit design minimizes unwanted signal interference, reduces processing delays, and enables relatively fast response execution [@Luo2021].
-Moreover, the presence of multiple divergent feedforward circuits suggests that slight variations in the synchronization and phase delay of ciliary beating (within the range of a few hundred milliseconds) may have evolved as a mechanism for generating behavioral diversity.
-
-From a broader neuroanatomical perspective, nervous system structures across various animals have been shaped through selective optimization to adapt to environmental constraints [@Bullmore2012; @Perin2011].
-For example, in nematodes, the spatial arrangement of ganglia is optimized to minimize the total length of neural wiring [@White1986].
-Similarly, in ctenophores, understanding how transmission via syncytial neurons and synaptic connections has been selectively optimized at the circuit level will provide crucial insights into the evolutionary scaling of neural networks from early nervous systems to complex brains [@Farnworth2024].
-In particular, this research may help elucidate how circuit modularization and layered structures emerged over evolutionary time.
+This architecture can be interpreted as a recurrent information loop, in which output from the ANN circuit modulates ciliary activity, and the resulting ciliary state is in turn relayed back to the ANN via bridge cells.
+Such a structure suggests a loop-based coordination mechanism, where the central circuit adapts its timing based on feedback from its target.
+Although the electrophysiological properties of bridge cells remain to be elucidated, their recursive integration into the ANN circuit indicates that they may play a central role in maintaining the robustness and flexibility of endogenous ciliary rhythms.
 
@@
 
-![**Supplemental figure**\
+![**Figure 1—figure supplement 1**\
 (A)...](figures/Supplemental_figure.png)
+
+![**Figure 2—figure supplement 1**\
+(A)...](figures/Supplemental_figure.png)

### 2025-05-06 manuscript/Manuscript.qmd
conclusion

@@
 
+### Conclusion
+
+How do differences in balancer cilia coordination across anatomical planes shape behavior?
+In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances.
+Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization—but these remain hypotheses.
+
+Our findings reveal a circuit composed of ANN neurons and bridge cells that coordinates ciliary beating rhythms in space and time without sensory input, and features a recursive architecture in which ciliary outputs are relayed back into the circuit via bridge cells.
+Such a configuration suggests that even with a limited number of cells, the system is capable of flexible and robust rhythm control.
+
+As Jagar et al.
+(2010) once described, this organ evokes the metaphor of a “ciliary brain.” The network we describe here similarly orchestrates output through dynamic, self-regulating coordination.
+These dynamics, visualized through live imaging of the entire cell ensemble, offer a window into the core functions of a nervous system—not in complexity, but in principle.
+
 ## Materials and Methods

### 2025-05-06 manuscript/Manuscript.qmd
ref

@@
 In this study, we generated the first high-resolution connectome of a ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of five-day-old *Mnemiopsis leidyi* cydippids using volumetric electron microscopy.
-Among these, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve nets (ANNs)—which exhibit multiple nuclei and a morphology distinct from the “beads-on-a-string” organization reported for the subepithelial nerve net (SNN) in one-day-old cydippids.
+Among these, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve nets (ANNs)—which exhibit multiple nuclei and a morphology distinct from the “beads-on-a-string” organization reported for the subepithelial nerve net (SNN) in one-day-old cydippids [@Burkhardt2023; @Sachkova2021].
 Instead of forming discrete nodal swellings, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion, suggesting a unique neuron subtype specific to the aboral organ.
@@
 These findings indicate that ANNs are unlikely to serve as motor drivers in the conventional sense.
-Rather, they may be involved in coordinating or modulating the timing and dynamics of ciliary activity.
+Rather, they may be involved in coordinating or modulating the timing and dynamics of ciliary activity [@keijzer2013; @dewiljes2015; @Veraszto2017].
 
@@
 Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs.
-inhibitory) of these neurons remains unresolved.
+inhibitory) of these neurons remains unresolved [@Ryan2013; @Moroz2014].
 However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.
@@
 Strikingly, this organization resembles circuit motifs found in other animals.
-For instance, in Platynereis dumerilii larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands, while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields.
-Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits.
+For instance, in *Platynereis dumerilii* larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands [@Veraszto2017], while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields [@Veraszto2017; @calderón2024].
+Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015;@roberts2022].
 
 Taken together, our findings suggest that the ANN network represents a non-motor, coordination-based neural system—one whose primary function is to tune and synchronize locomotor effectors rather than activate them directly.
-This departs fundamentally from reflex-arc models of early nervous systems and instead supports the idea that coordination, not command, may have been a foundational role of ancestral neurons.
+This departs fundamentally from reflex-arc models of early nervous systems [@arendt2021; @jékely2010] and instead supports the idea that coordination, not command, may have been a foundational role of ancestral neurons.
 
@@
 Originally described morphologically by Tamm et al.
-(2002), bridge cells extend thin projections toward the base of the balancer cilia.
-Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes.
+(2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @Tamm_2014b].
+Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024].
 These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed toward the ANN circuit.
 
-As revealed by our high-speed imaging, ciliary beat frequency (CBF) exhibited rhythmic changes over time that differed between anatomical planes, indicating spatially distinct coordination dynamics.
+As revealed by our high-speed imaging, ciliary beat frequency (CBF) exhibited rhythmic changes over time that differed between anatomical planes (Figures 4E and F), indicating spatially distinct coordination dynamics.
 These CBF rhythms are likely generated by the ANN circuit, while bridge cells may function to stabilize or adjust this rhythmic output in accordance with the real-time state of the cilia.
@@
 This architecture can be interpreted as a recurrent information loop, in which output from the ANN circuit modulates ciliary activity, and the resulting ciliary state is in turn relayed back to the ANN via bridge cells.
-Such a structure suggests a loop-based coordination mechanism, where the central circuit adapts its timing based on feedback from its target.
-Although the electrophysiological properties of bridge cells remain to be elucidated, their recursive integration into the ANN circuit indicates that they may play a central role in maintaining the robustness and flexibility of endogenous ciliary rhythms.
+Such a structure suggests a loop-based coordination mechanism [@arshavsky2003; @selverston2010], where the central circuit adapts its timing based on feedback from its target.
+Although the electrophysiological properties of bridge cells remain to be elucidated, their recursive integration into the ANN circuit indicates that they may play a central role in maintaining the robustness and flexibility of endogenous ciliary rhythms [@kennedy2024].
 
@@
 How do differences in balancer cilia coordination across anatomical planes shape behavior?
-In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances.
-Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization—but these remain hypotheses.
+In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances \[[@waggett1999; @colin2010; @courtney2020]\].
+Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.
 
@@
 As Jagar et al.
-(2010) once described, this organ evokes the metaphor of a “ciliary brain.” The network we describe here similarly orchestrates output through dynamic, self-regulating coordination.
+(2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011].
+The network we describe here similarly orchestrates output through dynamic, self-regulating coordination.
 These dynamics, visualized through live imaging of the entire cell ensemble, offer a window into the core functions of a nervous system—not in complexity, but in principle.
@@
 Additionally, we appended the quadrant number to the cell type name after an underscore ("\_") to indicate the cell-body’s location.
-For example, cells located in the first quadrant are named with "_Q1." If a cell is situated between the first and second quadrants, we appended "Q1Q2" to the cell-type name.
+For example, cells located in the first quadrant are named with "\_Q1." If a cell is situated between the first and second quadrants, we appended "Q1Q2" to the cell-type name.
 

### 2025-05-06 manuscript/Manuscript.qmd
supplement figure

@@
 ![**Figure 1—figure supplement 1**\
-(A)...](figures/Supplemental_figure.png)
+(A)...](figures/Figure1_Supplement1.png)
 
 ![**Figure 2—figure supplement 1**\
-(A)...](figures/Supplemental_figure.png)
+(A)...](figures/Figure2_Supplement1.png)

### 2025-05-06 manuscript/Manuscript.qmd
MS

@@
 Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioral patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system.
-Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig; @Hernandez-Nicaise; @Jager].
-These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022].
+Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011].
+These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022].
 Notably, early single-cell transcriptomic studies failed to identify canonical neuronal clusters using conventional markers [@Sebe-Pedros2018].
 It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were subsequently identified [@Sachkova2021; @Hayakawa2022].
-In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@Moroz].
+In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015].
 This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems.
@@
 To address the lack of structural understanding of how ctenophores control behavior, we focused on a unique sensory structure—the aboral organ.
-This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@Horiddge1965; @Tamm1980].
+This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@HORRIDGE1965; @Tamm1980].
 Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez-Nicaise1968; @Jager2011].
 A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency.
-Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@Horridge1965; @Tamm1982; @Tamm2014].
+Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @Tamm_2014b].
 Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997].
@@
 
-Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F) [@Martindale1999].
+Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999].
 The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999].
@@
 Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs.
-inhibitory) of these neurons remains unresolved [@Ryan2013; @Moroz2014].
+inhibitory) of these neurons remains unresolved [@Ryan2013;@Moroz2014].
 However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.
@@
 For instance, in *Platynereis dumerilii* larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands [@Veraszto2017], while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields [@Veraszto2017; @calderón2024].
-Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015;@roberts2022].
+Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015; @roberts2022].
 
@@
 How do differences in balancer cilia coordination across anatomical planes shape behavior?
-In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances \[[@waggett1999; @colin2010; @courtney2020]\].
+In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances [@waggett1999; @colin2010; @courtney2020].
 Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.

### 2025-05-06 manuscript/Manuscript.tex
MS

--- 2024-09-18 manuscript/Manuscript.tex
@@
 \usepackage[lmargin=0.8in,rmargin=0.8in,tmargin=0.8in,bmargin=0.8in]{geometry}
-\ifLuaTeX
-  \usepackage{luacolor}
-  \usepackage[soul]{lua-ul}
-\else
-  \usepackage{soul}
-
-\fi
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
-
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{241,243,245}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
-\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
-\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\makeatother
 
@@
 \hypersetup{
-  pdftitle={Cell-type diversity and connectome of the ctenophore apical organ},
+  pdftitle={Neural Connectome of the Ctenophore Statocyst},
   colorlinks=true,
@@
 
-\title{Cell-type diversity and connectome of the ctenophore apical
-organ}
+\title{Neural Connectome of the Ctenophore Statocyst}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{\textbf{Kei Jokura}\(^{1,2,3,4,5*}\), \textbf{Sanja
+Jasek}\(^{1,2,6}\), \textbf{Lara Kewalow}\(^{6}\), \textbf{Pawel
+Burkhardt}\(^{7}\), \textbf{Gáspár Jékely}\(^{1,2,6,*}\) \(^{1}\) Living
+Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom
+\(^{2}\) Biosciences, Faculty of Health and Life Sciences, University of
+Exeter, Exeter EX4 4QD, UK \(^{3}\) Grass Laboratory, Marine Biological
+Laboratory, Woods Hole, MA 02543, USA \(^{4}\) Exploratory Research
+Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan
+\(^{5}\) National Institute for Basic Biology (NIBB), Okazaki, 444-8585,
+Japan \(^{6}\) Heidelberg University, Centre for Organismal Studies
+(COS), 69120 Heidelberg, Germany \(^{7}\) Michael Sars Centre,
+University of Bergen, Norway \textbf{\(^{*}\) Correspondence:}
+\href{mailto:jokura@nibb.ac.jp}{\nolinkurl{jokura@nibb.ac.jp}},
+\href{mailto:gaspar.jekely@cos.uni-heidelberg.de}{\nolinkurl{gaspar.jekely@cos.uni-heidelberg.de}}}
 \author{}
@@
 \setstretch{1.2}
-\hfill\break
-
-Kei Jokura\textsuperscript{1,3,*}, Sanja Jasek\textsuperscript{1,2},
-Lara Kewalow\textsuperscript{2}, Pawel Burkhardt\textsuperscript{4},
-Gáspár Jékely\textsuperscript{1,2,*}\\
-
-\textsuperscript{1}Living Systems Institute, University of Exeter,
-Exeter, EX4 4QD, United Kingdom\\
-\textsuperscript{2}Heidelberg University, Centre for Organismal Studies
-(COS), 69120 Heidelberg, Germany\\
-\textsuperscript{3}Exploratory Research Center on Life and Living
-Systems (ExCELLS), National Institutes of Natural Sciences, Okazaki,
-Japan\\
-\textsuperscript{4}Michael Sars Centre, University of Bergen, Norway\\
-\textsuperscript{*}Correspondence: jokura@nibb.ac.jp,
-gaspar.jekely@cos.uni-heidelberg.de
-
 \subsection{Abstract}\label{abstract}
 
-\textbf{text in bold} \emph{italic} \ul{underline}
+Ctenophores possess a unique gravity receptor (statocyst) in their
+aboral organ formed by four clusters of ciliated balancer cells that
+collectively support a statolyth. During reorientation, differential
+load on the balancer cilia leads to altered beating of the ciliated comb
+rows to elicit turns. To study the neural bases of gravity sensing, we
+imaged by volume electron microscopy (vEM) the aboral organ of the
+ctenophore \emph{Mnemiopsis leidyi}. We reconstructed 909 cells,
+including syncytial neurons that form a nerve net. The syncytial neurons
+synapse on the balancer cells and also form reciprocal connection with
+the bridge cells that span the statocyst. High-speed imaging revealed
+that balancer cilia beat and arrest in a coordinated manner but with
+differences between the sagittal and tentacular planes of the animal,
+reflecting nerve-net organisation. Our results suggest a coordinating
+rather than effector function for the nerve net and inform our
+understanding of the diversity of nervous-systems organisation across
+animals.
 
@@
 
-First descriptions of AO: (Aronova, 1974)
+Ctenophores, or comb jellies, are gelatinous marine animals that
+actively swim by beating rows of fused cilia known as comb plates. These
+coordinated ciliary movements not only generate propulsion but also
+allow precise control of body orientation and directional changes within
+the water column. Despite lacking a centralized nervous system,
+ctenophores exhibit sophisticated and dynamic behavioral patterns and
+respond to a broad range of external stimuli, suggesting the presence of
+an evolutionarily unique and functionally integrated sensorimotor
+system. Several components of the ctenophore nervous system have been
+described, including a subepithelial nerve net (SNN), mesogleal neurons,
+tentacular nerves, and elements of nervous system in the aboral organ
+(Hernandez-Nicaise, 1984; Hertwig, 1880; Jager et al., 2010). These
+neural structures are characterized by unique features such as a
+syncytial architecture (Burkhardt et al., 2023) and an extensive
+repertoire of lineage-specific neuropeptides not found in other
+metazoans (Hayakawa et al., 2022; Sachkova et al., 2021). Notably, early
+single-cell transcriptomic studies failed to identify canonical neuronal
+clusters using conventional markers (Sebé-Pedrós et al., 2018). It was
+only through the use of pro-neuropeptide-based markers that distinct
+neural cell types were subsequently identified (Hayakawa et al., 2022;
+Sachkova et al., 2021). In parallel, phylogenomic and chromosomal
+synteny analyses have suggested that ctenophores may represent the
+sister group to all other animals (Li et al., 2021; Ryan et al., 2013;
+Schultz et al., 2023; Whelan et al., 2015), raising the possibility that
+their nervous system evolved independently from those of other metazoans
+(Moroz, 2015). This hypothesis positions ctenophores as a key lineage
+for re-evaluating the origins and diversity of neural systems. However,
+without a detailed understanding of how ctenophore neurons are spatially
+organized and interconnected, it remains unclear how their nervous
+system gives rise to behavior. Clarifying this relationship between
+neural architecture and behavioral control is essential to uncovering
+the logic of ctenophore neural function.
 
-The lobate ctenophore \emph{Mnemiopsis}
+To address the lack of structural understanding of how ctenophores
+control behavior, we focused on a unique sensory structure---the aboral
+organ. This gravity-sensing organ contains a statocyst supported by four
+mechanosensory balancer cells, which act as pacemakers for comb plate
+beating and posture control (HORRIDGE, 1965; Tamm, 1980).
+Ultrastructural and immunohistochemical studies have revealed that the
+aboral organ contains diverse sensory and neural components, including a
+distinct neural structure termed the deep nerve net (Aronova, 1974;
+HERNANDEZ-NICAISE, 1968; Jager et al., 2010). A key feature of balancer
+cilia is their bidirectional, mechanoresponsive control of beat
+frequency. Depending on the direction of deflection and the animal's
+geotactic state, balancer cilia may either increase or decrease their
+activity, indicating a dynamic response system not fixed like vertebrate
+hair cells (HORRIDGE, 1965; Tamm, 1982; Tamm, 2014a). Previous studies
+have implicated membrane potential, intracellular calcium concentration,
+and synaptic input in modulating this response (Lowe, 1997). In
+particular, deflection-induced excitation can be abolished by calcium
+channel inhibitors or calcium-free conditions (Lowe, 1997), while high
+extracellular K⁺ depolarizes and activates cilia in the absence of
+mechanical stimuli (Lowe, 1997). Local calcium application further
+suggests that Ca²⁺ influx at the ciliary base may be required for the
+excitatory response (Lowe, 1997). These findings support the hypothesis
+that neural modulation plays a critical role in shaping balancer cell
+output (Lowe, 1997).
 
-The lithocytes have a large calcium carbonate concretion and\ldots{}
-(Tamm, 2014a) Living lithocytes are transported along the balancer cilia
-to the tip of the statocyst (Noda and Tamm, 2014) The projections of a
-group of bridge cells located on two sides of the apical organ forms an
-arch and connects the balancers along the tentacular plane (Tamm and
-Tamm, 2002).
-
-The easiest way is to use the command line to add new refs to the .bib
-file:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\#curl {-}LH "Accept: application/x{-}bibtex" https://doi.org/10.7554/eLife.91258.1 \textgreater{}\textgreater{} references.bib}
-\end{Highlighting}
-\end{Shaded}
+To investigate how balancer cilia are modulated within the aboral organ,
+we combined volume electron microscopy and high-speed imaging in the
+ctenophore \emph{Mnemiopsis leidyi}. Through connectomic reconstruction,
+we characterized the synaptic architecture of the aboral nerve net, and
+functional imaging revealed correlated activity patterns in balancer
+cilia. These results suggest that the aboral nerve net plays a central
+role in integrating sensory input and coordinating motor output.
 
@@
 \subsubsection{\texorpdfstring{Volume EM reconstruction of the
-\emph{Mnemiopsis} apical
-organ}{Volume EM reconstruction of the Mnemiopsis apical organ}}\label{volume-em-reconstruction-of-the-mnemiopsis-apical-organ}
+\emph{Mnemiopsis} aboral
+organ}{Volume EM reconstruction of the Mnemiopsis aboral organ}}\label{volume-em-reconstruction-of-the-mnemiopsis-aboral-organ}
 
-\begin{itemize}
-\tightlist
-\item
-  specimen
-\item
-  volume, resolution, fixation, quality
-\item
-  number of sections
-\item
-  skeletonisation in Catmaid
-\item
-  number of cells
-\item
-  general anatomy, regions, quadrants
-\end{itemize}
+For vEM analysis, we used a five-day-old cydippid larva of
+\emph{Mnemiopsis leidyi}. To obtain high-quality ultrastructural
+preservation, we used high pressure freezing followed by freeze
+substitution and resin embedding (Epon). The specimen was sectioned from
+the aboral tip and we collected approximately 1,000 ultra-thin (50 nm)
+serial sections as ribbons. The sections were imaged in a Zeiss Gemini
+500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only
+the region containing the aboral organ. Subsequently, 620 of these
+images were stitched and aligned in TrakEM2 (Cardona et al., 2012). The
+final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We
+skeletonised and annotated all cells in this dataset in CATMAID
+(Saalfeld et al., 2009), resulting in a reconstruction of 909 cells with
+nuclei and cell bodies intact. Most of the cells, including the
+balancers and the aboral nerve net neurons are fully contained within
+the volume, due to the compact and mostly self-contained organisation of
+the aboral organ.
+
+In cells containing cilia, we also traced cilia along their length and
+annotated all basal bodies. For neuronal skeletonization, nodes were
+placed to interconnect the profiles of the same neuron's processes
+across layers, extending the skeleton until all branches were fully
+traced. All nodes relevant to synaptic sites were tagged, and skeletons
+were named and assigned multi-level annotations. As described later,
+some neurons formed loop-like structures (anastomosed neurons), wherein
+separated branches often rejoined either the main branch or smaller
+branches (Burkhardt et al., 2023). In such cases, branch nodes were
+placed near the closest existing node and annotated accordingly (CATMAID
+only supports skeleton trees). The entire skeletonized volume was
+composed of 134,591 nodes. 88 fragments could not be attached to
+somata-associated skeletons. Most of these fragments represent short
+skeletal branches that could not be traced beyond gaps or low-quality
+layers.
+
+Next, we divided the entire aboral organ into four broad quadrants
+(Q1-Q4) to facilitate grouping the identified cells (Figure 1G)
+(Martindale and Henry, 1999). The general body plan of ctenophores, when
+viewed from the aboral side, exhibits biradial symmetry around the anal
+pores (Martindale and Henry, 1999). This symmetry corresponds to the
+four blastomeres present at the four-cell stage during early embryonic
+development (Martindale and Henry, 1999). We assigned each of the 909
+skeletonised cells to one of the four quadrants.
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure1.png}
+{\centering \includegraphics{figures/Figure1.png}
 
@@
 
-\caption{\textbf{Figure 1. Title fig 1} pic of a larva, boxed region of
-AO, schematic, catmaid pic with slice, dimensions, quadrants, regions,
-(A) legend (B) legend.}
+\caption{\textbf{Figure 1. Morphology of the aboral organ in
+\emph{Mnemiopsis leidyi}}\\
+(A) Whole-body image of a 5-day-old \emph{M. leidyi} cydippid larva in
+lateral view. The boxed region indicates the aboral organ (ao). Scale
+bar: 100 µm.\\
+(B) Schematic diagram of the aboral view of a cydippid larva. The
+two-tone coloration represents the biradial symmetry of the body. For
+convenience, the body is divided into four quadrants, designated as the
+first (Q1) through fourth (Q4) quadrants. The two primary viewing angles
+are referred to as the sagittal plane (S) and the tentacular plane (T).
+Abbreviations: ao, aboral organ; cg, ciliated groove; cr, comb plates;
+tb, tentacle bulbs.\\
+(C) Aboral views of the aboral organ highlighting its spatial
+organization. The left panel presents a schematic representation of the
+aboral organ, illustrating the four body quadrants (Q1--Q4), color-coded
+consistently with panel (B). The right panel shows a corresponding
+differential interference contrast (DIC) image from the same
+perspective. The region enclosed by the dotted line delineates the
+boundary of the aboral organ. The sagittal (S) and tentacular (T) arrows
+indicate the viewing directions corresponding to those planes. Scale
+bar: 50 µm. bal, balancer; cg, ciliated grooves; li, lithocyte.\\
+(D) Lateral views of the aboral organ in two orthogonal planes captured
+using DIC microscopy. The left panel shows a lateral view in the
+sagittal (S) plane, while the right panel displays a view in the
+tentacular (T) plane. Balancer cilia (bal) and lithocytes (li) are
+enclosed within the dome (do). Scale bar: 50 µm. A--O indicates the
+aboral--oral body axis.\\
+(E) Schematic representations of the aboral organ in sagittal and
+tentacular planes. The left panel shows a lateral schematic in the
+sagittal plane (S), and the right panel shows the same in the tentacular
+plane (T), corresponding to the views in panel (D). Colors match the
+quadrant scheme shown in panels (B) and (C).\\
+(F) Example of cell tracing using the collaborative annotation toolkit
+CATMAID for large-scale electron microscopy image datasets. The
+spherical objects indicate nuclear positions, while the lines represent
+the traced cell centers. Scale bar: 10 µm.\\
+(G) Three-dimensional reconstruction of cells composing the aboral
+organ, displayed from different perspectives. The left panel presents an
+aboral view of the aboral organ. The middle panel shows a sagittal plane
+view, while the right panel provides a tentacular plane view. Cells are
+color-coded according to their respective quadrants. Lithocytes (li) are
+represented as three gray spheres, while balancers (bal) are depicted as
+gray lines. Scale bar: 25 µm.}
 
@@
 
-\subsubsection{Classification of cell
-types}\label{classification-of-cell-types}
+\subsubsection{An aboral synaptic nerve net of syncytial
+neurons}\label{an-aboral-synaptic-nerve-net-of-syncytial-neurons}
 
-\begin{itemize}
-\tightlist
-\item
-  ultrastructural criteria (cilia, vesicles etc)
-\item
-  number of types, cells per type - table
-\end{itemize}
+Classical neural staining techniques did not provide clear images of
+neurons at the aboral organ. However, ultrastructural studies revealed
+morphological evidence for neural elements, including synapses that are
+located on the epithelial floor of the aboral organ---a region where
+epithelial cells are densely packed and form a floor-like structure
+(Hernandez-Nicaise, 1973; Horridge and Mackay, 1964). In our dataset, we
+identified synapses based on the previously described pre-synaptic triad
+morphology consisting of a single layer of vesicles, a cisterna of
+smooth endoplasmic reticulum and an associated mitochondrion. At
+synaptic sites, we marked mitochondria as synaptic nodes (orange in
+CATMAID) and connected this node to the nearest node in postsynaptic
+cells across the regions where synaptic vesicles aligned with the
+presynaptic cell's membrane (light blue arrows in CATMAID)
+(Supplementary Figure S2A). We could not detect specialized
+post-synaptic structures, in agreement with previous studies
+(Hernandez-Nicaise, 1973). Synapses were identified as either monoadic
+or polyadic, with one pre-synaptic neuron forming connections with one
+or multiple post-synaptic cells (Supplementary Figure S2B).
+
+We reconstructed three aboral nerve-net (ANN) neurons, each with a
+syncytial morphology (Figure 2A and B). Each ANN had multiple nuclei
+(ranging from two to five per neuron), with anastomosing membranes
+(Supplementary Figure S2C). These neurons are distinct from the
+syncytial subepithelial nerve net (SNN) neurons with blebbed morphology
+previously identified by vEM in the body wall of \emph{Mnemiopsis}
+(Burkhardt et al., 2023). Our results reveal a second type of syncytial
+neurons in ctenophores (Supplementary Figure S2D).
+
+Furthermore, ANNs are also different from other neurons previously
+reported by EM, including mesogleal neurons and ciliated sensory cells
+that synapse on the SNN (types 1-4)(Burkhardt et al., 2023). Based on
+their distinct size and position, we classified the ANNs into two types.
+The first type is a single large neuron (ANN\_Q1-4) spans all four
+quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has
+150 presynaptic sites. The second type consists of two neurons
+(ANN\_Q1Q2 and ANN\_Q3Q4), each spanning two quadrants and containing
+two nuclei. ANN\_Q1Q2 had 119 presynaptic structures, whereas ANN\_Q3Q4
+had 68 presynaptic structures.
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{figures/Figure_network.png}
+{\centering \includegraphics{figures/Figure2.png}
 
@@
 
-\caption{\textbf{Figure 2. Title fig 2} Balancer complex: balancer,
-lithocyte, groove 2 types, bridge - EM pics, (A) legend (B) legend.}
+\caption{\textbf{Figure 2. Organisation of the aboral synaptic nerve
+net}\\
+(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4),
+which spans all four quadrants and contains multiple nuclei (blue). The
+spheres represent the positions of individual nuclei. The left panel
+shows a aboral view of the aboral organ, the middle panel presents a
+sagittal plane view, and the right panel provides a tentacular plane
+view.\\
+(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each
+spanning two quadrants and containing multiple nuclei. The spheres
+indicate the positions of the nuclei. The left panel shows a aboral view
+of the aboral organ, the middle panel presents a sagittal plane view,
+and the right panel provides a tentacular plane view.\\
+(C) Graph showing the number of mitochondria per cell and the proportion
+of mitochondria involved in synapse formation for each cell type. Green
+represents mitochondria associated with synapses, while gray represents
+mitochondria not involved in synapse formation. ANN has a significantly
+higher number of mitochondria compared to other cell types, with most of
+them forming synapses. Except for lithocytes, each cell type has a
+sample size of at least three cells. The abbreviations are as follows:
+bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv
+(dense vesicle cells), imc (intra-multiciliated cells), la (lamellate
+bodies), li (lithocytes), pl (plumose), and ef (epithelial floor
+cells).\\
+(D) Localization of mitochondria within ANN Q1-4. Red indicates
+mitochondria associated with the presynaptic triad structures, yellow
+marks mitochondria containing synaptic vesicles but lacking a clearly
+defined presynaptic triad, blue represents mitochondria with unclear
+synaptic vesicles, and black denotes mitochondria where no synaptic
+vesicles were identified.\\
+(E) Representative electron micrographs of presynaptic triad structures
+observed in the dataset and their schematic diagrams. The left diagram
+illustrates synaptic projections from the center ANN (ANN Q1-4) to the
+lateral ANN (ANN Q3Q4), while the right diagram shows synaptic
+projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN
+Q1-4). The abbreviations are as follows: mi (mitochondrion), er
+(endoplasmic reticulum), and sv (synaptic vesicles).\\
+(F) Plot showing the three-dimensional positions of synapses. Synapses
+from the center ANN to the lateral ANN are shown in magenta, while
+synapses from the lateral ANN to the center ANN are shown in cyan. Blue
+dots indicate the locations of autapses within ANN Q1-4.}
 
@@
 
-``the larval bridge has not yet developed forks at its ends, as seen in
-adult Mnemiopsis bridges''(Tamm and Tamm, 2002)
+\subsubsection{Synaptic connectome of the gravisensory
+organ}\label{synaptic-connectome-of-the-gravisensory-organ}
 
-Figure3 - dome, bicil, monocil, no-cil, plumose, dense vesicle cells
-(rename), gland? cells, oligo/multiciliated cells
+Our synapse annotation revealed synapses between the ANN neurons and
+synapses that ANN\_Q1-4 forms on itself (autapses). The ANN neurons also
+form synapses on the gravity-sensing balancer cells.
 
-Plumose cells (Hernandez-Nicaise, 1984) may be pressure sensors (Tamm,
-2014b)
+Balancer cells are monociliated cells with long motile cilia that form
+four bundles of compound cilia, one in each quadrant. These bundles come
+together at the centre of the aboral organ to support the cellular mass
+of the statolyth. The bending of the cilia and the position of somata
+differed clearly when viewed laterally from the sagittal or the
+tentacular plane (Figure 3B). In each quadrant, there were between 28-37
+balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28). Each cell contained 3
+to 10 mitochondria.
 
-Figure4 - lamellate body complex, LB, intra-multi-ciliated, relationship
-to balancer
+ANN\_Q1-4 formed synapses on balancer cells in all four quadrants (on 7
+cells in Q1, 11 cells in Q2, 6 cells in Q3 and 10 cells in Q4) while
+ANN\_Q1Q2 and ANN\_Q3Q4 synapsed to balancer cells in their respective
+quadrants (ANN\_Q1Q2 on 6/37 cells in Q1 and 8/32 cells in Q2; ANN\_Q3Q4
+on 1/32 cells in Q3 and 5/28 cells in Q4). Some balancer cells received
+inputs from both ANN\_Q1-4 and either of ANN\_Q1Q2 or ANN\_Q3Q4. While
+previous studies suggested the presence of afferent synapses from
+balancer cells to neurons (Hernandez-Nicaise, 1974), our data did not
+reveal any presynaptic sites in balancer cells.
 
-Figure5 - nerve net, different fragments/cells/syntytia, EM, relation to
-entire organ
+The second group of cells that form synaptic contact with the ANN were
+the bridge cells. Bridge cells, first described by Tamm et al.~in 2002
+(Tamm and Tamm, 2002), are characterized by bundles of elongated
+processes filled with microtubules that arch over the epithelial layer,
+resembling a bridge. Their somata are located at the base of the paired
+balancer-cell clusters along the tentacle surface and extend across the
+sagittal plane towards the base of the opposite balancer cells. Bridge
+cells form two distinct cell groups across the sagittal plane, in the
+Q1Q2 and Q3Q4 regions. In the Q1Q2 region, we identified 14 bridge
+cells, in Q3Q4, 12 cells.
 
-Figure 6 - cilia? renderings of basal bodies, centrioles, length,
-axoneme structure (table), table about cell types, number of cilia etc.
+Bridge cells had presynaptic sites with the typical presynaptic triad
+structure near 30\% of their mitochondria. These bridge-cell synapses
+were formed on ANN neurons or other bridge cells. Bridge cells in the
+Q1Q2 region formed synapses on ANN\_Q1Q2 (3 cells) or on both ANN\_Q1Q2
+and ANN\_Q1-4 (2 cells). Bridge cells in Q3Q4 synapsed on ANN\_Q3Q4 (1
+cell) or ANN\_Q1-4 (6 cells).
 
-\subsection{Synapses}\label{synapses}
+Nearly all bridge cells (25/26) also received synaptic input from ANNs.
+Bridge cells in Q1Q2 received inputs from ANN\_Q1Q2 (11 cells),
+ANN\_Q1-4 (1 cell), or both (2 cells). In Q3Q4, synapses on bridge cells
+were from ANN\_Q3Q4 (1 cell), ANN\_Q1-4 (7 cells), or both (1 cell).
 
-``synaptic regions with the characteristic presynaptic triad mor-
-phology of ctenophore nerves (i.e., a single layer of vesicles, smooth
-ER sac, and closely apposed mito- chondria (Hernandez-Nicaise, 1973;
-Horridge and Mackay, 1964) Hernandez- Nicaise, 1973)''
+In both The Q1Q2 and the Q3Q4 regions, bridge cells also formed synapses
+with adjacent bridge cells. However, we found no synapses across the
+sagittal plane to bridge cells in the opposite region.
 
-Figure 7 - synapses, reconstruction, annotation, numbers, distribution
-of partners, example EM (figure supplement - lots of EM), mitochondria
-(Fig suppl - mitochondrion stats - Sanja)
+To analyze the synaptic connectivity graph of the balancer organ and the
+ANN, we grouped cells of the same type and in the same region into a
+single node, summed the number of synapses, and laid out the network to
+reflect the anatomy of the four quadrants (Figure 3H). The network is
+characterised by feedback connections among ANN neurons and between ANN
+and bridge cells with a clear rotational (or mirror) symmetry. Q1Q2 and
+Q3Q4 also form separate local circuits centered around their regional
+ANN neurons while ANN\_Q1-4 mediates global connectivity across the
+entire balancer organ. Notably, there were no synapses from the
+mechanosensory balancer cells to the ANN or from ANN to motor cells
+(e.g.~ciliary groove), contrary to our initial expectation of a
+sensory-motor or input-output model of balancer function.
 
-Gap juntions? Innexins are expressed in the apical organ in specific
-patches of cells in each quadrant (Ortiz et al., 2023)
+The ANN neurons also formed synapses on other cell types in the aboral
+organ, including the dense-vesicle cells, EF cells and several
+non-ciliated, monociliated or biciliated cells. These cell types and
+synaptic connections are outside the gravisensory organ and will be
+described elsewhere.
 
-Ctenophore synapses (Hernandez-Nicaise, 1973)
+\begin{figure}[H]
 
-We found no synapses between bridge and balancer, in agreement with Tamm
-(Tamm and Tamm, 2002).
+{\centering \includegraphics{figures/Figure3.png}
 
-Figure 8 - connectome cell-type based, quadrant based, nerve net
-syntycia, bridge, quadrants (Fig suppl - cell-based network)
+}
 
-Tamm\&Tamm observed synapses from neurites to bridge cells (2002). and
-in Mnemiopsis larvae observed synapses from bridge cells onto other
-cells
+\caption{\textbf{Figure 3. Reconstruction of the balancers and bridges
+that receives synapses from the ANN.}\\
+(A) Graph showing the number of synaptic outputs from the ANNs (ANN
+Q1-4: blue, ANN Q1Q2: orange, ANN Q3Q4: pink) circuit for each cell
+type. The balancer and bridge cell types receive relatively high outputs
+from the ANNs circuit. ANN represents the number of synapses within the
+ANNs circuit. The abbreviations are as follows: bal (balancer), brg
+(bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells),
+imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes),
+pl (plumose), and ef (epithelial floor cells).\\
+(B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink,
+and Q3Q4 in orange) and balancer ciliated cells (light blue). The
+thickness of the arrows and the numbers correspond to the number of
+synapses. Synaptic structures from the same neuron targeting the same
+balancer cell are grouped into hexagons, with the number of cells added
+to the label.\\
+(C) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4
+quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while
+the Q3Q4-side bridge cells (6 cells) are shown in xxx color. The
+morphology of individual bridge cells extending across opposite quadrant
+regions is depicted. The spheres represent the positions of individual
+nuclei. The left panel shows an aboral view of the aboral organ, the
+middle panel presents a sagittal plane view, and the right panel
+provides a tentacular plane view.\\
+(D) Mitochondrial localization within bridge cells and associated
+presynaptic triad structures. Red indicates mitochondria associated with
+presynaptic triad structures, yellow marks mitochondria containing
+synaptic vesicles but lacking a clearly defined presynaptic triad, blue
+represents mitochondria with unclear synaptic vesicles, and black
+denotes mitochondria where no synaptic vesicles were identified. The
+left panel shows a dorsal view of the aboral organ, the middle panel
+presents a sagittal plane view, and the right panel provides a
+tentacular plane view.\\
+(E) Synaptic connections from ANN neurons to balancer ciliated cells.
+The positions of synapses from ANNs to balancers (magenta) are
+indicated. The left panel presents an aboral view, while the right panel
+shows a lateral view and a tentacular plane view. Balancer ciliated
+cells are depicted in light gray.\\
+(F) Synaptic connections between ANNs and bridge cells. The positions of
+synapses from ANNs to bridge cells (magenta) and from bridge cells to
+ANNs (light blue) are indicated. The left panel shows an aboral view,
+while the right panel presents a tentacular plane lateral view. Bridge
+cells are shown in light gray.\\
+(G) Connectivity matrix of the gravity-sensing neural circuit. Columns
+represent presynaptic cell groups, while rows represent postsynaptic
+cell groups. The numbers and varying shades of blue correspond to the
+number of synapses.\\
+(H) Complete connectivity map of the gravity-sensing neural circuit.
+Cells belonging to the same group are enclosed in hexagons, and the
+number of cells is added to their labels. The thickness of the arrows
+and the numerical values indicate the number of synapses. ANN Q1-4 are
+shown in blue, ANN Q1Q2 in orange, ANNs in Q3Q4 in pink, balancer cell
+groups in gray, and bridge cell groups in yellow.}
 
-Inserting Figures
+\end{figure}%
 
-You can add your figures into the rendered document. We saved the
-figures into /manuscript/figures or /manuscript/figure\_supplements and
-can insert them from there. We use knitr::include\_graphics for this.
-The title and legend can also be edited, as will as the width of the
-output figure.
+\subsubsection{Dynamics of balancer cilia imply a coordinating function
+for the nerve
+net}\label{dynamics-of-balancer-cilia-imply-a-coordinating-function-for-the-nerve-net}
 
-\subsection{Equations}\label{equations}
+To investigate the regulation of balancer function, we carried out a
+high-speed video-microscope analysis of balancer cilia. Previous studies
+(Lowe, 1997; Tamm, 1982, 1980) have established that balancer cilia
+function as mechanoreceptors, with their beating frequency modulated by
+inclination. Tamm also suggested that differences in statolith
+morphology and the shape of balancer cilia between the tentacular and
+sagittal planes could lead to different forces exerted on cilia by the
+statolith (Sidney L. Tamm, 2015a; Tamm, 2014b). To further explore this,
+we used a tilted microscope with a vertical stage where we mounted
+immobilised cydippid larvae with their aboral-oral axis aligned in
+different orientations relative to the gravity vector. We then compared
+larvae oriented in different angles and either with their sagittal or
+tentacular plane parallel to the sample stage.
 
-Equations can also be inserted, Insert -\textgreater{} Display Math:
+In larvae with their sagittal plane facing the objective, we could
+compared balancer-cilia movements between Q1(4) and Q2(3). In other
+larvae oriented in the tentacular plane, we could simultaneously image
+Q1(3) and Q2(4). We used a high-speed camera and recorded at 100 fps for
+2 minutes (12,000 frames). To analyse ciliary beating, we selected
+regions of interest (ROIs) in areas where brightness changes indicated
+ciliary beating. Ciliary beating was both manually quantified and
+plotted as kymographs.
 
-\[
-\bar{X} = \frac{\sum_{i=1}^{n} x_{i}}{n}
-\]
+During the two-minute recordings, balancer cilia could beat fast, slow,
+or exhibit abrupt stops (arrest) and start moving again (re-beat).
+Occasionally, large body-contractions moved the entire cydippid out of
+frame, and data from these episods were excluded. We focused on three
+metrics for inter-quadrant comparison: (1) the timing of ciliary arrest,
+(2) the timing of re-beat after an arrested phase, and (3) ciliary beat
+frequency.
 
-\subsection{Sourcing code and working with
-variable}\label{sourcing-code-and-working-with-variable}
+We found that arrest and re-beat events were synchronized between
+balancers across the sagittal plane. However, while the timing of
+re-beat was also near-simultaneous along the tentacular plane, arrest
+timing was offset by up to 2.16 seconds (Figure 4D). Overall, these data
+reveal that arrests are only coordinated between Q1-Q2 and Q3-Q4,
+whereas re-beat is coordinated over all four quadrants.
 
-The `analysis/scripts/statistics\_for\_paper.R' script is sourced and it
-runs but the output is not included in the knitted output. But we can
-access the variables defined in the sourced script simply by adding ` r
-var\_name ` between ` backticks, in this case max\_PRC value is (now
-this number comes from our sourced script).
+As shown in Figure 4E, beat frequencies of left and right balancers
+fluctuated substantially in both orientations. However, while left-right
+activity was largely synchronized in sagittal pairs (Figure 4E, left),
+tentacular pairs (Figure 4E, right) showed strikingly periodic
+alternations between left and right balancer activity. This pattern
+suggested an offset in timing during arrest or re-beat events between
+tentacular balancers.
 
-If we update the data, the script can recalculate the variable we want
-to refer to in the text and update the number.
+To quantify this, we calculated a rolling Pearson correlation between
+left and right balancer activity in each plane. The resulting mean
+correlations were significantly higher in the tentacular plane compared
+to the sagittal plane (Figure 4F; t-test, p = 0.014). This result
+indicates that tentacular balancers beat in a more coordinated,
+mirror-symmetric manner, whereas sagittal balancers exhibit greater
+variability between sides.
+
+Evaluating these data in light of the circuit diagram suggests that
+shared ANN inputs (ANN\_Q1Q2 or ANN\_Q3Q4) to a pair of balancers along
+the sagittal plane may underlie their synchronized arrests. In contrast,
+in the tentacular plane, separate ANNs innervate the balancer pairs. At
+the same time, the ANN\_Q1-4 neuron synapses on all four balancers,
+hinting at a neural substrate for their synchronised re-beat.
+
+Notably, bridge neurons associated with each quadrant region (Q1Q2 or
+Q3Q4) extend their processes across the midline, forming connections
+that span between opposite quadrant domains when viewed along the
+tentacular plane. This morphology suggests a potential role for bridge
+neurons in modulating inter-quadrant coordination. By linking ANN
+activity across orthogonal planes, these neurons may contribute to the
+dynamic regulation of CBF patterns across the whole system.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{figures/Figure4.png}
+
+}
+
+\caption{\textbf{Figure 4. Differences in Balancer Cilia Movement
+Control Due to Variations in the Neural Circuit Pathways}\\
+(A) Schematic diagram of differential interference contrast (DIC)
+microscopy setup for imaging balancer cilia movement. The microscope was
+tilted 90 degrees so that the stage was positioned vertically. A
+monochrome CMOS camera sensitive to near-infrared (NIR) light was used,
+synchronized with an 850 nm strobe light source. The movement of the
+balancer cilia was recorded using a 40× objective lens.\\
+(B) Representative DIC images of the aboral organ viewed along the
+sagittal (left) and tentacular (right) planes. Arrowheads indicate the
+balancer cilia selected for kymograph-based analysis in each
+orientation.\\
+(C) Representative kymographs showing arrest and re-beat events of
+balancer cilia. Left and right balancer cilia were simultaneously
+recorded in either the sagittal (left) or tentacular plane (right).
+Arrowheads mark the onset of arrest (open) and re-beat (filled);
+horizontal arrows represent 1-second intervals.\\
+(D) Boxplots showing time differences in the onset of arrest (white
+boxes) and re-beat (gray boxes) between left and right balancer cilia.
+Colors of the box outlines indicate the imaging plane: sagittal (blue)
+or tentacular (orange). Each dot represents a single larva.\\
+(E) Bar plots showing ciliary beat frequency (CBF) dynamics of left and
+right balancers (top and bottom, respectively) recorded from sagittal
+(left) and tentacular (right) planes. Positive and negative values
+represent opposite sides within each plane.\\
+(F) Mean rolling Pearson correlation of left--right balancer CBF
+calculated with a 20-frame window. Each dot represents an individual
+larva. Balancers on the tentacular plane showed significantly higher
+correlation than those on the sagittal plane (t-test, p = 0.014).}
+
+\end{figure}%
+
+\subsection{Discussion}\label{discussion}
+
+\subsubsection{Functional Nerve Net for Ciliary
+Coordination}\label{functional-nerve-net-for-ciliary-coordination}
+
+In this study, we generated the first high-resolution connectome of a
+ctenophore nervous system by reconstructing approximately 900 cells from
+the aboral organ of five-day-old \emph{Mnemiopsis leidyi} cydippids
+using volumetric electron microscopy. Among these, we identified a
+previously undescribed type of syncytial neuron---here referred to as
+aboral nerve nets (ANNs)---which exhibit multiple nuclei and a
+morphology distinct from the ``beads-on-a-string'' organization reported
+for the subepithelial nerve net (SNN) in one-day-old cydippids
+(Burkhardt et al., 2023; Sachkova et al., 2021). Instead of forming
+discrete nodal swellings, ANNs extend through intercellular spaces,
+weaving between neighboring cells in a smooth, interdigitating fashion,
+suggesting a unique neuron subtype specific to the aboral organ.
+
+Remarkably, ANNs form no synaptic output to classical effector organs
+such as muscles or comb plates. All observed output synapses were
+directed to balancer ciliary cells. Moreover, the projection patterns of
+individual ANNs were asymmetric, selectively targeting different regions
+of the balancer system. These findings indicate that ANNs are unlikely
+to serve as motor drivers in the conventional sense. Rather, they may be
+involved in coordinating or modulating the timing and dynamics of
+ciliary activity (Keijzer et al., 2013; Verasztó et al., 2017; Wiljes et
+al., 2015).
+
+Notably, anatomical differences in ANN projection patterns across
+sagittal and tentacular planes corresponded with plane-specific
+differences in balancer cilia dynamics. This correlation supports the
+idea that individual ANNs may play opposing or complementary roles in
+regulating ciliary beating---potentially exerting excitatory or
+inhibitory effects to shape spatiotemporal activity patterns. Because
+known neurotransmitter-related genes show little homology in
+ctenophores, the functional polarity (excitatory vs.~inhibitory) of
+these neurons remains unresolved (Moroz et al., 2014; Ryan et al.,
+2013). However, the combination of circuit architecture and behavioral
+outputs provides the first functional insight into the role of neurons
+in a ctenophore.
+
+Strikingly, this organization resembles circuit motifs found in other
+animals. For instance, in \emph{Platynereis dumerilii} larvae, a single
+cholinergic motor neuron synchronously halts multiple ciliary bands
+(Verasztó et al., 2017), while a pair of serotonergic neurons project
+axons that cross at the midline in a chiasm-like fashion, enabling
+reciprocal modulation of left and right ciliary fields (Calderón et al.,
+2024; Verasztó et al., 2017). Although the morphology differs, these
+similarities point to convergent evolution of distributed ciliary
+control circuits(Moroz, 2015; Roberts et al., 2022).
+
+Taken together, our findings suggest that the ANN network represents a
+non-motor, coordination-based neural system---one whose primary function
+is to tune and synchronize locomotor effectors rather than activate them
+directly. This departs fundamentally from reflex-arc models of early
+nervous systems (Arendt, 2021; Jékely, 2010) and instead supports the
+idea that coordination, not command, may have been a foundational role
+of ancestral neurons.
+
+\begin{figure}[H]
+
+{\centering \includegraphics{figures/Figure5.png}
+
+}
+
+\caption{\textbf{Figure 5. Comparison of Neural Circuits Regulating
+Ciliary Movement}\\
+(A) Each neuron represented by a circle is color-coded to indicate
+putative homologous cell types involved in ciliary movement regulation.
+Ciliated cells are shown in gray. Synapses are indicated by arrows, with
+magenta representing synapses that induce ciliary arrest and blue
+representing synapses that induce ciliary re-beat or an increase in
+ciliary beating frequency.\\
+(Left) Gravity-sensing neural circuit of \emph{M. leidyi}, as revealed
+in this study. Bridge cells (yellow squares) are suggested to be
+electrically coupled (indicated by yellow zigzag lines), implying their
+potential involvement in feedback mechanisms between neurons and
+ciliated cells.\\
+(Righr) Neural circuit regulating prototroch ciliary movement in
+\emph{P. dumerilii}. Serotonergic neurons (Ser-h1 neurons, blue)
+activate ciliary movement, while cholinergic neurons (MC neurons,
+magenta) induce ciliary arrest. Arrows directed toward ciliated cells in
+the head prototroch are color-coded accordingly.}
+
+\end{figure}%
+
+\subsubsection{Bridge Cells as Feedback Regulators of Ciliary
+Rhythms}\label{bridge-cells-as-feedback-regulators-of-ciliary-rhythms}
+
+In our volumetric reconstruction, bridge cells were identified as the
+only cell type providing synaptic input to the ANN network, positioning
+them as key upstream regulators of the ciliary coordination system in
+the aboral organ. Originally described morphologically by Tamm et
+al.~(2002), bridge cells extend thin projections toward the base of the
+balancer cilia (Tamm, 2014a; Tamm and Tamm, 2002). Based on this spatial
+relationship, they may receive information about the physical state of
+the cilia---such as mechanical load or phase---or possibly local field
+potential (LFP) changes (Jurisch-Yaksi et al., 2024; Sheu et al., 2022).
+These features suggest that bridge cells may act as transducers,
+converting local ciliary signals into neural inputs directed toward the
+ANN circuit.
+
+As revealed by our high-speed imaging, ciliary beat frequency (CBF)
+exhibited rhythmic changes over time that differed between anatomical
+planes (Figures 4E and F), indicating spatially distinct coordination
+dynamics. These CBF rhythms are likely generated by the ANN circuit,
+while bridge cells may function to stabilize or adjust this rhythmic
+output in accordance with the real-time state of the cilia. In this way,
+bridge cells could provide feedback about the execution target---i.e.,
+the balancer cilia---back into the coordination circuit, serving as a
+core element of a self-regulating rhythm control system.
+
+This architecture can be interpreted as a recurrent information loop, in
+which output from the ANN circuit modulates ciliary activity, and the
+resulting ciliary state is in turn relayed back to the ANN via bridge
+cells. Such a structure suggests a loop-based coordination mechanism
+(Arshavsky, 2003; Selverston, 2010), where the central circuit adapts
+its timing based on feedback from its target. Although the
+electrophysiological properties of bridge cells remain to be elucidated,
+their recursive integration into the ANN circuit indicates that they may
+play a central role in maintaining the robustness and flexibility of
+endogenous ciliary rhythms (Kennedy and Weissbourd, 2024).
+
+\subsubsection{Conclusion}\label{conclusion}
+
+How do differences in balancer cilia coordination across anatomical
+planes shape behavior? In \emph{Mnemiopsis}, the animal swims upright
+with its mouth facing upward, and performs 90-degree rotational turns
+when hunting prey or responding to disturbances (Colin et al., 2010;
+Courtney et al., 2020; Waggett, 1999). Large rotational maneuvers may
+involve balancer cells in the tentacular plane, while sagittal-plane
+balancers may contribute to postural stabilization (Sidney L. Tamm,
+2015b)---but these remain hypotheses.
+
+Our findings reveal a circuit composed of ANN neurons and bridge cells
+that coordinates ciliary beating rhythms in space and time without
+sensory input, and features a recursive architecture in which ciliary
+outputs are relayed back into the circuit via bridge cells. Such a
+configuration suggests that even with a limited number of cells, the
+system is capable of flexible and robust rhythm control.
+
+As Jagar et al.~(2011) once described, this organ evokes the metaphor of
+a ``ciliary brain'' (Jager et al., 2010). The network we describe here
+similarly orchestrates output through dynamic, self-regulating
+coordination. These dynamics, visualized through live imaging of the
+entire cell ensemble, offer a window into the core functions of a
+nervous system---not in complexity, but in principle.
 
@@
 
+\subsubsection{Specimen Preparation, Volume Electron Microscopy, and
+Image
+Processing}\label{specimen-preparation-volume-electron-microscopy-and-image-processing}
+
+Larvae of \emph{Mnemiopsis leidyi} (five days old) were cryofixed using
+a high-pressure freezing apparatus (HPM Live μ, CryoCapCell,
+https://www.cryocapcell.com/technologies/high-pressure-freezer) and
+immediately transferred to liquid nitrogen for storage. The frozen
+samples were processed in a substitution medium containing 2\% (w/v)
+osmium tetroxide and 0.5\% uranyl acetate in acetone, using a
+cryo-substitution device (EM AFS-2, Leica). Cryo-substitution was
+carried out by gradually raising the temperature from -90°C to -70°C
+over 4 hours, then returning to -20°C over 2 hours, and finally to room
+temperature over 2 hours. The samples were then embedded in epoxy resin
+(EMbed 812, Electron Microscopy Sciences). Serial ultrathin sections of
+50 nm thickness were prepared using a Leica UC7 ultramicrotome and a 45º
+DiATOME diamond knife. To enhance section adhesion and improve
+hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO
+Glass, UQG Optics) was treated with air glow discharge using the PELCO
+easiGlow system (Ted Pella, Inc.), rendering the surface negatively
+charged. Section ribbons were collected on prepared glass slides, gently
+dried to ensure proper stretching, and firmly adhered to the glass
+surface. The sections were stained with UranyLess and lead citrate
+(Reynolds) using airless staining procedures. The glass slides were
+mounted on a STEM-specific stage (Zeiss) using Copper Foil EMI Shielding
+Tape (3M). Imaging was performed using a Gemini SEM 500 (Zeiss),
+equipped with SmartSEM and Atlas 5 imaging software (Zeiss), using the
+Inlens detector. A total of 620 serial sections, each 50 nm thick, were
+analyzed. The imaging resolution was 2.8 nm/pixel, and the acceleration
+voltage was set to 1.5 kV. The imaging time for each tile was 54 minutes
+and 21 seconds, with a pixel size of 2.8 nm, a tile size of 91.8 × 91.8
+µm (32768 × 32768 pixels), and a dwell time of 3.0 µs.
+
+\subsubsection{Image-Stack Alignment and Export for
+CATMAID}\label{image-stack-alignment-and-export-for-catmaid}
+
+To process the image stack, we utilized the TrakEM2 plugin of FIJI
+(ImageJ) (version 2.0.0-rc-15/1.49k / Java 1.6.0\_24 (64-bit) -- 2014).
+A project was created, and all TIFF images were imported using the
+``import sequence as grid'' function. Subsequently, the following
+filters were applied sequentially: Invert, Equalize Histogram, and
+Gaussian Blur.
+
+The alignment process consisted of three stages, each progressively
+refining the spatial accuracy (rigid, affine, elastic).
+
+Initially, a rigid alignment was run with the following parameters:
+least squares mode (linear feature correspondences), encompassing the
+entire layer range with the first layer as the reference. Only visible
+images were used, without propagation. The alignment was executed with
+an initial Gaussian blur of 1.6 pixels, three steps per scale octave, a
+minimum image size of 512 pixels, and a maximum of 2048 pixels.
+Additional parameters included a feature descriptor size of 8,
+orientation bins of 8, and a closest ratio of 0.92. The alignment
+allowed clearing the cache, using 32 feature-extraction threads, a
+maximal alignment error of 100 pixels, a minimal inlier ratio of 0.20,
+and a minimum of 12 inliers. The expected and desired transformations
+were set to rigid, with testing multiple hypotheses (tolerance: 5.00
+pixels) and considering up to 5 neighboring layers, giving up after 5
+failures. Regularization was done with a maximal iteration of 1000, a
+maximal plateau width of 200, and a rigid lambda of 0.10.
+
+Next we applied an affine alignment step with similar parameters, except
+the expected and desired transformations were set to affine. The minimal
+image size was reduced to 64 pixels, while the other parameters
+(Gaussian blur, feature descriptor size, inliers, and testing
+hypotheses) remained unchanged to ensure consistent processing.
+
+Finally, we ran two iterations of elastic alignment to fine-tune the
+spatial data. Key parameters included a block-matching layer scale of
+0.05, a search radius of 200 pixels, a block radius of 2000 pixels, and
+a resolution of 60. Correlation filters were employed with a minimal
+PMCC r of 0.10, a maximal curvature ratio of 1000, and a maximal
+second-best r/best r of 0.90. A local smoothness filter was applied with
+the approximate local transformation set to affine, a local region sigma
+of 1000 pixels, and an absolute maximal local displacement of 10 pixels
+(relative maximal displacement: 3.00). Pre-aligned layers were tested
+for up to 4 neighboring layers. The elastic alignment used a rigid
+approximation, maximal iterations of 3000, a plateau width of 200,
+spring mesh stiffness of 0.01, and a maximal stretch of 2000 pixels. A
+legacy optimizer was employed to enhance performance.
+
+After each alignment stage, the project was saved as an XML file under a
+unique name to preserve iterative progress. Finally, the images were
+exported from FIJI using TrakEM2 in a format compatible with CATMAID.
+
+\subsubsection{Neuron Tracing, Synapse Annotation, and
+Review}\label{neuron-tracing-synapse-annotation-and-review}
+
+For skeletonisation, annotation and tagging we used CATMAID installed on
+a local server (Saalfeld et al., 2009; Schneider-Mizell et al., 2016).
+To mark the locations of cell bodies, we placed tags at the approximate
+center of each nucleus within the dataset. At each nuclear center, the
+radius of the single node was adjusted according to the size of the cell
+body in that specific layer. All skeletons were rooted at their
+respective cell bodies, and the root nodes were tagged as ``soma.''
+Synapses were identified based on four key structural features: the cell
+membrane, synaptic vesicles, the endoplasmic reticulum, and
+mitochondria. Most synapses could be verified across consecutive
+sections, ensuring accurate annotation and connectivity mapping.
+
+\subsubsection{Cell-type Nomenclature, Annotations and Connectome
+Analysis}\label{cell-type-nomenclature-annotations-and-connectome-analysis}
+
+We assigned each cell a specific cell-type name based on its category
+(balancer, bridge, bristle, ciliated grooves, dense vesicle cell, dome,
+epithelial floor (ef), intracellular multiciliated cells (imc),
+lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a
+total of 12 cell types. Some cells only had more general features and
+were classified based on the presence/absence and number of cilia into
+four ciliated cell types:(biciliated (biC), monociliated (monoC),
+multiciliated (multiC), non-ciliated (non-C)). Additionally, we appended
+the quadrant number to the cell type name after an underscore (``\_'')
+to indicate the cell-body's location. For example, cells located in the
+first quadrant are named with ``\_Q1.'' If a cell is situated between
+the first and second quadrants, we appended ``Q1Q2'' to the cell-type
+name.
+
+We further distinguished cells of the same type within the same region
+by serial numbering. This way each cell in the volume has a unique name
+string. Each cell was also assigned multiple annotations, which can be
+utilized to query the database via CATMAID or the CATMAID API (e.g.,
+using the R catmaid package (Bates et al., 2020)). These annotations
+provide a structured and precise framework for identifying and analyzing
+specific cells, facilitating robust data integration and retrieval from
+the dataset.
+
+\subsubsection{Imaging the Activity of Balancer
+Cilia}\label{imaging-the-activity-of-balancer-cilia}
+
+For ciliary imaging, we used the cydippid stage of \emph{M. leidyi} at
+five days post-fertilization. A coverslip with a thin layer of Vaseline
+applied to its two edges was gently placed over a glass slide. Filtered
+natural seawater and the cydippids were introduced beneath the
+coverslip. The orientation of the larvae was adjusted by carefully
+sliding the coverslip, and slight pressure was then applied to
+immobilize the larvae.
+
+Balancer cilia movement was imaged using a differential interference
+contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a
+high-sensitivity monochrome CMOS camera optimized for the near-infrared
+range (UI-3360CP-NIR-GL Rev.2, iDS), and a 40x glycerine-immersion
+objective lens (LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27). To
+stabilize the statolith, the microscope was tilted by 90° to achieve
+vertical stage alignment. Illumination was provided by a custom-made NIR
+LED strobe system (850 nm) synchronized with the camera's exposure
+signals. Image acquisition was controlled using BohNavi software.
+Recordings were acquired at 640 × 480 pixels, 100 frames per second
+(fps), for a duration of 2 minutes, using 0.05 ms pulse illumination
+from the LED.
+
+For beat frequency analysis, kymographs were generated using the Multi
+Kymograph plugin in Fiji. One-second segments were manually extracted
+from the kymographs, and ciliary beat frequencies were counted visually
+for each ROI. These manually determined frequencies were then used for
+subsequent computational analysis.
+
+To process the beat frequency time series data, left and right cilia
+traces were extracted, filtered to remove missing or non-finite values,
+and assigned frame indices. Data were saved as RDS objects for
+downstream analysis. Mean values were plotted as bar graphs with the
+left and right cilia shown in positive and negative directions,
+respectively.
+
+To assess coordination between left and right cilia, we computed rolling
+Pearson correlation coefficients over a window of 20 frames. For each
+dataset, we excluded windows with insufficient valid data or no
+variation. Time series of correlation coefficients were plotted and
+summarized as mean values. These mean correlations were then compared
+across planes (sagittal vs.~tentacular) using Welch's t-test. The
+resulting statistics were visualized as boxplots with jittered data
+points.
+
+\subsubsection{Data and Code
+Availability}\label{data-and-code-availability}
+
+The EM image stacks, including all traces and annotations, are available
+at https://catmaid.jekelylab.ex.ac.uk. The dataset encompasses all EM
+images (in JPG format), skeletons, meshes, node tags, connectors, and
+annotations. Additionally, we provide all R scripts used for data
+acquisition and figure generation (Jokura et al., 2025). All plots,
+figures (including anatomical renderings), and figure layouts can be
+fully reproduced using the provided R scripts. While the scripts are
+mostly organized by figure, some general-purpose scripts---for tasks
+such as loading libraries, accessing CATMAID data, and defining common
+functions---are shared across multiple figures.
+
 \subsection{Acknowledgements}\label{acknowledgements}
 
-We would like to thank the Jekely lab for the R project template
-(\url{https://github.com/JekelyLab/new_paper_template}) we used to write
-this paper. This work was funded by \ldots{} JSPS\ldots{} ERC.. Others
-\ldots{}
+This work was supported by the Japan Society for the Promotion of
+Science (JSPS) Overseas Research Fellowships, the Grass Foundation, the
+Kavli Foundation, and the Marine Biological Laboratory (MBL). This
+project also received funding from the European Research Council (ERC)
+under the European Union's Horizon 2020 research and innovation
+programme (grant agreement No.~101020792). We thank Dr.~Réza Shahidi for
+sectioning and imaging, and Paulina Cherek for high-pressure freezing
+and sample preparation. We are also grateful to Dr.~Chris Bjornsson, the
+MBL Central Microscopy Facility, Dr.~Shoji A. Baba, and Dr.~Kogiku Shiba
+(Shimoda Marine Research Center) for their assistance with microscopy.
 
-\subsection{References}\label{references}
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
 
@@
 \begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-arendt2021}
+Arendt D. 2021. Elementary nervous systems. \emph{Philosophical
+Transactions of the Royal Society B: Biological Sciences}
+\textbf{376}:20200347.
+doi:\href{https://doi.org/10.1098/rstb.2020.0347}{10.1098/rstb.2020.0347}
+
 \bibitem[\citeproctext]{ref-aronova1974electron}
@@
 
+\bibitem[\citeproctext]{ref-arshavsky2003}
+Arshavsky Y. 2003. Cellular and network properties in the functioning of
+the nervous system: from central pattern generators to cognition.
+\emph{Brain Research Reviews} \textbf{41}:229--267.
+doi:\href{https://doi.org/10.1016/s0165-0173(02)00249-7}{10.1016/s0165-0173(02)00249-7}
+
+\bibitem[\citeproctext]{ref-Bates2020}
+Bates AS, Manton JD, Jagannathan SR, Costa M, Schlegel P, Rohlfing T,
+Jefferis GS. 2020. The natverse, a versatile toolbox for combining and
+analysing neuroanatomical data. \emph{eLife} \textbf{9}.
+doi:\href{https://doi.org/10.7554/elife.53350}{10.7554/elife.53350}
+
+\bibitem[\citeproctext]{ref-Burkhardt2023}
+Burkhardt P, Colgren J, Medhus A, Digel L, Naumann B, Soto-Angel JJ,
+Nordmann E-L, Sachkova MY, Kittelmann M. 2023. Syncytial nerve net in a
+ctenophore adds insights on the evolution of nervous systems.
+\emph{Science} \textbf{380}:293--297.
+doi:\href{https://doi.org/10.1126/science.ade5645}{10.1126/science.ade5645}
+
+\bibitem[\citeproctext]{ref-calderuxf3n2024}
+Calderón LAB, Shahidi R, Jékely G. 2024.
+\href{http://dx.doi.org/10.7554/eLife.94306.2}{Mechanism of barotaxis in
+marine zooplankton}.
+
+\bibitem[\citeproctext]{ref-cardona2012}
+Cardona A, Saalfeld S, Schindelin J, Arganda-Carreras I, Preibisch S,
+Longair M, Tomancak P, Hartenstein V, Douglas RJ. 2012. TrakEM2 Software
+for Neural Circuit Reconstruction. \emph{PLoS ONE} \textbf{7}:e38011.
+doi:\href{https://doi.org/10.1371/journal.pone.0038011}{10.1371/journal.pone.0038011}
+
+\bibitem[\citeproctext]{ref-colin2010}
+Colin SP, Costello JH, Hansson LJ, Titelman J, Dabiri JO. 2010. Stealth
+predation and the predatory success of the invasive ctenophore
+{\emph{Mnemiopsis leidyi}}. \emph{Proceedings of the National Academy of
+Sciences} \textbf{107}:17223--17227.
+doi:\href{https://doi.org/10.1073/pnas.1003170107}{10.1073/pnas.1003170107}
+
+\bibitem[\citeproctext]{ref-courtney2020}
+Courtney A, Merces GOT, Pickering M. 2020.
+\href{http://dx.doi.org/10.1101/2020.05.25.114744}{Characterising the
+behaviour of the ctenophore{\emph{Pleurobrachia pileus}}in a laboratory
+aquaculture system}.
+
+\bibitem[\citeproctext]{ref-Hayakawa2022}
+Hayakawa E, Guzman C, Horiguchi O, Kawano C, Shiraishi A, Mohri K, Lin
+M-F, Nakamura R, Nakamura R, Kawai E, Komoto S, Jokura K, Shiba K,
+Shigenobu S, Satake H, Inaba K, Watanabe H. 2022. Mass spectrometry of
+short peptides reveals common features of metazoan peptidergic neurons.
+\emph{Nature Ecology \& Evolution} \textbf{6}:1438--1448.
+doi:\href{https://doi.org/10.1038/s41559-022-01835-7}{10.1038/s41559-022-01835-7}
+
 \bibitem[\citeproctext]{ref-Hernandez_Nicaise_1984}
@@
 doi:\href{https://doi.org/10.1007/978-3-642-51593-4_9}{10.1007/978-3-642-51593-4\_9}
+
+\bibitem[\citeproctext]{ref-Hernandez-Nicaise1974}
+Hernandez-Nicaise M-L. 1974. Ultrastructural evidence for a
+sensory-motor neuron in Ctenophora. \emph{Tissue and Cell}
+\textbf{6}:43--47.
+doi:\href{https://doi.org/10.1016/0040-8166(74)90021-4}{10.1016/0040-8166(74)90021-4}
 
@@
 
+\bibitem[\citeproctext]{ref-Hernandez-Nicaise1968}
+HERNANDEZ-NICAISE M-L. 1968. Specialized Connexions between Nerve Cells
+and Mesenchymal Cells in Ctenophores. \emph{Nature}
+\textbf{217}:1075--1076.
+doi:\href{https://doi.org/10.1038/2171075a0}{10.1038/2171075a0}
+
+\bibitem[\citeproctext]{ref-Hertwig1880}
+Hertwig R. 1880. Ueber den bau der ctenophoren. \emph{Jenaische
+Zeitschrift f{ü}r Naturwissenschaft} \textbf{14}:393.
+
+\bibitem[\citeproctext]{ref-HORRIDGE1965}
+HORRIDGE GA. 1965. RELATIONS BETWEEN NERVES AND CILIA IN CTENOPHORES.
+\emph{American Zoologist} \textbf{5}:357--375.
+doi:\href{https://doi.org/10.1093/icb/5.3.357}{10.1093/icb/5.3.357}
+
 \bibitem[\citeproctext]{ref-Horridge_1964}
@@
 
-\bibitem[\citeproctext]{ref-Noda_2014}
-Noda N, Tamm SL. 2014. Lithocytes are transported along the ciliary
-surface to build the statolith of ctenophores. \emph{Current Biology}
-\textbf{24}:R951--R952.
-doi:\href{https://doi.org/10.1016/j.cub.2014.08.045}{10.1016/j.cub.2014.08.045}
+\bibitem[\citeproctext]{ref-Jager2011}
+Jager M, Chiori R, Alié A, Dayraud C, Quéinnec E, Manuel M. 2010. New
+insights on ctenophore neural anatomy: Immunofluorescence study in
+{\emph{Pleurobrachia pileus}} (Müller, 1776). \emph{Journal of
+Experimental Zoology Part B: Molecular and Developmental Evolution}
+\textbf{316B}:171--187.
+doi:\href{https://doi.org/10.1002/jez.b.21386}{10.1002/jez.b.21386}
 
-\bibitem[\citeproctext]{ref-Ortiz_2023}
-Ortiz J, Bobkov YV, DeBiasse MB, Mitchell DG, Edgar A, Martindale MQ,
-Moss AG, Babonis LS, Ryan JF. 2023. Independent innexin radiation shaped
-signaling in ctenophores. \emph{Molecular Biology and Evolution}
-\textbf{40}.
-doi:\href{https://doi.org/10.1093/molbev/msad025}{10.1093/molbev/msad025}
+\bibitem[\citeproctext]{ref-juxe9kely2010}
+Jékely G. 2010. Origin and early evolution of neural circuits for the
+control of ciliary locomotion. \emph{Proceedings of the Royal Society B:
+Biological Sciences} \textbf{278}:914--922.
+doi:\href{https://doi.org/10.1098/rspb.2010.2027}{10.1098/rspb.2010.2027}
+
+\bibitem[\citeproctext]{ref-jurisch-yaksi2024}
+Jurisch-Yaksi N, Wachten D, Gopalakrishnan J. 2024. The neuronal cilium
+{\textendash} a highly diverse and dynamic organelle involved in sensory
+detection and neuromodulation. \emph{Trends in Neurosciences}
+\textbf{47}:383--394.
+doi:\href{https://doi.org/10.1016/j.tins.2024.03.004}{10.1016/j.tins.2024.03.004}
+
+\bibitem[\citeproctext]{ref-keijzer2013}
+Keijzer F, Duijn M van, Lyon P. 2013. What nervous systems do: early
+evolution, input{\textendash}output, and the skin brain thesis.
+\emph{Adaptive Behavior} \textbf{21}:67--85.
+doi:\href{https://doi.org/10.1177/1059712312465330}{10.1177/1059712312465330}
+
+\bibitem[\citeproctext]{ref-kennedy2024}
+Kennedy A, Weissbourd B. 2024. Dynamics of neural activity in early
+nervous system evolution. \emph{Current Opinion in Behavioral Sciences}
+\textbf{59}:101437.
+doi:\href{https://doi.org/10.1016/j.cobeha.2024.101437}{10.1016/j.cobeha.2024.101437}
+
+\bibitem[\citeproctext]{ref-Li2021}
+Li Y, Shen X-X, Evans B, Dunn CW, Rokas A. 2021. Rooting the Animal Tree
+of Life. \emph{Molecular Biology and Evolution} \textbf{38}:4322--4333.
+doi:\href{https://doi.org/10.1093/molbev/msab170}{10.1093/molbev/msab170}
+
+\bibitem[\citeproctext]{ref-Lowe1997}
+Lowe B. 1997. The role of Ca2+ in deflection-induced excitation of
+motile, mechanoresponsive balancer cilia in the ctenophore statocyst.
+\emph{Journal of experimental biology} \textbf{200}:1593--1606.
+
+\bibitem[\citeproctext]{ref-Martindale1999}
+Martindale MQ, Henry JQ. 1999. Intracellular Fate Mapping in a Basal
+Metazoan, the Ctenophore Mnemiopsis leidyi, Reveals the Origins of
+Mesoderm and the Existence of Indeterminate Cell Lineages.
+\emph{Developmental Biology} \textbf{214}:243--257.
+doi:\href{https://doi.org/10.1006/dbio.1999.9427}{10.1006/dbio.1999.9427}
+
+\bibitem[\citeproctext]{ref-moroz2015}
+Moroz LL. 2015. Convergent evolution of neural systems in ctenophores.
+\emph{Journal of Experimental Biology} \textbf{218}:598--611.
+doi:\href{https://doi.org/10.1242/jeb.110692}{10.1242/jeb.110692}
+
+\bibitem[\citeproctext]{ref-Moroz2014}
+Moroz LL, Kocot KM, Citarella MR, Dosung S, Norekian TP, Povolotskaya
+IS, Grigorenko AP, Dailey C, Berezikov E, Buckley KM, Ptitsyn A,
+Reshetov D, Mukherjee K, Moroz TP, Bobkova Y, Yu F, Kapitonov VV, Jurka
+J, Bobkov YV, Swore JJ, Girardo DO, Fodor A, Gusev F, Sanford R, Bruders
+R, Kittler E, Mills CE, Rast JP, Derelle R, Solovyev VV, Kondrashov FA,
+Swalla BJ, Sweedler JV, Rogaev EI, Halanych KM, Kohn AB. 2014. The
+ctenophore genome and the evolutionary origins of neural systems.
+\emph{Nature} \textbf{510}:109--114.
+doi:\href{https://doi.org/10.1038/nature13400}{10.1038/nature13400}
+
+\bibitem[\citeproctext]{ref-roberts2022}
+Roberts RJV, Pop S, Prieto-Godino LL. 2022. Evolution of central neural
+circuits: state of the art and perspectives. \emph{Nature Reviews
+Neuroscience} \textbf{23}:725--743.
+doi:\href{https://doi.org/10.1038/s41583-022-00644-y}{10.1038/s41583-022-00644-y}
+
+\bibitem[\citeproctext]{ref-Ryan2013}
+Ryan JF, Pang K, Schnitzler CE, Nguyen A-D, Moreland RT, Simmons DK,
+Koch BJ, Francis WR, Havlak P, Smith SA, Putnam NH, Haddock SHD, Dunn
+CW, Wolfsberg TG, Mullikin JC, Martindale MQ, Baxevanis AD. 2013. The
+Genome of the Ctenophore {\emph{Mnemiopsis leidyi}} and Its Implications
+for Cell Type Evolution. \emph{Science} \textbf{342}.
+doi:\href{https://doi.org/10.1126/science.1242592}{10.1126/science.1242592}
+
+\bibitem[\citeproctext]{ref-Saalfeld2009}
+Saalfeld S, Cardona A, Hartenstein V, Tomančák P. 2009. CATMAID:
+collaborative annotation toolkit for massive amounts of image data.
+\emph{Bioinformatics} \textbf{25}:1984--1986.
+doi:\href{https://doi.org/10.1093/bioinformatics/btp266}{10.1093/bioinformatics/btp266}
+
+\bibitem[\citeproctext]{ref-Sachkova2021}
+Sachkova MY, Nordmann E-L, Soto-Àngel JJ, Meeda Y, Górski B, Naumann B,
+Dondorp D, Chatzigeorgiou M, Kittelmann M, Burkhardt P. 2021.
+Neuropeptide repertoire and 3D anatomy of the ctenophore nervous system.
+\emph{Current Biology} \textbf{31}:5274--5285.e6.
+doi:\href{https://doi.org/10.1016/j.cub.2021.09.005}{10.1016/j.cub.2021.09.005}
+
+\bibitem[\citeproctext]{ref-Schneider-Mizell2016}
+Schneider-Mizell CM, Gerhard S, Longair M, Kazimiers T, Li F, Zwart MF,
+Champion A, Midgley FM, Fetter RD, Saalfeld S, Cardona A. 2016.
+Quantitative neuroanatomy for connectomics in Drosophila. \emph{eLife}
+\textbf{5}.
+doi:\href{https://doi.org/10.7554/elife.12059}{10.7554/elife.12059}
+
+\bibitem[\citeproctext]{ref-Schultz2023}
+Schultz DT, Haddock SHD, Bredeson JV, Green RE, Simakov O, Rokhsar DS.
+2023. Ancient gene linkages support ctenophores as sister to other
+animals. \emph{Nature} \textbf{618}:110--117.
+doi:\href{https://doi.org/10.1038/s41586-023-05936-6}{10.1038/s41586-023-05936-6}
+
+\bibitem[\citeproctext]{ref-Sebe-Pedros2018}
+Sebé-Pedrós A, Chomsky E, Pang K, Lara-Astiaso D, Gaiti F, Mukamel Z,
+Amit I, Hejnol A, Degnan BM, Tanay A. 2018. Early metazoan cell type
+diversity and the evolution of multicellular gene regulation.
+\emph{Nature Ecology \& Evolution} \textbf{2}:1176--1188.
+doi:\href{https://doi.org/10.1038/s41559-018-0575-6}{10.1038/s41559-018-0575-6}
+
+\bibitem[\citeproctext]{ref-selverston2010}
+Selverston AI. 2010. Invertebrate central pattern generator circuits.
+\emph{Philosophical Transactions of the Royal Society B: Biological
+Sciences} \textbf{365}:2329--2345.
+doi:\href{https://doi.org/10.1098/rstb.2009.0270}{10.1098/rstb.2009.0270}
+
+\bibitem[\citeproctext]{ref-sheu2022}
+Sheu S-H, Upadhyayula S, Dupuy V, Pang S, Deng F, Wan J, Walpita D,
+Pasolli HA, Houser J, Sanchez-Martinez S, Brauchi SE, Banala S, Freeman
+M, Xu CS, Kirchhausen T, Hess HF, Lavis L, Li Y, Chaumont-Dubel S,
+Clapham DE. 2022. A serotonergic axon-cilium synapse drives nuclear
+signaling to alter chromatin accessibility. \emph{Cell}
+\textbf{185}:3390--3407.e18.
+doi:\href{https://doi.org/10.1016/j.cell.2022.07.026}{10.1016/j.cell.2022.07.026}
+
+\bibitem[\citeproctext]{ref-Tamm1982}
+Tamm S. 1982. Electrical conduction and behaviour in" simple"
+invertebrates. \emph{(No Title)}.
+
+\bibitem[\citeproctext]{ref-Tamm1980}
+Tamm S. 1980. Cilia and ctenophores. \emph{Oceanus} \textbf{23}:50--59.
+
+\bibitem[\citeproctext]{ref-Tamm2015}
+Tamm Sidney L. 2015a. Functional consequences of the asymmetric
+architecture of the ctenophore statocyst. \emph{The Biological Bulletin}
+\textbf{229}:173--184.
+
+\bibitem[\citeproctext]{ref-tamm2015}
+Tamm Sidney L. 2015b. Functional Consequences of the Asymmetric
+Architecture of the Ctenophore Statocyst. \emph{The Biological Bulletin}
+\textbf{229}:173--184.
+doi:\href{https://doi.org/10.1086/bblv229n2p173}{10.1086/bblv229n2p173}
 
 \bibitem[\citeproctext]{ref-Tamm_2014}
-Tamm SL. 2014a. Formation of the statolith in the ctenophore mnemiopsis
+Tamm SL. 2014b. Formation of the statolith in the ctenophore mnemiopsis
 leidyi. \emph{The Biological Bulletin} \textbf{227}:7--18.
@@
 \bibitem[\citeproctext]{ref-Tamm_2014b}
-Tamm SL. 2014b. Cilia and the life of ctenophores. \emph{Invertebrate
+Tamm SL. 2014a. Cilia and the life of ctenophores. \emph{Invertebrate
 Biology} \textbf{133}:1--46.
@@
 
+\bibitem[\citeproctext]{ref-Veraszto2017}
+Verasztó C, Ueda N, Bezares-Calderón LA, Panzera A, Williams EA, Shahidi
+R, Jékely G. 2017. Ciliomotor circuitry underlying whole-body
+coordination of ciliary activity in the Platynereis larva. \emph{eLife}
+\textbf{6}.
+doi:\href{https://doi.org/10.7554/elife.26000}{10.7554/elife.26000}
+
+\bibitem[\citeproctext]{ref-waggett1999}
+Waggett R. 1999. Capture mechanisms used by the lobate ctenophore,
+mnemiopsis leidyi, preying on the copepod acartia tonsa. \emph{Journal
+of Plankton Research} \textbf{21}:2037--2052.
+doi:\href{https://doi.org/10.1093/plankt/21.11.2037}{10.1093/plankt/21.11.2037}
+
+\bibitem[\citeproctext]{ref-Whelan2015a}
+Whelan NV, Kocot KM, Moroz LL, Halanych KM. 2015. Error, signal, and the
+placement of Ctenophora sister to all other animals. \emph{Proceedings
+of the National Academy of Sciences} \textbf{112}:5773--5778.
+doi:\href{https://doi.org/10.1073/pnas.1503453112}{10.1073/pnas.1503453112}
+
+\bibitem[\citeproctext]{ref-dewiljes2015}
+Wiljes OO de, Elburg RAJ van, Biehl M, Keijzer FA. 2015. Modeling
+spontaneous activity across an excitable epithelium: Support for a
+coordination scenario of early neural evolution. \emph{Frontiers in
+Computational Neuroscience} \textbf{9}.
+doi:\href{https://doi.org/10.3389/fncom.2015.00110}{10.3389/fncom.2015.00110}
+
 \end{CSLReferences}
 
+\newpage
+
+\subsection*{Supplementary material}\label{supplementary-material}
+\addcontentsline{toc}{subsection}{Supplementary material}
+
+\begin{figure}[H]
+
+{\centering \includegraphics{figures/Figure1_Supplement1.png}
+
+}
+
+\caption{\textbf{Figure 1---figure supplement 1}\\
+(A)\ldots{}}
+
+\end{figure}%%
+\begin{figure}[H]
+
+{\centering \includegraphics{figures/Figure2_Supplement1.png}
+
+}
+
+\caption{\textbf{Figure 2---figure supplement 1}\\
+(A)\ldots{}}
+
+\end{figure}%
+
 \end{document}

### 2025-05-16 manuscript/Manuscript.qmd
MS, refs: update

@@
 subtitle: |
-  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewalow**$^{6}$, **Pawel Burkhardt**$^{7}$, **Gáspár Jékely**$^{1,2,6,*}$  
+  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,7}$, **Lara Kewalow**$^{7}$, **Pawel Burkhardt**$^{6}$, **Gáspár Jékely**$^{1,2,7,*}$  
   <br><br>  
@@
   <br> 
-  $^{6}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  
+  $^{6}$ Michael Sars Centre, University of Bergen, Norway
   <br>
-  $^{7}$ Michael Sars Centre, University of Bergen, Norway
+  $^{7}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  
   <br><br>  
-  **$^{*}$ Correspondence:**  
-  [jokura@nibb.ac.jp](mailto:jokura@nibb.ac.jp),  
-  [gaspar.jekely@cos.uni-heidelberg.de](mailto:gaspar.jekely@cos.uni-heidelberg.de)
+  **$^{*}$ Correspondence** <br>   
+  ![](pictures/emails.png){height=20}<br>  
 mainfont: Arial
@@
 High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation.
-Our results suggest a coordinating rather than effector function for the nerve net and inform our understanding of the diversity of nervous-systems organisation across animals.
+Our results suggest a coordinating rather than sensory-motor function for the nerve net and inform our understanding of the diversity of nervous-systems organisation across animals.
 
@@
 Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioral patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system.
-Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011].
+Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise1973a; @Jager2011].
 These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022].
@@
 This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@HORRIDGE1965; @Tamm1980].
-Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez-Nicaise1968; @Jager2011].
+Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011].
 A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency.
-Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @Tamm_2014b].
+Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2015].
 Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997].
@@
 
-Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999].
+Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F) [@Martindale1999].
 The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999].
@@
 Classical neural staining techniques did not provide clear images of neurons at the aboral organ.
-However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise_1973].
+However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a].
 In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion.
 At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Supplementary Figure S2A).
-We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise_1973].
+We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise1973b].
 Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Supplementary Figure S2B).
@@
 Some balancer cells received inputs from both ANN_Q1-4 and either of ANN_Q1Q2 or ANN_Q3Q4.
-While previous studies suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez-Nicaise1974], our data did not reveal any presynaptic sites in balancer cells.
+While previous studies suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez_Nicaise1974], our data did not reveal any presynaptic sites in balancer cells.
 
@@
 Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination.
-Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014; @Tamm2015].
+Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@tamm2014b; @tamm2015].
 To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector.
@@
 Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs.
-inhibitory) of these neurons remains unresolved [@Ryan2013;@Moroz2014].
+inhibitory) of these neurons remains unresolved [@Ryan2013; @moroz2015].
 However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.
@@
 For instance, in *Platynereis dumerilii* larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands [@Veraszto2017], while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields [@Veraszto2017; @calderón2024].
-Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015; @roberts2022].
+Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015;@roberts2022].
 
@@
 Originally described morphologically by Tamm et al.
-(2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @Tamm_2014b].
+(2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @Tamm_2014c].
 Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024].
@@
 How do differences in balancer cilia coordination across anatomical planes shape behavior?
-In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances [@waggett1999; @colin2010; @courtney2020].
+In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances \[[@waggett1999; @colin2010; @courtney2020]\].
 Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.
@@
 
-As Jagar et al.
+As Jager et al.
 (2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011].

### 2025-05-16 manuscript/Manuscript.qmd
MS: fix ref fields

--- 2025-05-06 manuscript/Manuscript.qmd
@@
   <br><br>  
-  **$^{*}$ Correspondence:**  
-  [jokura@nibb.ac.jp](mailto:jokura@nibb.ac.jp),  
-  [gaspar.jekely@cos.uni-heidelberg.de](mailto:gaspar.jekely@cos.uni-heidelberg.de)
+    **$^{*}$ Correspondence:**  ![](pictures/emails.png){width=30em}
 mainfont: Arial
@@
 This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@HORRIDGE1965; @Tamm1980].
-Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez-Nicaise1968; @Jager2011].
+Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011].
 A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency.
-Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @Tamm_2014b].
+Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b].
 Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997].
@@
 Classical neural staining techniques did not provide clear images of neurons at the aboral organ.
-However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise_1973].
+However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a].
 In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion.
 At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Supplementary Figure S2A).
-We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise_1973].
+We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise1973a].
 Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Supplementary Figure S2B).
@@
 Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination.
-Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014; @Tamm2015].
+Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014c; @tamm2015].
 To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector.
@@
 Originally described morphologically by Tamm et al.
-(2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @Tamm_2014b].
+(2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @tamm2014b].
 Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024].

### 2025-05-16 manuscript/Manuscript.qmd
MS: edits, fig legends, refs

@@
 High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation.
-Our results suggest a coordinating rather than effector function for the nerve net and inform our understanding of the diversity of nervous-systems organisation across animals.
+Our results suggest a coordinating rather than sensory-motor function for the nerve net and inform our understanding of the diversity of nervous-system organisation across animals.
 
@@
 These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column.
-Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioral patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system.
+Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioural patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system.
 Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011].
 These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022].
-Notably, early single-cell transcriptomic studies failed to identify canonical neuronal clusters using conventional markers [@Sebe-Pedros2018].
-It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were subsequently identified [@Sachkova2021; @Hayakawa2022].
+Notably, early genomic and single-cell transcriptomic studies failed to identify conserved neuronal markers in ctenophores [@Moroz2014; @Sebe-Pedros2018].
+It was only through the use of pro-neuropeptide-based markers that distinct neural cell types were molecularly identified [@Sachkova2021; @Hayakawa2022].
 In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015].
 This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems.
-However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system gives rise to behavior.
-Clarifying this relationship between neural architecture and behavioral control is essential to uncovering the logic of ctenophore neural function.
+However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system controls behaviour.
+Clarifying this relationship between neural architecture and behavioural control is essential to uncovering the logic of ctenophore neural function.
 
-To address the lack of structural understanding of how ctenophores control behavior, we focused on a unique sensory structure—the aboral organ.
-This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for comb plate beating and posture control [@HORRIDGE1965; @Tamm1980].
+To address the lack of structural, circuit-level understanding of the ctenophore nervous system, we focused on a unique sensory structure—the aboral organ.
+This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for the beating of comb plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980].
 Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011].
 A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency.
-Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system not fixed like vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b].
+Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed one like in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b].
 Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997].
@@
 Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997].
-These findings support the hypothesis that neural modulation plays a critical role in shaping balancer cell output [@Lowe1997].
+These findings support the hypothesis that neural modulation plays a critical role in shaping balancer-cell output [@Lowe1997].
 
 To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy and high-speed imaging in the ctenophore *Mnemiopsis leidyi*.
-Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, and functional imaging revealed correlated activity patterns in balancer cilia.
-These results suggest that the aboral nerve net plays a central role in integrating sensory input and coordinating motor output.
+Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cell.
+By functional imaging, we uncovered correlated activity patterns in balancer cilia across the four quandrants of the apical organ.
+Our results suggest that the aboral nerve net functions in coordinating the activity of balancer cilia, rather than fulfilling a feedforward sensory-motor function.
 
@@
 We imaged only the region containing the aboral organ.
-Subsequently, 620 of these images were stitched and aligned in TrakEM2 [@cardona2012].
+Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012].
 The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm.
-We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009], resulting in a reconstruction of 909 cells with nuclei and cell bodies intact.
+We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009], resulting in a reconstruction of 909 cells that had intact nuclei and cell bodies.
 Most of the cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.
@@
 
-Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999].
+Next, we divided the entire aboral organ into four quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1G) [@Martindale1999].
 The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores [@Martindale1999].
@@
 We assigned each of the 909 skeletonised cells to one of the four quadrants.
+<!-- check number of cells, 927 in figure -->
 
@@
 Colors match the quadrant scheme shown in panels (B) and (C).\
-(F) Example of cell tracing using the collaborative annotation toolkit CATMAID for large-scale electron microscopy image datasets.
-The spherical objects indicate nuclear positions, while the lines represent the traced cell centers.
-Scale bar: 10 µm.\
-(G) Three-dimensional reconstruction of cells composing the aboral organ, displayed from different perspectives.
-The left panel presents an aboral view of the aboral organ.
-The middle panel shows a sagittal plane view, while the right panel provides a tentacular plane view.
-Cells are color-coded according to their respective quadrants.
-Lithocytes (li) are represented as three gray spheres, while balancers (bal) are depicted as gray lines.
+(F) Example of cell tracing in CATMAID. The spherical objects indicate nuclear positions, while the lines represent the traced cell centers. Scale bar: 10 µm.\
+(G) Morphological rendering of cells composing the aboral organ displayed in aboral view (left pannel), sagittal plane view (middle panel) and tentacular plane view (right panel).
+Cells are color-coded according to quadrants.
+Lithocytes (li) are represented as three gray spheres, balancers (bal) are depicted as gray lines.
 Scale bar: 25 µm.](figures/Figure1.png)
@@
 Based on their distinct size and position, we classified the ANNs into two types.
-The first type is a single large neuron (ANN_Q1-4) spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei.
+The first type is a single large neuron (ANN_Q1-4) that spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei.
 This neuron has 150 presynaptic sites.
@@
 ![**Figure 2. Organisation of the aboral synaptic nerve net**\
-(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (blue).
-The spheres represent the positions of individual nuclei.
-The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
-(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei.
-The spheres indicate the positions of the nuclei.
-The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
-(C) Graph showing the number of mitochondria per cell and the proportion of mitochondria involved in synapse formation for each cell type.
-Green represents mitochondria associated with synapses, while gray represents mitochondria not involved in synapse formation.
-ANN has a significantly higher number of mitochondria compared to other cell types, with most of them forming synapses.
-Except for lithocytes, each cell type has a sample size of at least three cells.
-The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
-(D) Localization of mitochondria within ANN Q1-4.
+(A) Morphological rendering of the large aboral nerve-net neuron ANN_Q1-4 spanning all four quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel).
+Spheres indicate the positions of nuclei.\
+(B) Morphological rendering of the two smaller ANN neurons ANN_Q1Q2 (pink) and ANN_Q3Q4 (orange), each spanning two quadrants, in aboral view (left), sagittal view (middle panel) and tentacular view (right panel).
+Spheres indicate the positions of nuclei.\
+(C) Number of mitochondria per cell for each cell type.
+The number of synapse-associated mitochondria is shown in green, the number of mitochondia outside synapses is shown in grey.
+N > 2 cells for each cell type except lithocytes.
+Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cells; imc, intra-multiciliated cells; la, lamellate bodies; li, lithocytes; pl, plumose; ef, epithelial floor cells.\
+(D) Position of mitochondria within ANN_Q1-4.
 Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
@@
 The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4).
-The abbreviations are as follows: mi (mitochondrion), er (endoplasmic reticulum), and sv (synaptic vesicles).\
-(F) Plot showing the three-dimensional positions of synapses.
+Abbreviations: mi, mitochondrion; er, endoplasmic reticulum; sv, synaptic vesicles.\
+(F) Position of synapses.
 Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan.
-Blue dots indicate the locations of autapses within ANN Q1-4.](figures/Figure2.png)
+Blue dots indicate the locations of autapses within ANN_Q1-4.](figures/Figure2.png)
 
@@
 
-The second group of cells that form synaptic contact with the ANN were the bridge cells.
-Bridge cells, first described by Tamm et al.
-in 2002 [@Tamm_2002], are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge.
+The second group of cells that formed synaptic contacts with the ANN were the bridge cells.
+Bridge cells, first described in 2002 [@Tamm_2002], are characterized by bundles of elongated processes filled with microtubules that arch over the epithelial layer, resembling a bridge.
 Their somata are located at the base of the paired balancer-cell clusters along the tentacle surface and extend across the sagittal plane towards the base of the opposite balancer cells.
@@
 
-In both The Q1Q2 and the Q3Q4 regions, bridge cells also formed synapses with adjacent bridge cells. However, we found no synapses across the sagittal plane to bridge cells in the opposite region.
+In both the Q1Q2 and the Q3Q4 regions, bridge cells also formed synapses with adjacent bridge cells. However, we found no synapses across the sagittal plane to bridge cells in the opposite region.
 
@@
 
-![**Figure 3. Reconstruction of the balancers and bridges that receives synapses from the ANN.**\
-(A) Graph showing the number of synaptic outputs from the ANNs (ANN Q1-4: blue, ANN Q1Q2: orange, ANN Q3Q4: pink) circuit for each cell type.
-The balancer and bridge cell types receive relatively high outputs from the ANNs circuit.
-ANN represents the number of synapses within the ANNs circuit.
-The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
-(B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink, and Q3Q4 in orange) and balancer ciliated cells (light blue).
+![**Figure 3. Synaptic connectivity of the balancer and and bridge cells with the ANN.**\
+(A) Number of synaptic inputs from the ANNs (ANN_Q1-4: blue, ANN_Q1Q2: orange, ANN_Q3Q4: pink) to each cell type.
+Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cell; imc, intra-multiciliated cell; la, lamellate body; li, lithocyte; pl, plumose cell; ef, epithelial floor cell.\
+<!-- check order of panels, change here or change figure -->
+(B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink, and Q3Q4 in orange) and balancer cells (light blue).
 The thickness of the arrows and the numbers correspond to the number of synapses.
 Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.\
-(C) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants.
+(C) Morphological rendering of bridge cells spanning the Q1Q2 and Q3Q4 quadrants.
 The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color.
@@
 
-In larvae with their sagittal plane facing the objective, we could compared balancer-cilia movements between Q1(4) and Q2(3).
+In larvae with their sagittal plane facing the objective, we could compare balancer-cilia movements between Q1(4) and Q2(3).
 In other larvae oriented in the tentacular plane, we could simultaneously image Q1(3) and Q2(4).
@@
 
-![**Figure 4. Differences in Balancer Cilia Movement Control Due to Variations in the Neural Circuit Pathways**\
-(A) Schematic diagram of differential interference contrast (DIC) microscopy setup for imaging balancer cilia movement.
+![**Figure 4. Analysis of ciliary beating, arrests and re-beat across balancers in the four quadrants**\
+(A) Schematic diagram of the differential interference contrast (DIC) microscopy setup used to image the movement of balancer cilia.
 The microscope was tilted 90 degrees so that the stage was positioned vertically.
-A monochrome CMOS camera sensitive to near-infrared (NIR) light was used, synchronized with an 850 nm strobe light source.
-The movement of the balancer cilia was recorded using a 40× objective lens.\
+We used a 40x objective lens and a monochrome CMOS camera sensitive to near-infrared (NIR) light that was synchronized with an 850 nm strobe light source.\
 (B) Representative DIC images of the aboral organ viewed along the sagittal (left) and tentacular (right) planes.
@@
 Arrowheads mark the onset of arrest (open) and re-beat (filled); horizontal arrows represent 1-second intervals.\
-(D) Boxplots showing time differences in the onset of arrest (white boxes) and re-beat (gray boxes) between left and right balancer cilia.
+(D) Boxplots showing time differences in the onset of arrest (white boxes) and re-beat (grey boxes) between left and right balancer cilia.
 Colors of the box outlines indicate the imaging plane: sagittal (blue) or tentacular (orange).
 Each dot represents a single larva.\
-(E) Bar plots showing ciliary beat frequency (CBF) dynamics of left and right balancers (top and bottom, respectively) recorded from sagittal (left) and tentacular (right) planes.
+(E) Bar plots showing ciliary beat frequency (CBF) dynamics of left and right balancers (top and bottom, respectively) recorded from the sagittal (left) or the tentacular (right) planes.
 Positive and negative values represent opposite sides within each plane.\
@@
 
-### Functional Nerve Net for Ciliary Coordination
+### An Aboral Nerve Net for Ciliary Coordination
 
-In this study, we generated the first high-resolution connectome of a ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of five-day-old *Mnemiopsis leidyi* cydippids using volumetric electron microscopy.
-Among these, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve nets (ANNs)—which exhibit multiple nuclei and a morphology distinct from the “beads-on-a-string” organization reported for the subepithelial nerve net (SNN) in one-day-old cydippids [@Burkhardt2023; @Sachkova2021].
+In this study, we generated the first high-resolution connectome of the ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of a five-day-old *Mnemiopsis leidyi* cydippid by volumetric electron microscopy.
+Among the reconstructed cells, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve net (ANNs)—with multiple nuclei and a morphology distinct from thesubepithelial nerve net (SNN) with a “beads-on-a-string” organization reported in one-day-old cydippids [@Burkhardt2023; @Sachkova2021].
 Instead of forming discrete nodal swellings, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion, suggesting a unique neuron subtype specific to the aboral organ.
@@
 inhibitory) of these neurons remains unresolved [@Ryan2013;@Moroz2014].
-However, the combination of circuit architecture and behavioral outputs provides the first functional insight into the role of neurons in a ctenophore.
+However, the combination of circuit architecture and behavioural outputs provides the first functional insight into the role of neurons in a ctenophore.
 
@@
 
-How do differences in balancer cilia coordination across anatomical planes shape behavior?
+How do differences in balancer cilia coordination across anatomical planes shape behaviour?
 In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances [@waggett1999; @colin2010; @courtney2020].

### 2025-05-22 manuscript/Manuscript.qmd
MS: discussion edits

@@
 In this study, we generated the first high-resolution connectome of the ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of a five-day-old *Mnemiopsis leidyi* cydippid by volumetric electron microscopy.
-Among the reconstructed cells, we identified a previously undescribed type of syncytial neuron—here referred to as aboral nerve net (ANNs)—with multiple nuclei and a morphology distinct from thesubepithelial nerve net (SNN) with a “beads-on-a-string” organization reported in one-day-old cydippids [@Burkhardt2023; @Sachkova2021].
-Instead of forming discrete nodal swellings, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion, suggesting a unique neuron subtype specific to the aboral organ.
+Among the reconstructed cells, we identified a previously undescribed type of syncytial neuron---here referred to as aboral nerve net (ANNs)---with multiple nuclei.
+The ANN has a morphology distinct from that of the subepithelial nerve net (SNN) reported for one-day-old cydippids [@Burkhardt2023; @Sachkova2021].
+Instead of forming discrete nodal swellings or a “beads-on-a-string” organization typical for the SNN, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion.
+The presence of two SNN-type neurons more laterally in the aboral organ in our dataset (Figure 2--figure supplement1) further underline the distinct identity of ANNs.
 
-Remarkably, ANNs form no synaptic output to classical effector organs such as muscles or comb plates.
+ANNs form no synaptic output to classical effector organs such as muscles or comb plates.
 All observed output synapses were directed to balancer ciliary cells.
@@
 These findings indicate that ANNs are unlikely to serve as motor drivers in the conventional sense.
-Rather, they may be involved in coordinating or modulating the timing and dynamics of ciliary activity [@keijzer2013; @dewiljes2015; @Veraszto2017].
+Rather, they may be involved in coordinating or modulating the timing and dynamics of the activity of the sensory-motile balancer cilia, possibly modulating their responsiveness to statocyst load [@keijzer2013; @dewiljes2015; @Veraszto2017].
 
-Notably, anatomical differences in ANN projection patterns across sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics.
-This correlation supports the idea that individual ANNs may play opposing or complementary roles in regulating ciliary beating—potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns.
+What could be the function of the aboral nerve net?
+We found that the anatomical differences in the innervation patterns of ANNs across the sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics.
+This suggests that individual ANNs may play opposing or complementary roles in regulating ciliary beating---potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns.
 Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs.

### 2025-05-22 analysis/scripts/Figure1.R
Fig1: fine tune spacing, panel legends

@@ -44,167 +44,49 @@ close3d()
 
 # assemble figure -------------------------------------------------------------
 
-# read pics
-#panel_balancer <- ggdraw() + draw_image(readPNG("manuscript/pictures/balancer.png")) +
-#  draw_label("balancer cells", x = 0.1, y = 0.98, size = 10, fontface = "plain") +
-#  draw_label("aboral view", x = 0.1, y = 0.86, color = "black", size = 8) +
-#  draw_label("lateral view of S plane", x = 0.45, y = 0.86, color = "black", size = 8) +
-#  draw_label("lateral view of T plane", x = 0.75, y = 0.86, color = "black", size = 8) +
-#  draw_line(x = c(0.85, 0.95), y = c(0.1, 0.1), color = "black", size = 0.5) +
-#  draw_label(expression(paste("25 ", mu, " m")), x = 0.9, y = 0.14, color = "black", size = 7, hjust = 0.5) +
-#  draw_label("S", x = 0.325, y = 0.16, size = 6, color = "black", hjust = 0.5) +
-#  geom_segment(aes(x = 0.25, y = 0.16, xend = 0.31, yend = 0.16),
-#    color = "black",
-#    arrow = arrow(ends = "both", type = "closed", length = unit(0.1, "cm")),
-#    lineend = "butt",
-#    linejoin = "mitre",
-#    arrow.fill = "black", linewidth = 0.175
-#  ) +
-#  draw_label("T", x = 0.28, y = 0.05, size = 6, color = "black", hjust = 0.5) +
-#  geom_segment(aes(x = 0.28, y = 0.08, xend = 0.28, yend = 0.24),
-#    color = "black",
-#    arrow = arrow(ends = "both", type = "closed", length = unit(0.1, "cm")),
-#    lineend = "butt",
-#    linejoin = "mitre",
-#    arrow.fill = "black", linewidth = 0.175
-#  ) +
-#  draw_label("A", x = 0.66, y = 0.25, size = 6, color = "black", hjust = 0.5) +
-#  draw_label("O", x = 0.66, y = 0.05, size = 6, color = "black", hjust = 0.5) +
-#  geom_segment(aes(x = 0.66, y = 0.09, xend = 0.66, yend = 0.21),
-#    color = "black",
-#    arrow = arrow(ends = "both", type = "closed", length = unit(0.1, "cm")),
[…]

### 2025-05-23 manuscript/Manuscript.qmd
MS: edits, add full connectome code, source_data folder

@@
 However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a].
-In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion.
-At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Supplementary Figure S2A).
+In our dataset, we identified 428 synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion.
+At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Figure 2--figure supplement 1A).
 We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise1973a].
-Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Supplementary Figure S2B).
+Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Figure 2--figure supplement 1B).
 
@@
 The number of synapse-associated mitochondria is shown in green, the number of mitochondia outside synapses is shown in grey.
-N > 2 cells for each cell type except lithocytes.
+N \> 2 cells for each cell type except lithocytes.
 Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cells; imc, intra-multiciliated cells; la, lamellate bodies; li, lithocytes; pl, plumose; ef, epithelial floor cells.\
@@
 Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cell; imc, intra-multiciliated cell; la, lamellate body; li, lithocyte; pl, plumose cell; ef, epithelial floor cell.\
-<!-- check order of panels, change here or change figure -->
-(B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink, and Q3Q4 in orange) and balancer cells (light blue).
+<!-- check order of panels, change here or change figure --> (B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink, and Q3Q4 in orange) and balancer cells (light blue).
 The thickness of the arrows and the numbers correspond to the number of synapses.
@@
 ANNs form no synaptic output to classical effector organs such as muscles or comb plates.
-All observed output synapses were directed to balancer ciliary cells.
-Moreover, the projection patterns of individual ANNs were asymmetric, selectively targeting different regions of the balancer system.
-These findings indicate that ANNs are unlikely to serve as motor drivers in the conventional sense.
+All observed output synapses were directed to balancer ciliary cells, bridge cells or other sensory cells.
+This suggests that ANNs are unlikely to serve as motor drivers in the conventional sense.
 Rather, they may be involved in coordinating or modulating the timing and dynamics of the activity of the sensory-motile balancer cilia, possibly modulating their responsiveness to statocyst load [@keijzer2013; @dewiljes2015; @Veraszto2017].
@@
 What could be the function of the aboral nerve net?
-We found that the anatomical differences in the innervation patterns of ANNs across the sagittal and tentacular planes corresponded with plane-specific differences in balancer cilia dynamics.
-This suggests that individual ANNs may play opposing or complementary roles in regulating ciliary beating---potentially exerting excitatory or inhibitory effects to shape spatiotemporal activity patterns.
-Because known neurotransmitter-related genes show little homology in ctenophores, the functional polarity (excitatory vs.
-inhibitory) of these neurons remains unresolved [@Ryan2013;@Moroz2014].
+We found differences in the innervation patterns of ANNs across the sagittal and tentacular planes.
+The two smaller ANNs targeted a pair of balancers (either Q1-Q2 or Q3-Q4) across the sagittal plane.
+This innervation pattern corresponded with plane-specific differences in the dynamics of balancer cilia.
+Across the sagittal but not the tentacular plane, the closures of balancer cilia were near-simultaneous, within less than one beat phase.
+In contrast, the large ANN innervated balancers in all four quadrants and could account for the near-simultaneous re-beat (within less than one beat phase) of the four balancers.
+The neurotransmitter identity and activation dynamics of these neurons remains unresolved.
 However, the combination of circuit architecture and behavioural outputs provides the first functional insight into the role of neurons in a ctenophore.
 
-Strikingly, this organization resembles circuit motifs found in other animals.
-For instance, in *Platynereis dumerilii* larvae, a single cholinergic motor neuron synchronously halts multiple ciliary bands [@Veraszto2017], while a pair of serotonergic neurons project axons that cross at the midline in a chiasm-like fashion, enabling reciprocal modulation of left and right ciliary fields [@Veraszto2017; @calderón2024].
-Although the morphology differs, these similarities point to convergent evolution of distributed ciliary control circuits[@moroz2015; @roberts2022].
+The functional organisation of the ANN circuit is reminiscent of the ciliomotor circuit in the larvae of the marine annelid *Platynereis dumerilii*.
+Here, the activation of cholinergic motor neurons leads to the synchronous arrest of cilia on multiple multiciliary bands [@Veraszto2017].
+This is antagonised by serotonergic midline-crossing neurons that drive ciliary re-beat [@Veraszto2017; @calderon2024].
+The ctenophore balancer-circuit is the second synaptically mapped ciliomotor circuit that convergently evolved analogous control mechanisms to the annelid and other cilimotor systems [@moroz2015; @Marinkovi2020; @roberts2022].
 
-Taken together, our findings suggest that the ANN network represents a non-motor, coordination-based neural system—one whose primary function is to tune and synchronize locomotor effectors rather than activate them directly.
-This departs fundamentally from reflex-arc models of early nervous systems [@arendt2021; @jékely2010] and instead supports the idea that coordination, not command, may have been a foundational role of ancestral neurons.
+Overall, our findings suggest that the ANN network has a coordinating function to synchronize the activity of balancer cilia thereby possibly fine-tuning their activity or resetting their sensitivity.
+This departs fundamentally from reflex-arc models of simple nervous systems [@arendt2021; @jekely2010] and instead supports the idea that coordination, not command, may have been a foundational role of ancestral neurons [@keijzer2013].
+Such an organisation may have been retained in modern ctenophores ever since they evolved a pelagic lifestyle and a balancer [@stanley1983].
 
 ![**Figure 5. Comparison of Neural Circuits Regulating Ciliary Movement**\
-(A) Each neuron represented by a circle is color-coded to indicate putative homologous cell types involved in ciliary movement regulation.
+(A) Neurons are represented by circles, color-coded to indicate analogous functions in ciliary control.
 Ciliated cells are shown in gray.
 Synapses are indicated by arrows, with magenta representing synapses that induce ciliary arrest and blue representing synapses that induce ciliary re-beat or an increase in ciliary beating frequency.\
-(Left) Gravity-sensing neural circuit of *M.
-leidyi*, as revealed in this study.
-Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying their potential involvement in feedback mechanisms between neurons and ciliated cells.\
-(Righr) Neural circuit regulating prototroch ciliary movement in *P.
+(Left) Neural circuit of the *M.
+leidyi* gravi-sensory organ.
+Bridge cells (yellow squares) are suggested to be electrically coupled (indicated by yellow zigzag lines), implying a potential involvement in feedback mechanisms between neurons and ciliated cells.\
+(Right) Neural circuit regulating prototroch ciliary movement in *P.
 dumerilii*.
-Serotonergic neurons (Ser-h1 neurons, blue) activate ciliary movement, while cholinergic neurons (MC neurons, magenta) induce ciliary arrest.
-Arrows directed toward ciliated cells in the head prototroch are color-coded accordingly.](figures/Figure5.png)
+Serotonergic neurons (Ser-h1 neurons, blue) activate ciliary movement, while a cholinergic neuron (MC neuron, magenta) induces ciliary arrest.](figures/Figure5.png)
 
@@
 Originally described morphologically by Tamm et al.
-(2002), bridge cells extend thin projections toward the base of the balancer cilia [@Tamm_2002; @tamm2014b].
-Based on this spatial relationship, they may receive information about the physical state of the cilia—such as mechanical load or phase—or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024].
-These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed toward the ANN circuit.
+(2002), bridge cells extend thin projections towards the base of the balancer cilia [@Tamm_2002; @tamm2014b].
+Based on this spatial relationship, they may receive information about the physical state of the cilia---such as mechanical load or phase---or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024].
+These features suggest that bridge cells may act as transducers, converting local ciliary signals into neural inputs directed towards the ANN circuit.
 
@@
 These CBF rhythms are likely generated by the ANN circuit, while bridge cells may function to stabilize or adjust this rhythmic output in accordance with the real-time state of the cilia.
-In this way, bridge cells could provide feedback about the execution target—i.e., the balancer cilia—back into the coordination circuit, serving as a core element of a self-regulating rhythm control system.
+This way, bridge cells may provide feedback about the state of their cilia to the coordination circuit, serving as a core element of a self-regulating rhythm-control system.
 
@@
 Such a structure suggests a loop-based coordination mechanism [@arshavsky2003; @selverston2010], where the central circuit adapts its timing based on feedback from its target.
-Although the electrophysiological properties of bridge cells remain to be elucidated, their recursive integration into the ANN circuit indicates that they may play a central role in maintaining the robustness and flexibility of endogenous ciliary rhythms [@kennedy2024].
+Although the electrophysiological properties of bridge cells remain to be elucidated, their recursive integration into the ANN circuit indicates that they may function in maintaining the robustness and flexibility of endogenous ciliary rhythms [@kennedy2024].
 
@@
 In *Mnemiopsis*, the animal swims upright with its mouth facing upward, and performs 90-degree rotational turns when hunting prey or responding to disturbances [@waggett1999; @colin2010; @courtney2020].
-Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015]—but these remain hypotheses.
+Large rotational maneuvers may involve balancer cells in the tentacular plane, while sagittal-plane balancers may contribute to postural stabilization [@tamm2015].
 
-Our findings reveal a circuit composed of ANN neurons and bridge cells that coordinates ciliary beating rhythms in space and time without sensory input, and features a recursive architecture in which ciliary outputs are relayed back into the circuit via bridge cells.
+Our findings reveal a circuit composed of ANN neurons and bridge cells that coordinates ciliary beating rhythms in space and time without sensory input, and features a recursive architecture in which ciliary outputs may feed back to the circuit via bridge cells.
 Such a configuration suggests that even with a limited number of cells, the system is capable of flexible and robust rhythm control.
@@
 (2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011].
-The network we describe here similarly orchestrates output through dynamic, self-regulating coordination.
-These dynamics, visualized through live imaging of the entire cell ensemble, offer a window into the core functions of a nervous system—not in complexity, but in principle.
+The network we describe here orchestrates output through dynamic, self-regulating coordination.
+Future work could uncover this dynamics, visualized through live imaging of the entire cell ensemble, to offer a window into the core functions of a unique nervous system.
 
@@
 
-![**Figure 1—figure supplement 1**\
-(A)...](figures/Figure1_Supplement1.png)
+![**Figure 1—figure supplement 1. Cell-type composition of the ctenophore aboral organ**\
+Morphological rendering of skeletonised cell types, (A) balancer cells (B) bridge cells, (C) bristle cells, (D) ciliated groove cells, (E) dense vesicle cells, (F) dome cells, (G) intra-multiciliated cells, (H) lamellate bodies, (I) lithocytes, (J) plumose cells, (K) aboral nerve-net neurons (L) epithelial floor cells.](figures/Figure1_Supplement1.png)
+
+![**Figure 1—figure supplement 2. Balancers with all other cells in the ctenophore aboral organ**\
+Morphological rendering of skeletonised cells with the balancer cells shown in orange.](figures/Figure1_Supplement2.png)
 

### 2025-05-23 analysis/scripts/connectome_from_CATMAID.R

@@ -1,0 +1,161 @@
+# code to generate the ctenophore apical organ connectome graph based on the CATMAID database
+# Gaspar Jekely 2023-2025
+
+# source packages and functions ------------------------------------------------
+source("analysis/scripts/packages_and_functions.R")
+
+# get all synapses from CATMAID
+all_syn_connectors <- catmaid_fetch(
+  path = "35/connectors/",
+  body = list(
+    relation_type = "presynaptic_to",
+    relation_type = "postsynaptic_to",
+    with_partners = "true"
+  )
+)
+length(all_syn_connectors$connectors)
+
+connector_ids <- unlist(unname(sapply(all_syn_connectors$connectors, "[[", 1)))
+connectivity <- catmaid_get_connectors(connector_ids, pid = 35)
+length(connector_ids)
+dim(connectivity)[1]
+connectivity
+
+#create graph
+edges <- tibble(
+  pre = as.character(connectivity |>
+            select(pre) |>
+            pull()),
+  post = as.character(connectivity |>
+             select(post) |>
+             pull()
+           )
+)
+  
+conn.tb <- tbl_graph(edges = edges, directed = TRUE)
+conn.tb
+
+# retrieve node names, simplify and filter the graph --------------------------------------------------------
+
+# add weighted degree to nodes
+conn.tb.str <- conn.tb %>%
+  activate(nodes) %>%
+  mutate(strength = centrality_degree(
+    weights = NULL,
+    mode = "all",
+    loops = TRUE,
+    normalized = FALSE
+  ))
+
+# test if graph is simple (no multi-edges)
+is_simple(as.igraph(conn.tb.str))
+
+# add edge weight of 1
+conn.tb.str <- set_edge_attr(conn.tb.str, "weight", value = 1)
+conn.tb.str
+
+# merge edges by converting to simple and summing edge weights
+conn.tb.str.sum <- conn.tb.str %>%
+  activate(edges) %>%
+  convert(to_simple) %>%
+  mutate(weight = map_dbl(.orig_data, ~ sum(.x$weight))) %>%
+  select(-.orig_data, -.tidygraph_edge_index) %>%
+  activate(nodes) %>%
+  select(-.tidygraph_node_index)
+
+conn.tb.str.sum
+is_simple(as.igraph(conn.tb.str.sum))
+components(conn.tb.str.sum)
+
+# filter by node strength
[…]

### 2025-05-28 analysis/scripts/connectome_from_CATMAID.R
packages: define by list

@@ -21,31 +21,34 @@ length(connector_ids)
 dim(connectivity)[1]
 connectivity
 
-#create graph
+# create graph
 edges <- tibble(
   pre = as.character(connectivity |>
-            select(pre) |>
-            pull()),
+    select(pre) |>
+    pull()),
   post = as.character(connectivity |>
-             select(post) |>
-             pull()
-           )
+    select(post) |>
+    pull())
 )
-  
+
 conn.tb <- tbl_graph(edges = edges, directed = TRUE)
 conn.tb
 
 # retrieve node names, simplify and filter the graph --------------------------------------------------------
 
 # add weighted degree to nodes
-conn.tb.str <- conn.tb %>%
-  activate(nodes) %>%
-  mutate(strength = centrality_degree(
-    weights = NULL,
-    mode = "all",
-    loops = TRUE,
-    normalized = FALSE
-  ))
+add_node_wdegree <- function(graph.tb){
+  graph.tb|>
+    activate(nodes) |>
+    mutate(strength = centrality_degree(
+      weights = NULL,
+      mode = "all",
+      loops = TRUE,
+      normalized = FALSE
+    ))
+}
+  
+conn.tb.str <- add_node_wdegree(conn.tb)
 
 # test if graph is simple (no multi-edges)
 is_simple(as.igraph(conn.tb.str))
@@ -55,13 +58,22 @@ conn.tb.str <- set_edge_attr(conn.tb.str, "weight", value = 1)
 conn.tb.str
 
 # merge edges by converting to simple and summing edge weights
-conn.tb.str.sum <- conn.tb.str %>%
-  activate(edges) %>%
-  convert(to_simple) %>%
-  mutate(weight = map_dbl(.orig_data, ~ sum(.x$weight))) %>%
-  select(-.orig_data, -.tidygraph_edge_index) %>%
-  activate(nodes) %>%
-  select(-.tidygraph_node_index)
+
+conn.tb.str.sum <- conn.tb.str |>
+  activate(edges) |>
+  group_by(from, to) |>
+  mutate(weight = sum(weight)) |>
+  distinct() |>
+  activate(nodes)
+
+# merge edges by converting to simple and summing edge weights
+#conn.tb.str.sum <- conn.tb.str %>%
+#  activate(edges) %>%
+#  convert(to_simple) %>%
[…]

### 2025-05-28 analysis/scripts/packages_and_functions.R

@@ -1,28 +1,19 @@
 rm(list = ls(all.names = TRUE)) #will clear all objects includes hidden objects.
 gc() #free up memory and report the memory usage.
 
-# load some packages
-library(catmaid)
-library(plyr)
-library(tidyverse)
-library(cowplot)
-library(png)
-library(igraph)
-library(networkD3)
-library(visNetwork)
-library(webshot2)
-library(patchwork)
-library(RColorBrewer)
-library(igraph)
-library(tidygraph)
-library(av)
-library(jpeg)
-library(magick)
-library(ggbeeswarm)
-library(scales)
-library(ggsignif)
-library(rstatix)
+#package list
+pkgs = c(
+  "catmaid", "plyr", "tidyverse",
+  "cowplot", "png", "igraph", 
+  "networkD3", "visNetwork","webshot2", 
+  "patchwork", "RColorBrewer", "tidygraph", 
+  "av", "jpeg", "magick",
+  "ggbeeswarm", "scales", "ggsignif", 
+  "rstatix", "purrr"
+  )
 
+# load packages
+inst = lapply(pkgs, library, character.only = TRUE) 
 
 # catmaid connection, needs username, password AND token - weird!
 {

### 2025-05-28 manuscript/Manuscript.tex

@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
 
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
 \usepackage{bookmark}
@@
 University of Bergen, Norway \textbf{\(^{*}\) Correspondence:}
-\href{mailto:jokura@nibb.ac.jp}{\nolinkurl{jokura@nibb.ac.jp}},
-\href{mailto:gaspar.jekely@cos.uni-heidelberg.de}{\nolinkurl{gaspar.jekely@cos.uni-heidelberg.de}}}
+\includegraphics[width=30em,height=\textheight,keepaspectratio]{pictures/emails.png}}
 \author{}
@@
 reflecting nerve-net organisation. Our results suggest a coordinating
-rather than effector function for the nerve net and inform our
-understanding of the diversity of nervous-systems organisation across
+rather than sensory-motor function for the nerve net and inform our
+understanding of the diversity of nervous-system organisation across
 animals.
@@
 the water column. Despite lacking a centralized nervous system,
-ctenophores exhibit sophisticated and dynamic behavioral patterns and
+ctenophores exhibit sophisticated and dynamic behavioural patterns and
 respond to a broad range of external stimuli, suggesting the presence of
@@
 metazoans (Hayakawa et al., 2022; Sachkova et al., 2021). Notably, early
-single-cell transcriptomic studies failed to identify canonical neuronal
-clusters using conventional markers (Sebé-Pedrós et al., 2018). It was
-only through the use of pro-neuropeptide-based markers that distinct
-neural cell types were subsequently identified (Hayakawa et al., 2022;
-Sachkova et al., 2021). In parallel, phylogenomic and chromosomal
-synteny analyses have suggested that ctenophores may represent the
-sister group to all other animals (Li et al., 2021; Ryan et al., 2013;
-Schultz et al., 2023; Whelan et al., 2015), raising the possibility that
-their nervous system evolved independently from those of other metazoans
-(Moroz, 2015). This hypothesis positions ctenophores as a key lineage
-for re-evaluating the origins and diversity of neural systems. However,
-without a detailed understanding of how ctenophore neurons are spatially
-organized and interconnected, it remains unclear how their nervous
-system gives rise to behavior. Clarifying this relationship between
-neural architecture and behavioral control is essential to uncovering
-the logic of ctenophore neural function.
-
-To address the lack of structural understanding of how ctenophores
-control behavior, we focused on a unique sensory structure---the aboral
-organ. This gravity-sensing organ contains a statocyst supported by four
-mechanosensory balancer cells, which act as pacemakers for comb plate
-beating and posture control (HORRIDGE, 1965; Tamm, 1980).
-Ultrastructural and immunohistochemical studies have revealed that the
-aboral organ contains diverse sensory and neural components, including a
-distinct neural structure termed the deep nerve net (Aronova, 1974;
-HERNANDEZ-NICAISE, 1968; Jager et al., 2010). A key feature of balancer
-cilia is their bidirectional, mechanoresponsive control of beat
-frequency. Depending on the direction of deflection and the animal's
-geotactic state, balancer cilia may either increase or decrease their
-activity, indicating a dynamic response system not fixed like vertebrate
-hair cells (HORRIDGE, 1965; Tamm, 1982; Tamm, 2014a). Previous studies
-have implicated membrane potential, intracellular calcium concentration,
-and synaptic input in modulating this response (Lowe, 1997). In
-particular, deflection-induced excitation can be abolished by calcium
-channel inhibitors or calcium-free conditions (Lowe, 1997), while high
+genomic and single-cell transcriptomic studies failed to identify
+conserved neuronal markers in ctenophores (Moroz et al., 2014;
+Sebé-Pedrós et al., 2018). It was only through the use of
+pro-neuropeptide-based markers that distinct neural cell types were
+molecularly identified (Hayakawa et al., 2022; Sachkova et al., 2021).
+In parallel, phylogenomic and chromosomal synteny analyses have
+suggested that ctenophores may represent the sister group to all other
+animals (Li et al., 2021; Ryan et al., 2013; Schultz et al., 2023;
+Whelan et al., 2015), raising the possibility that their nervous system
+evolved independently from those of other metazoans (Moroz, 2015). This
+hypothesis positions ctenophores as a key lineage for re-evaluating the
+origins and diversity of neural systems. However, without a detailed
+understanding of how ctenophore neurons are spatially organized and
+interconnected, it remains unclear how their nervous system controls
+behaviour. Clarifying this relationship between neural architecture and
+behavioural control is essential to uncovering the logic of ctenophore
+neural function.
+
+To address the lack of structural, circuit-level understanding of the
+ctenophore nervous system, we focused on a unique sensory
+structure---the aboral organ. This gravity-sensing organ contains a
+statocyst supported by four mechanosensory balancer cells, which act as
+pacemakers for the beating of comb plate cilia and thereby control body
+orientation (Horridge, 1965; Tamm, 1980). Ultrastructural and
+immunohistochemical studies have revealed that the aboral organ contains
+diverse sensory and neural components, including a distinct neural
+structure termed the deep nerve net (Aronova, 1974; Hernandez-Nicaise,
+1968; Jager et al., 2010). A key feature of balancer cilia is their
+bidirectional, mechanoresponsive control of beat frequency. Depending on
+the direction of deflection and the animal's geotactic state, balancer
+cilia may either increase or decrease their activity, indicating a
+dynamic response system, not a fixed one like in vertebrate hair cells
+(Horridge, 1965; Tamm, 1982; Tamm, 2014a). Previous studies have
+implicated membrane potential, intracellular calcium concentration, and
+synaptic input in modulating this response (Lowe, 1997). In particular,
+deflection-induced excitation can be abolished by calcium channel
+inhibitors or calcium-free conditions (Lowe, 1997), while high
 extracellular K⁺ depolarizes and activates cilia in the absence of
@@
 excitatory response (Lowe, 1997). These findings support the hypothesis
-that neural modulation plays a critical role in shaping balancer cell
+that neural modulation plays a critical role in shaping balancer-cell
 output (Lowe, 1997).
@@
 ctenophore \emph{Mnemiopsis leidyi}. Through connectomic reconstruction,
-we characterized the synaptic architecture of the aboral nerve net, and
-functional imaging revealed correlated activity patterns in balancer
-cilia. These results suggest that the aboral nerve net plays a central
-role in integrating sensory input and coordinating motor output.
+we characterized the synaptic architecture of the aboral nerve net,
+revealing synaptic inputs to the balancer cell. By functional imaging,
+we uncovered correlated activity patterns in balancer cilia across the
+four quandrants of the apical organ. Our results suggest that the aboral
+nerve net functions in coordinating the activity of balancer cilia,
+rather than fulfilling a feedforward sensory-motor function.
 
@@
 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only
-the region containing the aboral organ. Subsequently, 620 of these
-images were stitched and aligned in TrakEM2 (Cardona et al., 2012). The
+the region containing the aboral organ. Subsequently, images from 620
+layers were stitched and aligned in TrakEM2 (Cardona et al., 2012). The
 final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We
 skeletonised and annotated all cells in this dataset in CATMAID
-(Saalfeld et al., 2009), resulting in a reconstruction of 909 cells with
-nuclei and cell bodies intact. Most of the cells, including the
+(Saalfeld et al., 2009), resulting in a reconstruction of 909 cells that
+had intact nuclei and cell bodies. Most of the cells, including the
 balancers and the aboral nerve net neurons are fully contained within
@@
 
-Next, we divided the entire aboral organ into four broad quadrants
-(Q1-Q4) to facilitate grouping the identified cells (Figure 1G)
-(Martindale and Henry, 1999). The general body plan of ctenophores, when
-viewed from the aboral side, exhibits biradial symmetry around the anal
-pores (Martindale and Henry, 1999). This symmetry corresponds to the
-four blastomeres present at the four-cell stage during early embryonic
+Next, we divided the entire aboral organ into four quadrants (Q1-Q4) to
+facilitate grouping the identified cells (Figure 1G) (Martindale and
+Henry, 1999). The general body plan of ctenophores, when viewed from the
+aboral side, exhibits biradial symmetry around the anal pores
+(Martindale and Henry, 1999). This symmetry corresponds to the four
+blastomeres present at the four-cell stage during early embryonic
 development (Martindale and Henry, 1999). We assigned each of the 909
@@
 
-{\centering \includegraphics{figures/Figure1.png}
+{\centering \pandocbounded{\includegraphics[keepaspectratio]{figures/Figure1.png}}
 
@@
 quadrant scheme shown in panels (B) and (C).\\
-(F) Example of cell tracing using the collaborative annotation toolkit
-CATMAID for large-scale electron microscopy image datasets. The
-spherical objects indicate nuclear positions, while the lines represent
-the traced cell centers. Scale bar: 10 µm.\\
-(G) Three-dimensional reconstruction of cells composing the aboral
-organ, displayed from different perspectives. The left panel presents an
-aboral view of the aboral organ. The middle panel shows a sagittal plane
-view, while the right panel provides a tentacular plane view. Cells are
-color-coded according to their respective quadrants. Lithocytes (li) are
-represented as three gray spheres, while balancers (bal) are depicted as
-gray lines. Scale bar: 25 µm.}
+(F) Example of cell tracing in CATMAID. The spherical objects indicate
+nuclear positions, while the lines represent the traced cell centers.
+Scale bar: 10 µm.\\
+(G) Morphological rendering of cells composing the aboral organ
+displayed in aboral view (left pannel), sagittal plane view (middle
+panel) and tentacular plane view (right panel). Cells are color-coded
+according to quadrants. Lithocytes (li) are represented as three gray
+spheres, balancers (bal) are depicted as gray lines. Scale bar: 25 µm.}
 
@@
 (Hernandez-Nicaise, 1973; Horridge and Mackay, 1964). In our dataset, we
-identified synapses based on the previously described pre-synaptic triad
-morphology consisting of a single layer of vesicles, a cisterna of
+identified 428 synapses based on the previously described pre-synaptic
+triad morphology consisting of a single layer of vesicles, a cisterna of
 smooth endoplasmic reticulum and an associated mitochondrion. At
@@
 cells across the regions where synaptic vesicles aligned with the
-presynaptic cell's membrane (light blue arrows in CATMAID)
-(Supplementary Figure S2A). We could not detect specialized
-post-synaptic structures, in agreement with previous studies
-(Hernandez-Nicaise, 1973). Synapses were identified as either monoadic
-or polyadic, with one pre-synaptic neuron forming connections with one
-or multiple post-synaptic cells (Supplementary Figure S2B).
+presynaptic cell's membrane (light blue arrows in CATMAID) (Figure
+2--figure supplement 1A). We could not detect specialized post-synaptic
+structures, in agreement with previous studies (Hernandez-Nicaise,
+1973). Synapses were identified as either monoadic or polyadic, with one
+pre-synaptic neuron forming connections with one or multiple
+post-synaptic cells (Figure 2--figure supplement 1B).
 
@@
 their distinct size and position, we classified the ANNs into two types.
-The first type is a single large neuron (ANN\_Q1-4) spans all four
+The first type is a single large neuron (ANN\_Q1-4) that spans all four
 quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has
@@
 
-{\centering \includegraphics{figures/Figure2.png}
+{\centering \pandocbounded{\includegraphics[keepaspectratio]{figures/Figure2.png}}
 
@@
 net}\\
-(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4),
-which spans all four quadrants and contains multiple nuclei (blue). The
-spheres represent the positions of individual nuclei. The left panel
-shows a aboral view of the aboral organ, the middle panel presents a
-sagittal plane view, and the right panel provides a tentacular plane
-view.\\
-(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each
-spanning two quadrants and containing multiple nuclei. The spheres
-indicate the positions of the nuclei. The left panel shows a aboral view
-of the aboral organ, the middle panel presents a sagittal plane view,
-and the right panel provides a tentacular plane view.\\
-(C) Graph showing the number of mitochondria per cell and the proportion
-of mitochondria involved in synapse formation for each cell type. Green
-represents mitochondria associated with synapses, while gray represents
-mitochondria not involved in synapse formation. ANN has a significantly
-higher number of mitochondria compared to other cell types, with most of
-them forming synapses. Except for lithocytes, each cell type has a
-sample size of at least three cells. The abbreviations are as follows:
-bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv
-(dense vesicle cells), imc (intra-multiciliated cells), la (lamellate
-bodies), li (lithocytes), pl (plumose), and ef (epithelial floor
-cells).\\
-(D) Localization of mitochondria within ANN Q1-4. Red indicates
+(A) Morphological rendering of the large aboral nerve-net neuron
+ANN\_Q1-4 spanning all four quadrants, in aboral view (left), sagittal
+view (middle panel) and tentacular view (right panel). Spheres indicate
+the positions of nuclei.\\
+(B) Morphological rendering of the two smaller ANN neurons ANN\_Q1Q2
+(pink) and ANN\_Q3Q4 (orange), each spanning two quadrants, in aboral
+view (left), sagittal view (middle panel) and tentacular view (right
+panel). Spheres indicate the positions of nuclei.\\
+(C) Number of mitochondria per cell for each cell type. The number of
+synapse-associated mitochondria is shown in green, the number of
+mitochondia outside synapses is shown in grey. N \textgreater{} 2 cells
+for each cell type except lithocytes. Abbreviations: bal, balancer; brg,
+bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cells; imc,
+intra-multiciliated cells; la, lamellate bodies; li, lithocytes; pl,
+plumose; ef, epithelial floor cells.\\
+(D) Position of mitochondria within ANN\_Q1-4. Red indicates
 mitochondria associated with the presynaptic triad structures, yellow
@@
 projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN
-Q1-4). The abbreviations are as follows: mi (mitochondrion), er
-(endoplasmic reticulum), and sv (synaptic vesicles).\\
-(F) Plot showing the three-dimensional positions of synapses. Synapses
-from the center ANN to the lateral ANN are shown in magenta, while
-synapses from the lateral ANN to the center ANN are shown in cyan. Blue
-dots indicate the locations of autapses within ANN Q1-4.}
+Q1-4). Abbreviations: mi, mitochondrion; er, endoplasmic reticulum; sv,
+synaptic vesicles.\\
+(F) Position of synapses. Synapses from the center ANN to the lateral
+ANN are shown in magenta, while synapses from the lateral ANN to the
+center ANN are shown in cyan. Blue dots indicate the locations of
+autapses within ANN\_Q1-4.}
 
@@
 
-The second group of cells that form synaptic contact with the ANN were
-the bridge cells. Bridge cells, first described by Tamm et al.~in 2002
-(Tamm and Tamm, 2002), are characterized by bundles of elongated
-processes filled with microtubules that arch over the epithelial layer,
-resembling a bridge. Their somata are located at the base of the paired
-balancer-cell clusters along the tentacle surface and extend across the
-sagittal plane towards the base of the opposite balancer cells. Bridge
-cells form two distinct cell groups across the sagittal plane, in the
-Q1Q2 and Q3Q4 regions. In the Q1Q2 region, we identified 14 bridge
-cells, in Q3Q4, 12 cells.
+The second group of cells that formed synaptic contacts with the ANN
+were the bridge cells. Bridge cells, first described in 2002 (Tamm and
+Tamm, 2002), are characterized by bundles of elongated processes filled
+with microtubules that arch over the epithelial layer, resembling a
+bridge. Their somata are located at the base of the paired balancer-cell
+clusters along the tentacle surface and extend across the sagittal plane
+towards the base of the opposite balancer cells. Bridge cells form two
+distinct cell groups across the sagittal plane, in the Q1Q2 and Q3Q4
+regions. In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12
+cells.
 
@@
 
-In both The Q1Q2 and the Q3Q4 regions, bridge cells also formed synapses
+In both the Q1Q2 and the Q3Q4 regions, bridge cells also formed synapses
 with adjacent bridge cells. However, we found no synapses across the
@@
 
-{\centering \includegraphics{figures/Figure3.png}
+{\centering \pandocbounded{\includegraphics[keepaspectratio]{figures/Figure3.png}}
 
@@
 
-\caption{\textbf{Figure 3. Reconstruction of the balancers and bridges
-that receives synapses from the ANN.}\\
-(A) Graph showing the number of synaptic outputs from the ANNs (ANN
-Q1-4: blue, ANN Q1Q2: orange, ANN Q3Q4: pink) circuit for each cell
-type. The balancer and bridge cell types receive relatively high outputs
-from the ANNs circuit. ANN represents the number of synapses within the
-ANNs circuit. The abbreviations are as follows: bal (balancer), brg
-(bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells),
-imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes),
-pl (plumose), and ef (epithelial floor cells).\\
+\caption{\textbf{Figure 3. Synaptic connectivity of the balancer and and
+bridge cells with the ANN.}\\
+(A) Number of synaptic inputs from the ANNs (ANN\_Q1-4: blue, ANN\_Q1Q2:
+orange, ANN\_Q3Q4: pink) to each cell type. Abbreviations: bal,
+balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense
+vesicle cell; imc, intra-multiciliated cell; la, lamellate body; li,
+lithocyte; pl, plumose cell; ef, epithelial floor cell.\\
 (B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink,
-and Q3Q4 in orange) and balancer ciliated cells (light blue). The
-thickness of the arrows and the numbers correspond to the number of
-synapses. Synaptic structures from the same neuron targeting the same
-balancer cell are grouped into hexagons, with the number of cells added
-to the label.\\
-(C) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4
+and Q3Q4 in orange) and balancer cells (light blue). The thickness of
+the arrows and the numbers correspond to the number of synapses.
+Synaptic structures from the same neuron targeting the same balancer
+cell are grouped into hexagons, with the number of cells added to the
+label.\\
+(C) Morphological rendering of bridge cells spanning the Q1Q2 and Q3Q4
 quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while
@@
 sagittal planes could lead to different forces exerted on cilia by the
-statolith (Sidney L. Tamm, 2015a; Tamm, 2014b). To further explore this,
-we used a tilted microscope with a vertical stage where we mounted
-immobilised cydippid larvae with their aboral-oral axis aligned in
-different orientations relative to the gravity vector. We then compared
-larvae oriented in different angles and either with their sagittal or
-tentacular plane parallel to the sample stage.
+statolith (Tamm, 2015, 2014b). To further explore this, we used a tilted
+microscope with a vertical stage where we mounted immobilised cydippid
+larvae with their aboral-oral axis aligned in different orientations
+relative to the gravity vector. We then compared larvae oriented in
+different angles and either with their sagittal or tentacular plane
+parallel to the sample stage.
 
 In larvae with their sagittal plane facing the objective, we could
-compared balancer-cilia movements between Q1(4) and Q2(3). In other
+compare balancer-cilia movements between Q1(4) and Q2(3). In other
 larvae oriented in the tentacular plane, we could simultaneously image
@@
 
-{\centering \includegraphics{figures/Figure4.png}
+{\centering \pandocbounded{\includegraphics[keepaspectratio]{figures/Figure4.png}}
 
@@
 
-\caption{\textbf{Figure 4. Differences in Balancer Cilia Movement
-Control Due to Variations in the Neural Circuit Pathways}\\
-(A) Schematic diagram of differential interference contrast (DIC)
-microscopy setup for imaging balancer cilia movement. The microscope was
-tilted 90 degrees so that the stage was positioned vertically. A
-monochrome CMOS camera sensitive to near-infrared (NIR) light was used,
-synchronized with an 850 nm strobe light source. The movement of the
-balancer cilia was recorded using a 40× objective lens.\\
+\caption{\textbf{Figure 4. Analysis of ciliary beating, arrests and
+re-beat across balancers in the four quadrants}\\
+(A) Schematic diagram of the differential interference contrast (DIC)
+microscopy setup used to image the movement of balancer cilia. The
+microscope was tilted 90 degrees so that the stage was positioned
+vertically. We used a 40x objective lens and a monochrome CMOS camera
+sensitive to near-infrared (NIR) light that was synchronized with an 850
+nm strobe light source.\\
 (B) Representative DIC images of the aboral organ viewed along the
@@
 (D) Boxplots showing time differences in the onset of arrest (white
-boxes) and re-beat (gray boxes) between left and right balancer cilia.
+boxes) and re-beat (grey boxes) between left and right balancer cilia.
 Colors of the box outlines indicate the imaging plane: sagittal (blue)
@@
 (E) Bar plots showing ciliary beat frequency (CBF) dynamics of left and
-right balancers (top and bottom, respectively) recorded from sagittal
-(left) and tentacular (right) planes. Positive and negative values
-represent opposite sides within each plane.\\
+right balancers (top and bottom, respectively) recorded from the
+sagittal (left) or the tentacular (right) planes. Positive and negative
+values represent opposite sides within each plane.\\
 (F) Mean rolling Pearson correlation of left--right balancer CBF
@@
 
-\subsubsection{Functional Nerve Net for Ciliary
-Coordination}\label{functional-nerve-net-for-ciliary-coordination}
-
-In this study, we generated the first high-resolution connectome of a
+\subsubsection{An Aboral Nerve Net for Ciliary
+Coordination}\label{an-aboral-nerve-net-for-ciliary-coordination}
+
+In this study, we generated the first high-resolution connectome of the
 ctenophore nervous system by reconstructing approximately 900 cells from
-the aboral organ of five-day-old \emph{Mnemiopsis leidyi} cydippids
-using volumetric electron microscopy. Among these, we identified a
-previously undescribed type of syncytial neuron---here referred to as
-aboral nerve nets (ANNs)---which exhibit multiple nuclei and a
-morphology distinct from the ``beads-on-a-string'' organization reported
-for the subepithelial nerve net (SNN) in one-day-old cydippids
-(Burkhardt et al., 2023; Sachkova et al., 2021). Instead of forming
-discrete nodal swellings, ANNs extend through intercellular spaces,
-weaving between neighboring cells in a smooth, interdigitating fashion,
-suggesting a unique neuron subtype specific to the aboral organ.
-
-Remarkably, ANNs form no synaptic output to classical effector organs
-such as muscles or comb plates. All observed output synapses were
-directed to balancer ciliary cells. Moreover, the projection patterns of
-individual ANNs were asymmetric, selectively targeting different regions
-of the balancer system. These findings indicate that ANNs are unlikely
-to serve as motor drivers in the conventional sense. Rather, they may be
-involved in coordinating or modulating the timing and dynamics of
-ciliary activity (Keijzer et al., 2013; Verasztó et al., 2017; Wiljes et
-al., 2015).
-
-Notably, anatomical differences in ANN projection patterns across
-sagittal and tentacular planes corresponded with plane-specific
-differences in balancer cilia dynamics. This correlation supports the
-idea that individual ANNs may play opposing or complementary roles in
-regulating ciliary beating---potentially exerting excitatory or
-inhibitory effects to shape spatiotemporal activity patterns. Because
-known neurotransmitter-related genes show little homology in
-ctenophores, the functional polarity (excitatory vs.~inhibitory) of
-these neurons remains unresolved (Moroz et al., 2014; Ryan et al.,
-2013). However, the combination of circuit architecture and behavioral
-outputs provides the first functional insight into the role of neurons
-in a ctenophore.
-
-Strikingly, this organization resembles circuit motifs found in other
-animals. For instance, in \emph{Platynereis dumerilii} larvae, a single
-cholinergic motor neuron synchronously halts multiple ciliary bands
-(Verasztó et al., 2017), while a pair of serotonergic neurons project
-axons that cross at the midline in a chiasm-like fashion, enabling
-reciprocal modulation of left and right ciliary fields (Calderón et al.,
-2024; Verasztó et al., 2017). Although the morphology differs, these
-similarities point to convergent evolution of distributed ciliary
-control circuits(Moroz, 2015; Roberts et al., 2022).
-
-Taken together, our findings suggest that the ANN network represents a
-non-motor, coordination-based neural system---one whose primary function
-is to tune and synchronize locomotor effectors rather than activate them
-directly. This departs fundamentally from reflex-arc models of early
-nervous systems (Arendt, 2021; Jékely, 2010) and instead supports the
-idea that coordination, not command, may have been a foundational role
-of ancestral neurons.
+the aboral organ of a five-day-old \emph{Mnemiopsis leidyi} cydippid by
+volumetric electron microscopy. Among the reconstructed cells, we
+identified a previously undescribed type of syncytial neuron---here
+referred to as aboral nerve net (ANNs)---with multiple nuclei. The ANN
+has a morphology distinct from that of the subepithelial nerve net (SNN)
+reported for one-day-old cydippids (Burkhardt et al., 2023; Sachkova et
+al., 2021). Instead of forming discrete nodal swellings or a
+``beads-on-a-string'' organization typical for the SNN, ANNs extend
+through intercellular spaces, weaving between neighboring cells in a
+smooth, interdigitating fashion. The presence of two SNN-type neurons
+more laterally in the aboral organ in our dataset (Figure 2--figure
+supplement1) further underline the distinct identity of ANNs.
+
+ANNs form no synaptic output to classical effector organs such as
+muscles or comb plates. All observed output synapses were directed to
+balancer ciliary cells, bridge cells or other sensory cells. This
+suggests that ANNs are unlikely to serve as motor drivers in the
+conventional sense. Rather, they may be involved in coordinating or
+modulating the timing and dynamics of the activity of the sensory-motile
+balancer cilia, possibly modulating their responsiveness to statocyst
+load (Keijzer et al., 2013; Verasztó et al., 2017; Wiljes et al., 2015).
+
+What could be the function of the aboral nerve net? We found differences
+in the innervation patterns of ANNs across the sagittal and tentacular
+planes. The two smaller ANNs targeted a pair of balancers (either Q1-Q2
+or Q3-Q4) across the sagittal plane. This innervation pattern
+corresponded with plane-specific differences in the dynamics of balancer
+cilia. Across the sagittal but not the tentacular plane, the closures of
+balancer cilia were near-simultaneous, within less than one beat phase.
+In contrast, the large ANN innervated balancers in all four quadrants
+and could account for the near-simultaneous re-beat (within less than
+one beat phase) of the four balancers. The neurotransmitter identity and
+activation dynamics of these neurons remains unresolved. However, the
+combination of circuit architecture and behavioural outputs provides the
+first functional insight into the role of neurons in a ctenophore.
+
+The functional organisation of the ANN circuit is reminiscent of the
+ciliomotor circuit in the larvae of the marine annelid \emph{Platynereis
+dumerilii}. Here, the activation of cholinergic motor neurons leads to
+the synchronous arrest of cilia on multiple multiciliary bands (Verasztó
+et al., 2017). This is antagonised by serotonergic midline-crossing
+neurons that drive ciliary re-beat (Calderón et al., 2024; Verasztó et
+al., 2017). The ctenophore balancer-circuit is the second synaptically
+mapped ciliomotor circuit that convergently evolved analogous control
+mechanisms to the annelid and other cilimotor systems (Marinković et
+al., 2019; Moroz, 2015; Roberts et al., 2022).
+
+Overall, our findings suggest that the ANN network has a coordinating
+function to synchronize the activity of balancer cilia thereby possibly
+fine-tuning their activity or resetting their sensitivity. This departs
+fundamentally from reflex-arc models of simple nervous systems (Arendt,
+2021; Jékely, 2010) and instead supports the idea that coordination, not
+command, may have been a foundational role of ancestral neurons (Keijzer
+et al., 2013). Such an organisation may have been retained in modern
+ctenophores ever since they evolved a pelagic lifestyle and a balancer
+(Stanley and Stürmer, 1983).
 
@@
 
-{\centering \includegraphics{figures/Figure5.png}
+{\centering \pandocbounded{\includegraphics[keepaspectratio]{figures/Figure5.png}}
 
@@
 Ciliary Movement}\\
-(A) Each neuron represented by a circle is color-coded to indicate
-putative homologous cell types involved in ciliary movement regulation.
-Ciliated cells are shown in gray. Synapses are indicated by arrows, with
-magenta representing synapses that induce ciliary arrest and blue
-representing synapses that induce ciliary re-beat or an increase in
-ciliary beating frequency.\\
-(Left) Gravity-sensing neural circuit of \emph{M. leidyi}, as revealed
-in this study. Bridge cells (yellow squares) are suggested to be
-electrically coupled (indicated by yellow zigzag lines), implying their
-potential involvement in feedback mechanisms between neurons and
-ciliated cells.\\
-(Righr) Neural circuit regulating prototroch ciliary movement in
+(A) Neurons are represented by circles, color-coded to indicate
+analogous functions in ciliary control. Ciliated cells are shown in
+gray. Synapses are indicated by arrows, with magenta representing
+synapses that induce ciliary arrest and blue representing synapses that
+induce ciliary re-beat or an increase in ciliary beating frequency.\\
+(Left) Neural circuit of the \emph{M. leidyi} gravi-sensory organ.
+Bridge cells (yellow squares) are suggested to be electrically coupled
+(indicated by yellow zigzag lines), implying a potential involvement in
+feedback mechanisms between neurons and ciliated cells.\\
+(Right) Neural circuit regulating prototroch ciliary movement in
 \emph{P. dumerilii}. Serotonergic neurons (Ser-h1 neurons, blue)
-activate ciliary movement, while cholinergic neurons (MC neurons,
-magenta) induce ciliary arrest. Arrows directed toward ciliated cells in
-the head prototroch are color-coded accordingly.}
+activate ciliary movement, while a cholinergic neuron (MC neuron,
+magenta) induces ciliary arrest.}
 
@@
 the aboral organ. Originally described morphologically by Tamm et
-al.~(2002), bridge cells extend thin projections toward the base of the
+al.~(2002), bridge cells extend thin projections towards the base of the
 balancer cilia (Tamm, 2014a; Tamm and Tamm, 2002). Based on this spatial
@@
 These features suggest that bridge cells may act as transducers,
-converting local ciliary signals into neural inputs directed toward the
+converting local ciliary signals into neural inputs directed towards the
 ANN circuit.
@@
 while bridge cells may function to stabilize or adjust this rhythmic
-output in accordance with the real-time state of the cilia. In this way,
-bridge cells could provide feedback about the execution target---i.e.,
-the balancer cilia---back into the coordination circuit, serving as a
-core element of a self-regulating rhythm control system.
+output in accordance with the real-time state of the cilia. This way,
+bridge cells may provide feedback about the state of their cilia to the
+coordination circuit, serving as a core element of a self-regulating
+rhythm-control system.
 
@@
 their recursive integration into the ANN circuit indicates that they may
-play a central role in maintaining the robustness and flexibility of
-endogenous ciliary rhythms (Kennedy and Weissbourd, 2024).
+function in maintaining the robustness and flexibility of endogenous
+ciliary rhythms (Kennedy and Weissbourd, 2024).
 
@@
 How do differences in balancer cilia coordination across anatomical
-planes shape behavior? In \emph{Mnemiopsis}, the animal swims upright
+planes shape behaviour? In \emph{Mnemiopsis}, the animal swims upright
 with its mouth facing upward, and performs 90-degree rotational turns
@@
 involve balancer cells in the tentacular plane, while sagittal-plane
-balancers may contribute to postural stabilization (Sidney L. Tamm,
-2015b)---but these remain hypotheses.
+balancers may contribute to postural stabilization (Tamm, 2015).
 
@@
 sensory input, and features a recursive architecture in which ciliary
-outputs are relayed back into the circuit via bridge cells. Such a
+outputs may feed back to the circuit via bridge cells. Such a
 configuration suggests that even with a limited number of cells, the
@@
 a ``ciliary brain'' (Jager et al., 2010). The network we describe here
-similarly orchestrates output through dynamic, self-regulating
-coordination. These dynamics, visualized through live imaging of the
-entire cell ensemble, offer a window into the core functions of a
-nervous system---not in complexity, but in principle.
+orchestrates output through dynamic, self-regulating coordination.
+Future work could uncover this dynamics, visualized through live imaging
+of the entire cell ensemble, to offer a window into the core functions
+of a unique nervous system.
 
@@
 
-\bibitem[\citeproctext]{ref-calderuxf3n2024}
+\bibitem[\citeproctext]{ref-calderon2024}
 Calderón LAB, Shahidi R, Jékely G. 2024.
@@
 
-\bibitem[\citeproctext]{ref-Hernandez-Nicaise1974}
+\bibitem[\citeproctext]{ref-Hernandez_Nicaise1974}
 Hernandez-Nicaise M-L. 1974. Ultrastructural evidence for a
@@
 
-\bibitem[\citeproctext]{ref-Hernandez_Nicaise_1973}
-Hernandez-Nicaise M-L. 1973. The nervous system of ctenophores III.
-Ultrastructure of synapses. \emph{Journal of Neurocytology}
-\textbf{2}:249--263.
-doi:\href{https://doi.org/10.1007/bf01104029}{10.1007/bf01104029}
-
-\bibitem[\citeproctext]{ref-Hernandez-Nicaise1968}
-HERNANDEZ-NICAISE M-L. 1968. Specialized Connexions between Nerve Cells
+\bibitem[\citeproctext]{ref-Hernandez_Nicaise1973a}
+Hernandez-Nicaise M-L. 1973. Le Système Nerveux des Cténaires.
+\emph{Zeitschrift für Zellforschung und Mikroskopische Anatomie}
+\textbf{137}:223--250.
+doi:\href{https://doi.org/10.1007/BF00307432}{10.1007/BF00307432}
+
+\bibitem[\citeproctext]{ref-Hernandez_Nicaise1968}
+Hernandez-Nicaise M-L. 1968. Specialized Connexions between Nerve Cells
 and Mesenchymal Cells in Ctenophores. \emph{Nature}
@@
 \bibitem[\citeproctext]{ref-HORRIDGE1965}
-HORRIDGE GA. 1965. RELATIONS BETWEEN NERVES AND CILIA IN CTENOPHORES.
+Horridge GA. 1965. Relations between nerves and cilia in ctenophores.
 \emph{American Zoologist} \textbf{5}:357--375.
@@
 
-\bibitem[\citeproctext]{ref-juxe9kely2010}
+\bibitem[\citeproctext]{ref-jekely2010}
 Jékely G. 2010. Origin and early evolution of neural circuits for the
@@
 \emph{Journal of experimental biology} \textbf{200}:1593--1606.
+
+\bibitem[\citeproctext]{ref-Marinkovi2020}
+Marinković M, Berger J, Jékely G. 2019. Neuronal coordination of motile
+cilia in locomotion and feeding. \emph{Philosophical Transactions of the
+Royal Society B: Biological Sciences} \textbf{375}:20190165.
+doi:\href{https://doi.org/10.1098/rstb.2019.0165}{10.1098/rstb.2019.0165}
 
@@
 
+\bibitem[\citeproctext]{ref-stanley1983}
+Stanley GD, Stürmer W. 1983. The first fossil ctenophore from the Lower
+Devonian of West Germany. \emph{Nature} \textbf{303}:518--520.
+doi:\href{https://doi.org/10.1038/303518a0}{10.1038/303518a0}
+
 \bibitem[\citeproctext]{ref-Tamm1982}
@@
 
-\bibitem[\citeproctext]{ref-Tamm2015}
-Tamm Sidney L. 2015a. Functional consequences of the asymmetric
-architecture of the ctenophore statocyst. \emph{The Biological Bulletin}
-\textbf{229}:173--184.
-
 \bibitem[\citeproctext]{ref-tamm2015}
-Tamm Sidney L. 2015b. Functional Consequences of the Asymmetric
-Architecture of the Ctenophore Statocyst. \emph{The Biological Bulletin}
+Tamm SL. 2015. Functional Consequences of the Asymmetric Architecture of
+the Ctenophore Statocyst. \emph{The Biological Bulletin}
 \textbf{229}:173--184.
@@
 
-\bibitem[\citeproctext]{ref-Tamm_2014}
-Tamm SL. 2014b. Formation of the statolith in the ctenophore mnemiopsis
+\bibitem[\citeproctext]{ref-tamm2014b}
+Tamm SL. 2014a. Formation of the statolith in the ctenophore mnemiopsis
 leidyi. \emph{The Biological Bulletin} \textbf{227}:7--18.
@@
 
-\bibitem[\citeproctext]{ref-Tamm_2014b}
-Tamm SL. 2014a. Cilia and the life of ctenophores. \emph{Invertebrate
+\bibitem[\citeproctext]{ref-Tamm_2014c}
+Tamm SL. 2014b. Cilia and the life of ctenophores. \emph{Invertebrate
 Biology} \textbf{133}:1--46.
@@
 
-{\centering \includegraphics{figures/Figure1_Supplement1.png}
+{\centering \pandocbounded{\includegraphics[keepaspectratio]{figures/Figure1_Supplement1.png}}
 
@@
 
-\caption{\textbf{Figure 1---figure supplement 1}\\
-(A)\ldots{}}
+\caption{\textbf{Figure 1---figure supplement 1. Cell-type composition
+of the ctenophore aboral organ}\\
+Morphological rendering of skeletonised cell types, (A) balancer cells
+(B) bridge cells, (C) bristle cells, (D) ciliated groove cells, (E)
+dense vesicle cells, (F) dome cells, (G) intra-multiciliated cells, (H)
+lamellate bodies, (I) lithocytes, (J) plumose cells, (K) aboral
+nerve-net neurons (L) epithelial floor cells.}
 
@@
 
-{\centering \includegraphics{figures/Figure2_Supplement1.png}
+{\centering \pandocbounded{\includegraphics[keepaspectratio]{figures/Figure1_Supplement2.png}}
+
+}
+
+\caption{\textbf{Figure 1---figure supplement 2. Balancers with all
+other cells in the ctenophore aboral organ}\\
+Morphological rendering of skeletonised cells with the balancer cells
+shown in orange.}
+
+\end{figure}%%
+\begin{figure}[H]
+
+{\centering \pandocbounded{\includegraphics[keepaspectratio]{figures/Figure2_Supplement1.png}}
 

### 2025-06-01 analysis/scripts/Video1.R
put videos

@@ -34,7 +34,7 @@ balancer_conn <- connectors(balancer)
 presyn_balancer <- subset(balancer_conn, prepost == 0)
 postsyn_balancer <- subset(balancer_conn, prepost == 1)
 
-# plot cells ----------------
+# plot cells -------------------------------------------------------------------
 
 nopen3d() # opens a pannable 3d window
 mfrow3d(1, 2)  #defines the two scenes
@@ -260,12 +260,12 @@ for (l in 1:90){
 close3d()
 
 
-#read png files and write video
+#read png files and write video --------------------------------------------------
 av::av_encode_video(
   paste('videoframes/', list.files("videoframes/", '*.png'), 
         sep = ""),
   framerate = 10,
-  output = 'manuscript/videos/Video1.mp4'
+  output = 'manuscript/videos/supplemental video1.mp4'
   )
 
 unlink("videoframes", recursive = T)

### 2025-06-01 analysis/scripts/Video2.R

@@ -175,7 +175,7 @@ av::av_encode_video(
   paste('videoframes/', list.files("videoframes/", '*.png'), 
         sep = ""),
   framerate = 10,
-  output = 'manuscript/videos/Video2.mp4'
+  output = 'manuscript/videos/supplemental video2.mp4'
 )
 
 # delete temp folder -----------------

### 2025-06-02 manuscript/Manuscript.qmd
edit Supplementary material legends
ed supplement legends
edit legends of supplement figure

@@
 
-![**Figure 1—figure supplement 2. Balancers with all other cells in the ctenophore aboral organ**\
-Morphological rendering of skeletonised cells with the balancer cells shown in orange.](figures/Figure1_Supplement2.png)
+![**Figure 1—figure supplement 2. 3D reconstruction of all cells comprising the aboral organ**\
+(A) Morphological rendering of skeletonised cells in the ctenophore aboral organ.
+Each color represents a distinct cell type.
+The balancer cells are shown in orange.
+Some cell groups lack distinguishing subcellular features required for identification and are shown in white.](figures/Figure1_Supplement2.png)
 
-![**Figure 2—figure supplement 1**\
-(A)...](figures/Figure2_Supplement1.png)
+![**Figure 2—figure supplement 1. Ultrastructural features and synaptic organization of the ANN, with comparison to the subepithelial nerve net (SNN)**\
+(A) Registration and annotation of synapses in the actual cellular traces.
+Pink dots indicate the approximate center of each individual cell, marked within its plasma membrane boundary.
+Mitochondria involved in synaptic contacts are labeled with orange dots.
+Red arrows point from these mitochondria to the presynaptic cell body, marking presynaptic structures.
+Cyan arrows point from the orange-labeled mitochondria to the corresponding postsynaptic cell.\
+(B) Electron microscopy (EM) images showing representative presynaptic structures.
+The left panel shows a monadic synapse, while the right panel shows a polyadic (byadic) synapse, in which a single presynaptic site contacts two postsynaptic cells.\
+(C) Nuclei within two different types of ANN cells are shown in purple in a single EM section.
+On the left, the ANN_Q1-4 cell extends widely across the aboral organ and appears as a continuous structure in 3D, despite having nuclei located distantly from each other.
+In contrast, ANN_Q1Q2 (right) contains two adjacent nuclei situated close together, suggesting a more localized distribution.\
+(D) Three-dimensional reconstruction of the subepithelial nerve net (SNN).
+The SNN extends broadly, enveloping the aboral organ.
+Some portions of the network remain as unconnected fragments due to incomplete reconstruction.\
+(E) Morphological comparison between the SNN (left) and ANN (right) in EM images.
+The SNN shows a "beads-on-a-string" morphology, with discrete bead-like swellings connected by faint electron-dense fibrous structures.
+In contrast, the ANN exhibits a more diffuse distribution, often appearing to interdigitate between surrounding cells.
+Relatively large mitochondria are frequently observed in the ANN processes.](figures/Figure2_Supplement1.png)
+
+![**Figure 4—figure supplement 1**\
+(A)...](figures/Figure4_Supplement1.png)
+
+![**supplement video1**\
+A...]
+
+![**supplement video2**\
+A...]
+
+![**supplement video3**\
+A...]
+
+![**supplement video4**\
+A...]
+
+![**supplement video5**\
+A...]
+
+![**supplement video6**\
+A...]

### 2025-06-02 analysis/scripts/Figure2_figure_supplement1.R

@@ -64,8 +64,23 @@ close3d()
 
 # assemble figure -------------------------------------------------------------
 
-panel_SNN_3d <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/plot_SNN.png"))
-panel_ANN_SNN_EM <- ggdraw() + draw_image(readPNG("manuscript/pictures/compaarison_EM_ANN_SNN.png"))
+panel_SNN_3d <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/plot_SNN.png")) +
+  draw_label("subepithelial nerve net (SNN)", x = 0.5, y = 1.05, color = Okabe_Ito[5], size = 10, hjust = 0.5) +
+  draw_label("aboral view", x = 0.18, y = 0.95, color = "black", size = 8, hjust = 0.5) +
+  draw_label("lateral view", x = 0.67, y = 0.95, color = "black", size = 8, hjust = 0.5) +
+  draw_label("Q1", x = 0.3, y = 0.9, color = "black", size = 9, hjust = 0.5, alpha = 0.5) +
+  draw_label("Q2", x = 0.06, y = 0.79, color = "black", size = 9, hjust = 0.5, alpha = 0.5) +
+  draw_label("Q3", x = 0.03, y = 0.08, color = "black", size = 9, hjust = 0.5, alpha = 0.5) +
+  draw_label("Q4", x = 0.275, y = 0.11, color = "black", size = 9, hjust = 0.5, alpha = 0.5) +
+  draw_label("sagittal plane", x = 0.5, y = 0.86, color = "black", size = 7.5, hjust = 0.5) +
+  draw_label("tentacular plane", x = 0.83, y = 0.86, color = "black", size = 7.5, hjust = 0.5) +
+  draw_label("A", x = 0.67, y = 0.85, size = 7, color = "black", hjust = 0.5) +
+  draw_label("O", x = 0.67, y = 0.65, size = 7, color = "black", hjust = 0.5) +
+  draw_line(x = c(0.67, 0.67), y = c(0.69, 0.81), color = "black", linewidth = 0.65) +
+  draw_label(expression(paste("25 ", mu, "m")), x = 0.95, y = 0.1, color = "black", size = 7, hjust = 0.5) +
+  draw_line(x = c(0.91, 0.99), y = c(0.05, 0.05), color = "black", linewidth = 0.7)
+
[…]

### 2025-06-02 manuscript/Manuscript.qmd
add video legends

@@
 
-![**Figure 4—figure supplement 1**\
-(A)...](figures/Figure4_Supplement1.png)
+![**Figure 4—figure supplement 1.Ciliary beat frequency (CBF) dynamics of balancer cells in additional samples**\
+Bar plots showing CBF dynamics of left and right balancers (top and bottom panels, respectively) recorded from the sagittal (A–C) and tentacular (D–F) planes.
+Positive and negative values indicate opposing sides within each imaging plane.
+These data correspond to additional samples not shown in the main figure and support the results presented in Figure 4E.](figures/Figure4_Supplement1.png)
 
-![**supplement video1**\
-A...]
+![**Video 1.** A...]
 
-![**supplement video2**\
-A...]
+![**Video 2.** A...]
 
-![**supplement video3**\
-A...]
+![**Video 3.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
+The white line overlaid on the kymographs indicates the current time point shown in the video.
+This example captures a ciliary arrest event in the sagittal plane.]
 
-![**supplement video4**\
-A...]
+![**Video 4.**This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
+The white line overlaid on the kymographs indicates the current time point shown in the video.
+This example captures a ciliary re-beat event in the sagittal plane.]
 
-![**supplement video5**\
-A...]
+![**Video 5.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
+The white line overlaid on the kymographs indicates the current time point shown in the video.
+This example captures a ciliary arrest event in the tentacular plane.]
 
-![**supplement video6**\
-A...]
+![**Video 6.**This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
+The white line overlaid on the kymographs indicates the current time point shown in the video.
+This example captures a ciliary re-beat event in the tentacular plane.]

### 2025-06-02 analysis/scripts/Video1.R

@@ -196,6 +196,7 @@ plot3d(
 
 
 next3d(clear=FALSE)
+
 plot_multinucleated_cell(SSN_Q1Q2,
                          lwd = 2, alpha = 1, col = Okabe_Ito[6])
 plot_multinucleated_cell(SSN_Q3Q4,

### 2025-06-02 manuscript/Manuscript.qmd
edit video 1

@@
 
-![**Video 1.** A...]
+![**Video 1.** This video shows a rotating 3D rendering of the aboral organ, featuring three distinct cell types—balancer cells, bridge cells, and components of the apical nerve net (ANN).
+Cells of each type are subdivided and color-coded according to their spatial regions.
+A lithocyte is shown in gray as a reference point, and the outline of the aboral organ is indicated in white.]
 
-![**Video 2.** A...]
+![**Video 2.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
+The white line overlaid on the kymographs indicates the current time point shown in the video.
+This example captures a ciliary arrest event in the sagittal plane.]
 
@@
 The white line overlaid on the kymographs indicates the current time point shown in the video.
-This example captures a ciliary arrest event in the sagittal plane.]
+This example captures a ciliary re-beat event in the sagittal plane.]
 
-![**Video 4.**This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+![**Video 4.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
 The white line overlaid on the kymographs indicates the current time point shown in the video.
-This example captures a ciliary re-beat event in the sagittal plane.]
+This example captures a ciliary arrest event in the tentacular plane.]
 
@@
 The white line overlaid on the kymographs indicates the current time point shown in the video.
-This example captures a ciliary arrest event in the tentacular plane.]
-
-![**Video 6.**This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
-The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
-The white line overlaid on the kymographs indicates the current time point shown in the video.
 This example captures a ciliary re-beat event in the tentacular plane.]

### 2025-06-02 manuscript/Manuscript.qmd
html

--- 2025-06-02 manuscript/Manuscript.qmd
@@
 Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cell; imc, intra-multiciliated cell; la, lamellate body; li, lithocyte; pl, plumose cell; ef, epithelial floor cell.\
-<!-- check order of panels, change here or change figure --> (B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink, and Q3Q4 in orange) and balancer cells (light blue).
-The thickness of the arrows and the numbers correspond to the number of synapses.
-Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.\
+(B) Morphological rendering of balancer cells, with each cell color-coded by quadrant: Q1 (yellow), Q2 (blue), Q3 (orange), and Q4 (magenta).
+Each sphere represents the position of an individual nucleus.
+The fine projections extending from the cells represent traced trajectories that follow the cell body and continue into a single cilium from the point where it emerges.
+Views are shown from three orthogonal perspectives: aboral view (left), sagittal plane (middle), and tentacular plane (right).
+The spatial arrangement highlights the quadrant-specific organization of balancer clusters within the aboral organ.
+Scale bar: 25 µm.\
 (C) Morphological rendering of bridge cells spanning the Q1Q2 and Q3Q4 quadrants.
@@
 
-![**Video 1.** A...]
+**Video 1.** This video shows a rotating 3D rendering of the aboral organ, featuring three distinct cell types—balancer cells, bridge cells, and components of the apical nerve net (ANN).
+Cells of each type are subdivided and color-coded according to their spatial regions.
+A lithocyte is shown in gray as a reference point, and the outline of the aboral organ is indicated in white.
 
-![**Video 2.** A...]
-
-![**Video 3.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+**Video 2.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
 The white line overlaid on the kymographs indicates the current time point shown in the video.
-This example captures a ciliary arrest event in the sagittal plane.]
+This example captures a ciliary arrest event in the sagittal plane.
 
-![**Video 4.**This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+**Video 3.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
 The white line overlaid on the kymographs indicates the current time point shown in the video.
-This example captures a ciliary re-beat event in the sagittal plane.]
+This example captures a ciliary re-beat event in the sagittal plane.
 
-![**Video 5.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+**Video 4.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
 The white line overlaid on the kymographs indicates the current time point shown in the video.
-This example captures a ciliary arrest event in the tentacular plane.]
+This example captures a ciliary arrest event in the tentacular plane.
 
-![**Video 6.**This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+**Video 5.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
 The white line overlaid on the kymographs indicates the current time point shown in the video.
-This example captures a ciliary re-beat event in the tentacular plane.]
+This example captures a ciliary re-beat event in the tentacular plane.

### 2025-06-02 manuscript/Manuscript.qmd
add Luciola-Math fond, style.css and update yaml

@@
     **$^{*}$ Correspondence:**  ![](pictures/emails.png){width=30em}
-mainfont: Arial
+mainfont: Luciola-Math
 monofont: DejaVu Sans Mono #Times New Roman
@@
     embed-resources: true
+    css: style.css 
   pdf:

### 2025-06-02 manuscript/Manuscript.qmd
ms edits, add refs

@@
 In parallel, phylogenomic and chromosomal synteny analyses have suggested that ctenophores may represent the sister group to all other animals [@Ryan2013; @Whelan2015a; @Li2021; @Schultz2023], raising the possibility that their nervous system evolved independently from those of other metazoans [@moroz2015].
-This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems.
+This hypothesis positions ctenophores as a key lineage for re-evaluating the origins and diversity of neural systems [@Burkhardt_2021].
 However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system controls behaviour.
@@
 To address the lack of structural, circuit-level understanding of the ctenophore nervous system, we focused on a unique sensory structure—the aboral organ.
-This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for the beating of comb plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980].
+This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for the beating of comb-plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980].
 Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011].
@@
 Local calcium application further suggests that Ca²⁺ influx at the ciliary base may be required for the excitatory response [@Lowe1997].
-These findings support the hypothesis that neural modulation plays a critical role in shaping balancer-cell output [@Lowe1997].
+These findings support the hypothesis that ionic currents and potential neural inputs play critical roles in shaping balancer-cell output [@Lowe1997].
 
 To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy and high-speed imaging in the ctenophore *Mnemiopsis leidyi*.
-Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cell.
+Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cells.
 By functional imaging, we uncovered correlated activity patterns in balancer cilia across the four quandrants of the apical organ.
@@
 The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm.
-We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009], resulting in a reconstruction of 909 cells that had intact nuclei and cell bodies.
-Most of the cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.
+We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016], resulting in a reconstruction of 972 cells that had intact nuclei and cell bodies, 899 of which formed the aboral organ (AO).
+Most of the 899 AO cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.
 
@@
 The entire skeletonized volume was composed of 134,591 nodes.
-88 fragments could not be attached to somata-associated skeletons.
+73 fragments could not be attached to somata-associated skeletons.
 Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.
@@
 This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999].
-We assigned each of the 909 skeletonised cells to one of the four quadrants.
+We assigned each of the 899 skeletonised cells to one of the four quadrants (Figure 1G).
 <!-- check number of cells, 927 in figure -->

### 2025-06-03 manuscript/Manuscript.qmd
remove fragments from full connectome, add fig to ms

@@
 Some balancer cells received inputs from both ANN_Q1-4 and either of ANN_Q1Q2 or ANN_Q3Q4.
-While previous studies suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez_Nicaise1974], our data did not reveal any presynaptic sites in balancer cells.
+While previous studies suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez_Nicaise1974], our data did not reveal any presynaptic sites in balancer cells (Figure 3 -- figure supplement 1).
 
@@
 Bridge cells form two distinct cell groups across the sagittal plane, in the Q1Q2 and Q3Q4 regions.
-In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12 cells.
+In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12 cells (Figure 3 -- figure supplement 1).
 
@@
 
+![**Figure 3—figure supplement 1. Full synaptic connectome of the ctenophore apical organ**\
+Full synaptic graph showing individual cells as nodes connected by synaptic contacts.
+Arrows point from the presynaptic to the postsynaptic cell.
+Edge thickness is proportional to the number of synapses.](figures/Figure3_Supplement1.png)
+
 ![**Figure 4—figure supplement 1.Ciliary beat frequency (CBF) dynamics of balancer cells in additional samples**\

### 2025-06-03 analysis/scripts/connectome_from_CATMAID.R

@@ -31,8 +31,8 @@ edges <- tibble(
     pull())
 )
 
-conn.tb <- tbl_graph(edges = edges, directed = TRUE)
-conn.tb
+conn_tb <- tbl_graph(edges = edges, directed = TRUE)
+conn_tb
 
 # retrieve node names, simplify and filter the graph --------------------------------------------------------
 
@@ -48,18 +48,18 @@ add_node_wdegree <- function(graph.tb){
     ))
 }
   
-conn.tb.str <- add_node_wdegree(conn.tb)
+conn_tb_str <- add_node_wdegree(conn_tb)
 
 # test if graph is simple (no multi-edges)
-is_simple(as.igraph(conn.tb.str))
+is_simple(as.igraph(conn_tb_str))
 
 # add edge weight of 1
-conn.tb.str <- set_edge_attr(conn.tb.str, "weight", value = 1)
-conn.tb.str
+conn_tb_str <- set_edge_attr(conn_tb_str, "weight", value = 1)
+conn_tb_str
 
 # merge edges by converting to simple and summing edge weights
 
-conn.tb.str.sum <- conn.tb.str |>
+conn_tb_str_sum <- conn_tb_str |>
   activate(edges) |>
   group_by(from, to) |>
   mutate(weight = sum(weight)) |>
@@ -67,7 +67,7 @@ conn.tb.str.sum <- conn.tb.str |>
   activate(nodes)
 
 # merge edges by converting to simple and summing edge weights
-#conn.tb.str.sum <- conn.tb.str %>%
+#conn_tb_str_sum <- conn_tb_str %>%
 #  activate(edges) %>%
 #  convert(to_simple) %>%
 #  mutate(weight = map_dbl(.orig_data, ~ sum(.x$weight))) %>%
@@ -75,41 +75,49 @@ conn.tb.str.sum <- conn.tb.str |>
 #  activate(nodes) %>%
 #  select(-.tidygraph_node_index)
 
-conn.tb.str.sum
-is_simple(as.igraph(conn.tb.str.sum))
-components(conn.tb.str.sum)
+conn_tb_str_sum
+is_simple(as.igraph(conn_tb_str_sum))
+components(conn_tb_str_sum)
 
 # filter by node strength
-conn.tb.str.sum.filt <- conn.tb.str.sum %>%
+conn_tb_str_sum_filt <- conn_tb_str_sum %>%
   activate(nodes) %>%
   filter(strength > 0)
[…]

### 2025-06-03 analysis/scripts/Figure3_figure_supplement1.R
rename connectome graph file

@@ -1,5 +1,5 @@
 # code to generate the ctenophore apical organ connectome graph based on the CATMAID database
-# Gaspar Jekely 2023-2025
+# Gaspar Jekely 2025
 
 # source packages and functions ------------------------------------------------
 source("analysis/scripts/packages_and_functions.R")
@@ -17,9 +17,6 @@ length(all_syn_connectors$connectors)
 
 connector_ids <- unlist(unname(sapply(all_syn_connectors$connectors, "[[", 1)))
 connectivity <- catmaid_get_connectors(connector_ids, pid = 35)
-length(connector_ids)
-dim(connectivity)[1]
-connectivity
 
 # create graph
 edges <- tibble(
@@ -32,13 +29,12 @@ edges <- tibble(
 )
 
 conn_tb <- tbl_graph(edges = edges, directed = TRUE)
-conn_tb
 
 # retrieve node names, simplify and filter the graph --------------------------------------------------------
 
 # add weighted degree to nodes
-add_node_wdegree <- function(graph.tb){
-  graph.tb|>
+add_node_wdegree <- function(graph.tb) {
+  graph.tb |>
     activate(nodes) |>
     mutate(strength = centrality_degree(
       weights = NULL,
@@ -47,7 +43,7 @@ add_node_wdegree <- function(graph.tb){
       normalized = FALSE
     ))
 }
-  
+
 conn_tb_str <- add_node_wdegree(conn_tb)
 
 # test if graph is simple (no multi-edges)
@@ -55,10 +51,8 @@ is_simple(as.igraph(conn_tb_str))
 
 # add edge weight of 1
 conn_tb_str <- set_edge_attr(conn_tb_str, "weight", value = 1)
-conn_tb_str
 
 # merge edges by converting to simple and summing edge weights
-
 conn_tb_str_sum <- conn_tb_str |>
   activate(edges) |>
   group_by(from, to) |>
@@ -66,16 +60,7 @@ conn_tb_str_sum <- conn_tb_str |>
   distinct() |>
   activate(nodes)
 
-# merge edges by converting to simple and summing edge weights
-#conn_tb_str_sum <- conn_tb_str %>%
-#  activate(edges) %>%
-#  convert(to_simple) %>%
-#  mutate(weight = map_dbl(.orig_data, ~ sum(.x$weight))) %>%
-#  select(-.orig_data, -.tidygraph_edge_index) %>%
-#  activate(nodes) %>%
[…]

### 2025-06-05 analysis/scripts/Figure2_figure_supplement1.R
Fig 2 fig suppl1: add pic to plot SNN with ANN

@@ -7,57 +7,75 @@ source("analysis/scripts/packages_and_functions.R")
 # load Subepithelial nerve net -------------------------------------------------
 
 SNN <- read_smooth_neuron("celltype:SNN")
+ANN <- read_smooth_neuron("celltype:SSN")
 
 # 3d plot Subepithelial nerve net -----------------------------------------------
 
 close3d()
 # 3d plotting of cells
-nopen3d() 
-mfrow3d(1, 3)  
-#define the size of the rgl window, the view and zoom
-par3d(windowRect = c(0, 0, 1200, 350))
+nopen3d()
+mfrow3d(1, 3)
+# define the size of the rgl window, the view and zoom
+par3d(windowRect = c(0, 0, 2400, 700))
 
-#plot aboral view
-plot_multinucleated_cell(SNN, lwd = 1.25, alpha = 0.8, col = Okabe_Ito[5])
+# plot aboral view
+plot_multinucleated_cell(SNN, lwd = 1.5, alpha = 1, col = Okabe_Ito[2])
 plot3d(outline, add = TRUE, alpha = 0.025, col = Okabe_Ito[8])
 
-#aboral view
+# aboral view
 aboral()
 par3d(zoom = 0.7)
 
-#move to next panel in rgl window
-next3d(clear=F)
+# move to next panel in rgl window
+next3d(clear = F)
 
 
 
-#plot lateral view of Sagittal plane
-plot_multinucleated_cell(SNN, lwd = 1.25, alpha = 0.8, col = Okabe_Ito[5])
+# plot lateral view of Sagittal plane
+plot_multinucleated_cell(SNN, lwd = 1.5, alpha = 1, col = Okabe_Ito[2])
 plot3d(outline, add = TRUE, alpha = 0.025, col = Okabe_Ito[8])
 
-#lateral view of Sagittal plane
+# lateral view of Sagittal plane
 sagittal()
 par3d(zoom = 0.7)
 
-#move to next panel in rgl window
-next3d(clear=F)
+# move to next panel in rgl window
+next3d(clear = F)
 
 
 
-#plot lateral view of Tentacular plane
-plot_multinucleated_cell(SNN, lwd = 1.25, alpha = 0.8, col = Okabe_Ito[5])
+# plot lateral view of Tentacular plane
[…]

### 2025-06-05 manuscript/Manuscript.qmd
ms edit

@@
 subtitle: |
-  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewalow**$^{6}$, **Pawel Burkhardt**$^{7}$, **Gáspár Jékely**$^{1,2,6,*}$  
+  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewelow**$^{6}$, **Pawel Burkhardt**$^{7}$, **Gáspár Jékely**$^{1,2,6,*}$  
   <br><br>  
@@
 We thank Dr.
-Réza Shahidi for sectioning and imaging, and Paulina Cherek for high-pressure freezing and sample preparation.
+Réza Shahidi for sectioning and imaging, Paulina Cherek for high-pressure freezing and sample preparation, and Iva Verbanac for help with tracing.
 We are also grateful to Dr.

### 2025-06-05 manuscript/Manuscript.qmd
ms: edit video legends

@@
 
-**Video 1.** This video shows a rotating 3D rendering of the aboral organ, featuring three distinct cell types—balancer cells, bridge cells, and components of the apical nerve net (ANN).
+**Video 1.** Morphological rendering of the aboral organ, featuring three distinct cell types—balancer cells, bridge cells, and components of the apical nerve net (ANN).
 Cells of each type are subdivided and color-coded according to their spatial regions.
@@
 
-**Video 2.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+**Video 2.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary bundles.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
@@
 
-**Video 3.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+**Video 3.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary bundles.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
@@
 
-**Video 4.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+**Video 4.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary clusters.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
@@
 
-**Video 5.** This video shows high-speed imaging of balancer ciliary beating along with kymographs from two individual balancer cilia.
+**Video 5.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary clusters.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).

### 2025-06-05 manuscript/Manuscript.qmd
update video and scale bars etc

@@
 The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4).
-Abbreviations: mi, mitochondrion; er, endoplasmic reticulum; sv, synaptic vesicles.\
+Abbreviations: mi, mitochondrion; er, endoplasmic reticulum; sv, synaptic vesicles.
+Scale bar: 500 nm.\
 (F) Position of synapses.
@@
 (B) Representative DIC images of the aboral organ viewed along the sagittal (left) and tentacular (right) planes.
-Arrowheads indicate the balancer cilia selected for kymograph-based analysis in each orientation.\
+Arrowheads indicate the balancer cilia selected for kymograph-based analysis in each orientation.
+Scale bar: 25 µm.\
 (C) Representative kymographs showing arrest and re-beat events of balancer cilia.
@@
 Some portions of the network remain as unconnected fragments due to incomplete reconstruction.\
-(E) Morphological comparison between the SNN (left) and ANN (right) in EM images.
-The SNN shows a "beads-on-a-string" morphology, with discrete bead-like swellings connected by faint electron-dense fibrous structures.
-In contrast, the ANN exhibits a more diffuse distribution, often appearing to interdigitate between surrounding cells.
+(E) Morphological comparison between the ANN in EM images. The SNN shows a "beads-on-a-string" morphology, with discrete bead-like swellings connected by faint electron-dense fibrous structures. \
+(F) Three-dimensional reconstruction of the SNN and ANNs.\
+(G) Morphological comparison between the SNN in EM images.
+The ANN exhibits a more diffuse distribution, often appearing to interdigitate between surrounding cells.
 Relatively large mitochondria are frequently observed in the ANN processes.](figures/Figure2_Supplement1.png)
@@
 
-**Video 2.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary bundles.
+**Video 2.** Visualization of balancer cilia extending from basal bodies, marked by red dots.
+Each red dot indicates the position of a basal body, from which the cilium projects distally.
+A lithocyte is shown in gray as a reference point, and the outline of the aboral organ is indicated in white.
+
+**Video 3.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary bundles.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
@@
 
-**Video 3.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary bundles.
+**Video 4.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary bundles.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
@@
 
-**Video 4.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary clusters.
+**Video 5.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary clusters.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).
@@
 
-**Video 5.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary clusters.
+**Video 6.** High-speed imaging of balancer ciliary beating along with kymographs from two individual balancer ciliary clusters.
 The recording was acquired at 100 frames per second and played back at 30 frames per second (0.3×).

### 2025-06-05 analysis/scripts/Figure2_figure_supplement1.R

@@ -95,10 +95,18 @@ panel_SNN_3d <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/plot_
   draw_label("A", x = 0.67, y = 0.85, size = 7, color = "black", hjust = 0.5) +
   draw_label("O", x = 0.67, y = 0.65, size = 7, color = "black", hjust = 0.5) +
   draw_line(x = c(0.67, 0.67), y = c(0.69, 0.81), color = "black", linewidth = 0.65) +
-  draw_label(expression(paste("25 ", mu, "m")), x = 0.95, y = 0.1, color = "black", size = 7, hjust = 0.5) +
-  draw_line(x = c(0.91, 0.99), y = c(0.05, 0.05), color = "black", linewidth = 0.7)
+  draw_label(expression(paste("25 ", mu, "m")), x = 0.95, y = 0.1, color = "black", size = 7.5, hjust = 0.5) +
+  draw_line(x = c(0.91, 0.99), y = c(0.05, 0.05), color = "black", linewidth = 0.9)
 
-panel_ANN_SNN_EM <- ggdraw() + draw_image(readPNG("manuscript/pictures/comparison_ANN_SNN.png"))
+panel_SNN_ANN_3d <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/plot_SNN_ANN.png")) +
+  draw_label("SNN", x = 0.25, y = 1, color = "skyblue", size = 10, hjust = 0.5) +
+  draw_label("ANN Q1-4", x = 0.35, y = 1, color = Okabe_Ito[5], size = 10, hjust = 0.5) +
+  draw_label("ANN Q1Q2", x = 0.5, y = 1, color = Okabe_Ito[6], size = 10, hjust = 0.5) +
+  draw_label("ANN Q3Q4", x = 0.65, y = 1, color = Okabe_Ito[7], size = 10, hjust = 0.5)
+
+panel_SNN_EM <- ggdraw() + draw_image(readPNG("manuscript/pictures/EM_SNN.png"))
+
+panel_ANN_EM <- ggdraw() + draw_image(readPNG("manuscript/pictures/EM_ANN.png"))
 
 panel_CAT_pic <- ggdraw() + draw_image(readPNG("manuscript/pictures/figure_sup2_1.png"))
 
[…]

### 2025-06-05 analysis/scripts/Video2.R

@@ -7,7 +7,7 @@ source("analysis/scripts/packages_and_functions.R")
 # load cell types to plot individually -------------------------------------
 
 balancer <- read_smooth_neuron("celltype:balancer")
-lithocyte <- read_smooth_neuron("celltype:lithocyte")
+lithocyte <- read_smooth_neuron("statolith")
 
 balancer_Q1 <- read_smooth_neuron(get_skids_with_annot(pid = 35, c("celltype:balancer", "Q1")))
 balancer_Q2 <- read_smooth_neuron(get_skids_with_annot(pid = 35, c("celltype:balancer", "Q2")))
@@ -82,12 +82,12 @@ plot_cells <- function(){
   
   plot3d(
     outline,
-    add = T, alpha = 0.05, col = "grey50"
+    add = T, alpha = 0.05, col = "grey80"
   )
   plot3d(
     lithocyte,
-    soma = T, lwd = 1, add = T,
-    alpha = 0.6, col = Okabe_Ito[8]
+    soma = T, lwd = 0, add = T,
+    alpha = 0.6, col = "grey50"
   )
   # input from SSN to balancer
   plot3d(
@@ -175,7 +175,7 @@ av::av_encode_video(
   paste('videoframes/', list.files("videoframes/", '*.png'), 
         sep = ""),
   framerate = 10,
-  output = 'manuscript/videos/supplemental video2.mp4'
+  output = 'manuscript/videos/Video2.mp4'
 )
 
 # delete temp folder -----------------

### 2025-06-06 manuscript/Manuscript.qmd
ms update

--- 2025-06-05 manuscript/Manuscript.qmd
@@
 subtitle: |
-  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewelow**$^{6}$, **Pawel Burkhardt**$^{7}$, **Gáspár Jékely**$^{1,2,6,*}$  
+  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Niederhuas**$^{6}$, **Pawel Burkhardt**$^{7}$, **Gáspár Jékely**$^{1,2,6,*}$  
   <br><br>  

### 2025-06-16 manuscript/Manuscript.qmd
MS: edits based on comments

@@
 The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm.
-We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016], resulting in a reconstruction of 972 cells that had intact nuclei and cell bodies, 899 of which formed the aboral organ (AO).
-Most of the 899 AO cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.
+We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016], resulting in a reconstruction of 972 cells that had intact nuclei and cell bodies, 898 of which formed the aboral organ (AO).
+Most of the 898 AO cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.
 
-In cells containing cilia, we also traced cilia along their length and annotated all basal bodies.
+In cells containing cilia, we also traced cilia along their length and annotated all basal bodies (Video 2.
+and not shown).
 For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced.
@@
 
-Classical neural staining techniques did not provide clear images of neurons at the aboral organ.
-However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a].
+Immunofuorescent stainings in *Pleurobrachia pileus* revealed a nerve net at the base of the aboral organ that seemed continuous with the polygonal nerve net of the surrounding epithelium [@Jager2011].
+Ultrastructural studies revealed synapses near the floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a].
 In our dataset, we identified 428 synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion.
@@
 These neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023].
-Our results reveal a second type of syncytial neurons in ctenophores (Supplementary Figure S2D).
+ANN neurons also lack dense core vesicles, structures that are abundant throughout the cell bodies and neurites of SNN neurons [@Burkhardt2023].
+Our results reveal a second type of syncytial neurons in ctenophores (Figure 2--figure supplement C-G).
 
@@
 The first type is a single large neuron (ANN_Q1-4) that spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei.
-This neuron has 150 presynaptic sites.
+This neuron has 150 presynaptic sites (Figure 2C-F).
 The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and containing two nuclei.
-ANN_Q1Q2 had 119 presynaptic structures, whereas ANN_Q3Q4 had 68 presynaptic structures.
+ANN_Q1Q2 had 119 presynapses, whereas ANN_Q3Q4 had 68 presynaptic structures (Figure 2C-F).
 
@@
 To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector.
-We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage.
+We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage (Figure 4A).
 
@@
 
-![**Figure 2—figure supplement 1. Ultrastructural features and synaptic organization of the ANN, with comparison to the subepithelial nerve net (SNN)**\
+![**Figure 2—figure supplement 1. Ultrastructural features and synaptic organization of the ANN, in comparison to the subepithelial nerve net (SNN)**\
 (A) Registration and annotation of synapses in the actual cellular traces.
@@
 On the left, the ANN_Q1-4 cell extends widely across the aboral organ and appears as a continuous structure in 3D, despite having nuclei located distantly from each other.
-In contrast, ANN_Q1Q2 (right) contains two adjacent nuclei situated close together, suggesting a more localized distribution.\
+In contrast, ANN_Q1Q2 (right) has two adjacent nuclei.\
 (D) Three-dimensional reconstruction of the subepithelial nerve net (SNN).

### 2025-06-23 manuscript/Manuscript.qmd
MS: edit
Fig 1: fix n cells, ms edit

@@
 subtitle: |
-  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Niederhuas**$^{6}$, **Pawel Burkhardt**$^{7}$, **Gáspár Jékely**$^{1,2,6,*}$  
+  **Kei Jokura**$^{1,2,3,4,5*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Niederhaus**$^{6}$, **Pawel Burkhardt**$^{7}$, **Gáspár Jékely**$^{1,2,6,*}$  
   <br><br>  

### 2025-06-23 analysis/scripts/Figure1.R

@@ -46,10 +46,11 @@ close3d()
 
 panel_catmaid_overview <- ggdraw() + draw_image(readPNG("manuscript/pictures/overview.png")) +
   draw_label("serial EM volume", x = 0.5, y = 0.99, size = 10, fontface = "plain", hjust = 0.5) +
-  draw_label("619 sections", x = 0.025, y = 0.81, color = "black", size = 9, hjust = 0) +
-  draw_label("927 cells", x = 0.025, y = 0.75, color = "black", size = 9, hjust = 0) +
+  draw_label("619 sections", x = 0.025, y = 0.86, color = "black", size = 9, hjust = 0) +
+  draw_label("972 cells", x = 0.025, y = 0.8, color = "black", size = 9, hjust = 0) +
+  draw_label("898 AO cells", x = 0.025, y = 0.74, color = "black", size = 9, hjust = 0) +
   draw_line(x = c(0.75, 0.9), y = c(0.05, 0.05), color = "black", size = 1) +
-  draw_label(expression(paste("10 ", mu, "m")), x = 0.825, y = 0.09, color = "black", size = 8, hjust = 0.5)
+  draw_label(expression(paste("10 ", mu, " m")), x = 0.825, y = 0.09, color = "black", size = 8, hjust = 0.5)
 
 
 panel_quadrants <- ggdraw() + draw_image(readPNG("manuscript/pictures/quadrants.png")) +
@@ -61,7 +62,7 @@ panel_quadrants <- ggdraw() + draw_image(readPNG("manuscript/pictures/quadrants.
   draw_label("lateral view", x = 0.66, y = 0.99, color = "black", size = 10, hjust = 0.5) +
   draw_label("sagittal plane", x = 0.5, y = 0.95, color = "black", size = 10, hjust = 0.5) +
   draw_label("tentacular plane", x = 0.83, y = 0.95, color = "black", size = 10, hjust = 0.5)  +
-  draw_label(expression(paste("25 ", mu, "m")), x = 0.95, y = 0.1, color = "black", size = 10, hjust = 0.5) +
+  draw_label(expression(paste("25 ", mu, " m")), x = 0.95, y = 0.1, color = "black", size = 10, hjust = 0.5) +
   draw_line(x = c(0.91, 0.99), y = c(0.05, 0.05), color = "black", linewidth = 1) +
   draw_label("A", x = 0.66, y = 0.25, size = 10, color = "black", hjust = 0.5) +
[…]

### 2025-06-24 manuscript/Manuscript.qmd
MS: edits of synapse description

@@
 To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*.
-We reconstructed 909 cells, including syncytial neurons that form a nerve net.
+We reconstructed 972 cells, including syncytial neurons that form a nerve net.
 The syncytial neurons synapse on the balancer cells and also form reciprocal connection with the bridge cells that span the statocyst.
@@
 However, without a detailed understanding of how ctenophore neurons are spatially organized and interconnected, it remains unclear how their nervous system controls behaviour.
-Clarifying this relationship between neural architecture and behavioural control is essential to uncovering the logic of ctenophore neural function.
+Clarifying this relationship between neural architecture and behavioural control is essential to uncovering ctenophore neural function.
 
 To address the lack of structural, circuit-level understanding of the ctenophore nervous system, we focused on a unique sensory structure—the aboral organ.
-This gravity-sensing organ contains a statocyst supported by four mechanosensory balancer cells, which act as pacemakers for the beating of comb-plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980].
+This gravity-sensing organ contains a statocyst formed by lithocytes and four clusters of mechanosensory balancer cells, which act as pacemakers for the beating of comb-plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980].
 Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011].
+
 A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency.
@@
 
-In cells containing cilia, we also traced cilia along their length and annotated all basal bodies (Video 2.
-and not shown).
+In cells containing cilia, we also traced cilia along their length and annotated all basal bodies (Video 2 and not shown).
 For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced.
@@
 This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development [@Martindale1999].
-We assigned each of the 899 skeletonised cells to one of the four quadrants (Figure 1G).
-<!-- check number of cells, 927 in figure -->
+We assigned each of the 898 skeletonised AO cells to one of the four quadrants (Figure 1G).
 
@@
 We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology (Figure 2A and B).
-Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Supplementary Figure S2C).
-These neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023].
+Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Figure 2--figure supplement 1C).
+ANN neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023].
 ANN neurons also lack dense core vesicles, structures that are abundant throughout the cell bodies and neurites of SNN neurons [@Burkhardt2023].
-Our results reveal a second type of syncytial neurons in ctenophores (Figure 2--figure supplement C-G).
+In our dataset, we also reconstructed 10 SNN neurons.
+These cells showed a blebbed morphology, contained several dense core vesicles and were located at the periphery of the AO (Figure 2--figure supplement 1).
+Our results reveal a second type of syncytial neuron in ctenophores (Figure 2--figure supplement C-G).
 
@@
 ANN_Q1Q2 had 119 presynapses, whereas ANN_Q3Q4 had 68 presynaptic structures (Figure 2C-F).
+SNN neurons had no or only few synapses.
 
@@
 
-Our synapse annotation revealed synapses between the ANN neurons and synapses that ANN_Q1-4 forms on itself (autapses). The ANN neurons also form synapses on the gravity-sensing balancer cells.
+Our synapse annotation revealed several types of synaptic connections within the gravisensory organ.
+We identified synapses between different ANN neurons and synapses that ANN_Q1-4 forms on itself (autapses).
+The ANN neurons also form synapses on the gravity-sensing balancer cells and the bridge cells.
+Bridge cells, but not balancer cells, also synapse on the ANN.
 
@@
 In each quadrant, there were between 28-37 balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28).
-Each cell contained 3 to 10 <!--# check --> mitochondria.
+Each cell contained 3 to 10 mitochondria.
 
@@
 
-Bridge cells had presynaptic sites with the typical presynaptic triad structure near 30% of their mitochondria.
+Nearly all bridge cells (25/26) received synaptic input from ANNs.
+Bridge cells in Q1Q2 received inputs from ANN_Q1Q2 (11 cells), ANN_Q1-4 (1 cell), or both (2 cells).
+In Q3Q4, synapses on bridge cells were from ANN_Q3Q4 (1 cell), ANN_Q1-4 (7 cells), or both (1 cell).
+
+Bridge cells also had presynaptic sites with the typical presynaptic triad structure near 30% of their mitochondria.
 These bridge-cell synapses were formed on ANN neurons or other bridge cells.
@@
 Bridge cells in Q3Q4 synapsed on ANN_Q3Q4 (1 cell) or ANN_Q1-4 (6 cells).
-
-Nearly all bridge cells (25/26) also received synaptic input from ANNs.
-Bridge cells in Q1Q2 received inputs from ANN_Q1Q2 (11 cells), ANN_Q1-4 (1 cell), or both (2 cells).
-In Q3Q4, synapses on bridge cells were from ANN_Q3Q4 (1 cell), ANN_Q1-4 (7 cells), or both (1 cell).
 
@@
 
-The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, EF cells and several non-ciliated, monociliated or biciliated cells.
-These cell types and synaptic connections are outside the gravisensory organ and will be described elsewhere.
+The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, EF cells and several non-ciliated, monociliated or biciliated cells (Figure 3--figures supplement 1).
+These cell types and synaptic connections are outside the gravisensory organ and are not considered further here.
 

### 2025-06-24 analysis/scripts/Figure2.R
rename bristle cells to large_glanular_cells and update tables, fig2

@@ -326,7 +326,7 @@ df <- df %>%
 rename_map <- c(
   "balancer" = "bal", 
   "bridge" = "brg", 
-  "bristle" = "bsl", 
+  "large_granular_cell" = "lgc", 
   "Cgroove" = "cg",
   "dense_vesicle" = "dv", 
   "dome" = "do", 
@@ -361,7 +361,7 @@ mean_summary <- summary_df %>%
 
 
 # Align the number of cells (n) of each celltype with celltype_order
-celltype_order <- c("bal", "brg", "bsl", "cg", "dv", "do", "imc", "la", "li", "pl", "ANN", "ef")
+celltype_order <- c("bal", "brg", "lgc", "cg", "dv", "do", "imc", "la", "li", "pl", "ANN", "ef")
 
 n_counts <- summary_df %>%
   group_by(celltype) %>%

### 2025-06-24 analysis/scripts/Figure3.R
bristle to larage glanular rename, update Fig3

@@ -101,12 +101,12 @@ SSN_downstream <- SSN_downstream %>%
 
 
 # bar plot
-all_celltypes <- c("balancer", "bridge", "bristle", "Cgroove", "dense_vesicle", "dome", 
+all_celltypes <- c("balancer", "bridge", "large_glanular_cell", "Cgroove", "dense_vesicle", "dome", 
                    "intra-multi-ciliated", "lamellate", "lithocyte", "plumose", "SSN", 
                    "epithelial_floor")
 
 label_mapping <- c(
-  "balancer" = "bal", "bridge" = "brg", "bristle" = "bsl", 
+  "balancer" = "bal", "bridge" = "brg", "large_glanular_cell" = "lgc", 
   "Cgroove" = "cg", 
   "dense_vesicle" = "dv", "dome" = "do", 
   "intra-multi-ciliated" = "imc", "lamellate" = "la", "lithocyte" = "li", 

### 2025-06-24 manuscript/Manuscript.qmd
MS and fig: change bristle to large granular cell

@@
 N \> 2 cells for each cell type except lithocytes.
-Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cells; imc, intra-multiciliated cells; la, lamellate bodies; li, lithocytes; pl, plumose; ef, epithelial floor cells.\
+Abbreviations: bal, balancer; brg, bridge; lgc, large granular cell; cg, ciliated groove; dv, dense vesicle cells; imc, intra-multiciliated cells; la, lamellate bodies; li, lithocytes; pl, plumose; ef, epithelial floor cells.\
 (D) Position of mitochondria within ANN_Q1-4.
@@
 (A) Number of synaptic inputs from the ANNs (ANN_Q1-4: blue, ANN_Q1Q2: orange, ANN_Q3Q4: pink) to each cell type.
-Abbreviations: bal, balancer; brg, bridge; bsl, bristle; cg, ciliated groove; dv, dense vesicle cell; imc, intra-multiciliated cell; la, lamellate body; li, lithocyte; pl, plumose cell; ef, epithelial floor cell.\
+Abbreviations: bal, balancer; brg, bridge; lgc, large granular cell; cg, ciliated groove; dv, dense vesicle cell; imc, intra-multiciliated cell; la, lamellate body; li, lithocyte; pl, plumose cell; ef, epithelial floor cell.\
 (B) Morphological rendering of balancer cells, with each cell color-coded by quadrant: Q1 (yellow), Q2 (blue), Q3 (orange), and Q4 (magenta).
@@
 
-We assigned each cell a specific cell-type name based on its category (balancer, bridge, bristle, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types.
+We assigned each cell a specific cell-type name based on its category (balancer, bridge, large granular, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types.
 Some cells only had more general features and were classified based on the presence/absence and number of cilia into four ciliated cell types:(biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)).
@@
 ![**Figure 1—figure supplement 1. Cell-type composition of the ctenophore aboral organ**\
-Morphological rendering of skeletonised cell types, (A) balancer cells (B) bridge cells, (C) bristle cells, (D) ciliated groove cells, (E) dense vesicle cells, (F) dome cells, (G) intra-multiciliated cells, (H) lamellate bodies, (I) lithocytes, (J) plumose cells, (K) aboral nerve-net neurons (L) epithelial floor cells.](figures/Figure1_Supplement1.png)
+Morphological rendering of skeletonised cell types, (A) balancer cells (B) bridge cells, (C) large granular cells, (D) ciliated groove cells, (E) dense vesicle cells, (F) dome cells, (G) intra-multiciliated cells, (H) lamellate bodies, (I) lithocytes, (J) plumose cells, (K) aboral nerve-net neurons (L) epithelial floor cells.](figures/Figure1_Supplement1.png)
 

### 2025-06-24 analysis/scripts/Figure1_figure_supplement1.R

@@ -193,7 +193,7 @@ panel_balancer <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/plo
 panel_bridge <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/plot_bridge.png")) +
   draw_label("bridge cells", x = 0.5, y = 0.95, size = 8, fontface="bold", hjust = 0.5)
 panel_bristle <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/plot_bristle.png")) +
-  draw_label("bristle cells", x = 0.5, y = 0.95, size = 8, fontface="bold", hjust = 0.5)
+  draw_label("large granular cells", x = 0.5, y = 0.95, size = 8, fontface="bold", hjust = 0.5)
 panel_Cgroove <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/plot_Cgroove.png")) +
   draw_label("ciliated groove cells", x = 0.5, y = 0.95, size = 8, fontface="bold", hjust = 0.5)
 panel_dense_vesicle <- ggdraw() + draw_image(readPNG("manuscript/pictures/3d_plot/plot_dense_vesicle.png")) +

### 2025-06-25 manuscript/Manuscript.qmd
MS: reference all supllementary figures and videos in main text

@@
 Most of the 898 AO cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.
+We classified these cells into cell types (Figure 1--figure supplement 1, Figure 1--figure supplement 2).
 
@@
 These cells showed a blebbed morphology, contained several dense core vesicles and were located at the periphery of the AO (Figure 2--figure supplement 1).
-Our results reveal a second type of syncytial neuron in ctenophores (Figure 2--figure supplement C-G).
+Our results reveal a second type of syncytial neuron in ctenophores (Figure 2--figure supplement 1 C-G).
 
@@
 
-Balancer cells are monociliated cells with long motile cilia that form four bundles of compound cilia, one in each quadrant.
+Balancer cells are monociliated cells with long motile cilia that form four bundles of compound cilia, one in each quadrant (Video 2).
 These bundles come together at the centre of the aboral organ to support the cellular mass of the statolyth.
@@
 Bridge cells form two distinct cell groups across the sagittal plane, in the Q1Q2 and Q3Q4 regions.
-In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12 cells (Figure 3 -- figure supplement 1).
+In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12 cells (Figure 3 -- figure supplement 1, Video 1).
 
@@
 
-The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, EF cells and several non-ciliated, monociliated or biciliated cells (Figure 3--figures supplement 1).
+The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, epithelial floor cells and several non-ciliated, monociliated or biciliated cells (Figure 3--figure supplement 1).
 These cell types and synaptic connections are outside the gravisensory organ and are not considered further here.
@@
 We found that arrest and re-beat events were synchronized between balancers across the sagittal plane.
-However, while the timing of re-beat was also near-simultaneous along the tentacular plane, arrest timing was offset by up to 2.16 seconds (Figure 4D).
+However, while the timing of re-beat was also near-simultaneous along the tentacular plane, arrest timing was offset by up to 2.16 seconds (Figure 4D, Videos 3-6).
 Overall, these data reveal that arrests are only coordinated between Q1-Q2 and Q3-Q4, whereas re-beat is coordinated over all four quadrants.
 
-As shown in Figure 4E, beat frequencies of left and right balancers fluctuated substantially in both orientations.
+As shown in Figure 4E (and Figure 4--figure supplement 1), beat frequencies of left and right balancers fluctuated substantially in both orientations.
 However, while left-right activity was largely synchronized in sagittal pairs (Figure 4E, left), tentacular pairs (Figure 4E, right) showed strikingly periodic alternations between left and right balancer activity.
@@
 
-![**Figure 4—figure supplement 1.Ciliary beat frequency (CBF) dynamics of balancer cells in additional samples**\
+![**Figure 4—figure supplement 1. Ciliary beat frequency (CBF) dynamics of balancer cells in additional samples**\
 Bar plots showing CBF dynamics of left and right balancers (top and bottom panels, respectively) recorded from the sagittal (A–C) and tentacular (D–F) planes.

### 2025-06-25 analysis/scripts/Figure1_figure_supplement1.R
Fig 1 suppl1: Change bristle cell name, clean up code

@@ -4,33 +4,65 @@
 source("analysis/scripts/packages_and_functions.R")
 
 
-# load all cell types to plot individually -------------------------------------
-#twelve celltypes
-balancer <- read_smooth_neuron("celltype:balancer")
-bridge <- read_smooth_neuron("celltype:bridge")
-bristle <- read_smooth_neuron("celltype:bristle")
-Cgroove <- read_smooth_neuron("celltype:Cgroove")
-dense_vesicle <- read_smooth_neuron("celltype:dense_vesicle")
-dome <- read_smooth_neuron("celltype:dome")
-intra_multi_ciliated <- read_smooth_neuron("celltype:intra-multi-ciliated")
-lamellate <- read_smooth_neuron("celltype:lamellate")
-lithocyte <- read_smooth_neuron("celltype:lithocyte")
-plumose <- read_smooth_neuron("celltype:plumose")
-SSN <- read_smooth_neuron("celltype:SSN")
-epithelial_floor <- read_smooth_neuron("celltype:epithelial_floor")
+# show just a subset of celltypes, just the interesting ones
 
-with_soma <- read_smooth_neuron("with_soma")
+celltype_map <- c(
+  "balancer" = "balancer cells",
+  "bridge" = "bridge cells",
+  "large_granular_cell" = "large granular cells",
+  "Cgroove" = "ciliated groove cells",
+  "dense_vesicle" = "dense vesicle cells",
+  "dome" = "dome cells",
+  "intra-multi-ciliated" = "intracellular multiciliated cells",
+  "lamellate" = "lamellate bodies",
+  "lithocyte" = "lithocytes",
+  "plumose" = "plumose cells",
+  "SSN" = "aboral nerve net neurons",
+  "epithelial_floor" = "epithelial floor cells"
+)
 
-# function individual plot process --------------------------------------------------------
+# celltypes short, because these are not all celltypes in the animal
+celltypes_short <- names(celltype_map)
 
-plot_neuron_views <- function(neuron_name, neuron_data) {
+
+skids_with_soma <- catmaid_get_label_stats(pid=35) %>%
+  filter(labelName == "soma") %>%
+  select(skeletonID) %>%
+  pull() %>%
+  unique()
+
[…]

### 2025-07-08 manuscript/Manuscript.qmd
edit main text (Results, Discussion and Materials and Methods)

@@
 For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*.
-To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon).
-The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons.
-The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy).
+To ensure optimal ultrastructural preservation, the specimen was prepared by high-pressure freezing, followed by freeze substitution and Epon embedding.
+From the aboral tip, we collected \~1,000 ultra-thin (50 nm) serial sections as ribbons, targeting only the region containing the aboral organ.
+These were imaged using a Zeiss Gemini 500 SEM at a resolution of 2.0 nm/pixel (xy).
 We imaged only the region containing the aboral organ.
@@
 The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm.
-We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016], resulting in a reconstruction of 972 cells that had intact nuclei and cell bodies, 898 of which formed the aboral organ (AO).
-Most of the 898 AO cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.
+We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016].
+We reconstructed 972 cells with intact nuclei and somata.
+Among these, 898 were located within the aboral organ (AO), including balancers and the aboral nerve net neurons.
+Due to the compact and mostly self-contained organization of the AO, these cells were largely contained within the imaged volume.
 We classified these cells into cell types (Figure 1--figure supplement 1, Figure 1--figure supplement 2).
@@
 All nodes relevant to synaptic sites were tagged, and skeletons were named and assigned multi-level annotations.
-As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023].
-In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees).
+As described later, some neurons formed loop-like structures (anastomosed neurons), in which separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023].
+Since CATMAID only supports skeleton trees, branch nodes in such cases were placed near the closest existing node and annotated accordingly.
 The entire skeletonized volume was composed of 134,591 nodes.
@@
 
-Immunofuorescent stainings in *Pleurobrachia pileus* revealed a nerve net at the base of the aboral organ that seemed continuous with the polygonal nerve net of the surrounding epithelium [@Jager2011].
-Ultrastructural studies revealed synapses near the floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a].
-In our dataset, we identified 428 synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion.
-At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Figure 2--figure supplement 1A).
-We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise1973a].
-Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Figure 2--figure supplement 1B).
+Previous immunofluorescent staining in *Pleurobrachia pileus* suggested the presence of a nerve net at the base of the aboral organ, possibly continuous with the surrounding epithelial nerve net [@Jager2011].
+Electron microscopy studies later revealed synapses near the floor of the aboral organ, a region where densely packed epithelial cells form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a].
 
-We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology (Figure 2A and B).
-Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Figure 2--figure supplement 1C).
-ANN neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023].
-ANN neurons also lack dense core vesicles, structures that are abundant throughout the cell bodies and neurites of SNN neurons [@Burkhardt2023].
-In our dataset, we also reconstructed 10 SNN neurons.
-These cells showed a blebbed morphology, contained several dense core vesicles and were located at the periphery of the AO (Figure 2--figure supplement 1).
-Our results reveal a second type of syncytial neuron in ctenophores (Figure 2--figure supplement 1 C-G).
+In our volume EM dataset, we identified 428 synapses based on the canonical presynaptic triad morphology: a cluster of synaptic vesicles, a smooth endoplasmic reticulum cisterna, and a mitochondrion.
+At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane, as indicated by light blue arrows in CATMAID (Figure 2--figure supplement 1A).
+No specialized postsynaptic densities were observed, consistent with prior reports [@Hernandez_Nicaise1973a].
+Synapses were either monoadic (one postsynaptic target) or polyadic (multiple postsynaptic targets) (Figure 2--figure supplement 1B).
+
+We reconstructed three aboral nerve-net (ANN) neurons, each exhibiting a syncytial morphology characterized by anastomosing membranes and multiple nuclei (ranging from two to five) (Figure 2A and B, Figure 2--figure supplement 1C).
+These ANN neurons are morphologically distinct from the previously described syncytial subepithelial nerve net (SNN) neurons in the body wall of *Mnemiopsis*, which show a blebbed morphology and contain abundant dense core vesicles [@Burkhardt2023].
+Our results identify a second type of syncytial neuron in *Mnemiopsis*.
+In addition to the previously characterized SNN neurons in the body wall, we observed ANN neurons within the aboral organ.
+Unlike SNNs, which exhibit a blebbed morphology and contain numerous dense core vesicles, ANN neurons are smooth-surfaced and lack dense vesicles.
+ANN neurons were located centrally within the aboral organ, while SNN neurons were reconstructed near its periphery (Figure 2--figure supplement 1C–G).
 
 Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023].
-Based on their distinct size and position, we classified the ANNs into two types.
-The first type is a single large neuron (ANN_Q1-4) that spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei.
-This neuron has 150 presynaptic sites (Figure 2C-F).
-The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and containing two nuclei.
-ANN_Q1Q2 had 119 presynapses, whereas ANN_Q3Q4 had 68 presynaptic structures (Figure 2C-F).
-SNN neurons had no or only few synapses.
+Based on soma size and spatial distribution, we classified ANN neurons into two types.
+The first is a single large neuron, ANN_Q1-4, spanning all four quadrants and containing four (possibly five) nuclei.
+This neuron formed 150 presynaptic sites.
+The second type includes two smaller neurons, ANN_Q1Q2 and ANN_Q3Q4, each confined to two adjacent quadrants and containing two nuclei.
+ANN_Q1Q2 had 119 presynapses, while ANN_Q3Q4 had 68 (Figure 2C–F).
+In contrast, SNN neurons exhibited either no synaptic contacts or only a few.
 
@@
 
-Our synapse annotation revealed several types of synaptic connections within the gravisensory organ.
+Synapse annotation revealed multiple types of connections involving ANN neurons within the gravisensory organ, including reciprocal ANN–bridge cell connections, ANN autapses, and outputs to balancer cells.
 We identified synapses between different ANN neurons and synapses that ANN_Q1-4 forms on itself (autapses).
@@
 Some balancer cells received inputs from both ANN_Q1-4 and either of ANN_Q1Q2 or ANN_Q3Q4.
-While previous studies suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez_Nicaise1974], our data did not reveal any presynaptic sites in balancer cells (Figure 3 -- figure supplement 1).
+Notably, we found no presynaptic sites in balancer cells (Figure 3 -- figure supplement 1), contradicting earlier suggestions of afferent signaling from balancer cells to neurons [@Hernandez_Nicaise1974].
 
@@
 In the Q1Q2 region, we identified 14 bridge cells, in Q3Q4, 12 cells (Figure 3 -- figure supplement 1, Video 1).
+
+In contrast to balancer cells, bridge cells participated in reciprocal synaptic interactions with ANN neurons, forming both pre- and postsynaptic connections.
 
@@
 
-To analyze the synaptic connectivity graph of the balancer organ and the ANN, we grouped cells of the same type and in the same region into a single node, summed the number of synapses, and laid out the network to reflect the anatomy of the four quadrants (Figure 3H).
-The network is characterised by feedback connections among ANN neurons and between ANN and bridge cells with a clear rotational (or mirror) symmetry.
-Q1Q2 and Q3Q4 also form separate local circuits centered around their regional ANN neurons while ANN_Q1-4 mediates global connectivity across the entire balancer organ.
-Notably, there were no synapses from the mechanosensory balancer cells to the ANN or from ANN to motor cells (e.g.
-ciliary groove), contrary to our initial expectation of a sensory-motor or input-output model of balancer function.
+To examine the overall synaptic organization of the balancer organ, we grouped cells of the same type and region into network nodes, summed the synaptic counts, and mapped the network onto the four-quadrant anatomy (Figure 3H).
+The resulting connectivity graph revealed feedback loops among ANN neurons and between ANN and bridge cells, displaying rotational symmetry.
+Local sub-circuits centered around ANN_Q1Q2 and ANN_Q3Q4 were evident, while ANN_Q1-4 formed global connections across quadrants.
+
+Surprisingly, and contrary to the expected sensory–motor or input–output model of balancer function, we found no synapses from the mechanosensory balancer cells to ANN neurons, nor from ANN neurons to known motor effectors such as ciliated groove cells.
 
@@
 
-To investigate the regulation of balancer function, we carried out a high-speed video-microscope analysis of balancer cilia.
+To investigate whether balancer cilia are coordinated by neural inputs, we performed high-speed video microscopy to analyze their activity patterns across body axes.
 Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination.
@@
 
-During the two-minute recordings, balancer cilia could beat fast, slow, or exhibit abrupt stops (arrest) and start moving again (re-beat).
+During the two-minute recordings, balancer cilia could beat fast, slow, or exhibit abrupt stops of beating ("arrest") followed by re-initiation ("re-beat").
 Occasionally, large body-contractions moved the entire cydippid out of frame, and data from these episods were excluded.
@@
 As shown in Figure 4E (and Figure 4--figure supplement 1), beat frequencies of left and right balancers fluctuated substantially in both orientations.
-However, while left-right activity was largely synchronized in sagittal pairs (Figure 4E, left), tentacular pairs (Figure 4E, right) showed strikingly periodic alternations between left and right balancer activity.
-This pattern suggested an offset in timing during arrest or re-beat events between tentacular balancers.
+However, while left–right activity was largely synchronized in sagittal pairs (Figure 4E, left), tentacular pairs (Figure 4E, right) displayed strikingly periodic alternations between left and right balancer activity.
+This pattern suggests a consistent temporal offset between the two sides, possibly arising from differences in the timing of arrest or re-beat events.
 
 To quantify this, we calculated a rolling Pearson correlation between left and right balancer activity in each plane.
-The resulting mean correlations were significantly higher in the tentacular plane compared to the sagittal plane (Figure 4F; t-test, p = 0.014).
+The resulting mean correlations were significantly higher in the tentacular plane compared to the sagittal plane (Figure 4F; t-test, *p* = 0.014).
 This result indicates that tentacular balancers beat in a more coordinated, mirror-symmetric manner, whereas sagittal balancers exhibit greater variability between sides.
@@
 This morphology suggests a potential role for bridge neurons in modulating inter-quadrant coordination.
-By linking ANN activity across orthogonal planes, these neurons may contribute to the dynamic regulation of CBF patterns across the whole system.
+By linking ANN activity across orthogonal planes, these neurons may contribute to the dynamic regulation of CBF patterns by facilitating information flow across orthogonal planes, thus supporting coordinated activity throughout the aboral organ.
 
@@
 Each dot represents an individual larva.
-Balancers on the tentacular plane showed significantly higher correlation than those on the sagittal plane (t-test, p = 0.014).](figures/Figure4.png)
+Balancers on the tentacular plane showed significantly higher correlation than those on the sagittal plane (t-test, *p* = 0.014).](figures/Figure4.png)
 
@@
 The ANN has a morphology distinct from that of the subepithelial nerve net (SNN) reported for one-day-old cydippids [@Burkhardt2023; @Sachkova2021].
+A recent study also identified a distinct subtype of SNN neurons in the aboral region, termed “condensed” SNN, which display a more compact morphology compared to body wall SNNs [@Ferraioli2025].
+However, the ANN neurons described here are morphologically and anatomically distinct from both the body wall SNN and the condensed SNN.
 Instead of forming discrete nodal swellings or a “beads-on-a-string” organization typical for the SNN, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion.
@@
 
-Our findings reveal a circuit composed of ANN neurons and bridge cells that coordinates ciliary beating rhythms in space and time without sensory input, and features a recursive architecture in which ciliary outputs may feed back to the circuit via bridge cells.
+Our findings reveal a circuit composed of ANN neurons and bridge cells that coordinates ciliary beating rhythms in space and time without direct sensory input, and features a recursive architecture in which ciliary outputs may feed back to the circuit via bridge cells.
 Such a configuration suggests that even with a limited number of cells, the system is capable of flexible and robust rhythm control.
@@
 The network we describe here orchestrates output through dynamic, self-regulating coordination.
-Future work could uncover this dynamics, visualized through live imaging of the entire cell ensemble, to offer a window into the core functions of a unique nervous system.
+Future work could uncover these dynamics, visualized through live imaging of the entire cell ensemble, to offer a window into the core functions of a unique nervous system.
 
@@
 The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica).
+
 Cryo-substitution was carried out by gradually raising the temperature from -90°C to -70°C over 4 hours, then returning to -20°C over 2 hours, and finally to room temperature over 2 hours.
 The samples were then embedded in epoxy resin (EMbed 812, Electron Microscopy Sciences).
+
 Serial ultrathin sections of 50 nm thickness were prepared using a Leica UC7 ultramicrotome and a 45º DiATOME diamond knife.
-To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the surface negatively charged.
+A conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.) to enhance section adhesion and improve hydrophilicity.
+This process rendered the glass surface negatively charged.
+
 Section ribbons were collected on prepared glass slides, gently dried to ensure proper stretching, and firmly adhered to the glass surface.
@@
 The glass slides were mounted on a STEM-specific stage (Zeiss) using Copper Foil EMI Shielding Tape (3M).
+
 Imaging was performed using a Gemini SEM 500 (Zeiss), equipped with SmartSEM and Atlas 5 imaging software (Zeiss), using the Inlens detector.
@@
 
-We assigned each cell a specific cell-type name based on its category (balancer, bridge, large granular, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types.
-Some cells only had more general features and were classified based on the presence/absence and number of cilia into four ciliated cell types:(biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)).
-Additionally, we appended the quadrant number to the cell type name after an underscore ("\_") to indicate the cell-body’s location.
-For example, cells located in the first quadrant are named with "\_Q1." If a cell is situated between the first and second quadrants, we appended "Q1Q2" to the cell-type name.
+We assigned each cell a specific cell-type name based on its morphological or functional category (e.g., balancer, bridge, large granular, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, and ANN), resulting in a total of 12 primary cell types.
+
+In addition, some cells lacked distinctive features and were instead categorized into one of four general ciliated types, based on the presence and number of cilia: biciliated (biC), monociliated (monoC), multiciliated (multiC), or non-ciliated (non-C).
+
+To indicate spatial positioning, we appended the quadrant number to the cell-type name using an underscore.
+For instance, cells located in the first quadrant were labeled "\_Q1," while those situated between two quadrants were labeled with both (e.g., "Q1Q2").
+
+Cells of the same type and in the same region were further distinguished by serial numbering, ensuring each cell has a unique identifier.
 
@@
 A coverslip with a thin layer of Vaseline applied to its two edges was gently placed over a glass slide.
-Filtered natural seawater and the cydippids were introduced beneath the coverslip.
-The orientation of the larvae was adjusted by carefully sliding the coverslip, and slight pressure was then applied to immobilize the larvae.
+Filtered natural seawater and cydippids were placed beneath the coverslip.
+The orientation of the larvae was adjusted by carefully sliding the coverslip, and the larvae were gently immobilized by applying slight pressure.
+
+To ensure consistency in gravitational input, we included only larvae whose aboral–oral axis was oriented nearly perpendicular to the gravitational vector (tilted by no more than 20°).
+This minimized the potential influence of larval tilt on left–right differences in balancer cilia activity, as all balancers received comparable stimulation from the statolith.
 
 Balancer cilia movement was imaged using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a high-sensitivity monochrome CMOS camera optimized for the near-infrared range (UI-3360CP-NIR-GL Rev.2, iDS), and a 40x glycerine-immersion objective lens (LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27).
-To stabilize the statolith, the microscope was tilted by 90° to achieve vertical stage alignment.
+To ensure consistent statolith positioning during imaging, the microscope was tilted by 90° to achieve vertical stage alignment.
 Illumination was provided by a custom-made NIR LED strobe system (850 nm) synchronized with the camera’s exposure signals.
@@
 For beat frequency analysis, kymographs were generated using the Multi Kymograph plugin in Fiji.
-One-second segments were manually extracted from the kymographs, and ciliary beat frequencies were counted visually for each ROI.
+One-second segments were manually extracted from the kymographs, and ciliary beat frequencies were visually counted for each manually defined region of interest (ROI).
 These manually determined frequencies were then used for subsequent computational analysis.
 
-To process the beat frequency time series data, left and right cilia traces were extracted, filtered to remove missing or non-finite values, and assigned frame indices.
+Traces of left and right cilia were extracted, cleaned to remove missing or non-finite values, and indexed by frame.
 Data were saved as RDS objects for downstream analysis.
@@
 Baba, and Dr.
-Kogiku Shiba (Shimoda Marine Research Center) for their assistance with microscopy.
+Kogiku Shiba for their assistance with microscopy.
+We also thank Dr.
+Sidney L.
+Tamm and Dr.
+Steven H.
+D.
+Haddock for their valuable feedback on the manuscript.
 
@@
 Some portions of the network remain as unconnected fragments due to incomplete reconstruction.\
-(E) Morphological comparison between the ANN in EM images. The SNN shows a "beads-on-a-string" morphology, with discrete bead-like swellings connected by faint electron-dense fibrous structures. \
+(E) Morphological comparison between the ANN in EM images. The SNN shows a "beads-on-a-string" morphology, with discrete bead-like swellings connected by faint electron-dense fibrous structures.\
 (F) Three-dimensional reconstruction of the SNN and ANNs.\

### 2025-07-08 manuscript/Manuscript.qmd
edit again

@@
 To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector.
+In all cases, only larvae whose aboral–oral axis was approximately parallel to the gravity vector (mouth facing downward within a 0–20° deviation) were included for analysis.
+This ensured that all balancer cilia experienced comparable gravitational input from the statolith, allowing for an unbiased comparison across quadrants.
 We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage (Figure 4A).

### 2025-07-09 manuscript/Manuscript.qmd
ed method

@@
 Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioural patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system.
-Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011].
+Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of the nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011].
 These neural structures are characterized by unique features such as a syncytial architecture [@Burkhardt2023] and an extensive repertoire of lineage-specific neuropeptides not found in other metazoans [@Sachkova2021; @Hayakawa2022].
@@
 This gravity-sensing organ contains a statocyst formed by lithocytes and four clusters of mechanosensory balancer cells, which act as pacemakers for the beating of comb-plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980].
+In the natural environment, mostly ctnophores can freely adjust their posture within the water column and exhibit both positively and negatively gravitactic orientations.
+This behavioral flexibility is likely supported by the aboral organ, which functions as a gravity sensor and pacemaker for ciliary movement.
+In this study, we focused on the orientation in which the aboral organ faces upward ("balancer-up" posture), as this configuration facilitates intuitive interpretation of balance-like functions and matches the setup used in high-speed imaging experiments.
 Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011].
@@
 
+This orientation corresponds to a “balancer-up” posture, in which the aboral organ faces upward.
+We selected this configuration not only to ensure consistent stimulation from the statolith across individuals, but also because it facilitates intuitive interpretation of balancer function—as if balancing a stone on top.
+Moreover, this posture matches the orientation used in electron microscopy and was technically advantageous for high-speed imaging under vertical-stage alignment.
+Using a consistent orientation throughout minimized potential confusion when comparing ultrastructural and functional data.
+
 Balancer cilia movement was imaged using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a high-sensitivity monochrome CMOS camera optimized for the near-infrared range (UI-3360CP-NIR-GL Rev.2, iDS), and a 40x glycerine-immersion objective lens (LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27).

### 2025-07-09 manuscript/Manuscript.qmd
ed main text

@@
 These coordinated ciliary movements not only generate propulsion but also allow precise control of body orientation and directional changes within the water column.
-Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioural patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system.
+Despite lacking a centralized nervous system, ctenophores exhibit sophisticated and dynamic behavioural patterns and respond to a broad range of external stimuli, suggesting the presence of an evolutionarily unique and functionally integrated sensorimotor system coordinating locomotion and posture.
 Several components of the ctenophore nervous system have been described, including a subepithelial nerve net (SNN), mesogleal neurons, tentacular nerves, and elements of the nervous system in the aboral organ [@Hertwig1880; @Hernandez_Nicaise_1984; @Jager2011].
@@
 This gravity-sensing organ contains a statocyst formed by lithocytes and four clusters of mechanosensory balancer cells, which act as pacemakers for the beating of comb-plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980].
-In the natural environment, mostly ctnophores can freely adjust their posture within the water column and exhibit both positively and negatively gravitactic orientations.
+In the natural environment, most ctenophores can freely adjust their posture within the water column and exhibit both positively and negatively gravitactic orientations.
 This behavioral flexibility is likely supported by the aboral organ, which functions as a gravity sensor and pacemaker for ciliary movement.
@@
 A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency.
-Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed one like in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b].
+Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed-direction response as seen in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b].
 Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997].

### 2025-07-09 manuscript/Manuscript.qmd
MS: edits of intro

--- 2025-07-08 manuscript/Manuscript.qmd
@@
 A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency.
-Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed one like in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b].
+Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system [@HORRIDGE1965; @Tamm1982; @tamm2014b].
 Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997].
@@
 
-To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy and high-speed imaging in the ctenophore *Mnemiopsis leidyi*.
+To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy (vEM) and high-speed imaging in the ctenophore *Mnemiopsis leidyi*.
 Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cells.

### 2025-07-10 manuscript/Manuscript.qmd
Ms: reorganised sections in introduction

--- 2025-07-09 manuscript/Manuscript.qmd
@@
 This gravity-sensing organ contains a statocyst formed by lithocytes and four clusters of mechanosensory balancer cells, which act as pacemakers for the beating of comb-plate cilia and thereby control body orientation [@HORRIDGE1965; @Tamm1980].
+A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency. 
+Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system [@HORRIDGE1965; @Tamm1982; @tamm2014b].
+
 In the natural environment, most ctenophores can freely adjust their posture within the water column and exhibit both positively and negatively gravitactic orientations.
@@
 In this study, we focused on the orientation in which the aboral organ faces upward ("balancer-up" posture), as this configuration facilitates intuitive interpretation of balance-like functions and matches the setup used in high-speed imaging experiments.
-Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011].
 
-A key feature of balancer cilia is their bidirectional, mechanoresponsive control of beat frequency.
-Depending on the direction of deflection and the animal’s geotactic state, balancer cilia may either increase or decrease their activity, indicating a dynamic response system, not a fixed-direction response as seen in vertebrate hair cells [@HORRIDGE1965; @Tamm1982; @tamm2014b].
-Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating this response [@Lowe1997].
+Previous studies have implicated membrane potential, intracellular calcium concentration, and synaptic input in modulating the balancer response [@Lowe1997].
 In particular, deflection-induced excitation can be abolished by calcium channel inhibitors or calcium-free conditions [@Lowe1997], while high extracellular K⁺ depolarizes and activates cilia in the absence of mechanical stimuli [@Lowe1997].
@@
 
-To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy and high-speed imaging in the ctenophore *Mnemiopsis leidyi*.
+Ultrastructural and immunohistochemical studies have revealed that the aboral organ contains diverse sensory and neural components, including a distinct neural structure termed the deep nerve net [@aronova1974electron; @Hernandez_Nicaise1968; @Jager2011].
+However, how this deep nerve net regulates balancer or other functions has remained elusive.
+
+To investigate how balancer cilia are modulated within the aboral organ, we combined volume electron microscopy (vEM) and high-speed imaging in the ctenophore *Mnemiopsis leidyi*. 
 Through connectomic reconstruction, we characterized the synaptic architecture of the aboral nerve net, revealing synaptic inputs to the balancer cells. 

### 2025-07-10 manuscript/Manuscript.qmd
Ms: edit results, discussion, add line breaks for easier overview of changes

@@
 From the aboral tip, we collected \~1,000 ultra-thin (50 nm) serial sections as ribbons, targeting only the region containing the aboral organ. 
-These were imaged using a Zeiss Gemini 500 SEM at a resolution of 2.0 nm/pixel (xy).
-We imaged only the region containing the aboral organ.
-Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012].
+These were imaged using a Zeiss Gemini 500 SEM at a pixel size (xy) of 2.0 nm. 
+We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. 
 The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. 
@@
 Since CATMAID only supports skeleton trees, branch nodes in such cases were placed near the closest existing node and annotated accordingly. 
-The entire skeletonized volume was composed of 134,591 nodes.
-73 fragments could not be attached to somata-associated skeletons.
+The entire skeletonized volume was composed of 134,591 nodes. 73 fragments could not be attached to somata-associated skeletons. 
 Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.
@@
 The second type includes two smaller neurons, ANN_Q1Q2 and ANN_Q3Q4, each confined to two adjacent quadrants and containing two nuclei. 
-ANN_Q1Q2 had 119 presynapses, while ANN_Q3Q4 had 68 (Figure 2C–F).
-In contrast, SNN neurons exhibited either no synaptic contacts or only a few.
+ANN_Q1Q2 had 119 presynapses, while ANN_Q3Q4 had 68 (Figure 2C–F). In contrast, SNN neurons exhibited either no synaptic contacts or only a few.
 
@@
 The bending of the cilia and the position of somata differed clearly when viewed laterally from the sagittal or the tentacular plane (Figure 3B). 
-In each quadrant, there were between 28-37 balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28).
-Each cell contained 3 to 10 mitochondria.
+In each quadrant, there were between 28-37 balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28). Each cell contained 3 to 10 mitochondria.
 
@@
 
-In both the Q1Q2 and the Q3Q4 regions, bridge cells also formed synapses with adjacent bridge cells. However, we found no synapses across the sagittal plane to bridge cells in the opposite region.
+In both the Q1Q2 and the Q3Q4 regions, bridge cells also formed synapses with adjacent bridge cells. 
+However, we found no synapses across the sagittal plane to bridge cells in the opposite region.
 
@@
 The ANN neurons also formed synapses on other cell types in the aboral organ, including the dense-vesicle cells, epithelial floor cells and several non-ciliated, monociliated or biciliated cells (Figure 3--figure supplement 1).
+None of these other cell types synapsed on the ANN.
 These cell types and synaptic connections are outside the gravisensory organ and are not considered further here.
@@
 
-To investigate whether balancer cilia are coordinated by neural inputs, we performed high-speed video microscopy to analyze their activity patterns across body axes.
+To investigate whether balancer cilia are coordinated by neural inputs, we did high-speed video microscopy recordings to analyze their activity patterns across body axes. 
 Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. 
@@
 This morphology suggests a potential role for bridge neurons in modulating inter-quadrant coordination. 
-By linking ANN activity across orthogonal planes, these neurons may contribute to the dynamic regulation of CBF patterns by facilitating information flow across orthogonal planes, thus supporting coordinated activity throughout the aboral organ.
+By linking ANN activity across orthogonal planes, these neurons may contribute to the dynamic regulation of CBF patterns by facilitating the spread of activity across orthogonal planes, thus supporting coordinated beating throughout the aboral organ.
 
@@
 The ANN has a morphology distinct from that of the subepithelial nerve net (SNN) reported for one-day-old cydippids [@Burkhardt2023; @Sachkova2021]. 
-A recent study also identified a distinct subtype of SNN neurons in the aboral region, termed “condensed” SNN, which display a more compact morphology compared to body wall SNNs [@Ferraioli2025].
-However, the ANN neurons described here are morphologically and anatomically distinct from both the body wall SNN and the condensed SNN.
+A parallel vEM study of a one-day-old cydippid also identified a distinct subtype of SNN neurons in the aboral region, termed “condensed” SNN, which display a more compact morphology compared to body wall SNNs [@ferraioli2025].
+However, the ANN neurons described here are morphologically and anatomically distinct from both the body wall SNN and the condensed SNN in the earlier stage. 
 Instead of forming discrete nodal swellings or a “beads-on-a-string” organization typical for the SNN, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion.
@@
 All observed output synapses were directed to balancer ciliary cells, bridge cells or other sensory cells. 
-This suggests that ANNs are unlikely to serve as motor drivers in the conventional sense.
+This suggests that ANNs are unlikely to serve as motor drivers (premotor neurons or motoneurons) in the conventional sense. 
 Rather, they may be involved in coordinating or modulating the timing and dynamics of the activity of the sensory-motile balancer cilia, possibly modulating their responsiveness to statocyst load [@keijzer2013; @dewiljes2015; @Veraszto2017].
@@
 In our volumetric reconstruction, bridge cells were identified as the only cell type providing synaptic input to the ANN network, positioning them as key upstream regulators of the ciliary coordination system in the aboral organ.
-Originally described morphologically by Tamm et al.
-(2002), bridge cells extend thin projections towards the base of the balancer cilia [@Tamm_2002; @tamm2014b].
+Originally described morphologically by Tamm et al. (2002), bridge cells extend thin projections towards the base of the balancer cilia [@Tamm_2002; @tamm2014b]. 
 Based on this spatial relationship, they may receive information about the physical state of the cilia---such as mechanical load or phase---or possibly local field potential (LFP) changes [@sheu2022; @jurisch-yaksi2024].
@@
 
-As Jagar et al.
-(2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011].
+As Jagar et al. (2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011]. 
 The network we describe here orchestrates output through dynamic, self-regulating coordination. 

### 2025-07-10 manuscript/Manuscript.qmd
add video7 and ed Method

@@
 
-To ensure consistency in gravitational input, we included only larvae whose aboral–oral axis was oriented nearly perpendicular to the gravitational vector (tilted by no more than 20°).
+To ensure consistency in gravitational input, we selected only larvae whose aboral–oral axis was oriented nearly perpendicular to the gravitational vector (within 20° tilt).
 This minimized the potential influence of larval tilt on left–right differences in balancer cilia activity, as all balancers received comparable stimulation from the statolith.
@@
 
-Balancer cilia movement was imaged using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a high-sensitivity monochrome CMOS camera optimized for the near-infrared range (UI-3360CP-NIR-GL Rev.2, iDS), and a 40x glycerine-immersion objective lens (LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27).
-To ensure consistent statolith positioning during imaging, the microscope was tilted by 90° to achieve vertical stage alignment.
-Illumination was provided by a custom-made NIR LED strobe system (850 nm) synchronized with the camera’s exposure signals.
-Image acquisition was controlled using BohNavi software.
+Because tilt angle strongly affects the beat frequency of balancer cilia via asymmetric mechanical loading by the statolith [@Tamm1980], our quantitative analysis was restricted to recordings that met the above orientation criterion (tilt < 20°).
+In this orientation, gravitational input from the statolith is approximately symmetrical across the left and right balancers, allowing meaningful comparisons of their activity.
+To illustrate how excessive tilt induces asymmetric stimulation, Video 7 presents two examples of larvae outside the permitted range: one tilted by ~135° in the sagittal plane and the other by ~60° in the tentacular plane.
+In both cases, strong asymmetries in balancer ciliary beating are evident, consistent with previous observations of gravity-induced geotactic responses [@Tamm1980].
+To illustrate how excessive tilt induces asymmetric stimulation, these examples demonstrate why we restricted our analysis to near-vertical orientations, where gravitational input is approximately balanced across the balancers.
+
+Balancer cilia movement was imaged using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a 40x glycerine-immersion objective (LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27).
+Images were acquired with a high-sensitivity monochrome CMOS camera (UI-3360CP-NIR-GL Rev.2, iDS), optimized for near-infrared detection.
+Illumination was provided by a custom-built 850 nm LED strobe system, synchronized with camera exposure using BohNavi software.
 Recordings were acquired at 640 × 480 pixels, 100 frames per second (fps), for a duration of 2 minutes, using 0.05 ms pulse illumination from the LED.
@@
 Data were saved as RDS objects for downstream analysis.
-Mean values were plotted as bar graphs with the left and right cilia shown in positive and negative directions, respectively.
+Mean values were plotted as bar graphs with left and right cilia plotted as positive and negative bars, respectively, to visually distinguish their activity.
 
@@
 This example captures a ciliary re-beat event in the tentacular plane.
+
+**Video 7.** High-speed imaging of balancer ciliary beating in the aboral organ tilted relative to gravity, illustrating gravity-induced asymmetry.
+The example on the left shows an aboral organ tilted by approximately 135° in the sagittal plane, while the example on the right shows one tilted by approximately 60° in the tentacular plane.
+For each case, the left and right balancer ciliary clusters are displayed separately.
+Ciliary beating is shown alongside kymographs extracted from two representative clusters.
+Recordings were acquired at 100 frames per second and are played back at 30 frames per second (0.3×).
+A white vertical line overlaid on each kymograph indicates the current frame shown in the video.
+These examples demonstrate asymmetrical balancer activity consistent with geotactic responses reported by Tamm (1980) [@Tamm1980].

### 2025-07-11 manuscript/Manuscript.qmd
Ms: Video 7 add reference in Results

@@
 We reconstructed 972 cells with intact nuclei and somata. 
-Among these, 898 were located within the aboral organ (AO), including balancers and the aboral nerve net neurons. 
+Among these, 898 were located within the aboral organ (AO), including balancers and the aboral nerve net neurons (Video 1). 
 Due to the compact and mostly self-contained organization of the AO, these cells were largely contained within the imaged volume.
@@
 Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014c; @tamm2015].
-To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector.
-In all cases, only larvae whose aboral–oral axis was approximately parallel to the gravity vector (mouth facing downward within a 0–20° deviation) were included for analysis. 
-This ensured that all balancer cilia experienced comparable gravitational input from the statolith, allowing for an unbiased comparison across quadrants.
-We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage (Figure 4A).
+To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector (Videos 3-7).
+We could confirm earlier finding that the beating frequency of balancer cilia is dependent on the inclination (Video 7).
+Our aim here was to focus on spontaneous ciliary beating under conditions of equal load from the statolith across quadrants.
+Therefore, we only analysed larvae whose aboral–oral axis was approximately parallel to the gravity vector (mouth facing downward within a 0–20° deviation). 
+We then compared larvae oriented either with their sagittal or tentacular plane parallel to the sample stage (Figure 4A).
 
@@
 
-This orientation corresponds to a “balancer-up” posture, in which the aboral organ faces upward.
+This orientation corresponds to a “balancer-up” orientation, in which the aboral organ faces upward.
 We selected this configuration not only to ensure consistent stimulation from the statolith across individuals, but also because it facilitates intuitive interpretation of balancer function—as if balancing a stone on top.
-Moreover, this posture matches the orientation used in electron microscopy and was technically advantageous for high-speed imaging under vertical-stage alignment.
+Moreover, this orientation matches the orientation used in electron microscopy and was technically advantageous for high-speed imaging under vertical-stage alignment.
 Using a consistent orientation throughout minimized potential confusion when comparing ultrastructural and functional data.
 
-Because tilt angle strongly affects the beat frequency of balancer cilia via asymmetric mechanical loading by the statolith [@Tamm1980], our quantitative analysis was restricted to recordings that met the above orientation criterion (tilt < 20°).
+Because the tilt angle strongly affects the beat frequency of balancer cilia via asymmetric mechanical loading by the statolith [@Tamm1980], our quantitative analysis was restricted to recordings that met the above orientation criterion (tilt < 20°).
 In this orientation, gravitational input from the statolith is approximately symmetrical across the left and right balancers, allowing meaningful comparisons of their activity.
@@
 In both cases, strong asymmetries in balancer ciliary beating are evident, consistent with previous observations of gravity-induced geotactic responses [@Tamm1980].
-To illustrate how excessive tilt induces asymmetric stimulation, these examples demonstrate why we restricted our analysis to near-vertical orientations, where gravitational input is approximately balanced across the balancers.
 

### 2025-07-11 README.Rmd
reorganise folders
Ms: add Jokura et al Zenodo ref, update Readme.Rmd

@@
 # About the project: 
-- this is a repository of an R project containing code and source data for the ctenophore apical organ connectome paper.
+- this is a repository of an R project containing code and source data for the paper **Neural Connectome of the Ctenophore Statocyst**.
 Contributed by [Jékely lab](https://www.cos.uni-heidelberg.de/en/research-groups/gaspar-jekely).
@@
 
-> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>
-
-Our pre-print is online here:
-
-> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+*Neural Connectome of the Ctenophore Statocyst*\
+Kei Jokura, Sanja Jasek, Lara Niederhaus, Pawel Burkhardt, Gáspár Jékely\
+bioRxiv 2025.06.26.661250; doi: <https://doi.org/10.1101/2025.06.26.661250>
 
@@
 
-Please cite this compendium as:
+Please cite the permanend Zenodo record for this compendium as:
 
-> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>
+Jokura, K., Jasek, S., & Jékely, G. (2025). \
+Code documentation for Neural Connectome of the Ctenophore Statocyst (v1.0). Zenodo.\
+https://doi.org/10.5281/zenodo.15861953
 
@@
 
-  - [:file\_folder: text](/text/Manuscript.qmd): Quarto source document
+  - [:file\_folder: manuscript](/manuscript/Manuscript.qmd): Quarto source document
     for manuscript. Includes text, references and inserts the latest version of the figures from the /figures folder. 
@@
   - [:file\_folder: data](analysis/data): Data used in the analysis.
-  - [:file\_folder: data](analysis/scripts): scripts used in the analysis.
+  - [:file\_folder: scripts](analysis/scripts): scripts used in the analysis.
   - [:file\_folder: figures](manuscript/figures): Plots and other
@@
   - [:file\_folder:
-    supplements](manuscript/supplements):
-    Supplementary materials including notes and other documents
-    prepared and collected during the analysis.
+    videos](manuscript/videos):
+    Videos.
   - [:file\_folder:
